### 2020-05-06 article/main.tex
ajout article

\documentclass[preprint]{iacrtrans}

\usepackage[utf8]{inputenc}
\usepackage{amssymb, amsmath, amsfonts, amscd}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{algorithm2e}
\usepackage{tikz}
\usepackage{cellspace}
\usepackage{multirow}
%\usepackage{parskip}
\usetikzlibrary{patterns}

%\usepackage[noend]{algpseudocode}

\usepackage[pdftex,bookmarks,bookmarksopen,bookmarksdepth=3]{hyperref}
\hypersetup{colorlinks=true,citecolor=red,linkcolor=red,urlcolor=black}

% knuth-style algos
\newcommand{\slug}{\hbox{\kern1.5pt\vrule width2.5pt height6pt depth1.5pt\kern1.5pt}}
\def\xskip{\hskip 7pt plus 3pt minus 4pt}
\newdimen\algindent
\newif\ifitempar \itempartrue % normally true unless briefly set false
\def\algindentset#1{\setbox0\hbox{{\bf #1.\kern.25em}}\algindent=\wd0\relax}
\def\algbegin #1 #2{\algindentset{#21}\alg #1 #2} % when steps all have 1 digit
\def\aalgbegin #1 #2{\algindentset{#211}\alg #1 #2} % when 10 or more steps
\def\aaalgbegin #1 #2{\algindentset{#2111}\alg #1 #2} % when 10 or more steps
\def\alg#1(#2). {\medbreak % Usage: \algbegin Algorithm A (algname). This...
  \noindent{\bf#1}({\it#2\/}).\xskip\ignorespaces}
\def\algstep#1.{\ifitempar\smallskip\noindent\else\itempartrue
  \hskip-\parindent\fi
  \hbox to\algindent{\bf\hfil #1.\kern.25em}%
  \hangindent=\algindent\hangafter=1\ignorespaces}
% end of borrowed macros

\title{Predicting the PCG PSeudo-Random Number Generator In Practice}

\author{Charles Bouillaguet\inst{1} \and Julia Sauvage\inst{2} \and Florette Martinez\inst{3}}

\institute{% 
University of Lille, France \\ 
\email{charles.bouillaguet@univ-lille.fr}
\and 
Sorbonne University \\
\email{julia.sauvage@etu.upmc.fr}
\and 
LIP6, CNRS, SU ? \\
\email{florette.martinez@lip6.fr}

}

\begin{document}
\maketitle

\keywords{keywords}

\begin{abstract}
  blablabla
\end{abstract}

\section{Introduction}

\section{The PCG Pseudo-Random Number Generator Family}

\section{Warm-Up: Prediction with Known Increment }

\begin{theorem}
  The algorithm described above works with probability.... in time ...
\end{theorem}

\section{Prediction with Unknown Increment}

\begin{theorem}
  The algorithm described above works with probability.... in time ...
\end{theorem}

\section{Implementation of the Predictor and Practical Results}

\section{Conclusion}

\bibliographystyle{alpha}
\bibliography{biblio}

%\appendix

\end{document}

### 2020-05-06 C/fonctions.c
up-to-date

@@ -24,6 +24,8 @@ void init_var_globales(){
     //multiplier a OK !
     a = (((pcg128_t) 2549297995355413924) << k) + ((pcg128_t) 4865540595714422341);
     
+    a_inv = (((pcg128_t) 566787436162029664) << k) + ((pcg128_t) 11001107174925446285ull);
+
     //increment c OK !
     c = (((pcg128_t)6364136223846793005) << k) + 1442695040888963407;
     
@@ -111,12 +113,14 @@ int solve(pcg128_t* S, const unsigned long long* X, const int* rot, const unsign
     double Yprim[nbiter];
     unsigned long long tmp3[nbiter];
     
+    // total : 6 XOR, 6 AND, 6 SHIFT, 3 ADD, 3 SUB, 3 u64->double
     for (int i = 0 ; i < nbiter ; i++) {
         Y[i] = (((sumPol[i] ^ X[i]) % (1 << known_low)) << known_up ) + (rot[i] ^ (X[i] >> (k - known_up)));
         Yprim[i] = (Y[i] - sumPolY[i]) % (1 << (known_low + known_up));    
     }
     
     double tmp2[nbiter];
+    // total : 9 MUL (double) + 10 ADD (double)
     for (int i=0 ; i<nbiter ; i++) {
         tmp2[i] = 0;
         for(int j=0 ; j<nbiter ; j++)
@@ -124,16 +128,20 @@ int solve(pcg128_t* S, const unsigned long long* X, const int* rot, const unsign
         tmp2[i] += 6755399441055744.0;
     }
  
+    // total : 6 SHIFT
     for(int i = 0 ; i < nbiter ; i++)
         tmp3[i] = light_crazy_round(tmp2[i]);
 
     unsigned long long Sprim0 = 0;
+    // total : 3 ADD, 3 MUL
     for(int j=0 ; j<nbiter ; j++)
         Sprim0 += Greduite[j] * tmp3[j];
 
+    // SHIFT, ADD
     unsigned long long Smod = (Sprim0 << known_low) + sumPol[0];
     S[0] = (((pcg128_t)(Smod ^ X[0])) << k) + ((pcg128_t) Smod);
 
+    // clocking PCG two times
     for (int i = 1 ; i < nbiter ; i++) {
         S[i] = S[i-1] * a + c;
         unsigned long long XX = S[i] ^ (S[i] >> 64);

### 2020-05-06 C/fonctions.h

@@ -11,6 +11,7 @@
 
 int nb_thread;
 pcg128_t a;
+pcg128_t a_inv;
 pcg128_t c;
 pcg128_t polC[nbiter];
 

### 2020-05-06 C/main.c

@@ -3,6 +3,8 @@
 #include <stdio.h>
 #include <time.h>
 
+// Ce truc marche avec GCC mais pas avec ICC. Il faut investiguer...
+
 int main()
 {
     
@@ -12,9 +14,13 @@ int main()
         
     /********** INPUT ***********/
     unsigned long long X[nbiter];
-    X[ 0] = 0x3c8673c55ead80bc;
-    X[ 1] = 0xfd27cfa6f84ef5f1;
-    X[ 2] = 0x68a098da45558d92;
+//     X[ 0] = 0x3c8673c55ead80bc;
+//     X[ 1] = 0xfd27cfa6f84ef5f1;
+//     X[ 2] = 0x68a098da45558d92;
+    // valeurs fournies par M. O'Neil.
+    X[ 0] = 0x6ec191a37a421087;
+    X[ 1] = 0xec140ace169176fc;
+    X[ 2] = 0x85994d489913af70;
 
     double t1 = wtime();
     unsigned long long done = 0;
@@ -26,7 +32,7 @@ int main()
         pcg128_t S[nbiter];
         pcg128_t polW[nbiter];
         unsigned long long urX[nbiter];
-        int rot[nbiter];
+        int rot[nbiter] = {0, 0, 0};
         unsigned long long sumPol[nbiter];
         unsigned long long sumPolY[nbiter];
         
@@ -53,8 +59,10 @@ int main()
             if (solve(S, urX, rot, sumPol, sumPolY)) {
                 printf("\nInternal state found (%.1fs)\n", wtime() - t1);
                 struct pcg_state_128 rng;
-                rng.state = S[0];
-                for(int i = 1 ; i < 10 ; i++)
+                rng.state = (S[0] - c) * a_inv; // S[-1]
+                pcg128_t seed = (rng.state - c - a*c) * a_inv;
+                printf("Seed : %016" PRIx64 " %016" PRIx64 "\n", (uint64_t) (seed >> 64), (uint64_t) seed);
+                for(int i = 0 ; i < 16 ; i++)
                     printf("X[%2d] = %016" PRIx64 "\n", i, pcg_oneseq_128_xsl_rr_64_random_r(&rng));
             }
         }

### 2020-05-06 Cunknown/fonctions.c

@@ -5,6 +5,7 @@
 #include "fonctions.h"
 
 pcg128_t a;
+pcg128_t a_inv;
 pcg128_t powA[nboutput];
 pcg128_t polA[nboutput];
 
@@ -26,6 +27,8 @@ void init_var_globales()
 {
     //multiplier a OK !
     a = (((pcg128_t) 2549297995355413924) << k) + ((pcg128_t) 4865540595714422341);
+
+    a_inv = (((pcg128_t) 566787436162029664) << k) + ((pcg128_t) 11001107174925446285);
 
     //increment polynome polC OK !
     polA[0] = 0;
@@ -99,6 +102,7 @@ static inline int checkY(const char* goodY, int i, u64 Y)
 static inline bool confirm(u64 Y0, u64 DS640, const struct task_t *task)
 {
 	/**** Confirmation du DS640 ****/
+    // <= 8 ADD, 4 MUL, 8 SHIFT, 8 AND
     for (int i = 0 ; i < nbtest ; i++) {
         u64 tmp2 = ((u64) polA[i + nbiter]) * DS640 + task->sumPolTest[i]; //ATTENTION cast pcg128_t
         u64 Yi1 = (Y0 + (tmp2 >> (k - known_low - known_up))) % (1 << (known_low + known_up)); //avec ou sans retenue OK!
@@ -135,12 +139,15 @@ bool solve_isgood(const struct task_t *task)
     for (int i = 0; i < nbiter; i++) //Y
         tmp[i] = task->tabTmp[i + nbiter * task->rot[i]];  
     
+    // ADD
     u64 Y0 = tmp[0] + task->sumPolY[0];
     
+    // 4 SUB, 4 AND, 4 u64->double
     double tmp3[nbiter - 1];
     for (int i = 0; i < nbiter - 1; i++)  //DY
         tmp3[i] = (tmp[i+1] - tmp[i]) % (1 << (known_low + known_up));
     
+    // 20 ADD (double), 16 MUL (double)
     double u[nbiter-1];
     for (int i = 0; i < nbiter - 1; i++) {
         u[i] = 0.0;
@@ -149,11 +156,13 @@ bool solve_isgood(const struct task_t *task)
         u[i] += 6755399441055744.0;
     }
 
+    // 8 SHIFT, 4 MUL, 4 ADD
     u64 DS640 = 0;
     for (int i = 0; i < nbiter-1; i++) {
[…]

### 2020-05-06 Cunknown/main.c

@@ -206,7 +206,7 @@ void master(const u64 (*X)[nboutput])
 	int active_ranks;
 	int done = 0;
 	MPI_Comm_size(MPI_COMM_WORLD, &active_ranks);
-	active_ranks *= 2; /* because of the hack that runs two "threads" per MPI rank */
+	active_ranks = 2*(active_ranks - 1); /* because of the hack that runs two "threads" per MPI rank */
 
 	while (active_ranks > 0) {
 		MPI_Status status;
@@ -255,7 +255,7 @@ void master(const u64 (*X)[nboutput])
  * OK, so a bug in Intel's MPI library prevents us from using one MPI rank
  * per hyperthread, so here we cheat.
  */
-void slave(const u64 (*X)[nboutput])
+void slave(const u64 (*X)[nboutput], int rank)
 {
 	int T = omp_get_max_threads();
 	for (int i = 0; i < T; i++)
@@ -283,8 +283,10 @@ void slave(const u64 (*X)[nboutput])
 				do_task(task_id[t], &task[t], X);
 		}
 		for (int i = 0; i < T; i++)
-			MPI_Bsend(&task_id[i], 1, MPI_INT, 0, tag_result, MPI_COMM_WORLD);
+			if (task_id[i] >= 0)
+				MPI_Bsend(&task_id[i], 1, MPI_INT, 0, tag_result, MPI_COMM_WORLD);
 	}
+	printf("MPI slave %d, I'm done!\n", rank);
 }
 
 
@@ -305,20 +307,20 @@ int main(int argc, char **argv)
 	init_var_globales();
 	
 	u64 X[nboutput];
-	X[ 0] = 0xd4166f4c3e02d10a;
-	X[ 1] = 0x1d1ceb21e7737101;
-	X[ 2] = 0xf8b90f473a5426d3;
-	X[ 3] = 0xe3a3b7babb2ad9ca;
-	X[ 4] = 0x0077f2c80987dd13;
-	X[ 5] = 0xf8ddaf2431548a13;
-	X[ 6] = 0x80935e041bbab85a;
-	X[ 7] = 0xbe0fde3939201c50;
-	X[ 8] = 0xe9604fdf6b2177b7;
-
+	X[ 0] = 0x46b9d66bb1e9fbf0;
+	X[ 1] = 0xc0b7382496f0e363;
[…]

### 2020-05-06 Cunknown/step2.py

@@ -244,54 +244,55 @@ def full_DS(X, W0, WC, rots):
 
 if __name__ == '__main__':
     X = [0] * 48
-    X[ 1] = 0xd4166f4c3e02d10a;
-    X[ 2] = 0x1d1ceb21e7737101;
-    X[ 3] = 0xf8b90f473a5426d3;
-    X[ 4] = 0xe3a3b7babb2ad9ca;
-    X[ 5] = 0x0077f2c80987dd13;
-    X[ 6] = 0xf8ddaf2431548a13;
-    X[ 7] = 0x80935e041bbab85a;
-    X[ 8] = 0xbe0fde3939201c50;
-    X[ 9] = 0xe9604fdf6b2177b7;
-    X[10] = 0x95d9cf24a229cedf;
-    X[11] = 0x0434a85418759293;
-    X[12] = 0x04d230c1debe7999;
-    X[13] = 0x83a3cd257d4d04b0;
-    X[14] = 0x13990a23037c13c4;
-    X[15] = 0xfbfaea2e50411202;
-    X[16] = 0x421a394a36baebf8;
-    X[17] = 0x5a878c4594ea7221;
-    X[18] = 0xbd37307bdd522b9c;
-    X[19] = 0x39af06b3e9b3ae10;
-    X[20] = 0x1d21f1cee77b8e2e;
-    X[21] = 0xe4bddad0aacaf420;
-    X[22] = 0x1009ed344cd7f2f4;
-    X[23] = 0xff287c5797cceb71;
-    X[24] = 0x85e8968b0d8ab49b;
-    X[25] = 0x69a9b821830862cc;
-    X[26] = 0xf9fe65ed23740aea;
-    X[27] = 0x47669184bc43d948;
-    X[28] = 0xe3f19b31915ae6d3;
-    X[29] = 0x5f3945718b6dac44;
-    X[30] = 0x49bfacfe8056b33c;
-    X[31] = 0xf2b358ceb722628f;
-    X[32] = 0xb4d1bf17f2c57b71;
-    X[33] = 0xb4300000ad802deb;
-    X[34] = 0xe3125e8022de888d;
-    X[35] = 0x2c7f6d404196ed4d;
-    X[36] = 0xb10490274ecbe897;
-    X[37] = 0xb04da8a406da3814;
-    X[38] = 0xbc70124be9a196c9;
-    X[39] = 0xf81a244f765141e6;
-    X[40] = 0x20413cda8442a149;
[…]

### 2020-05-06 Cunknown/step3.py

@@ -1,14 +1,9 @@
-import time
-import fpylll
-import os
-from itertools import product
-from math import log
-import random
 
 k = 64
 known_up  = 6
 known_low  = 13
 a = 2549297995355413924 * 2**64 + 4865540595714422341
+a_inv = 566787436162029664* 2**64 + 11001107174925446285
 nbiter = 5
 nboutput = 40
 N = 2**128
@@ -65,60 +60,60 @@ def sortiesGenerateur(S0, c, k):
 
 if __name__ == '__main__':
     X = [0] * 48
-    X[ 1] = 0xd4166f4c3e02d10a;
-    X[ 2] = 0x1d1ceb21e7737101;
-    X[ 3] = 0xf8b90f473a5426d3;
-    X[ 4] = 0xe3a3b7babb2ad9ca;
-    X[ 5] = 0x0077f2c80987dd13;
-    X[ 6] = 0xf8ddaf2431548a13;
-    X[ 7] = 0x80935e041bbab85a;
-    X[ 8] = 0xbe0fde3939201c50;
-    X[ 9] = 0xe9604fdf6b2177b7;
-    X[10] = 0x95d9cf24a229cedf;
-    X[11] = 0x0434a85418759293;
-    X[12] = 0x04d230c1debe7999;
-    X[13] = 0x83a3cd257d4d04b0;
-    X[14] = 0x13990a23037c13c4;
-    X[15] = 0xfbfaea2e50411202;
-    X[16] = 0x421a394a36baebf8;
-    X[17] = 0x5a878c4594ea7221;
-    X[18] = 0xbd37307bdd522b9c;
-    X[19] = 0x39af06b3e9b3ae10;
-    X[20] = 0x1d21f1cee77b8e2e;
-    X[21] = 0xe4bddad0aacaf420;
-    X[22] = 0x1009ed344cd7f2f4;
-    X[23] = 0xff287c5797cceb71;
-    X[24] = 0x85e8968b0d8ab49b;
-    X[25] = 0x69a9b821830862cc;
-    X[26] = 0xf9fe65ed23740aea;
-    X[27] = 0x47669184bc43d948;
-    X[28] = 0xe3f19b31915ae6d3;
-    X[29] = 0x5f3945718b6dac44;
-    X[30] = 0x49bfacfe8056b33c;
[…]

### 2020-05-06 challenges/pcg64-challenge.c

@@ -4,46 +4,34 @@
 
 #include "pcg_variants.h"
 
-#define DEBUG 1
-
 int main()
 {
-    // Seed the full-blown generator with external entropy
-    pcg128_t seeds[2];
-    
-    FILE *f = fopen("/dev/urandom", "r");
-    if (f == NULL) {
-        perror("Something went wrong when opening /dev/urandom");
-        exit(EXIT_FAILURE);
-    }
-    if (fread(seeds, sizeof(seeds), 1, f) != 1)  {
-        perror("Something went wrong when reading /dev/urandom");
-        exit(EXIT_FAILURE);
-    }
+	pcg128_t seeds[2];
 
-    printf("Seed[0] : %016" PRIx64 " %016" PRIx64 "\n", (uint64_t) (seeds[0] >> 64), (uint64_t) seeds[0]);
-    printf("Seed[1] : %016" PRIx64 " %016" PRIx64 "\n\n", (uint64_t) (seeds[1] >> 64), (uint64_t) seeds[1]);
+	FILE *f = fopen("/dev/urandom", "r");
+	if (f == NULL) {
+		perror("Something went wrong when opening /dev/urandom");
+		exit(EXIT_FAILURE);
+	}
+	if (fread(seeds, sizeof(seeds), 1, f) != 1) {
+		perror("Something went wrong when reading /dev/urandom");
+		exit(EXIT_FAILURE);
+	}
 
-    pcg64_random_t rng;
-    pcg64_srandom_r(&rng, seeds[0], seeds[1]);
-    
-    int known_low = 11;
-    if (DEBUG) {
-        printf("known_low = %d\n", known_low);
-        printf("W_c = %04llx\n", (unsigned long long) (rng.inc % (1 << known_low)));
-    }
+	printf("Seed[0] : %016" PRIx64 " %016" PRIx64 "\n", (uint64_t) (seeds[0] >> 64), (uint64_t) seeds[0]);
+	printf("Seed[1] : %016" PRIx64 " %016" PRIx64 "\n\n", (uint64_t) (seeds[1] >> 64), (uint64_t) seeds[1]);
 
-    printf("Predictor input:\n");
-    for (int i = 0; i < 48; i++) {
-        printf("X[%2d] = 0x%016" PRIx64 ";", i, pcg64_random_r(&rng));
-        if (DEBUG)
[…]

### 2020-05-06 challenges/pcg64s-challenge.c

@@ -6,30 +6,31 @@
 
 int main()
 {
-    // Seed the "single-stream" generator with external entropy
-    pcg128_t seed;    
-    FILE *f = fopen("/dev/urandom", "r");
-    if (f == NULL) {
-        perror("Something went wrong when opening /dev/urandom");
-        exit(EXIT_FAILURE);
-    }
-    if (fread(&seed, sizeof(seed), 1, f) != 1)  {
-        perror("Something went wrong when reading /dev/urandom");
-        exit(EXIT_FAILURE);
-    }
-    printf("Seed : %016" PRIx64 " %016" PRIx64 "\n\n", (uint64_t) (seed >> 64), (uint64_t) seed);
+	// Seed the "single-stream" generator with external entropy
+	pcg128_t seed;
+	FILE *f = fopen("/dev/urandom", "r");
+	if (f == NULL) {
+		perror("Something went wrong when opening /dev/urandom");
+		exit(EXIT_FAILURE);
+	}
+	if (fread(&seed, sizeof(seed), 1, f) != 1) {
+		perror("Something went wrong when reading /dev/urandom");
+		exit(EXIT_FAILURE);
+	}
+	seed = (((pcg128_t) 0x5b3d418b8b51364b) << 64) + ((pcg128_t) 0x74eed710223e763f);
+	printf("Seed : %016" PRIx64 " %016" PRIx64 "\n\n", (uint64_t) (seed >> 64), (uint64_t) seed);
 
-    pcg64s_random_t rng;
-    pcg64s_srandom_r(&rng, seed);
-    
-    printf("Predictor input:\n");
-    for (int i = 0; i < 3; i++)
-        printf("X[%2d] = 0x%016" PRIx64 ";\n", i, pcg64s_random_r(&rng));
-    printf("\n");
-    
-    printf("Remaining of the sequence (predictor output, in principle):\n");
-    for (int i = 3; i < 10; i++)
-        printf("X[%2d] = 0x%016" PRIx64 ";\n", i, pcg64s_random_r(&rng));
+	pcg64s_random_t rng;
+	pcg64s_srandom_r(&rng, seed);
 
-    return EXIT_SUCCESS;
+	printf("Predictor input:\n");
+	for (int i = 0; i < 3; i++)
+		printf("X[%2d] = 0x%016" PRIx64 ";\n", i, pcg64s_random_r(&rng));
+	printf("\n");
+
[…]

### 2020-05-08 article/main.tex
up-to-date
ajout du rapport de projet sans unknown increment (qui est juste un algo faux)

@@
 \hypersetup{colorlinks=true,citecolor=red,linkcolor=red,urlcolor=black}
+
+%Ceux qu'on utilisait dans le projet :
+
+\usepackage{algorithm}
+\usepackage{algorithmic}
 
@@
 
-\section{Introduction}
+\section{Introduction} %ce qu'on avait écrit pour le projet, peut-être pas incroyable...
+
+Today's world has become a data driven world.
+Having access unpredictable data is invaluable when considering security problem.
+For example, security keys have to be generated randomly whether we are talking about RSA keys for secure communication or a cipher key for encrypting data.
+For this purpose, random data can be generated by pseudo-random generators, that rely on a predetermined mathematical structure to output seemingly random data.
+However, by knowing the internal structure and state of the generator, it is possible to reconstruct the entire string of data.\\
+
+The goal of our project is to study the flaws and try to cryptanalyse pseudo-random generators from the Permuted Congruential Generator (PCG) family, designed by Melissa E.
+O'Neill.\cite{melissapaper} Indeed, predicting a sequence from these generators is said to be "challenging"\cite{melissaweb} by their author, but this statement is not backed by any studies.
+Furthermore, PCG generators are very close to another well known family of generators, the Truncated Linear Congruential Generators, for which a certain amount of viable prediction algorithms are known.
+The combination between being highly commended (it achieved 5th place out of 29 in a recent performance test survey\cite{survey}) and a well presented website\cite{melissaweb} giving an feeling of professionalism can lead to the impression that PCG is a sufficiently secure family of generators.
+Given the lack of studies on the subject, there is a problem if it is used in critical pieces of software.
+This is why we will attempt to find flaws in PCG random generators.\\
+
+Some attacks against PCG have been carried out by Sebastiano Vigna\cite{vignaweb}, but this didn't yield any significant results, given that he only tried to break an incomplete version in a fairly straightforward way.
+He worked on a very specific version of PCG.
+He considers the result without modulo, which is a big simplification.
+Moreover, he didn't study the 128 bits long generator, but only the 64 bits long\cite{vignacode}.\\
+
+We focused on the general version of PCG, which happens to be very similar to the generator implemented by the function \texttt{random} of the C library.
+Both are based on a well known family of generators, the Linear Congruential Generators (LCG), but with the PCG one also using a non linear function and a rotation on the output.\\
+
+We are looking for a possible way to reduce this generator to an LCG, or a truncated LCG, because there are known algorithms to predict them.
+Because we aren't studying the same exact version as Sebastiano Vigna, his findings aren't very relevant to us.\\
+
+We decided to study the default version of the largest available generator of the 0.94 version of the PCG family, which has an internal state of 128 bits, named PCG128.
+We will mathematically analyse the algorithm to uncover weaknesses, and if methods for the LCG family can be translated to the PCG family.
 
@@
 
+\subsection{LCG algorithm}
+
+The PCG algorithm we studied is based on the Linear Congruential Generators algorithm.
+The LCG algorithm is simply based on a congruential multiplication.
+This type of generator is based on four values: an internal state $S_{i}$, a multiplier $A$, an increment $C$ and the size $M$.
+Generally, to initialize a generator of a given size, a seed value is provided as the first internal state, and a default multiplier and increment are used.
+We will study the case where $M$ and $A$ are known and where $M$ is a power of 2.\\
+
+Calling the generator results in calculating the next internal state $S_{i+1} = A \times S_i + C \mod{M}$ and outputting $S_{i+1}$ as data.
+
+This type of generator is trivially predictable because, the internal state is given to us at each output, and in the event that $C$ is unknown, we can still deduce it with two consecutive outputs: $C = S_{i+1} - A \times S_i$.\\
+
+A better variant of the standard LCG is the Truncated LCG, where the output consists only of the higher bits of the internal state $S_i$.
+This way the entire internal state isn't accessible.
+Returning the lower bits is not interesting because as we can understand easily and will prove later, lower bits of a state only depend on the lower bits of the previous state.
+
+\subsection{PCG algorithm}
+
+The PCG algorithm we studied is based on the LCG algorithm. To start, here are a few definitions:
+\begin{itemize}
+    \item let $k$ be the size of our generator, in this case $k = 128$ bits;
+    \item let $M$ be the modulus with $M = 2^k$;
+    \item let $S_i$ be the internal state at the $i$-th iteration and $S$ be the list of $S_i$;
+    \item let $X_i$ be the output at the $i$-th iteration of the generator and $X$ be the list of $X_i$;
+    \item let $lowS_i$ be the $k/2$ least significant bits of $S_i$;
+    \item let $upS_i$ be the $k/2$ most significant bits of $S_i$;
+    \item let $rot_i$ be the 6 most significant bits of $S_i$;
+    \item let $nbiter$ be the number of consecutive outputs for prediction purposes.
+\end{itemize}
+
+The internal state can be redefined as:
+\begin{equation}
+    S_i = 2^{k/2} \times upS_i + lowS_i 
+\end{equation}
+
+The 6 most significant bits, $rot_i$, of the internal state are used as a value to determine the rotation in the PCG algorithm.
+Let \texttt{rotate} be the function that performs said rotation (Figure \ref{pcg128out}).\\
+The rotation is a simple exchange of bits, where a $k/2$ bit value is split into 2 parts: one of $rot_i$ bits and the other in $(k/2) - rot_i$ bits.
+The two parts are then swapped, and the new value is returned.
+
+This generator is initialised with a given $k$ bit internal state $S_0$ (of size 128 bits in the case of PCG128), and optionally with an $k$ bit increment $C$.
+If no increment is given, the algorithm will use a known default increment.
+For each output of the generator, the algorithm will perform the following calculations" (Figure \ref{pcg128out}):\\
+\begin{align}
+    X_{i} &= \mathtt{rotate}(upS_i\ \mathtt{XOR}\ lowS_i)\\
+    S_{i+1} &= A \times S_{i} + C \mod{2^k}
+\end{align}
+In the case of PCG128, $A$ is a default increment of size 125 bits prime with $2^k$.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.75\linewidth]{pictures/PCG128.png}
+    \caption{PCG128 : Output process}
+    \label{pcg128out}
+\end{figure}
+
+Given the nature of the generator, the following values are known at every step:
+\begin{itemize}
+    \item the size of the generator $k$;
+    \item the list of outputs $X$;
+    \item the multiplier $A$;
+    \item and eventually $C$, if the default increment was used. 
+\end{itemize}
+
+To break this generator, we had to differentiate between cases when the default $C$ was used or not. It was also interesting to note what was the minimum number of $X_i$ needed to reliably attack it.
+
+\subsection{Brute-Force Attack with known $C$}
+In the case where $C$ is known, if we find $S_0$, we can calculate every $S_i$ with the given formula, and predict all the following outputs of the generator.
+The naive way is by doing a comprehensive research of $upS_0$, and then calculating $lowS_0$ with the formula :\\
+\begin{equation}
+    lowS_0 = upS_0\ \mathtt{XOR}\ \mathtt{unrotate}(X_0)
+\end{equation}
+\texttt with {unrotate} being the inverse function of \texttt{rotate}.\\
+
+If our guessing of $upS_0$ is correct, we can obtain the correct $X_1$ with $S_1$ deduced from the initial $S_0$.
+The preceding is a naive method for predicting PCG generator in $2^{64}$ operations.
+But we can have other values that verify that test.
+This will not return $S_0$ but a very short list of possibles $S_0$, with the actual one in it.
+We just need to test the following outputs to find the real $S_0$.
+This is yet a very slow way to guess the internal state, so we need a better solution.
+
+\subsection{Brute-Force Attack with unknown $C$}
+
+In case $C$ is unknown, retrieving $S_0$ is harder. The naive algorithm gives us an answer within $2^{128+6}$ simple operations whit the following algorithm:
+\\
+\begin{algorithmic}[!h]
+\FOR{all possibilities of $lowC$ and $upS_0$}
+    \STATE \texttt{unrotate}($X_0$)
+    \STATE $lowS_0 = upS_0\ \mathtt{XOR}\ X_0$
+    \STATE $lowS_1 = lowC + lowS_0 \times A \mod{2^{k/2}}$
+    \FOR{all possibilities of $rot_1$}
+        \STATE $\mathtt{unrotate}(X_1)$
+        \STATE $upS_1 = lowS_1\ \mathtt{XOR}\ X_1$
+        \STATE $C = S_1 - A \times S_0 \mod{2^{k}}$
+        \STATE $S_2 = S_1 \times A + C$
+        \IF {$S_2$ is coherent with  $X_2$}
+            \STATE print($C$,$S_0$)
+        \ENDIF
+    \ENDFOR
+\ENDFOR
+\end{algorithmic}
+
+This algorithm reveals us $C$ and a list of outputs with $S_0$ in it.
+The entry is 256 bits long, and we used 3 64 bits long outputs of the generator.
+There will be about $2^{64}$ possible $S_0$.
+This is why we need a couple other outputs to get the real $S_0$.
+Sorting out which is the correct $S_0$ can be done afterwards, and won't affect the complexity.
+But given the complexity of this algorithm, it is even more unrealistic to solve problem in this fashion then the previous one where $C$ was known.
+
+\subsection{Reduction to a truncated LCG model}
+By exploiting the nature of this generator, it becomes apparent that some values can be easily guessed.
+We already saw that if we guess $rot_i$, then retrieving $X_i$ becomes a simple \texttt{XOR} operation, and in turn, if we have some knowledge of a particular bit of $lowS_i$ (respectively $upS_i$), we can deduce the equivalent bit from $upS_i$ (respectively $lowS_i$).\\
+
+Moreover, there is an interesting proprity concerning the least significant bits of the internal state.
+Indeed, if we define $lowC$ as the 64 least significant bits of $C$ ($lowC = C \mod{2^{k/2}})$, we have:
+
+\begin{align}
+    S_{i+1} &= A \times S_i + C \mod{2^k}\\
+    upS_{i+1} \times 2^{k/2} + lowS_{i+1} &= A \times upS_i \times 2^{k/2} + A \times lowS_i + C \mod{2^k} \\
+    lowS_{i+1} &= A \times lowS_i + lowC \mod 2^{k/2}
+\end{align}
+We can see that $lowS_{i+1}$ is independent from $upS_i$.
+That mean that knowing $lowS_i$, we can deduce all $lowS_j$ with $j>i$.
+This relationship is used in the Brute-Force attack with known $C$, in the previous section.
+We can see on (Figure \ref{pcg128goodbits})%figure
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.70\linewidth]{pictures/deducingPCG128.png}
+    \caption{PCG128 : Relevant bits of the outputs}
+    \label{pcg128goodbits}
+\end{figure}
+that we can extract a lot of information by brute forcing certain well chosen bits.
+We defined $known\_up$ (respectively $known\_low$) the number of upper (respectively lower) bits searched by brute force.
+\newpage
+
+\section{Prediction of truncated LCG algorithms}
+
+\subsection{Different solutions for Truncated Linear Congruential Algorithms}
+
+The truncated LCG are not used anymore because there exists well known solutions for predicting their data stream in polynomial time.\\
+
+The first article we found was written by Joan Boyar\cite{Boyar1989}.
+The subject of her study was a special case of truncated LCG, with $A$, $C$ and $M$ unknown and has access to a very large part of the internal state.
+This made it so that we couldn't use her findings, because guessing a large part of the internal state wouldn't be cost-effective.
+However, her study included a very broad state of the art, explaining the specifics of many algorithms, as well as describing in which circumstances they should be used.\\
+
+The algorithm of Frieze \textit{et al}\cite{Frieze} solves the case where $M$ and $A$ are known, $M$ being a power of two and, possibly with a significant portion of the internal state unknown.
+The known part is the upper bits of the state.
+This algorithm is coherent with our problem, this being the reason why we focused on this one in particular.
+However, this algorithm won't work for some particular multipliers.
+But this problem doesn't concern us, because the default multiplier of \texttt{PCG128} is not one of them.\\
+
+\subsection{Frieze \textit{et al} algorithm}
+
+The algorithm we studied uses lattices, and more specifically use the\break Lenstra–Lenstra–Lovász lattice basis reduction algorithm (LLL algorithm)\cite{LLLpastack}.
+A large part of this algorithm is very well explained on Stack Overflow\cite{LLLstack}.
+We will detail the entire algorithm here.\\
+
+A lattice in $\mathbb{R}^n$ is a subgroup of the additive group $(\mathbb{R}^n,+)$, isomorphic to $(\mathbb{Z}^n,+)$.
+The LLL algorithm is an approximation algorithm of the reduction of a lattice basis.
+It returns a reduced basis, with shorter vectors than the ones given.
+These vectors are a good approximation of the basis with the smallest determinant.\\
+
+To use this algorithm, we reduce the sequence of $S_i$ to a geometric sequence $S'$ and, given the relationship between $S$ and $S'$, we can retrieve $S$.
+There are different methods to achieve this result and we selected two of them in our optimised algorithms.
+Consider the following geometric sequence:
+\begin{equation}
+    S'_i=\alpha ^i \times S'_0 \mod{M}
+\end{equation}
+With $S'$ as the vector made of all the $S'_i$ of length $nbiter$, and $A$ blocked, the $S'$ vectors are the vectors of the lattice generated by the matrix :
+
+\begin{equation}
+L=
+\begin{pmatrix} 
+M & 0 & & \dots & 0\\
+\alpha & -1 & 0 & \dots & 0\\
+\alpha ^2 & 0 & -1 & \dots & 0\\
+\vdots & & & \ddots & \vdots\\
+\alpha ^{nbiter - 1} & 0 & 0 & \dots & -1
+\end{pmatrix}
+\end{equation}
+
+%Beau "ell" : $\ell$
+
+Let $Y$ be the most significant bits of the internal states of LCG, and $Z$ be the lower bits, with $S = Y + Z$. We already know $Y$ from the outputs and we are searching for $Z$. 
+\begin{itemize}
+    \item Let $Y'$ be the upper bits of $S'$;
+    \item Let $Z'$ be the lower bits;
+    \item Let $L'$  be the reduced matrix returned by the LLL algorithm for $L$ ($L' = LLL(L)$);
+\end{itemize}
+
+By definition of $L$, we have :
+\begin{align}
+    L \times S' &=0 \mod{M}\\
+    L' \times S' &= 0 \mod{M}\\
+    L' \times (Y' + Z') &= M \times \ell\ \text{($\ell$ unknown integer vector)}\\
+    L' \times Z' &= M \times \ell - L' \times Y'
+\end{align}
+We have $M = 2^k$. L' coefficients are small. We want to find the actual value of $\ell$.\\
+
+We suppose we have every coordinate of $L' \times Z'$ smaller than $M/2$.
+Then every coordinate of $M \times \ell - L' \times Y'$ is smaller than $M/2$ and $M \times \ell$ is the closest vector of $M$ multiples from $L' \times Z'$.
+We have $\ell$ is the closest integer vector from $(B \times y)/M$.\\
+
+The only unknown variable left is $Z'$ and :
+\begin{equation}
+    Z' = L'^{-1} \times (M \times k) - Y' 
+\end{equation}
+Now we know $Z'$, and the integer $S' \mod{M}$, and finally, $S \mod{M}$ the internal state we were looking for.\\
+
+\subsection{Application to our generator}
+The $L$ matrix for our problem is :
+\begin{equation}
+    L=
+\begin{pmatrix}
+1208925819614629174706176 & 0 &  0 &  0 &  0 &  0 &  0 &  0 &  0 & 0\\
+442210214475561085761093 & -1 &  0 &  0 &  0 & 0 &  0 &  0 &  0 &  0\\
+548907270092662740397721 & 0 & -1 &  0 &  0 & 0 &  0 &  0 &  0 &  0\\
+68528166612430993102141 & 0 &  0 & -1 &  0 & 0 &  0 &  0 &  0 &  0\\
+579003058182980434679665 & 0 &  0 &  0 & -1 & 0 &  0 &  0 &  0 &  0\\
+907267876411637163239285 & 0 &  0 &  0 &  0 & -1 &  0 &  0 &  0 &  0\\
+971193525585951128490121 & 0 &  0 &  0 &  0 & 0 & -1 &  0 &  0 &  0\\
+423043601822819931331309 & 0 &  0 &  0 &  0 & 0 &  0 & -1 &  0 &  0\\
+722352710310758831226849 & 0 &  0 &  0 &  0 & 0 &  0 &  0 & -1 &  0\\
+239218985093571541740965 & 0 &  0 &  0 &  0 & 0 &  0 &  0 &  0 & -1
+\end{pmatrix}
+\end{equation}
+Each coefficient is about 128 bit long. Here is the $L'$ (using the LLL algorithm of Sagemath \cite{sagemath}):
+\begin{equation}
+\begin{pmatrix}
+    L' =
+    77 & -105 & -115 &  17 & -108 & -77  & -7 &  50 & 107 &  17\\
+     50  & 80 & 9 & -90 &  138 & -104  & 26  & 90 & -17 & -134\\
+    14 & -42 & 130 & -67 & -31 & -172 & 122 & -76 & -32 &  14\\
+    -61 &  97 &  70 & 161  & -4 & -229 & -23 & -44  & -1 & -26\\
+    118  & 33 & -65 &  -3 &  83 & -176 & -187 & -70 &  25 &  34\\
+    111 & -151 & -73 & 115 &  36 & 200 &  79  & -1 & -14 & -82\\
+    94  & 15 & -18 & -36 & 142 & -47 &  32 & -188 & 161 & -127\\
+    126 & -10 & 129 & 173 & -71  & 61 & -137 &  69 & 7 &  77\\
+    174 & -62 &  58  & 39 &  79 & -133 &  93 & 115 &  24 & 181\\
+    48 & 240 &  -2 & 114 & -15 &  37 & 145 & -100 &  24 & -19\\
+\end{pmatrix}
+\end{equation}
+This algorithm is clearly really efficient, the coefficients of $L'$ are really small.
+Its bigger coefficient is $240$.
+Its smaller than $2^8 = 256$.
+In our case $M = 2^{64 + known\_low}$.
+In our tests, we had $known\_low = 16$.
+We need $Z'$ to be less than 72 bits long, which isn't hard to attain.
+
 \section{Warm-Up: Prediction with Known Increment }
+\subsection{Reduction to a geometric sequence and validation of our values}
+
+$C$ is known.
+We are looking for a geometric sequence $S'$ with which we can validate our brute-forced bits and extract $S_0$.
+Let $polC$ be the vector where $polC_i = \sum_{j=0}^{i} C \times A^j$ and $S'$ where $S'_i = S_i - polC_i$.
+We have :
+\begin{align}
+    S'_{i+1} &= S_{i+1} - polC_{i+1}\\
+    S'_{i+1} &= A \times S_i + C - polC_{i+1}\\
+    S'_{i+1} &= A \times S_i + C - (C + A \times C + ...+ A^{i} \times C)\\
+    S'_{i+1} &= A \times (S'_i + polC_i) - A \times (C + ... + A^{i-1} \times C)\\
+    S'_{i+1} &= A \times S'_i + A \times polC_i - A \times polC_i\\
+    S'_{i+1} &= A \times S'_i
+\end{align}
+
+We have a convenient $S'$ that can be used in the algorithm presented in the previous section and, there exist a simple bijection between $S'$ and $S$.
+However, its important to notice that $S'$ will not be computed modulo $M = 2^k$ but modulo $N = 2^{k/2 + known\_low}$, because we are doing an comprehensive research of the middle bits of $S$ and not the upper ones (except for the 6 bits of $rot$).
+This is not a problem, because when we have the values of $X$, and $lowS$, we can always obtain $upS$ and ultimately $S$.\\
+
+Using this bijection, we can validate if our chosen values are correct.
+Indeed, after finding $Z'$ through the LLL algorithm, we can compute $S'_0$ and $S'_1$ modulo $2^{k/2 + known\_low}$, then $S_0$ and $S_1$ modulo $2^{k/2 + known\_low}$, and finally $S_1$ and $S_0$.
+We verify if $S_1$ is coherent with $S_0$.
+If it is, then we have guessed the correct values for our generator, if not, we then consider this another value as our next test.
+
+\newpage
+
+\subsection{Pseudo-Algorithm}
+
+\begin{algorithm}[h!]
+\caption{Algorithm with known $C$}
+\begin{algorithmic}
+\REQUIRE $X$ the list of outputs
+\STATE computation of the $LLL$ matrix
+\STATE computation of $polC$
+\FOR{ all $rot_i$ for $i$ in range($nbiter$) }
+    \FOR{i in range($nbiter$)}
+        \STATE \texttt{unrotate}($X_i$)
+    \ENDFOR
+    \FOR{all $W_0$}
+        \STATE calculate the other $W_i$ with $W_0$ and $C$
+        \STATE calculate $Y$ using $W_i\ \mathtt{XOR}\ X_i$ and $rot_i\ \mathtt{XOR}\ X_i$
+        \STATE $W' = W - polC$ (we subtract only the corresponding bits of $C$)
+        \STATE $Y' = Y - polC$
+        \STATE calculation of $Z'$ due to LLL algorithm
+        \IF{$2^{k/2-known\_up}>Y'>=2^{known\_low}$}
+            \IF{$Z$ is correct}
+                \STATE return the corresponding $S_0$
+            \ENDIF
+        \ENDIF
+    \ENDFOR
+\ENDFOR
+\end{algorithmic}
+\end{algorithm}
+
+\subsection{Performance test results}
+Our code is implemented in Sagemath.
+This choice was made because of the development advantages it offered: to be able to work with arbitrarily large integers and have access to a built-in LLL function.\\
+
+The number of bits obtained by exhaustive research, as we have seen in the first section, is $nbiter \times known\_up + known\_low$.
+The most efficient way should be to only guess the lower bits ($known\_up = 0$).
+But we need the 6 bit value of $rot_i$ in our algorithm.
+So we kept $known\_up = 6$ and searched for the best equilibrium between $nbiter$ and $known\_low$.
+\\
+
+Moreover, we also need this algorithm to be reliable.
+That is a concern because we know that there is a rounding error caused by an Euclidian Division.
+Depending on the error, it is possible that the algorithm doesn't find the correct $Z$ even if $Y$ and $W$ are correct (with $W$ the lower bits of $S$ and $Y$ the middle bits of $S$ found by \texttt{XOR}).
+\\
+
+We tested the reliability and the efficiency of the algorithm. By trial and error, we ended up finding two interesting settings :
+\begin{itemize}
+    \item $known\_low = 16$ and $nbiter = 4$ : we have a $2^{40}$ brute-force algorithm.
+When supplying it with PCG128 outputs, the rate at which the algorithm detects it and obtains the correct $S_0$ is of 90\%.
+    \item $known\_low = 15$ and $nbiter = 4$ : we have a $2^{39}$ brute-force algorithm.
+When supplying it with PCG128 outputs, the rate at which the algorithm detects it and obtains the correct $S_0$ is of 70\%.
+\end{itemize}
+
+In any case, if we supply a random output, the algorithm will detect this and will never return a false $S_0$.
+When not finding $S_0$, the default answer is to disqualify the input as a PCG data steam.\\
+
+Unfortunately, we didn't get to run our entire solution in a real time environment, because with our personal equipment, and not fully optimized code, we estimated that it would have take more or less ten years to obtain an answer, which is obviously too long.
+The local environment at the university didn't provide the right tools (SageMath), which complexified the development process of sharing the computation load between multiple computers.
+However, there are some methods and coding improvements to reduce the time needed, as we will discuss in the next section.
 

### 2020-05-08 article/main.tex
écrit un bout d'intro et réorganisé des trucs

@@
 
-%Ceux qu'on utilisait dans le projet :
-
-\usepackage{algorithm}
 \usepackage{algorithmic}
@@
 % end of borrowed macros
+
+\newcommand{\todo}[1]{\textcolor{red}{TODO:[#1]}}
 
@@
 
-Today's world has become a data driven world.
-Having access unpredictable data is invaluable when considering security problem.
-For example, security keys have to be generated randomly whether we are talking about RSA keys for secure communication or a cipher key for encrypting data.
-For this purpose, random data can be generated by pseudo-random generators, that rely on a predetermined mathematical structure to output seemingly random data.
-However, by knowing the internal structure and state of the generator, it is possible to reconstruct the entire string of data.\\
+\todo{Ca, c'est du blabla...}
 
-The goal of our project is to study the flaws and try to cryptanalyse pseudo-random generators from the Permuted Congruential Generator (PCG) family, designed by Melissa E.
-O'Neill.\cite{melissapaper} Indeed, predicting a sequence from these generators is said to be "challenging"\cite{melissaweb} by their author, but this statement is not backed by any studies.
-Furthermore, PCG generators are very close to another well known family of generators, the Truncated Linear Congruential Generators, for which a certain amount of viable prediction algorithms are known.
-The combination between being highly commended (it achieved 5th place out of 29 in a recent performance test survey\cite{survey}) and a well presented website\cite{melissaweb} giving an feeling of professionalism can lead to the impression that PCG is a sufficiently secure family of generators.
-Given the lack of studies on the subject, there is a problem if it is used in critical pieces of software.
-This is why we will attempt to find flaws in PCG random generators.\\
+Pseudo-random generators (PRG) are well-studied primitives in symmetric
+cryptography. A PRG is an efficient deterministic algorithm that stretch a small
+random seed into a longer pseudo-random stream. To achieve cryptographic-grade
+pseudo-randomness, a PRG must ensure that the pseudo-random stream is
+computationally indistinguishable from a ``truly'' random sequence of bits by
+efficient adversaries. Alternatively, it is possible to define pseudo-randomness
+by asking that no efficient algorithm is capable of predicting the next
+pseudo-random bit with non-negligible accuracy. The two definitions are in fact
+equivalent.
 
-Some attacks against PCG have been carried out by Sebastiano Vigna\cite{vignaweb}, but this didn't yield any significant results, given that he only tried to break an incomplete version in a fairly straightforward way.
-He worked on a very specific version of PCG.
-He considers the result without modulo, which is a big simplification.
-Moreover, he didn't study the 128 bits long generator, but only the 64 bits long\cite{vignacode}.\\
+\todo{mentioner stream ciphers}
 
-We focused on the general version of PCG, which happens to be very similar to the generator implemented by the function \texttt{random} of the C library.
-Both are based on a well known family of generators, the Linear Congruential Generators (LCG), but with the PCG one also using a non linear function and a rotation on the output.\\
+Not all pseudo-random generators are of cryptographic strength. In some
+applications, it is not necessarily necessary: to be used in Monte-Carlo
+simulations or generate random choices in games, a relaxed, non-cryptographic
+notion of pseudo-randomness may be sufficient. This allows for faster
+algorithms. For instance, the \texttt{Math.Random} function in Google's V8
+open-source JavaScript engine uses the \textsf{XorShift128} generator, which is
+a Linear Feedback Shift Register tailored for high software speed. On 64-bit
+processors, it produces 64 pseudo-random bits with 3 shifts and 4 XORs
+instructions. The \textsf{python} standard library's \texttt{random} module uses
+the Mersenne Twister. The \textsf{C} library that comes along \texttt{gcc} (the
+\texttt{glibc}) uses a (poor) truncated linear congruential generator by default
+to implement the \texttt{rand} function.
 
-We are looking for a possible way to reduce this generator to an LCG, or a truncated LCG, because there are known algorithms to predict them.
-Because we aren't studying the same exact version as Sebastiano Vigna, his findings aren't very relevant to us.\\
+In the realm of non-cryptographic random generators, a PRG is deemed ``good
+enough'' it is passes \emph{some} efficient statistical tests --- whereas the
+cryptographic notion of pseudo-randomness asks that it passes \emph{any}
+efficient test. There are \textit{de facto} statistical test suites (\todo{les
+  mentionner}). The goal of designers then consists in designing the fastest
+possible generator that passes the day's favorite test suite.
 
-We decided to study the default version of the largest available generator of the 0.94 version of the PCG family, which has an internal state of 128 bits, named PCG128.
-We will mathematically analyse the algorithm to uncover weaknesses, and if methods for the LCG family can be translated to the PCG family.
+\todo{Mentionner Ferrenberg 92}
+
+The scientific computing community also realized that the need for fast
+\emph{parallel} random number generation could be satisfied by the use of block
+ciphers in counter mode~\cite{Salmon11}. The need for speed then leads to the
+use of weakened cryptographic primitives.
+
+In most cases, it is easy to see that a non-cryptographic PRG does not meet the
+cryptographic notion of pseudo-randomness, and there are few exceptions. In this
+paper, we study the \textsf{PCG} family of non-cryptographic pseudo-random
+generators~\cite{melissapaper,melissaweb}.
+
+\textsf{PCG} stands for ``Permuted Congruential Generator'': it essentially
+consists in applying a non-linear filtering function on top of a (fairly weak)
+linear congruential generator (in a way reminiscent to the venerable filtered
+LFSRs). The resulting combination is fast and passes current test suites. Its
+designer claimed that distinguishing the pseudo-random stream produced would be
+``challenging''. The \textsf{PCG} family contains many members, but we focus on
+its strongest member, named either \textsf{PCG64} or \textsf{PCG-XSL-RR}. It has
+a 128-bit internal state and produces 64 bits when clocked. It is the default
+pseudo-random number generator in the popular \textsf{NumPy} scientific
+computing package for \textsf{Python}.
+
+The internal state of the \textsf{PCG64} generator is made of a 128-bit
+``state'' and a 128-bit ``increment'', whose intended use is to provide several
+pseudo-random streams with the same seed (just as the initialization vectors do
+in stream ciphers). A default increment is provided in case the end-user just
+want one pseudo-random stream with a single 128-bit seed.
+
+\paragraph{Contribution.} We describe an algorithm that reconstructs the full
+internal state of the strongest member of the \textsf{PCG} family. This allows
+to predict the pseudo-random stream deterministically and clock the generator
+backwards. The original seeds can also easily be reconstructed. The algorithm is
+practical and we have executed it in practice. It follows that predicting the
+output of the \textsf{PCG} should not be considered challenging anymore.
+
+Our algorithm reconstruct the internal state using the ``guess-and-determine''
+technique: some bits of the internal state are guessed ; assuming the guesses
+are correct, some other information is computed ; a consistency check discards
+bad guesses early on ; then candidate internal states are computed and fully
+tested. The problem actually come in two distinct flavors.
+
+When the increment is known (for instance when it is the default value), a
+simplified prediction algorithm recovers the internal state from 192 bits of
+pseudo-random stream. The process runs in 20 CPU minutes (on a single core of a
+server processor). It guesses 36 bits of the internal state, then solves an
+instance of the \textsc{Closest Vector Problem} (CVP) in a 3-dimensional
+euclidean lattice. This requires about 50 arithmetic operations in total and
+reveals the entire internal state if the guesses are correct.
+
+When the increment is unknown, things are a bit more complicated. This is the
+default situation in \textsf{NumPy}, where both the state and the increment are
+initialized using an external source of entropy. In this case, our prediction
+algorithm requires 3072 bits of pseudo-random stream ; it guesses between 51 and
+55 bits, then for each guess it solves and instance of CVP in dimension 4 (using
+about 75 arithmetic operations). This recovers 64 more bits of information about
+the difference between two successive states, and this is enough to filter the
+bad guesses. This information can then be used in a subsequent and comparably
+inexpensive phase to recover the entire internal state. On average, the whole
+process requires a bit less than 20 000 CPU hours to complete.
+
+We implemented the algorithms in \textsf{C}, then asked the designer of the PCG
+family to send us ``challenge'' pseudo-random streams ; we ran our code and
+emailed back the seeds used to generate the challenge streams the next day.
+
+\paragraph{Related Work.} \todo{Mentionner Knuth et les LCG + Frieze + Boyar + Stern...}
+
+\todo{mentionner Vigna ?}
+
+% The combination between being highly commended (it achieved 5th place out of 29
+% in a recent performance test survey\cite{survey}) and a well presented
+% website\cite{melissaweb} giving an feeling of professionalism can lead to the
+% impression that PCG is a sufficiently secure family of generators.
+
+% Some attacks against PCG have been carried out by Sebastiano
+% Vigna\cite{vignaweb}, but this didn't yield any significant results, given that
+% he only tried to break an incomplete version in a fairly straightforward way. He
+% worked on a very specific version of PCG. He considers the result without
+% modulo, which is a big simplification. Moreover, he didn't study the 128 bits
+% long generator, but only the 64 bits long\cite{vignacode}.
 
 \section{The PCG Pseudo-Random Number Generator Family}
+
+\todo{cette section doit décrire précisément PCG.}
+
+\begin{quote}
+  In addition, one of the generators, PCG-XSL-RR(described in Section 6.3.3), is
+  explicitly designed to make any attempt at state reconstruction especially
+  difficult, using xor folding tominimize the amount of information about
+  internal state that leaks out.
+\end{quote}
+
+\begin{quote}
+  Although these properties make it extremely challenging for an agent observing
+  theexternal behavior of a program to guess the state of its random number
+  generator,
+\end{quote}
+
+% import numpy
+% print(numpy.random.default_rng())
+% ---> Generator(PCG64)
 
@@
 
+\section{Naive Brute-Force Reconstruction of the Internal State}
+
+In this section we describe the simplest possible prediction algorithm we could
+think of. They serve as ``baseline'' to evaluate the more refined predictors
+described below.
+
+\todo{Dans le cas où C est connu, c'est le truc de Vigna, non ?}
+
 \subsection{Brute-Force Attack with known $C$}
-In the case where $C$ is known, if we find $S_0$, we can calculate every $S_i$ with the given formula, and predict all the following outputs of the generator.
-The naive way is by doing a comprehensive research of $upS_0$, and then calculating $lowS_0$ with the formula :\\
+
+In the case where $C$ is known, if we find $S_0$, we can calculate every $S_i$
+with the given formula, and predict all the following outputs of the
+generator. The naive way is by doing a comprehensive research of $upS_0$, and
+then calculating $lowS_0$ with the formula :
 \begin{equation}
@@
 \end{equation}
-\texttt with {unrotate} being the inverse function of \texttt{rotate}.\\
+with \texttt {unrotate} being the inverse function of \texttt{rotate}.
 
-If our guessing of $upS_0$ is correct, we can obtain the correct $X_1$ with $S_1$ deduced from the initial $S_0$.
-The preceding is a naive method for predicting PCG generator in $2^{64}$ operations.
-But we can have other values that verify that test.
-This will not return $S_0$ but a very short list of possibles $S_0$, with the actual one in it.
-We just need to test the following outputs to find the real $S_0$.
-This is yet a very slow way to guess the internal state, so we need a better solution.
+If our guessing of $upS_0$ is correct, we can obtain the correct $X_1$ with
+$S_1$ deduced from the initial $S_0$. The preceding is a naive method for
+predicting PCG generator in $2^{64}$ operations. But we can have other values
+that verify that test. This will not return $S_0$ but a very short list of
+possibles $S_0$, with the actual one in it. We just need to test the following
+outputs to find the real $S_0$. This is yet a very slow way to guess the
+internal state, so we need a better solution.
 
@@
 
-In case $C$ is unknown, retrieving $S_0$ is harder. The naive algorithm gives us an answer within $2^{128+6}$ simple operations whit the following algorithm:
-\\
+In case $C$ is unknown, retrieving $S_0$ is harder. The naive algorithm gives us
+an answer within $2^{128+6}$ simple operations whit the following algorithm:
+
 \begin{algorithmic}[!h]
@@
 
-This algorithm reveals us $C$ and a list of outputs with $S_0$ in it.
-The entry is 256 bits long, and we used 3 64 bits long outputs of the generator.
-There will be about $2^{64}$ possible $S_0$.
-This is why we need a couple other outputs to get the real $S_0$.
-Sorting out which is the correct $S_0$ can be done afterwards, and won't affect the complexity.
-But given the complexity of this algorithm, it is even more unrealistic to solve problem in this fashion then the previous one where $C$ was known.
+This algorithm reveals us $C$ and a list of outputs with $S_0$ in it. The entry
+is 256 bits long, and we used 3 64 bits long outputs of the generator. There
+will be about $2^{64}$ possible $S_0$. This is why we need a couple other
+outputs to get the real $S_0$. Sorting out which is the correct $S_0$ can be
+done afterwards, and won't affect the complexity. But given the complexity of
+this algorithm, it is even more unrealistic to solve problem in this fashion
+then the previous one where $C$ was known.
 
-\subsection{Reduction to a truncated LCG model}
-By exploiting the nature of this generator, it becomes apparent that some values can be easily guessed.
-We already saw that if we guess $rot_i$, then retrieving $X_i$ becomes a simple \texttt{XOR} operation, and in turn, if we have some knowledge of a particular bit of $lowS_i$ (respectively $upS_i$), we can deduce the equivalent bit from $upS_i$ (respectively $lowS_i$).\\
+\section{Predicting Truncated Linear Congruential Generators}
 
-Moreover, there is an interesting proprity concerning the least significant bits of the internal state.
-Indeed, if we define $lowC$ as the 64 least significant bits of $C$ ($lowC = C \mod{2^{k/2}})$, we have:
+\todo{Comme on utilise à plusieurs reprise l'idée de se ramener à un LCG
+  tronqué, autant décrire la technique à base de réseaux un bon coup.}
 
-\begin{align}
-    S_{i+1} &= A \times S_i + C \mod{2^k}\\
-    upS_{i+1} \times 2^{k/2} + lowS_{i+1} &= A \times upS_i \times 2^{k/2} + A \times lowS_i + C \mod{2^k} \\
-    lowS_{i+1} &= A \times lowS_i + lowC \mod 2^{k/2}
-\end{align}
-We can see that $lowS_{i+1}$ is independent from $upS_i$.
-That mean that knowing $lowS_i$, we can deduce all $lowS_j$ with $j>i$.
-This relationship is used in the Brute-Force attack with known $C$, in the previous section.
-We can see on (Figure \ref{pcg128goodbits})%figure
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.70\linewidth]{pictures/deducingPCG128.png}
-    \caption{PCG128 : Relevant bits of the outputs}
-    \label{pcg128goodbits}
-\end{figure}
-that we can extract a lot of information by brute forcing certain well chosen bits.
-We defined $known\_up$ (respectively $known\_low$) the number of upper (respectively lower) bits searched by brute force.
-\newpage
+The algorithm of Frieze \textit{et al}\cite{Frieze} solves the case where $M$
+and $A$ are known, $M$ being a power of two and, possibly with a significant
+portion of the internal state unknown. The known part is the upper bits of the
+state. This algorithm is coherent with our problem, this being the reason why we
+focused on this one in particular. However, this algorithm won't work for some
+particular multipliers. But this problem doesn't concern us, because the default
+multiplier of \texttt{PCG128} is not one of them.
 
-\section{Prediction of truncated LCG algorithms}
+The first article we found was written by Joan Boyar\cite{Boyar1989}. The
+subject of her study was a special case of truncated LCG, with $A$, $C$ and $M$
+unknown and has access to a very large part of the internal state. This made it
+so that we couldn't use her findings, because guessing a large part of the
+internal state wouldn't be cost-effective. However, her study included a very
+broad state of the art, explaining the specifics of many algorithms, as well as
+describing in which circumstances they should be used.
 
-\subsection{Different solutions for Truncated Linear Congruential Algorithms}
-
-The truncated LCG are not used anymore because there exists well known solutions for predicting their data stream in polynomial time.\\
-
-The first article we found was written by Joan Boyar\cite{Boyar1989}.
-The subject of her study was a special case of truncated LCG, with $A$, $C$ and $M$ unknown and has access to a very large part of the internal state.
-This made it so that we couldn't use her findings, because guessing a large part of the internal state wouldn't be cost-effective.
-However, her study included a very broad state of the art, explaining the specifics of many algorithms, as well as describing in which circumstances they should be used.\\
-
-The algorithm of Frieze \textit{et al}\cite{Frieze} solves the case where $M$ and $A$ are known, $M$ being a power of two and, possibly with a significant portion of the internal state unknown.
-The known part is the upper bits of the state.
-This algorithm is coherent with our problem, this being the reason why we focused on this one in particular.
-However, this algorithm won't work for some particular multipliers.
-But this problem doesn't concern us, because the default multiplier of \texttt{PCG128} is not one of them.\\
+%\subsection{Different solutions for Truncated Linear Congruential Algorithms}
 
@@
 
-\section{Warm-Up: Prediction with Known Increment }
+\section{Predicting PCG with Known Increment}
+\subsection{Reduction to a truncated LCG model}
+By exploiting the nature of this generator, it becomes apparent that some values can be easily guessed.
+We already saw that if we guess $rot_i$, then retrieving $X_i$ becomes a simple \texttt{XOR} operation, and in turn, if we have some knowledge of a particular bit of $lowS_i$ (respectively $upS_i$), we can deduce the equivalent bit from $upS_i$ (respectively $lowS_i$).\\
+
+Moreover, there is an interesting proprity concerning the least significant bits of the internal state.
+Indeed, if we define $lowC$ as the 64 least significant bits of $C$ ($lowC = C \mod{2^{k/2}})$, we have:
+
+\begin{align}
+    S_{i+1} &= A \times S_i + C \mod{2^k}\\
+    upS_{i+1} \times 2^{k/2} + lowS_{i+1} &= A \times upS_i \times 2^{k/2} + A \times lowS_i + C \mod{2^k} \\
+    lowS_{i+1} &= A \times lowS_i + lowC \mod 2^{k/2}
+\end{align}
+We can see that $lowS_{i+1}$ is independent from $upS_i$.
+That mean that knowing $lowS_i$, we can deduce all $lowS_j$ with $j>i$.
+This relationship is used in the Brute-Force attack with known $C$, in the previous section.
+We can see on (Figure \ref{pcg128goodbits})%figure
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.70\linewidth]{pictures/deducingPCG128.png}
+    \caption{PCG128 : Relevant bits of the outputs}
+    \label{pcg128goodbits}
+\end{figure}
+that we can extract a lot of information by brute forcing certain well chosen bits.
+We defined $known\_up$ (respectively $known\_low$) the number of upper (respectively lower) bits searched by brute force.
+
 \subsection{Reduction to a geometric sequence and validation of our values}
@@
 
+[insert here] attack predicting algo description.
+
 \begin{theorem}
@@
 
-\section{Prediction with Unknown Increment}
+\section{Predicting PCG with Unknown Increment}
+
+[insert here] attack predicting algo description. Phase 1, Phase 2, Phase 3
 

### 2020-05-09 C/fonctions.c
speedup

@@ -56,27 +56,35 @@ double wtime()
 
 ////////////////Fonctions pour la récupération de S//////////////
 
+static u64 rot(u64 x, int i)
+{
+        return (x >> (64 - i)) | (x << i);
+}
 
-void rotate(unsigned long long* rX, unsigned long long* X,int* rot){ //pas verifié, repris de pcg_random
+void rotate(unsigned long long* rX, const unsigned long long* X, const int* rot)
+{ //pas verifié, repris de pcg_random
     for(int i = 0 ; i < nbiter ; i++)
         rX[i]= (X[i] >> rot[i]) | (X[i] << ((- rot[i]) & 63));
 }
 
-void unrotate(unsigned long long* urX, unsigned long long* X, int* rot){//pas verifié, repris de pcg_random
+void unrotate(u64* urX, const u64* X, const int* rot)
+{//pas verifié, repris de pcg_random
     int rot2[nbiter];
     for( int i = 0 ; i < nbiter ; i++)
         rot2[i] = (k - rot[i]) % k;
     rotate(urX, X, rot2);
 }
 
-void getPolW(pcg128_t *polW, unsigned long long W0){ //OK !
+void getPolW(pcg128_t *polW, u64 W0)
+{ //OK !
     polW[0] = W0;
     for(int i = 1 ; i < nbiter ; i++){
         polW[i] = polW[i-1] * a;
     }
 }
 
-void getSumPol(unsigned long long* sumPol,unsigned long long* sumPolY, pcg128_t* polW){
+void getSumPol(u64* sumPol, u64* sumPolY, const pcg128_t* polW)
+{
     pcg128_t sum;
     for(int i = 0 ; i < nbiter ; i++){
         sum = polC[i] + polW[i];
@@ -84,7 +92,6 @@ void getSumPol(unsigned long long* sumPol,unsigned long long* sumPolY, pcg128_t*
         sumPolY[i] = (sum >> (k - known_up)) % (1 << (known_low + known_up));
     }
 }
-
 
 /* cf. https://stackoverflow.com/questions/17035464/a-fast-method-to-round-a-double-to-a-32-bit-int-explained#comment61972557_17035583 */
 static inline long long crazy_round(double x)
@@ -106,47 +113,76 @@ static inline long long light_crazy_round(double x)
 }
 
 
-/* sumPol/sumPolY are constant over many iterations. X (unrotated) and rot vary each time. */
[…]

### 2020-05-09 C/fonctions.h

@@ -1,5 +1,8 @@
 #include <stdint.h>
+#include <stdbool.h>
+
 #include "pcg_oneseq.h"
+
 #include <omp.h>
 
 
@@ -15,20 +18,28 @@ pcg128_t a_inv;
 pcg128_t c;
 pcg128_t polC[nbiter];
 
-// extern unsigned long long Greduite[9];
-// extern double invG[9];
+
+typedef unsigned long long u64;
+
+struct task_t {
+        u64 sumPol[nbiter];
+        u64 sumPolY[nbiter];
+        u64 X[nbiter][64];
+        double Yprime[nbiter][64];
+};
+
 
 /***** Fonctions *****/
 void init_var_globales();
 double wtime();
 
-
 ////////////////Fonctions pour la récupération de S//////////////
-void rotate(unsigned long long* rX, unsigned long long* X,int* rot);
-void unrotate(unsigned long long* urX, unsigned long long* X, int* rot);
-void getPolW(pcg128_t *polW, unsigned long long W0);
+void rotate(unsigned long long* rX, const unsigned long long* X, const int* rot);
+void unrotate(u64* urX, const u64* X, const int* rot);
+void getPolW(pcg128_t *polW, u64 W0);
+void getSumPol(u64* sumPol, u64* sumPolY, const pcg128_t* polW);
 
-void getSumPol(unsigned long long* sumPol,unsigned long long* sumPolY, pcg128_t* polW);
+void setup_task(u64 W0, const u64 *X, struct task_t *task);
 
-int solve(pcg128_t* S, const unsigned long long* X, const int* rot, const unsigned long long* sumPol, const unsigned long long* sumPolY);
 
+bool solve(pcg128_t* S, const int* rot, const struct task_t *task);

### 2020-05-09 C/main.c

@@ -3,7 +3,21 @@
 #include <stdio.h>
 #include <time.h>
 
-// Ce truc marche avec GCC mais pas avec ICC. Il faut investiguer...
+// avec ICC, ce truc ne marche pas sur ppti-gpu-1 si on met -xHost (zmm?). Il faut investiguer...
+
+void result_found(pcg128_t S, double start)
+{
+	printf("\nInternal state found (%.1fs)\n", wtime() - start);
+	
+	struct pcg_state_128 rng;
+	rng.state = (S - c) * a_inv;
+	pcg128_t seed = (rng.state - c - a*c) * a_inv;
+
+	printf("Seed : %016" PRIx64 " %016" PRIx64 "\n", (uint64_t) (seed >> 64), (uint64_t) seed);
+	for(int i = 0 ; i < 16 ; i++)
+		printf("X[%2d] = %016" PRIx64 "\n", i, pcg_oneseq_128_xsl_rr_64_random_r(&rng));
+}
+
 
 int main()
 {
@@ -11,65 +25,52 @@ int main()
     /*  INITIALISATION DES PARAMETRES  */
     
     init_var_globales();
-        
+	
     /********** INPUT ***********/
     unsigned long long X[nbiter];
-//     X[ 0] = 0x3c8673c55ead80bc;
-//     X[ 1] = 0xfd27cfa6f84ef5f1;
-//     X[ 2] = 0x68a098da45558d92;
-    // valeurs fournies par M. O'Neil.
+
+    // challenge output given by M. O'Neil
     X[ 0] = 0x6ec191a37a421087;
     X[ 1] = 0xec140ace169176fc;
     X[ 2] = 0x85994d489913af70;
 
-    double t1 = wtime();
+    double start = wtime();
     unsigned long long done = 0;
     
     #pragma omp parallel for
-    for (unsigned long long W0 = 0; W0 < (1<<known_low) ; W0++){//(1<<known_low)//W0=209818  
-        
-        /*Variables privées*/
-        pcg128_t S[nbiter];
-        pcg128_t polW[nbiter];
-        unsigned long long urX[nbiter];
-        int rot[nbiter] = {0, 0, 0};
-        unsigned long long sumPol[nbiter];
-        unsigned long long sumPolY[nbiter];
-        
-        getPolW(polW, W0);
-        getSumPol(sumPol, sumPolY, polW);
-        
-        if (omp_get_thread_num() == 0 && (W0 % 64) == 0) {
[…]

### 2020-05-09 C/fonctions.c
C known : + de précalcul

@@ -116,44 +116,30 @@ static inline long long light_crazy_round(double x)
 void setup_task(u64 W0, const u64 *urX, struct task_t *task)
 {
         pcg128_t polW[nbiter]; // temporary byproduct
+        u64 sumPol[nbiter]; // temporary byproduct
+        u64 sumPolY[nbiter]; // temporary byproduct
         
         // initialise polW à partir de W0
         getPolW(polW, W0);
 
         // initialise sumPol et sumPolY à partir de polW
-        getSumPol(task->sumPol, task->sumPolY, polW);
+        getSumPol(sumPol, sumPolY, polW);
 
         for (int i = 0; i < nbiter; i++)
                 for (int j = 0; j < 64; j++) {
                         u64 X = rot(urX[i], j);
-                        u64 Y = (((task->sumPol[i] ^ X) % (1 << known_low)) << 6) + (j ^ (X >> 58));
-                        u64 Yprime = (Y - task->sumPolY[i]) % (1 << (known_low + 6));
+                        u64 Y = (((sumPol[i] ^ X) % (1 << known_low)) << 6) + (j ^ (X >> 58));
+                        u64 Yprime = (Y - sumPolY[i]) % (1 << (known_low + 6));
                         
                         task->X[i][j] = X;
                         task->Yprime[i][j] = (double) Yprime; // conversion en double
                 }
+        task->sumPol0 = sumPol[0];
 }
 
 
-/* sumPol/sumPolY are constant over many iterations. 
-   X ("de-rotated") and rot vary each time. 
-   returns true when a consistent solution is found. In that case S is set to the correct internal state.
-   */
 bool solve(pcg128_t* S, const int* rot, const struct task_t *task)
-{
-        // pour chaque 0 <= j < nbiter, 0 <= i < 64, X[i] = T[i, rot[i]]
-        // par conséquent, (rot[i] ^ (X[i] >> 58)) == T'[i, rot[i]]
-        // Yprim[i] == T[i, rot[i]]
-
-    // total : 6 XOR, 6 AND, 6 SHIFT, 3 ADD, 3 SUB, 3 u64->double
-    //u64 Y[nbiter];
-    // double Yprim[nbiter];
-    // for (int i = 0; i < nbiter; i++) {
-    //     u64 X = task->X[i][rot[i]];
[…]

### 2020-05-09 C/fonctions.h

@@ -22,10 +22,11 @@ pcg128_t polC[nbiter];
 typedef unsigned long long u64;
 
 struct task_t {
-        u64 sumPol[nbiter];
-        u64 sumPolY[nbiter];
+        // u64 sumPol[nbiter];
+        // u64 sumPolY[nbiter];
         u64 X[nbiter][64];
         double Yprime[nbiter][64];
+        u64 sumPol0;
 };
 
 

### 2020-05-09 C/fonctions.c
checkpoint pending MORE precomputation

@@ -140,24 +140,14 @@ void setup_task(u64 W0, const u64 *urX, struct task_t *task)
 
 bool solve(pcg128_t* S, const int* rot, const struct task_t *task)
 {  
-    double tmp2[nbiter];
     // total : 9 MUL (double) + 10 ADD (double)
-    for (int i = 0; i < nbiter; i++) {
-        tmp2[i] = 0;
-        for(int j = 0; j < nbiter; j++)
-            tmp2[i] += invG[i * nbiter + j] * task->Yprime[j][rot[j]];
-        tmp2[i] += 6755399441055744.0;
-    }
- 
-    // total : 6 SHIFT
-    u64 tmp3[nbiter];
-    for(int i = 0 ; i < nbiter ; i++)
-        tmp3[i] = light_crazy_round(tmp2[i]);
-
     u64 Sprim0 = 0;
-    // total : 3 ADD, 3 MUL
-    for(int j=0 ; j<nbiter ; j++)
-        Sprim0 += Greduite[j] * tmp3[j];
+    for (int i = 0; i < nbiter; i++) {
+        double tmp2 = 0;
+        for(int j = 0; j < nbiter; j++)
+            tmp2 += invG[i * nbiter + j] * task->Yprime[j][rot[j]];
+        Sprim0 += Greduite[i] * light_crazy_round(tmp2 + 6755399441055744.0);
+    }
 
     // SHIFT, ADD
     u64 Smod = (Sprim0 << known_low) + task->sumPol0;

### 2020-05-09 C/fonctions.h

@@ -22,8 +22,6 @@ pcg128_t polC[nbiter];
 typedef unsigned long long u64;
 
 struct task_t {
-        // u64 sumPol[nbiter];
-        // u64 sumPolY[nbiter];
         u64 X[nbiter][64];
         double Yprime[nbiter][64];
         u64 sumPol0;
@@ -41,6 +39,4 @@ void getPolW(pcg128_t *polW, u64 W0);
 void getSumPol(u64* sumPol, u64* sumPolY, const pcg128_t* polW);
 
 void setup_task(u64 W0, const u64 *X, struct task_t *task);
-
-
 bool solve(pcg128_t* S, const int* rot, const struct task_t *task);

### 2020-05-09 C/main.c

@@ -27,7 +27,7 @@ int main()
     init_var_globales();
 	
     /********** INPUT ***********/
-    unsigned long long X[nbiter];
+    u64 X[nbiter];
 
     // challenge output given by M. O'Neil
     X[ 0] = 0x6ec191a37a421087;
@@ -35,14 +35,13 @@ int main()
     X[ 2] = 0x85994d489913af70;
 
     double start = wtime();
-    unsigned long long done = 0;
+    u64 done = 0;
     
     #pragma omp parallel for
-    for (unsigned long long W0 = 0; W0 < (1<<known_low) ; W0++){
+    for (u64 W0 = 0; W0 < (1<<known_low) ; W0++){
 	
 	/*Variables privées*/
 	pcg128_t S[nbiter];
-	// u64 urX[nbiter];
 	int rot[nbiter] = {0, 0, 0};
 	struct task_t task;
 	setup_task(W0, X, &task);
@@ -53,7 +52,7 @@ int main()
 	}
 
 
-	for (int r = 0 ; r < 1<<(3*known_up) ; r++) {
+	for (int r = 0 ; r < 1 << (nbiter*known_up) ; r++) {
 	    /***** Modification de rot et unrotX *****/
 	    rot[0] = (rot[0] + 1) % k;
 	    int i = 0;
@@ -61,7 +60,6 @@ int main()
 		i++;
 		rot[i] = (rot[i] + 1) %k;
 	    }
-	    // unrotate(urX, X, rot);
 	    
 	    /***** Résolution *****/
 	    if (solve(S, rot, &task))

### 2020-05-09 C/fonctions.c
speed-up !

@@ -22,12 +22,12 @@ double invG[9] = {
 
 void init_var_globales(){
     //multiplier a OK !
-    a = (((pcg128_t) 2549297995355413924) << k) + ((pcg128_t) 4865540595714422341);
+    a = (((pcg128_t) 2549297995355413924) << 64) + ((pcg128_t) 4865540595714422341);
     
-    a_inv = (((pcg128_t) 566787436162029664) << k) + ((pcg128_t) 11001107174925446285ull);
+    a_inv = (((pcg128_t) 566787436162029664) << 64) + ((pcg128_t) 11001107174925446285ull);
 
     //increment c OK !
-    c = (((pcg128_t)6364136223846793005) << k) + 1442695040888963407;
+    c = (((pcg128_t)6364136223846793005) << 64) + 1442695040888963407;
     
     //nombre de threads 
     nb_thread = omp_get_max_threads();
@@ -41,7 +41,7 @@ void init_var_globales(){
     }
 
     for (int i = 0; i < 9; i++)
-        invG[i] *= 1ll << (k - known_low - known_up);
+        invG[i] *= 1ll << (64 - known_low - 6);
 }
 
 
@@ -61,19 +61,6 @@ static u64 rot(u64 x, int i)
         return (x >> (64 - i)) | (x << i);
 }
 
-void rotate(unsigned long long* rX, const unsigned long long* X, const int* rot)
-{ //pas verifié, repris de pcg_random
-    for(int i = 0 ; i < nbiter ; i++)
-        rX[i]= (X[i] >> rot[i]) | (X[i] << ((- rot[i]) & 63));
-}
-
-void unrotate(u64* urX, const u64* X, const int* rot)
-{//pas verifié, repris de pcg_random
-    int rot2[nbiter];
-    for( int i = 0 ; i < nbiter ; i++)
-        rot2[i] = (k - rot[i]) % k;
-    rotate(urX, X, rot2);
-}
 
 void getPolW(pcg128_t *polW, u64 W0)
 { //OK !
@@ -89,7 +76,7 @@ void getSumPol(u64* sumPol, u64* sumPolY, const pcg128_t* polW)
     for(int i = 0 ; i < nbiter ; i++){
         sum = polC[i] + polW[i];
         sumPol[i] = sum;
-        sumPolY[i] = (sum >> (k - known_up)) % (1 << (known_low + known_up));
[…]

### 2020-05-09 C/fonctions.h

@@ -7,7 +7,7 @@
 
 
 /***** Macro et Variables globales *****/
-#define k 64
+// #define k 64
 #define known_up 6
 #define known_low 18
 #define nbiter 3
@@ -23,7 +23,7 @@ typedef unsigned long long u64;
 
 struct task_t {
         u64 X[nbiter][64];
-        double Yprime[nbiter][64];
+        double Ginv_Yprime[nbiter][nbiter][64];
         u64 sumPol0;
 };
 
@@ -33,8 +33,6 @@ void init_var_globales();
 double wtime();
 
 ////////////////Fonctions pour la récupération de S//////////////
-void rotate(unsigned long long* rX, const unsigned long long* X, const int* rot);
-void unrotate(u64* urX, const u64* X, const int* rot);
 void getPolW(pcg128_t *polW, u64 W0);
 void getSumPol(u64* sumPol, u64* sumPolY, const pcg128_t* polW);
 

### 2020-05-09 C/main.c

@@ -52,13 +52,13 @@ int main()
 	}
 
 
-	for (int r = 0 ; r < 1 << (nbiter*known_up) ; r++) {
+	for (int r = 0 ; r < 1 << (nbiter*6) ; r++) {
 	    /***** Modification de rot et unrotX *****/
-	    rot[0] = (rot[0] + 1) % k;
+	    rot[0] = (rot[0] + 1) % 64;
 	    int i = 0;
 	    while (rot[i] == 0 && i < nbiter) {
 		i++;
-		rot[i] = (rot[i] + 1) %k;
+		rot[i] = (rot[i] + 1) % 64;
 	    }
 	    
 	    /***** Résolution *****/

### 2020-05-09 C/fonctions.c
more fun, more precomputation, more speed

@@ -81,15 +81,6 @@ void getSumPol(u64* sumPol, u64* sumPolY, const pcg128_t* polW)
 }
 
 /* cf. https://stackoverflow.com/questions/17035464/a-fast-method-to-round-a-double-to-a-32-bit-int-explained#comment61972557_17035583 */
-static inline long long crazy_round(double x)
-{
-    union { double d; long long l; } magic; 
-    magic.d = x + 6755399441055744.0; 
-    magic.l <<= 13; 
-    magic.l >>= 13;
-    return magic.l;
-}
-
 static inline long long light_crazy_round(double x)
 {
     union { double d; long long l; } magic; 
@@ -97,6 +88,11 @@ static inline long long light_crazy_round(double x)
     magic.l <<= 13; 
     magic.l >>= 13;
     return magic.l;
+}
+
+static inline long long crazy_round(double x)
+{
+    return light_crazy_round(x + 6755399441055744.0);
 }
 
 
@@ -129,21 +125,26 @@ void setup_task(u64 W0, const u64 *urX, struct task_t *task)
         task->sumPol0 = sumPol[0];
 }
 
+void refresh_task(const int *rot, struct task_t *task)
+{
+    for (int i = 0; i < nbiter; i++) {
+        task->tmp2[i] = 0;
+        for(int j = 1; j < nbiter; j++)
+            task->tmp2[i] += task->Ginv_Yprime[i][j][rot[j]];
+    }
+}
 
 bool solve(pcg128_t* S, const int* rot, const struct task_t *task)
 {  
-    // total : 9 MUL (double) + 10 ADD (double)
     u64 Sprim0 = 0;
     for (int i = 0; i < nbiter; i++) {
-        double tmp2 = 0;
-        for(int j = 0; j < nbiter; j++)
-            tmp2 += task->Ginv_Yprime[i][j][rot[j]];
-        Sprim0 += Greduite[i] * light_crazy_round(tmp2 + 6755399441055744.0);
+        double tmp2 = task->tmp2[i] + task->Ginv_Yprime[i][0][rot[0]];
+        Sprim0 += Greduite[i] * crazy_round(tmp2);
     }
 
     // SHIFT, ADD
     u64 Smod = (Sprim0 << known_low) + task->sumPol0;
-    S[0] = (((pcg128_t)(Smod ^ task->X[0][rot[0]])) << 64) + ((pcg128_t) Smod);
[…]

### 2020-05-09 C/fonctions.h

@@ -25,6 +25,7 @@ struct task_t {
         u64 X[nbiter][64];
         double Ginv_Yprime[nbiter][nbiter][64];
         u64 sumPol0;
+        double tmp2[nbiter];
 };
 
 
@@ -33,8 +34,9 @@ void init_var_globales();
 double wtime();
 
 ////////////////Fonctions pour la récupération de S//////////////
-void getPolW(pcg128_t *polW, u64 W0);
-void getSumPol(u64* sumPol, u64* sumPolY, const pcg128_t* polW);
+// void getPolW(pcg128_t *polW, u64 W0);
+// void getSumPol(u64* sumPol, u64* sumPolY, const pcg128_t* polW);
 
 void setup_task(u64 W0, const u64 *X, struct task_t *task);
+void refresh_task(const int *rot, struct task_t *task);
 bool solve(pcg128_t* S, const int* rot, const struct task_t *task);

### 2020-05-09 C/main.c

@@ -60,7 +60,9 @@ int main()
 		i++;
 		rot[i] = (rot[i] + 1) % 64;
 	    }
-	    
+	    if (i > 0)
+	    	refresh_task(rot, &task);
+
 	    /***** Résolution *****/
 	    if (solve(S, rot, &task))
 		result_found(S[0], start);

### 2020-05-13 article/main.tex
lattice+partie de preuve facile

@@
 
+\section{lattice geometric progression}
+
+One of the key point of the algorithm is, given a modulus \(M\) and a given common ration \(\alpha\), to find a modular geometric progression knowing only the most significant bits of its \(n\) first terms.
+
+We call \(L(\alpha,M,n)\) the set of n-tuples \(U = (u_0,\dots u_{n-1})\) satisfying :
+
+\[u_{i} = \alpha u_{i-1} \mod M \] for all \(i\) in \(\{1,\dots,n-1\}\).
+
+We easily notice that, if \(U\) and \(V\) are in \(L(\alpha,M,n)\), so are \(-U\) and \(U+V\), hence \(L(\alpha,M,n)\) is a lattice.
+
+The lattice \(L(\alpha,M,n)\) is spawned by the lines of the following matrix :
+
+\[G =\begin{pmatrix}
+1&\alpha&\alpha^2&\dots&\alpha^{n-1}\\
+0&M&0&\dots&0\\
+0&0&M&\dots&0\\
+\dots&\dots&\dots&\dots&\dots\\
+0&0&0&\dots&M\\
+\end{pmatrix}\]
+
+We want to find an element \(U\) in the lattice only knowing the most significant bits of each of its coordinates.
+We decompose \(U = U_1 + U_2\) where \(U_1\) contains only the upper bits and \(U2\) the lowers one.
+There is no reason for \(U_1\) to be in the lattice, but as \(U_2\) is small, we know \(U_1\) is \emph{close} to \(U\).
+So, to find \(U\), we will solve CVP(\(L(\alpha,M)\),\(U_1\)).
+
 \section{Predicting PCG with Known Increment}
@@
 
+\section{babairounding}
+
+We call \(S't\) the truncated vector : \(S'[\ell,64+\ell]\).
+
+We consider the case were we successfully guessed the 6 upper bits of each \(S_i\) and the \(\ell\) lower bits of \(S_0\). Then we know the \(6+\ell\) upper bits of \(S't\).
+
+(à détailer ?)
+
+We had define \(S'\) such that \(S'\) is in the lattice \(L(A,2^{128},n)\) and its \(\ell \) lower bits are 0. Then \(S't\) is in \(L(1,2^{64},n)\).
+
+If we consider the vector \(T'\) computed in step (numerodelastepenquestion), we notice it is close to \(S't\) in the sense of they have exactly the same \(6+\ell\) upper bits.
+So we want to recover \(S't\) by solving an approach \(CVP(L(A,2^{64},n),T')\).
+
+Now we need to prove that the \(pwet\) obtained with the Babai rounding is indeed \(S't\).
+
+We denote \(r = rounding(YG^{-1}) \).
+
+\begin{align*}
+\lVert S't - pwet \rVert &= \lVert S't - rG \rVert \\
+&= \lVert S't - (r-YG^{-1} + YG^{-1})G \rVert\\
+&= \lVert S't - Y + (r-YG^{-1})G \rVert\\
+&\leqslant \lVert S't - Y \rVert + \lVert(r-YG^{-1})\rVert \times \lVert G\rVert\\	
+\end{align*}
+
+By definition, \(r\) is the closest integer vector to \(YG^{-1}\).
+But as \(S't\) is an element of the lattice, \(S'tG^{-1}\) is an integer vector.
+Thus \(r-YG^{-1}\) is smaller than \(S'tG^{-1}-YG^{-1}\).
+
+Hence :
+\begin{align*}
+\lVert S't - pwet \rVert &\leqslant \lVert S't - Y \rVert + \lVert(r-YG^{-1})\rVert \times \lVert G\rVert\\	
+&\leqslant \lVert S't - Y \rVert + \lVert(S'tG^{-1}-YG^{-1})\rVert \times \lVert G\rVert\\	
+&\leqslant \lVert S't - Y \rVert + \lVert(S't-Y)\rVert \times \lVert G^{-1} \rVert  \times \lVert G\rVert\\
+& 	\leqslant \lVert S't - Y \rVert \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert )\\
+\end{align*}
+
+But, by construction of \(Y\), we have that \(\lVert S't - Y \rVert \leqslant \sqrt{n2^{2(58-l)}} \). So
+
+\[ \lVert S't - pwet \rVert \leqslant \sqrt{n2^{2(58-l)}}\times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert )\]\\
+
+Now all we need to do is find sets of parameters \((n,l)\) such that the following inequality is satisfied: 
+\[\sqrt{n2^{2(58-\ell)}} \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert) \leqslant \lVert SVP(L(A,2^{64},n))\rVert \].
+
+We notice that only the left term of the inequality depends on \(\ell\).
+
+Hence, some sets of parameters that satisfies this inequality are :
+\begin{itemize}
+	\item \(n = 3\) and \(\ell \geqslant 19\)
+	\item \(n = 4\) and \(\ell \geqslant 13\)
+	\item \(n = 5\) and \(\ell \geqslant 11\)
+	\item \(n = 6\) and \(\ell \geqslant 8\)
+\end{itemize}
+
 \subsection{Performance test results}

### 2020-05-13 article/main.tex
Rajout de l'algo C connu

@@
 
-\subsection{Brute-Force Attack with known $C$}
-
-In the case where $C$ is known, if we find $S_0$, we can calculate every $S_i$
-with the given formula, and predict all the following outputs of the
-generator. The naive way is by doing a comprehensive research of $upS_0$, and
-then calculating $lowS_0$ with the formula :
-\begin{equation}
-    lowS_0 = upS_0\ \mathtt{XOR}\ \mathtt{unrotate}(X_0)
-\end{equation}
-with \texttt {unrotate} being the inverse function of \texttt{rotate}.
-
-If our guessing of $upS_0$ is correct, we can obtain the correct $X_1$ with
-$S_1$ deduced from the initial $S_0$. The preceding is a naive method for
-predicting PCG generator in $2^{64}$ operations. But we can have other values
-that verify that test. This will not return $S_0$ but a very short list of
-possibles $S_0$, with the actual one in it. We just need to test the following
-outputs to find the real $S_0$. This is yet a very slow way to guess the
-internal state, so we need a better solution.
-
-\subsection{Brute-Force Attack with unknown $C$}
-
 In case $C$ is unknown, retrieving $S_0$ is harder. The naive algorithm gives us
@@
 
-%\subsection{Different solutions for Truncated Linear Congruential Algorithms}
+\section{C connu}
 
-\subsection{Frieze \textit{et al} algorithm}
-
-The algorithm we studied uses lattices, and more specifically use the\break Lenstra–Lenstra–Lovász lattice basis reduction algorithm (LLL algorithm)\cite{LLLpastack}.
-A large part of this algorithm is very well explained on Stack Overflow\cite{LLLstack}.
-We will detail the entire algorithm here.\\
-
-A lattice in $\mathbb{R}^n$ is a subgroup of the additive group $(\mathbb{R}^n,+)$, isomorphic to $(\mathbb{Z}^n,+)$.
-The LLL algorithm is an approximation algorithm of the reduction of a lattice basis.
-It returns a reduced basis, with shorter vectors than the ones given.
-These vectors are a good approximation of the basis with the smallest determinant.\\
-
-To use this algorithm, we reduce the sequence of $S_i$ to a geometric sequence $S'$ and, given the relationship between $S$ and $S'$, we can retrieve $S$.
-There are different methods to achieve this result and we selected two of them in our optimised algorithms.
-Consider the following geometric sequence:
-\begin{equation}
-    S'_i=\alpha ^i \times S'_0 \mod{M}
-\end{equation}
-With $S'$ as the vector made of all the $S'_i$ of length $nbiter$, and $A$ blocked, the $S'$ vectors are the vectors of the lattice generated by the matrix :
+\subsection{Réduction}
+Pour deviner l'état interne du générateur, une méthode consiste à procéder par brute-force.
+On devine les 64 bits supérieurs de $S_0$, on en déduit la rotation $r_0$ utilisée, on peut retrouver  les 64 bits inférieurs:
 
 \begin{equation}
-L=
+   S_0[0:64] = S_0[64:128] XOR \texttt{ROT(X_0 , 64 - r_0)} 
+\end{equation}
+
+Cette méthode demanderait de l'ordre de 2^64 calculs, ce qui n'est pas acceptable.
+Nous pouvons cependant utiliser cette équation pour déduire des informations sur l'état interne de S.
+En effet,en devinant par brute-force les 6 bits du haut de S, et ses \ell bits du bas, on récupère :
+% schéma "relevant bits of the outputs"
+
+On réduit $S$ à une série géométrique plus simple, .
+Notons $S'$ l'état interne auquel on retire sa composante en $C$, et en $S_0[0:\ell]$ (note de bas de page : le fait de retirer $S_0[0:\ell]$ n'est pas nécéssaire, mais nous permet de limiter la taille des instances utilisées à 64 bits, très utile pour la programmation en C).
+Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, de matrice génératrice :
+\begin{equation}
 \begin{pmatrix} 
-M & 0 & & \dots & 0\\
-\alpha & -1 & 0 & \dots & 0\\
-\alpha ^2 & 0 & -1 & \dots & 0\\
+1 & \alpha & \alpha ^2 & \dots & \alpha ^{nbiter - 1}\\
+0 & M & 0 & \dots & 0\\
+0 & 0 & M & \dots & 0\\
 \vdots & & & \ddots & \vdots\\
-\alpha ^{nbiter - 1} & 0 & 0 & \dots & -1
+0 & 0 & \dots & 0 & M
 \end{pmatrix}
 \end{equation}
+On
 
-%Beau "ell" : $\ell$
+\subsection{Algorithme}
+On calcule préalablement les matrices $G$ et $G^{-1}$, avec $G$ la matrice génératrice réduite du réseau blablabla par l'algorithme LLL.
+Pour tous les $r_i$ et les $S_0[0 : \ell]$ possibles:\\
+\begin{enumerate}
+\item On récupère le $T$ correspondant à nos suppositions.
+On calcule $T'$.
+($T'_i = (S_i[64 - 6 : 64 + \ell] - (a^i*S_0[0 : \ell] + C * \sum_{j = 0}^i a^j)[64 - 6 : 64 + \ell]) * 2^{64 - 6 - \ell}$)\\
+%schema T et T'
 
-Let $Y$ be the most significant bits of the internal states of LCG, and $Z$ be the lower bits, with $S = Y + Z$. We already know $Y$ from the outputs and we are searching for $Z$. 
-\begin{itemize}
-    \item Let $Y'$ be the upper bits of $S'$;
-    \item Let $Z'$ be the lower bits;
-    \item Let $L'$  be the reduced matrix returned by the LLL algorithm for $L$ ($L' = LLL(L)$);
-\end{itemize}
+\item We get the correspondant $S'[\ell : 64 + \ell]$ by using the babai rounding technique on $T'$ using $G$ and $G^{-1}$.
 
-By definition of $L$, we have :
-\begin{align}
-    L \times S' &=0 \mod{M}\\
-    L' \times S' &= 0 \mod{M}\\
-    L' \times (Y' + Z') &= M \times \ell\ \text{($\ell$ unknown integer vector)}\\
-    L' \times Z' &= M \times \ell - L' \times Y'
-\end{align}
-We have $M = 2^k$. L' coefficients are small. We want to find the actual value of $\ell$.\\
+\item We test our solution by computing $S_0$ and testing if the following internals states are coherent with the PCG outputs. If the results are coherents, we return S_0.
 
-We suppose we have every coordinate of $L' \times Z'$ smaller than $M/2$.
-Then every coordinate of $M \times \ell - L' \times Y'$ is smaller than $M/2$ and $M \times \ell$ is the closest vector of $M$ multiples from $L' \times Z'$.
-We have $\ell$ is the closest integer vector from $(B \times y)/M$.\\
-
-The only unknown variable left is $Z'$ and :
-\begin{equation}
-    Z' = L'^{-1} \times (M \times k) - Y' 
-\end{equation}
-Now we know $Z'$, and the integer $S' \mod{M}$, and finally, $S \mod{M}$ the internal state we were looking for.\\
-
-\subsection{Application to our generator}
-The $L$ matrix for our problem is :
-\begin{equation}
-    L=
-\begin{pmatrix}
-1208925819614629174706176 & 0 &  0 &  0 &  0 &  0 &  0 &  0 &  0 & 0\\
-442210214475561085761093 & -1 &  0 &  0 &  0 & 0 &  0 &  0 &  0 &  0\\
-548907270092662740397721 & 0 & -1 &  0 &  0 & 0 &  0 &  0 &  0 &  0\\
-68528166612430993102141 & 0 &  0 & -1 &  0 & 0 &  0 &  0 &  0 &  0\\
-579003058182980434679665 & 0 &  0 &  0 & -1 & 0 &  0 &  0 &  0 &  0\\
-907267876411637163239285 & 0 &  0 &  0 &  0 & -1 &  0 &  0 &  0 &  0\\
-971193525585951128490121 & 0 &  0 &  0 &  0 & 0 & -1 &  0 &  0 &  0\\
-423043601822819931331309 & 0 &  0 &  0 &  0 & 0 &  0 & -1 &  0 &  0\\
-722352710310758831226849 & 0 &  0 &  0 &  0 & 0 &  0 &  0 & -1 &  0\\
-239218985093571541740965 & 0 &  0 &  0 &  0 & 0 &  0 &  0 &  0 & -1
-\end{pmatrix}
-\end{equation}
-Each coefficient is about 128 bit long. Here is the $L'$ (using the LLL algorithm of Sagemath \cite{sagemath}):
-\begin{equation}
-\begin{pmatrix}
-    L' =
-    77 & -105 & -115 &  17 & -108 & -77  & -7 &  50 & 107 &  17\\
-     50  & 80 & 9 & -90 &  138 & -104  & 26  & 90 & -17 & -134\\
-    14 & -42 & 130 & -67 & -31 & -172 & 122 & -76 & -32 &  14\\
-    -61 &  97 &  70 & 161  & -4 & -229 & -23 & -44  & -1 & -26\\
-    118  & 33 & -65 &  -3 &  83 & -176 & -187 & -70 &  25 &  34\\
-    111 & -151 & -73 & 115 &  36 & 200 &  79  & -1 & -14 & -82\\
-    94  & 15 & -18 & -36 & 142 & -47 &  32 & -188 & 161 & -127\\
-    126 & -10 & 129 & 173 & -71  & 61 & -137 &  69 & 7 &  77\\
-    174 & -62 &  58  & 39 &  79 & -133 &  93 & 115 &  24 & 181\\
-    48 & 240 &  -2 & 114 & -15 &  37 & 145 & -100 &  24 & -19\\
-\end{pmatrix}
-\end{equation}
-This algorithm is clearly really efficient, the coefficients of $L'$ are really small.
-Its bigger coefficient is $240$.
-Its smaller than $2^8 = 256$.
-In our case $M = 2^{64 + known\_low}$.
-In our tests, we had $known\_low = 16$.
-We need $Z'$ to be less than 72 bits long, which isn't hard to attain.
-
-\section{lattice geometric progression}
-
-One of the key point of the algorithm is, given a modulus \(M\) and a given common ration \(\alpha\), to find a modular geometric progression knowing only the most significant bits of its \(n\) first terms.
-
-We call \(L(\alpha,M,n)\) the set of n-tuples \(U = (u_0,\dots u_{n-1})\) satisfying :
-
-\[u_{i} = \alpha u_{i-1} \mod M \] for all \(i\) in \(\{1,\dots,n-1\}\).
-
-We easily notice that, if \(U\) and \(V\) are in \(L(\alpha,M,n)\), so are \(-U\) and \(U+V\), hence \(L(\alpha,M,n)\) is a lattice.
-
-The lattice \(L(\alpha,M,n)\) is spawned by the lines of the following matrix :
-
-\[G =\begin{pmatrix}
-1&\alpha&\alpha^2&\dots&\alpha^{n-1}\\
-0&M&0&\dots&0\\
-0&0&M&\dots&0\\
-\dots&\dots&\dots&\dots&\dots\\
-0&0&0&\dots&M\\
-\end{pmatrix}\]
-
-We want to find an element \(U\) in the lattice only knowing the most significant bits of each of its coordinates.
-We decompose \(U = U_1 + U_2\) where \(U_1\) contains only the upper bits and \(U2\) the lowers one.
-There is no reason for \(U_1\) to be in the lattice, but as \(U_2\) is small, we know \(U_1\) is \emph{close} to \(U\).
-So, to find \(U\), we will solve CVP(\(L(\alpha,M)\),\(U_1\)).
-
-\section{Predicting PCG with Known Increment}
-\subsection{Reduction to a truncated LCG model}
-By exploiting the nature of this generator, it becomes apparent that some values can be easily guessed.
-We already saw that if we guess $rot_i$, then retrieving $X_i$ becomes a simple \texttt{XOR} operation, and in turn, if we have some knowledge of a particular bit of $lowS_i$ (respectively $upS_i$), we can deduce the equivalent bit from $upS_i$ (respectively $lowS_i$).\\
-
-Moreover, there is an interesting proprity concerning the least significant bits of the internal state.
-Indeed, if we define $lowC$ as the 64 least significant bits of $C$ ($lowC = C \mod{2^{k/2}})$, we have:
-
-\begin{align}
-    S_{i+1} &= A \times S_i + C \mod{2^k}\\
-    upS_{i+1} \times 2^{k/2} + lowS_{i+1} &= A \times upS_i \times 2^{k/2} + A \times lowS_i + C \mod{2^k} \\
-    lowS_{i+1} &= A \times lowS_i + lowC \mod 2^{k/2}
-\end{align}
-We can see that $lowS_{i+1}$ is independent from $upS_i$.
-That mean that knowing $lowS_i$, we can deduce all $lowS_j$ with $j>i$.
-This relationship is used in the Brute-Force attack with known $C$, in the previous section.
-We can see on (Figure \ref{pcg128goodbits})%figure
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.70\linewidth]{pictures/deducingPCG128.png}
-    \caption{PCG128 : Relevant bits of the outputs}
-    \label{pcg128goodbits}
-\end{figure}
-that we can extract a lot of information by brute forcing certain well chosen bits.
-We defined $known\_up$ (respectively $known\_low$) the number of upper (respectively lower) bits searched by brute force.
-
-\subsection{Reduction to a geometric sequence and validation of our values}
-
-$C$ is known.
-We are looking for a geometric sequence $S'$ with which we can validate our brute-forced bits and extract $S_0$.
-Let $polC$ be the vector where $polC_i = \sum_{j=0}^{i} C \times A^j$ and $S'$ where $S'_i = S_i - polC_i$.
-We have :
-\begin{align}
-    S'_{i+1} &= S_{i+1} - polC_{i+1}\\
-    S'_{i+1} &= A \times S_i + C - polC_{i+1}\\
-    S'_{i+1} &= A \times S_i + C - (C + A \times C + ...+ A^{i} \times C)\\
-    S'_{i+1} &= A \times (S'_i + polC_i) - A \times (C + ... + A^{i-1} \times C)\\
-    S'_{i+1} &= A \times S'_i + A \times polC_i - A \times polC_i\\
-    S'_{i+1} &= A \times S'_i
-\end{align}
-
-We have a convenient $S'$ that can be used in the algorithm presented in the previous section and, there exist a simple bijection between $S'$ and $S$.
-However, its important to notice that $S'$ will not be computed modulo $M = 2^k$ but modulo $N = 2^{k/2 + known\_low}$, because we are doing an comprehensive research of the middle bits of $S$ and not the upper ones (except for the 6 bits of $rot$).
-This is not a problem, because when we have the values of $X$, and $lowS$, we can always obtain $upS$ and ultimately $S$.\\
-
-Using this bijection, we can validate if our chosen values are correct.
-Indeed, after finding $Z'$ through the LLL algorithm, we can compute $S'_0$ and $S'_1$ modulo $2^{k/2 + known\_low}$, then $S_0$ and $S_1$ modulo $2^{k/2 + known\_low}$, and finally $S_1$ and $S_0$.
-We verify if $S_1$ is coherent with $S_0$.
-If it is, then we have guessed the correct values for our generator, if not, we then consider this another value as our next test.
-
-\newpage
-
-\subsection{Pseudo-Algorithm}
-
-\begin{algorithm}[h!]
-\caption{Algorithm with known $C$}
-\begin{algorithmic}
-\REQUIRE $X$ the list of outputs
-\STATE computation of the $LLL$ matrix
-\STATE computation of $polC$
-\FOR{ all $rot_i$ for $i$ in range($nbiter$) }
-    \FOR{i in range($nbiter$)}
-        \STATE \texttt{unrotate}($X_i$)
-    \ENDFOR
-    \FOR{all $W_0$}
-        \STATE calculate the other $W_i$ with $W_0$ and $C$
-        \STATE calculate $Y$ using $W_i\ \mathtt{XOR}\ X_i$ and $rot_i\ \mathtt{XOR}\ X_i$
-        \STATE $W' = W - polC$ (we subtract only the corresponding bits of $C$)
-        \STATE $Y' = Y - polC$
-        \STATE calculation of $Z'$ due to LLL algorithm
-        \IF{$2^{k/2-known\_up}>Y'>=2^{known\_low}$}
-            \IF{$Z$ is correct}
-                \STATE return the corresponding $S_0$
-            \ENDIF
-        \ENDIF
-    \ENDFOR
-\ENDFOR
-\end{algorithmic}
-\end{algorithm}
+\end{enumerate}
 
@@
 
-\subsection{Performance test results}
-Our code is implemented in Sagemath.
-This choice was made because of the development advantages it offered: to be able to work with arbitrarily large integers and have access to a built-in LLL function.\\
-
-The number of bits obtained by exhaustive research, as we have seen in the first section, is $nbiter \times known\_up + known\_low$.
-The most efficient way should be to only guess the lower bits ($known\_up = 0$).
-But we need the 6 bit value of $rot_i$ in our algorithm.
-So we kept $known\_up = 6$ and searched for the best equilibrium between $nbiter$ and $known\_low$.
-\\
-
-Moreover, we also need this algorithm to be reliable.
-That is a concern because we know that there is a rounding error caused by an Euclidian Division.
-Depending on the error, it is possible that the algorithm doesn't find the correct $Z$ even if $Y$ and $W$ are correct (with $W$ the lower bits of $S$ and $Y$ the middle bits of $S$ found by \texttt{XOR}).
-\\
-
-We tested the reliability and the efficiency of the algorithm. By trial and error, we ended up finding two interesting settings :
-\begin{itemize}
-    \item $known\_low = 16$ and $nbiter = 4$ : we have a $2^{40}$ brute-force algorithm.
-When supplying it with PCG128 outputs, the rate at which the algorithm detects it and obtains the correct $S_0$ is of 90\%.
-    \item $known\_low = 15$ and $nbiter = 4$ : we have a $2^{39}$ brute-force algorithm.
-When supplying it with PCG128 outputs, the rate at which the algorithm detects it and obtains the correct $S_0$ is of 70\%.
-\end{itemize}
-
-In any case, if we supply a random output, the algorithm will detect this and will never return a false $S_0$.
-When not finding $S_0$, the default answer is to disqualify the input as a PCG data steam.\\
-
-Unfortunately, we didn't get to run our entire solution in a real time environment, because with our personal equipment, and not fully optimized code, we estimated that it would have take more or less ten years to obtain an answer, which is obviously too long.
-The local environment at the university didn't provide the right tools (SageMath), which complexified the development process of sharing the computation load between multiple computers.
-However, there are some methods and coding improvements to reduce the time needed, as we will discuss in the next section.
-
 [insert here] attack predicting algo description.

### 2020-05-13 article/main.tex
Avec un peu moins d'erreurs

@@
 \begin{equation}
-   S_0[0:64] = S_0[64:128] XOR \texttt{ROT(X_0 , 64 - r_0)} 
+   S_0[0:64] = S_0[64:128] \texttt{ XOR } \texttt{ROT(X_0 , 64 - r_0)} \\
 \end{equation}
 
-Cette méthode demanderait de l'ordre de 2^64 calculs, ce qui n'est pas acceptable.
-Nous pouvons cependant utiliser cette équation pour déduire des informations sur l'état interne de S.
-En effet,en devinant par brute-force les 6 bits du haut de S, et ses \ell bits du bas, on récupère :
+Cette méthode demanderait de l'ordre de $2^{64}$ calculs, ce qui n'est pas acceptable.
+Nous pouvons cependant utiliser ce{tte équation pour déduire des informations sur l'état interne de S.
+En effet,en devinant par brute-force les 6 bits du haut de S, et ses $\ell$ bits du bas, on récupère :\\
 % schéma "relevant bits of the outputs"
 
-On réduit $S$ à une série géométrique plus simple, .
+On réduit $S$ à une série géométrique plus simple.
 Notons $S'$ l'état interne auquel on retire sa composante en $C$, et en $S_0[0:\ell]$ (note de bas de page : le fait de retirer $S_0[0:\ell]$ n'est pas nécéssaire, mais nous permet de limiter la taille des instances utilisées à 64 bits, très utile pour la programmation en C).
-Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, de matrice génératrice :
+Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, de matrice génératrice :\\
 \begin{equation}
+M =
 \begin{pmatrix} 
@@
 \end{equation}
-On
 

### 2020-05-15 article/main.tex
ça compile
Ajout d'une bonne partie du code C inconnu et d'une image

@@
 
-\section{Naive Brute-Force Reconstruction of the Internal State}
-
-In this section we describe the simplest possible prediction algorithm we could
-think of. They serve as ``baseline'' to evaluate the more refined predictors
-described below.
-
-\todo{Dans le cas où C est connu, c'est le truc de Vigna, non ?}
-
-In case $C$ is unknown, retrieving $S_0$ is harder. The naive algorithm gives us
-an answer within $2^{128+6}$ simple operations whit the following algorithm:
-
-\begin{algorithmic}[!h]
-\FOR{all possibilities of $lowC$ and $upS_0$}
-    \STATE \texttt{unrotate}($X_0$)
-    \STATE $lowS_0 = upS_0\ \mathtt{XOR}\ X_0$
-    \STATE $lowS_1 = lowC + lowS_0 \times A \mod{2^{k/2}}$
-    \FOR{all possibilities of $rot_1$}
-        \STATE $\mathtt{unrotate}(X_1)$
-        \STATE $upS_1 = lowS_1\ \mathtt{XOR}\ X_1$
-        \STATE $C = S_1 - A \times S_0 \mod{2^{k}}$
-        \STATE $S_2 = S_1 \times A + C$
-        \IF {$S_2$ is coherent with  $X_2$}
-            \STATE print($C$,$S_0$)
-        \ENDIF
-    \ENDFOR
-\ENDFOR
-\end{algorithmic}
-
-This algorithm reveals us $C$ and a list of outputs with $S_0$ in it. The entry
-is 256 bits long, and we used 3 64 bits long outputs of the generator. There
-will be about $2^{64}$ possible $S_0$. This is why we need a couple other
-outputs to get the real $S_0$. Sorting out which is the correct $S_0$ can be
-done afterwards, and won't affect the complexity. But given the complexity of
-this algorithm, it is even more unrealistic to solve problem in this fashion
-then the previous one where $C$ was known.
-
 \section{Predicting Truncated Linear Congruential Generators}
@@
 \begin{equation}
-   S_0[0:64] = S_0[64:128] \texttt{ XOR } \texttt{ROT(X_0 , 64 - r_0)} \\
+   S_0[0:64] = S_0[64:128] \texttt{ XOR } \texttt{ROT}(X_0 , 64 - r_0) \\
 \end{equation}
@@
 Cette méthode demanderait de l'ordre de $2^{64}$ calculs, ce qui n'est pas acceptable.
-Nous pouvons cependant utiliser ce{tte équation pour déduire des informations sur l'état interne de S.
-En effet,en devinant par brute-force les 6 bits du haut de S, et ses $\ell$ bits du bas, on récupère :\\
-% schéma "relevant bits of the outputs"
+Nous pouvons cependant utiliser cette équation pour déduire des informations sur l'état interne de S.
+En effet,en devinant par brute-force les 6 bits du haut de S, et ses $\ell$ bits du bas, on récupère :\\% schéma "relevant bits of the outputs"
 
@@
 \begin{enumerate}
-\item On récupère le $T$ correspondant à nos suppositions.
-On calcule $T'$.
-($T'_i = (S_i[64 - 6 : 64 + \ell] - (a^i*S_0[0 : \ell] + C * \sum_{j = 0}^i a^j)[64 - 6 : 64 + \ell]) * 2^{64 - 6 - \ell}$)\\
+\item On récupère le $T$ correspondant à nos suppositions. On calcule $T'$. ($T'_i = (S_i[64 - 6 : 64 + \ell] - (a^i*S_0[0 : \ell] + C * \sum_{j = 0}^i a^j)[64 - 6 : 64 + \ell]) * 2^{64 - 6 - \ell}$)
 %schema T et T'
+Notons cependant que les composantes en $C$ et $S[0:l]$ sont calculées en dehors de la boucle et stockées dans un tableau pour des raisons de performance.
+Ceci est fait de même dans tous les algorithmes suivants.
 
@@
 
-\item We test our solution by computing $S_0$ and testing if the following internals states are coherent with the PCG outputs. If the results are coherents, we return S_0.
+\item We test our solution by computing $S_0$ and testing if the following internals states are coherent with the PCG outputs. If the results are coherents, we return $S_0$.
 
@@
 \section{Predicting PCG with Unknown Increment}
+We can't use the same algorithm for the general case. We don't known $C$, so we used an other geometric reduction.
+$(\Delta S_{i+1}S_i)_i$ est un suite géométrique de raison $a$.
 
-[insert here] attack predicting algo description. Phase 1, Phase 2, Phase 3
+On calcule préalablement les matrices $G$ et $G^{-1}$.
+
+\subsection{Part 1 : Premier filtre}
+Pour tous les $r_i$, les $S_0[0 : \ell]$ et les $C[0 : \ell]$ possibles:
+\begin{enumerate}
+  \item On récupère le $T$  correspondant à nos suppositions. On calcule $\Delta T$ :
+  \begin{equation}
+    \Delta T_i = (S_{i+1}[58 : 64 + \ell] - S_i[58 : 64 + \ell] - (S_0[0:\ell] * (a^{i+1} - a^i) + C[0:\ell] * a^i )[58 : 64 + \ell]) * 2^{64 - l - 6}
+  \end{equation}
+  On peut donc aussi définir $\Delta T_i = T'_{i+1} - T'_{i}$ avec $T'$ dont on aurait soustrait la composante en $C[0:\ell]$ et non en $C$.
+
+  \item We get the correspondant $\Delta S_0S_1[0 : 64 + \ell]$ by using the babai rounding technique on $\Delta T$ using $G$ and $G^{-1}$.
+
+  \item On teste s'il existe pour l'itération suivante au moins une rotation $r_n$ cohérente avec le $\Delta T$ trouvé. Pratiquement, pour chacune des 64 $r_n$ possibles, on teste si :
+  \begin{align}
+     \Delta S_0S_n[0:l] &= S_0[64:64 + l] XOR \mathtt{ROT(X_n , 64 - r_n)[0:l]}\\
+     r_n &= S_0[58:64] XOR \mathtt{ROT(X_n , 64 - r_n)[58:64]}
+  \end{align}
+  Si un de ces égalité n'est pas vérifiée, la solution n'est pas retenue, on passe à la prochaine itération.
+
+  \item On répète la dernière opération encore 3 fois.
+\end{enumerate}
+Cette première partie de l'algorithme comporte l'énorme majorité du temps de calcul, c'est donc sur elle que doit se porter la recherche de performance.
+
+\subsection{Part 2 : Récupération de la différence complète et deuxième filtre}
+On calcule préalalement H matrice génératrice du réseau Lblablabla2 réduite par la méthode LLL.
+On a récupéré de la Part 1 une série de $S_0[0 : \ell]$, $r_i$ (i = 0, ..., n) et $C[0 : \ell]$ parmi lequels se trouve notre solution.
+\begin{enumerate}
+  \item recupérer toutes les $r_i$ possibles, pour $n \leq i \leq m$  par la même méthode que l'étape 3 de la Part 1, répétée $m-n$ fois. On garde en mémoire toutes les rotations possiles.
+  \item Pour chacune des $r$ possibles
+  \begin{enumerate}
+    \item We cumpute $\Delta r$ avec le même méthode que pour $\Delta T$ : 
+    \begin{equation}
+      \Delta r_i &= (S_{i+1}[122 : 128] - S_i[122:128] - (S_0[0:\ell] * (a^{i+1} - a^i) + C[0:\ell] * a^i )[122:128]) * 2^{64 - l - 6}
+    \end{equation}
+    \item We get the correspondant $\Delta S'_0S'_1$ ($S'$ dont on aurait retiré que ) computing $\texttt{CVP}(r * 2^{128 - \ell}, Lblablabla2)$.
+We use the fpylll python library CVP function.
+On stocke le $DS_0S_1$ correspondant.
+  \end{enumerate}
+\end{enumerate}
 

### 2020-05-15 article/main.tex
charles latex magic
Ajout du débbut de la partie "presentation de l'algo dans le cas le plus simple" et petites corrections

@@
 \begin{align}
-    X_{i} &= \mathtt{rotate}(upS_i\ \mathtt{XOR}\ lowS_i)\\
+    X_{i} &= \mathtt{rotate}(upS_i\ \oplus\ lowS_i)\\
     S_{i+1} &= A \times S_{i} + C \mod{2^k}
@@
 
-\section{Predicting Truncated Linear Congruential Generators}
+\section{Algorithm of Frieze \textit{et al}}
 
@@
   tronqué, autant décrire la technique à base de réseaux un bon coup.}
+Pour casser PCG, nous avons utilisé la méthode présentée par Alan M.
+Frieze, Johan Hastad, Ravi Kannan, Jeffray C.
+Lagarias et Adi Shamir \cite{Frieze}.
+Plaçons nous dans le cas simple du Lehmer's generator.
+It's a Linear Congruential Generator, a known multiplier $A$, a known modulus $2^128$ and no increment.
+The outputs are the 64 upper-bits of $S$.
+For $n \in \mathbb{N}$, the $n$ firsts internal states form a lattice.
+This lattice is generated by the matrix :
+\begin{equation}
+genMatrix =
+\begin{pmatrix} 
+1 & A & A^2 & \dots & A^{n- 1}\\
+0 & M & 0 & \dots & 0\\
+0 & 0 & M & \dots & 0\\
+\vdots & & & \ddots & \vdots\\
+0 & 0 & \dots & 0 & M
+\end{pmatrix}
+\end{equation}
 
+The idea of this method is to consider a known vector $T$, close to $S$ the internal state we are looking for, and to search for the closest vector in the 
 The algorithm of Frieze \textit{et al}\cite{Frieze} solves the case where $M$
@@
 \begin{equation}
-   S_0[0:64] = S_0[64:128] \texttt{ XOR } \texttt{ROT}(X_0 , 64 - r_0) \\
+   S_0[0:64] = S_0[64:128] \oplus \texttt{ROT}(X_0 , 64 - r_0) \\
 \end{equation}
@@
 \begin{equation}
-M =
+genMatrix =
 \begin{pmatrix} 
-1 & \alpha & \alpha ^2 & \dots & \alpha ^{nbiter - 1}\\
+1 & A & A ^2 & \dots & A ^{n- 1}\\
 0 & M & 0 & \dots & 0\\
@@
 \begin{enumerate}
-\item On récupère le $T$ correspondant à nos suppositions. On calcule $T'$. ($T'_i = (S_i[64 - 6 : 64 + \ell] - (a^i*S_0[0 : \ell] + C * \sum_{j = 0}^i a^j)[64 - 6 : 64 + \ell]) * 2^{64 - 6 - \ell}$)
+\item On récupère le $T$ correspondant à nos suppositions.
+On calcule $T'$.
+($T'_i = (S_i[64 - 6 : 64 + \ell] - (A^i*S_0[0 : \ell] + C * \sum_{j = 0}^i A^j)[64 - 6 : 64 + \ell])[64 - 6 : 64 + \ell] * 2^{64 - 6 - \ell}$)
 %schema T et T'
@@
   \begin{equation}
-    \Delta T_i = (S_{i+1}[58 : 64 + \ell] - S_i[58 : 64 + \ell] - (S_0[0:\ell] * (a^{i+1} - a^i) + C[0:\ell] * a^i )[58 : 64 + \ell]) * 2^{64 - l - 6}
+    \Delta T_i = (S_{i+1}[58 : 64 + \ell] - S_i[58 : 64 + \ell] - (S_0[0:\ell] * (A^{i+1} - A^i) + C[0:\ell] * A^i )[58 : 64 + \ell]) * 2^{64 - l - 6}
   \end{equation}
@@
   \begin{align}
-     \Delta S_0S_n[0:l] &= S_0[64:64 + l] XOR \mathtt{ROT(X_n , 64 - r_n)[0:l]}\\
-     r_n &= S_0[58:64] XOR \mathtt{ROT(X_n , 64 - r_n)[58:64]}
+     \Delta S_0S_n[0:l] &= S_0[64:64 + l] \oplus \mathtt{ROT(X_n , 64 - r_n)[0:l]}\\
+     r_n &= S_0[58:64] \oplus \mathtt{ROT(X_n , 64 - r_n)[58:64]}
   \end{align}
@@
   \begin{enumerate}
-    \item We cumpute $\Delta r$ avec le même méthode que pour $\Delta T$ : 
-    \begin{equation}
-      \Delta r_i &= (S_{i+1}[122 : 128] - S_i[122:128] - (S_0[0:\ell] * (a^{i+1} - a^i) + C[0:\ell] * a^i )[122:128]) * 2^{64 - l - 6}
-    \end{equation}
+    \item We cumpute $r'$ et $\Delta r avec le même méthode que pour $\Delta T$ : 
+    \begin{align}
+      r'_i = r_i - (A^i*S_0[0 : \ell] + C[0 : \ell] * \sum_{j = 0}^i A^j)[64 - 6 : 64 + \ell])[122 : 128]
+      \Delta r_i &= r_{i+1} - r_i
+    \end{align}
+
     \item We get the correspondant $\Delta S'_0S'_1$ ($S'$ dont on aurait retiré que ) computing $\texttt{CVP}(r * 2^{128 - \ell}, Lblablabla2)$.
@@
 On stocke le $DS_0S_1$ correspondant.
+
+    \item 
   \end{enumerate}

### 2020-05-15 Cunknown/fonctions.c
premier dessin tikz

@@ -26,9 +26,9 @@ double invG[16] =
 void init_var_globales()
 {
     //multiplier a OK !
-    a = (((pcg128_t) 2549297995355413924) << k) + ((pcg128_t) 4865540595714422341);
+    a = (((pcg128_t) 2549297995355413924) << 64) + ((pcg128_t) 4865540595714422341);
 
-    a_inv = (((pcg128_t) 566787436162029664) << k) + ((pcg128_t) 11001107174925446285);
+    a_inv = (((pcg128_t) 566787436162029664) << 64) + ((pcg128_t) 11001107174925446285);
 
     //increment polynome polC OK !
     polA[0] = 0;
@@ -38,7 +38,7 @@ void init_var_globales()
         powA[i] = powA[i-1] * a;
     }
     for (int i = 0; i < 16; i++)
-    	invG[i] *= 1ull << (k - known_up - known_low);
+    	invG[i] *= 1ull << (64 - 6 - known_low);
 }
 
 //////////////////// chrono //////////////////
@@ -51,20 +51,20 @@ double wtime()
 
 static inline u64 unrotate(u64 Xi, int i)
 {
-    return (Xi >> (k-i)) | (Xi << i);
+    return (Xi >> (64-i)) | (Xi << i);
 }
 
 char * setupGoodY()
 {
-	char* goodY = malloc((1<<(known_low + known_up)) * sizeof(char) * nbtest / 8);
-	for (u64 y = 0 ; y < nbtest * (1<< (known_low + known_up)) / 8 ; y++)
+	char* goodY = malloc((1<<(known_low + 6)) * sizeof(char) * nbtest / 8);
+	for (u64 y = 0 ; y < nbtest * (1<< (known_low + 6)) / 8 ; y++)
 	    goodY[y] = 0;
 	return goodY;
 }
 
 static inline void setbit(char *goodY, int i, u64 Y, int v)
 {
-    int idx = Y + i * (1 << (known_up + known_low));
+    int idx = Y + i * (1 << (6 + known_low));
 	int j = idx / 8;
 	int l = idx % 8;
 	if (v == 1)
@@ -77,10 +77,10 @@ void getGoodY(char* goodY, const u64 (*X)[nboutput], const u64* lowSumPol, int v
 {
 	for (int i = 0 ; i < nbtest ; i++){
[…]

### 2020-05-15 Cunknown/fonctions.h

@@ -11,8 +11,6 @@ typedef long long          i64;
 typedef unsigned long long u64;
 
 /***** Macro et Variables globales *****/
-#define k 64
-#define known_up 6
 #define known_low 11 //à relancer 3 fois, sur des X differents (X0 - X4 ,X5 - X9,...) puis lancer avec 12 et enfin 13
 #define nbiter 5
 #define nbtest 4

### 2020-05-15 article/main.tex
premier dessin tikz
Avec encore des corrections et quelques ajouts

@@
 Lagarias et Adi Shamir \cite{Frieze}.
-Plaçons nous dans le cas simple du Lehmer's generator.
+Plaçons nous dans le cas simple du Lehmer's generator\cite{Lehmer}.
 It's a Linear Congruential Generator, a known multiplier $A$, a known modulus $2^128$ and no increment.
-The outputs are the 64 upper-bits of $S$.
-For $n \in \mathbb{N}$, the $n$ firsts internal states form a lattice.
+The outputs $X$ are the 64 upper-bits of $S$.
+For $n \in \mathbb{N}$, the $n$ firsts internal states form a lattice we will name $\mathcal{L}(A,2^{128},n)$.
+\todo{rajouter ref Lehmer's generator}
+
 This lattice is generated by the matrix :
@@
 
-The idea of this method is to consider a known vector $T$, close to $S$ the internal state we are looking for, and to search for the closest vector in the 
-The algorithm of Frieze \textit{et al}\cite{Frieze} solves the case where $M$
-and $A$ are known, $M$ being a power of two and, possibly with a significant
-portion of the internal state unknown. The known part is the upper bits of the
-state. This algorithm is coherent with our problem, this being the reason why we
-focused on this one in particular. However, this algorithm won't work for some
-particular multipliers. But this problem doesn't concern us, because the default
-multiplier of \texttt{PCG128} is not one of them.
+The idea of this method is to consider a known vector $T$, close to $S$ the internal state we are looking for, and to search for the closest vector in $\mathcal{L}(A,2^{128},n)$.
+In the case of Lehmer's generator, we use $X * 2^{64}$ as an approximation of $S$.
 
-The first article we found was written by Joan Boyar\cite{Boyar1989}. The
-subject of her study was a special case of truncated LCG, with $A$, $C$ and $M$
-unknown and has access to a very large part of the internal state. This made it
-so that we couldn't use her findings, because guessing a large part of the
-internal state wouldn't be cost-effective. However, her study included a very
-broad state of the art, explaining the specifics of many algorithms, as well as
-describing in which circumstances they should be used.
+\todo{preuve que c'est bien le plus proche ici ?}
+
+Pour améliorer les performances de nos algorithmes les plus critiques, nous n'avons dans certains cas pas utilisé d'oracle CVP parfait, mais la méthode du babai-rounding.
+
+\todo{description babai rounding}
+\todo{et preuve ?}
 
@@
 Nous pouvons cependant utiliser cette équation pour déduire des informations sur l'état interne de S.
-En effet,en devinant par brute-force les 6 bits du haut de S, et ses $\ell$ bits du bas, on récupère :\\% schéma "relevant bits of the outputs"
+En effet,en devinant par brute-force les 6 bits du haut de S, et ses $\ell$ bits du bas, on récupère :
+
+\todo{schéma "relevant bits of the outputs"}
 
@@
 Notons $S'$ l'état interne auquel on retire sa composante en $C$, et en $S_0[0:\ell]$ (note de bas de page : le fait de retirer $S_0[0:\ell]$ n'est pas nécéssaire, mais nous permet de limiter la taille des instances utilisées à 64 bits, très utile pour la programmation en C).
-Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, de matrice génératrice :\\
+Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, nommé $\mathcal{L}(A,2^64,n)$.
+On se réduit donc au cas facile étudié dans la partie précédente.
 \begin{equation}
@@
 \subsection{Algorithme}
-On calcule préalablement les matrices $G$ et $G^{-1}$, avec $G$ la matrice génératrice réduite du réseau blablabla par l'algorithme LLL.
+On calcule préalablement les matrices $G$ et $G^{-1}$, avec $G$ la matrice génératrice réduite du réseau $\mathcal{L}(A,2^64,n)$ par l'algorithme LLL.
 Pour tous les $r_i$ et les $S_0[0 : \ell]$ possibles:\\
@@
 ($T'_i = (S_i[64 - 6 : 64 + \ell] - (A^i*S_0[0 : \ell] + C * \sum_{j = 0}^i A^j)[64 - 6 : 64 + \ell])[64 - 6 : 64 + \ell] * 2^{64 - 6 - \ell}$)
-%schema T et T'
+
+\todo{schema T et T'}
+
 Notons cependant que les composantes en $C$ et $S[0:l]$ sont calculées en dehors de la boucle et stockées dans un tableau pour des raisons de performance.
@@
 We can't use the same algorithm for the general case. We don't known $C$, so we used an other geometric reduction.
-$(\Delta S_{i+1}S_i)_i$ est un suite géométrique de raison $a$.
+$(\Delta S_{i+1}S_i)_i$ est un suite géométrique de raison $A$.
 
@@
   \end{equation}
-  On peut donc aussi définir $\Delta T_i = T'_{i+1} - T'_{i}$ avec $T'$ dont on aurait soustrait la composante en $C[0:\ell]$ et non en $C$.
+  Ou plus simplement $\Delta T_i = T'_{i+1} - T'_{i}$ avec $T'$ dont on aurait soustrait la composante en $C[0:\ell]$ et non en $C$.
 
@@
   \begin{align}
-     \Delta S_0S_n[0:l] &= S_0[64:64 + l] \oplus \mathtt{ROT(X_n , 64 - r_n)[0:l]}\\
+     \Delta S_0S_n[0:l] &= S_0[64:64 + \ell] \oplus \mathtt{ROT(X_n , 64 - r_n)[0:\ell]}\\
      r_n &= S_0[58:64] \oplus \mathtt{ROT(X_n , 64 - r_n)[58:64]}
@@
 \subsection{Part 2 : Récupération de la différence complète et deuxième filtre}
-On calcule préalalement H matrice génératrice du réseau Lblablabla2 réduite par la méthode LLL.
+On calcule préalalement H matrice génératrice du réseau $\mathcal{L}(A,2^{128 - \ell},m)$ réduite par la méthode LLL.
 On a récupéré de la Part 1 une série de $S_0[0 : \ell]$, $r_i$ (i = 0, ..., n) et $C[0 : \ell]$ parmi lequels se trouve notre solution.
@@
   \begin{enumerate}
-    \item We cumpute $r'$ et $\Delta r avec le même méthode que pour $\Delta T$ : 
+    \item We cumpute $r'$ et $\Delta r$ avec le même méthode que pour $\Delta T$ : 
     \begin{align}
@@
 
-    \item We get the correspondant $\Delta S'_0S'_1$ ($S'$ dont on aurait retiré que ) computing $\texttt{CVP}(r * 2^{128 - \ell}, Lblablabla2)$.
+    \item We get the correspondant $\Delta S'_0S'_1$ ($S'$ dont on aurait retiré que ) computing $\texttt{CVP}(r * 2^{128 - \ell}, \mathcal{L}(A,2^{128 - \ell},m))$.
 We use the fpylll python library CVP function.
@@
 
-    \item 
+    \item \todo{parler du problème de retenue étrange}
   \end{enumerate}

### 2020-05-15 article/main.tex

@@
 
+\begin{tikzpicture}[yscale=0.66]
+  \draw[red, use as bounding box] (-0.75, -3.75) rectangle (12.8, 2.25);
+
+  % S_i
+  \begin{scope}
+    % remplissage
+    \fill[fill=lightgray] (4.6, 0) rectangle +(2.4, 1);
+    \fill[fill=lightgray] (0, 0) rectangle +(0.6, 1);
+    \fill[fill=lightgray] (11, 0) rectangle +(1.8, 1);
+
+    % bordures
+    \draw[thick]  (0, 0) rectangle (12.8, 1);
+    \draw  (0.6, 0) rectangle +(0, 1);
+    \draw[thick]  (6.4, 0) rectangle +(0, 1);
+    \draw  (7.0, 0) rectangle +(0, 1);
+    \draw  (4.6, 0) rectangle +(0, 1);
+    \draw  (11, 0) rectangle +(0, 1);
+
+    % déco autour
+    \node at (-0.5, 0.5) {$S_i$};
+    \draw[<->] (0, 1.5) -- node[above] {64 bits} +(6.375, 0);
+    \draw[<->] (6.425, 1.5) -- node[above] {64 bits} +(6.375, 0);
+  \end{scope}
+
+  % T_i
+  \begin{scope}[xshift=4.6cm, yshift=-1.5cm]    
+    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
+    \draw[thick]  (0, 0) rectangle +(2.4, 1);
+    \draw[]  (1.8, 0) rectangle +(0, 1);
+    \path (2.4, 0) rectangle node {00000000 \dots 00000000} (6.4, 1);
+
+    % déco
+    \node at (-0.5, 0.5) {$T_i$};
+    \draw[<->] (0, -0.5) -- node[below] {$\ell$ bits} +(1.775, 0);
+    \draw[<->] (1.8, -0.5) -- node[below] {6} +(0.6, 0);
+    \draw[<->] (0, -1.5) -- node[below] {64 bits} +(6.375, 0);
+  \end{scope}
+
+  % flèches S_i --> T_i
+  \draw[ultra thick,->] (5.5, 0.5) -- +(0, -1.5);
+  \draw[ultra thick,->] (6.7, 0.5) -- +(0, -1.5);
+
+\end{tikzpicture}
+
 \subsection{Part 1 : Premier filtre}
@@
 
-    \item We get the correspondant $\Delta S'_0S'_1$ ($S'$ dont on aurait retiré que ) computing $\texttt{CVP}(r * 2^{128 - \ell}, \mathcal{L}(A,2^{128 - \ell},m))$.
+    \item We get the correspondant $\Delta S'_0S'_1$ ($S'$ dont on aurait retiré que la composante en $C[0 : \ell]$ et non en $C$ entier) computing $\texttt{CVP}(r * 2^{128 - \ell}, \mathcal{L}(A,2^{128 - \ell},m))$.
 We use the fpylll python library CVP function.
@@
     \item \todo{parler du problème de retenue étrange}
+
   \end{enumerate}
@@
 \end{document}
+
+% Charles' emacs magic commands
+%%% Local Variables:
+%%% TeX-command-extra-options: "-shell-escape"
+%%% ispell-local-dictionary: "english"
+%%% eval: (flyspell-mode 1)
+%%% eval: (reftex-mode 1)
+%%% End:

### 2020-05-15 article/main.tex
lattice en mieux

@@
 
+\section{lattice geometric progression}
+
+\subsection{The lattice}
+
+One of the key point of the algorithm is, given a modulus \(M\) and a given common ration \(\alpha\), to find a modular geometric progression knowing only the most significant bits of its \(n\) first terms.
+
+We call \(L(\alpha,M,n)\) the set of n-tuples \(U = (u_0,\dots u_{n-1})\) satisfying :
+
+\[u_{i} = \alpha u_{i-1} \mod M \] for all \(i\) in \(\{1,\dots,n-1\}\).
+
+We easily notice that, if \(U\) and \(V\) are in \(L(\alpha,M,n)\), so are \(-U\) and \(U+V\), hence \(L(\alpha,M,n)\) is a lattice.
+
+The lattice \(L(\alpha,M,n)\) is spawned by the lines of the following \(n\times n\) matrix :
+
+\[\begin{pmatrix}
+1&\alpha&\alpha^2&\dots&\alpha^{n-1}\\
+0&M&0&\dots&0\\
+0&0&M&\dots&0\\
+\dots&\dots&\dots&\dots&\dots\\
+0&0&0&\dots&M\\
+\end{pmatrix}\]
+
+We want to find an element \(U\) in the lattice only knowing the most significant bits of each of its coordinates.
+
+We decompose \(U = T + (T-U)\) where \(T\) contains only the upper bits and \(T-U\) the lowers one.
+There is no reason for \(T\) to be in the lattice, but as \(T-U\) is small, we know \(T\) is \emph{close} to \(U\).
+So, to find \(U\), we will solve CVP(\(L(\alpha,M,n)\),\(T\)).
+
+\subsection{A first way to solve the CVP : the Babai's rounding}
+
+We compute \(G\) the LLL-reduction of the generating matrix of \(L(\alpha,M,n)\).
+
+As \(T\) is not an element of the lattice, \(TG^{-1}\) is not an integer vector.
+
+We fix \(r = rounding(TG^{-1}) \), then \(rG\) is an element of the lattice. What can we say about the distance between \(rG\) and \(U\) ?
+
+\begin{align*}
+\lVert U - rG \rVert &= \lVert U - (r-TG^{-1} + TG^{-1})G \rVert\\
+&= \lVert U - T + (r-TG^{-1})G \rVert\\
+&\leqslant \lVert U - T \rVert + \lVert(r-TG^{-1})\rVert \times \lVert G\rVert\\	
+\end{align*}
+
+By definition, \(r\) is the closest integer vector to \(TG^{-1}\). But as \(U\) is an element of the lattice, \(UG^{-1}\) is an integer vector. Thus \(r-TG^{-1}\) is smaller than \(UG^{-1}-TG^{-1}\).
+
+Hence :
+\begin{align*}
+\lVert U - rG \rVert &\leqslant \lVert U - T \rVert + \lVert(r-TG^{-1})\rVert \times \lVert G\rVert\\	
+&\leqslant \lVert U - T \rVert + \lVert(UG^{-1}-TG^{-1})\rVert \times \lVert G\rVert\\	
+&\leqslant \lVert U - T \rVert + \lVert(U-Y)\rVert \times \lVert G^{-1} \rVert  \times \lVert G\rVert\\
+& 	\leqslant \lVert U - T \rVert \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert )\\
+\end{align*}
+
+If we can prove that the left side of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.
+
 \section{C connu}
@@
 
-\section{babairounding}
+\subsection{preuve cas facile}
 
-We call \(S't\) the truncated vector : \(S'[\ell,64+\ell]\).
+The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero.
 
-We consider the case were we successfully guessed the 6 upper bits of each \(S_i\) and the \(\ell\) lower bits of \(S_0\). Then we know the \(6+\ell\) upper bits of \(S't\).
+We call \(U\) the truncated vector : \(S'[\ell,64+\ell]\). We notice that \(U\) is in \(L(A,2^{64},n)\).
 
-(à détailer ?)
+We consider the case were we successfully guessed the 6 upper bits of each \(S_i\) and the \(\ell\) lower bits of \(S_0\). Then we know the \(6+\ell\) upper bits of \(U\).
 
-We had define \(S'\) such that \(S'\) is in the lattice \(L(A,2^{128},n)\) and its \(\ell \) lower bits are 0. Then \(S't\) is in \(L(1,2^{64},n)\).
+(maybe dessin)
 
-If we consider the vector \(T'\) computed in step (numerodelastepenquestion), we notice it is close to \(S't\) in the sense of they have exactly the same \(6+\ell\) upper bits.
-So we want to recover \(S't\) by solving an approach \(CVP(L(A,2^{64},n),T')\).
+We consider the vector \(T'\) computed in step (numerodelastepenquestion). To recover \(U\), we use the Babai's rounding explain in section (insererlasection) on \(T'\) and \(L(A,2^{64},n)\).
 
-Now we need to prove that the \(pwet\) obtained with the Babai rounding is indeed \(S't\).
+As \(U\) and \(T'\) have exactly the same \(6+\ell\) upper bits  know that \(\lVert U-T \rVert \leqslant \sqrt{n2^{2(58-l)}}\).
 
-We denote \(r = rounding(YG^{-1}) \).
+So, to be sure that our algorithm return the vector \(U\), we need to find sets of parameters \((n,\ell)\) such that :
 
-\begin{align*}
-\lVert S't - pwet \rVert &= \lVert S't - rG \rVert \\
-&= \lVert S't - (r-YG^{-1} + YG^{-1})G \rVert\\
-&= \lVert S't - Y + (r-YG^{-1})G \rVert\\
-&\leqslant \lVert S't - Y \rVert + \lVert(r-YG^{-1})\rVert \times \lVert G\rVert\\	
-\end{align*}
+\[\sqrt{n2^{2(58-\ell)}} \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert) \leqslant \lVert SVP(L(A,2^{64},n))\rVert \]
 
-By definition, \(r\) is the closest integer vector to \(YG^{-1}\).
-But as \(S't\) is an element of the lattice, \(S'tG^{-1}\) is an integer vector.
-Thus \(r-YG^{-1}\) is smaller than \(S'tG^{-1}-YG^{-1}\).
+\begin{itemize}
+	\item For \(n=3\)
 
-Hence :
-\begin{align*}
-\lVert S't - pwet \rVert &\leqslant \lVert S't - Y \rVert + \lVert(r-YG^{-1})\rVert \times \lVert G\rVert\\	
-&\leqslant \lVert S't - Y \rVert + \lVert(S'tG^{-1}-YG^{-1})\rVert \times \lVert G\rVert\\	
-&\leqslant \lVert S't - Y \rVert + \lVert(S't-Y)\rVert \times \lVert G^{-1} \rVert  \times \lVert G\rVert\\
-& 	\leqslant \lVert S't - Y \rVert \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert )\\
-\end{align*}
+	\(G = \begin{pmatrix}
+	-1241281756092&3827459685972&-728312298332\\
+	-5001120657083&-2117155768935&5479732607037\\
+	8655886039732&3303731088004&6319848582548\\
+	\end{pmatrix}\)
 
-But, by construction of \(Y\), we have that \(\lVert S't - Y \rVert \leqslant \sqrt{n2^{2(58-l)}} \). So
+	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 2.869065991049632\)
 
-\[ \lVert S't - pwet \rVert \leqslant \sqrt{n2^{2(58-l)}}\times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert )\]\\
+	\(\lVert SVP(G) \rVert = 4.08909120094543e{^12}\)
 
-Now all we need to do is find sets of parameters \((n,l)\) such that the following inequality is satisfied: 
-\[\sqrt{n2^{2(58-\ell)}} \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert) \leqslant \lVert SVP(L(A,2^{64},n))\rVert \].
+	Hence, the smallest l such that the inequality is satisfied is \(\ell = 19\) %17 avec borne CVP
 
-We notice that only the left term of the inequality depends on \(\ell\).
+	\item For \(n=4\)
 
-Hence, some sets of parameters that satisfies this inequality are :
-\begin{itemize}
-	\item \(n = 3\) and \(\ell \geqslant 19\)
-	\item \(n = 4\) and \(\ell \geqslant 13\)
-	\item \(n = 5\) and \(\ell \geqslant 11\)
-	\item \(n = 6\) and \(\ell \geqslant 8\)
+	\(G = \begin{pmatrix}
+	-186304953996472&-126056243766680&7937589136904&93078431381544\\
+	-216211368070119&99587582169277&-214303762177807&-1707551230219\\
+	110964501361298&-5646098666150&-268280113597118&149382085707466\\
+	131252974561432&-233919070109448&-98716819647784&-134620659538888\\
+	\end{pmatrix}\)
+
+	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 2.060717188416003\)
+
+	\(\lVert SVP(G) \rVert = 2.43569932844650e^{14}\)
+
+	Hence, the smallest l such that the inequality is satisfied is \(\ell = 13\) %11 avec borne CVP
+
+	\item For \(n=5\)
+
+	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 3.7692142154392747\)
+
+	\(\lVert SVP(G) \rVert = 1.71669732658884e^{15}\)
+
+	Hence, the smallest l such that the inequality is satisfied is \(\ell = 11\)  %9 avec borne CVP
+
+	\item For \(n=6\)
+
+	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 2.6925925483728537\)
+
+	\(\lVert SVP(G) \rVert =1.02550589786192e^{16}\)
+
+	Hence, the smallest l such that the inequality is satisfied is \(\ell = 8\) %7 avec borne CVP
 \end{itemize}

### 2020-05-15 article/main.tex
section 2

@@
 \hypersetup{colorlinks=true,citecolor=red,linkcolor=red,urlcolor=black}
+
+\DeclareMathSymbol{:}{\mathord}{operators}{"3A}
 
@@
 
-\todo{cette section doit décrire précisément PCG.}
+This section describes the \textsf{PCG64} non-cryptographic pseudo-random number
+generator (more precisely \textsf{PCG-XSL-RR} in the designer's terminology). It
+has an internal state of 128-bit, which operate as a linear congruential
+generator modulo~$2^{128}$. More precisely:
+\[
+  S_{i+1} = a S_i + c \bmod 2^{128},
+\]
+Where $a = 47026247687942121848144207491837523525$ is a fixed 126-bit
+constant. The first initial state $S_0$ is the seed of the generator. The
+increment $c$ may be specified by the user of the PRNG to produce different
+output streams with the same seed (just as the IV acts in a stream cipher). If
+no value of $c$ is specified, then a default increment is provided:
+$c = 117397592171526113268558934119004209487$.
 
-\begin{quote}
-  In addition, one of the generators, PCG-XSL-RR(described in Section 6.3.3), is
-  explicitly designed to make any attempt at state reconstruction especially
-  difficult, using xor folding tominimize the amount of information about
-  internal state that leaks out.
-\end{quote}
+Each time the PRNG is clocked, 64 output bits are extracted from the internal
+state using a non-linear function that makes use of data-dependent rotations, in
+a way reminiscent of the \textsf{RC5} block cipher~\cite{Rivest94}. The six most
+significant bits of the internal state encode a number $0 \leq r < 64$. The two
+64-bit halves of the internal state are XORed together, and this 64-bit result
+is rotated right by $r$ positions. The successive 64-bit outputs of the
+generator are $X_0, X_1, \dots$ where:
+\[
+  X_i =(S_i[0:64] \oplus S_i[64:128]) \ggg S_i[122:128].
+\]
 
-\begin{quote}
-  Although these properties make it extremely challenging for an agent observing
-  theexternal behavior of a program to guess the state of its random number
-  generator,
-\end{quote}
+Figure~\ref{pcg128out} summarizes the process. The overall design strategy is
+similar to that of a filtered LFSR: the successive states of a weak internal
+generator with a strong algebraic structure are ``filtered'' by a non-linear
+function.
 
-% import numpy
-% print(numpy.random.default_rng())
-% ---> Generator(PCG64)
+Updating the internal state requires a $128 \times 128 \rightarrow 128$
+multiplication operation. In fact, this can be done with two
+$64 \times 64 \rightarrow 64$ multiplications, one
+$64 \times 64 \rightarrow 128$ multiplication and two 64-bit additions. High-end
+desktop CPUs all implement these operations in hardware, so the generator is
+relatively fast on these platforms.
 
-\subsection{LCG algorithm}
+\begin{figure}
+  \begin{center}
+\begin{tikzpicture}[scale=0.66]
+%  \draw[red, use as bounding box] (-1.25, -6.75) rectangle (12.8, 2.25);
 
-The PCG algorithm we studied is based on the Linear Congruential Generators algorithm.
-The LCG algorithm is simply based on a congruential multiplication.
-This type of generator is based on four values: an internal state $S_{i}$, a multiplier $A$, an increment $C$ and the size $M$.
-Generally, to initialize a generator of a given size, a seed value is provided as the first internal state, and a default multiplier and increment are used.
-We will study the case where $M$ and $A$ are known and where $M$ is a power of 2.\\
+  % S_i
 
-Calling the generator results in calculating the next internal state $S_{i+1} = A \times S_i + C \mod{M}$ and outputting $S_{i+1}$ as data.
+    % bordures
+    \draw[very thick]  (0, 0) rectangle (12.8, 1);
+    \draw  (0.6, 0) rectangle +(0, 1);
+    \draw[very thick]  (6.4, 0) rectangle +(0, 1);
+    \foreach \i in {0, 1, ..., 128} \draw (\i/10, 0) -- +(0, 0.1);
 
-This type of generator is trivially predictable because, the internal state is given to us at each output, and in the event that $C$ is unknown, we can still deduce it with two consecutive outputs: $C = S_{i+1} - A \times S_i$.\\
+    % déco autour
+    \node[draw] at (6.4, 3) (update) {$\times a + c \bmod 2^{128}$};
+    \draw[thick,->] (12.8, 0.5) -- (13.8, 0.5) node[below right] {$S_i$} -- (13.8, 3) -- (update);
+    \draw (13.7, 2.5) -- +(0.2, -0.2);
+    \path (13.9, 2.5) node[anchor=west] {128};
 
-A better variant of the standard LCG is the Truncated LCG, where the output consists only of the higher bits of the internal state $S_i$.
-This way the entire internal state isn't accessible.
-Returning the lower bits is not interesting because as we can understand easily and will prove later, lower bits of a state only depend on the lower bits of the previous state.
+    \draw[thick,->] (update) -- (-2, 3) -- (-2, 0.5) node[below left] {$S_{i+1}$} -- (0, 0.5);
+    \draw (-2.1, 2.5) -- +(0.2, -0.2);
+    \path (-1.9, 2.5) node[anchor=west] {128};
 
-\subsection{PCG algorithm}
+    \node[font=\footnotesize,anchor=east] at (12.9, -0.33) {0};
+    \node[font=\footnotesize,anchor=east] at (6.5, -0.33) {64};
+    \node[font=\footnotesize,anchor=west] at (0.5, -0.33) {122};
+    \node[font=\footnotesize] at (-0.25, -0.33) {128};
 
-The PCG algorithm we studied is based on the LCG algorithm. To start, here are a few definitions:
-\begin{itemize}
-    \item let $k$ be the size of our generator, in this case $k = 128$ bits;
-    \item let $M$ be the modulus with $M = 2^k$;
-    \item let $S_i$ be the internal state at the $i$-th iteration and $S$ be the list of $S_i$;
-    \item let $X_i$ be the output at the $i$-th iteration of the generator and $X$ be the list of $X_i$;
-    \item let $lowS_i$ be the $k/2$ least significant bits of $S_i$;
-    \item let $upS_i$ be the $k/2$ most significant bits of $S_i$;
-    \item let $rot_i$ be the 6 most significant bits of $S_i$;
-    \item let $nbiter$ be the number of consecutive outputs for prediction purposes.
-\end{itemize}
+    \draw (6.4, -2) node (x) {$\bigoplus$};
 
-The internal state can be redefined as:
-\begin{equation}
-    S_i = 2^{k/2} \times upS_i + lowS_i 
-\end{equation}
+    \draw[->] (3.2, -0.1) |- (x);
+    \draw (3.1, -0.9) -- +(0.2, -0.2);
+    \path (3.3, -1) node[anchor=west] {64};
 
-The 6 most significant bits, $rot_i$, of the internal state are used as a value to determine the rotation in the PCG algorithm.
-Let \texttt{rotate} be the function that performs said rotation (Figure \ref{pcg128out}).\\
-The rotation is a simple exchange of bits, where a $k/2$ bit value is split into 2 parts: one of $rot_i$ bits and the other in $(k/2) - rot_i$ bits.
-The two parts are then swapped, and the new value is returned.
+    \draw[->] (9.6, -0.1) |- (x);
+    \draw (9.5, -0.9) -- +(0.2, -0.2);
+    \path (9.7, -1) node[anchor=west] {64};
 
-This generator is initialised with a given $k$ bit internal state $S_0$ (of size 128 bits in the case of PCG128), and optionally with an $k$ bit increment $C$.
-If no increment is given, the algorithm will use a known default increment.
-For each output of the generator, the algorithm will perform the following calculations" (Figure \ref{pcg128out}):\\
-\begin{align}
-    X_{i} &= \mathtt{rotate}(upS_i\ \oplus\ lowS_i)\\
-    S_{i+1} &= A \times S_{i} + C \mod{2^k}
-\end{align}
-In the case of PCG128, $A$ is a default increment of size 125 bits prime with $2^k$.
+    \node[minimum width=1.8cm] at (6.4, -4) (r) {$\ggg$};
+    \draw (5.8, -3.6) rectangle +(1.2, -0.8);
+    \draw[fill=black] (5.8, -4.1) -- (5.95, -4) -- (5.8, -3.9) -- cycle;
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.75\linewidth]{pictures/PCG128.png}
-    \caption{PCG128 : Output process}
+    \draw (0.3, 1.1) -- (0.3, 1.5) -- (-1, 1.5) -- (-1, -4) -- (5.8, -4);
+    \draw (-1.1, -2.9) -- +(0.2, -0.2);
+    \path (-0.9, -3) node[anchor=west] {6};
+
+    \draw[->] (x) -- +(0, -1.5);
+    \node at (6.4, -6.25) (xi) {$X_i$};
+    \draw[->] (6.4, -4.4) -- (xi);
+
+    \draw (6.3, -4.9) -- +(0.2, -0.2);
+    \path (6.5, -5) node[anchor=west] {64};
+  \end{tikzpicture}
+\end{center}
+\caption{\textsf{PCG64}: Internal state update and output process.}
 \label{pcg128out}
@@
 
-Given the nature of the generator, the following values are known at every step:
-\begin{itemize}
-    \item the size of the generator $k$;
-    \item the list of outputs $X$;
-    \item the multiplier $A$;
-    \item and eventually $C$, if the default increment was used. 
-\end{itemize}
-
-To break this generator, we had to differentiate between cases when the default $C$ was used or not. It was also interesting to note what was the minimum number of $X_i$ needed to reliably attack it.
-
 \section{Algorithm of Frieze \textit{et al}}
 
-\todo{Comme on utilise à plusieurs reprise l'idée de se ramener à un LCG
-  tronqué, autant décrire la technique à base de réseaux un bon coup.}
+\todo{à fusionnier avec la section suivante!}
 Pour casser PCG, nous avons utilisé la méthode présentée par Alan M.
@@
 
-\section{lattice geometric progression}
-
+\section{Reconstructing Truncated Geometric Sequences}
+\todo{à fusionnier avec la section précédente!}
 \subsection{The lattice}
@@
 
-\subsection{A first way to solve the CVP : the Babai's rounding}
+\subsection{A first way to solve the CVP : Babai's Round-off Algorithm}
 
@@
 
-By definition, \(r\) is the closest integer vector to \(TG^{-1}\). But as \(U\) is an element of the lattice, \(UG^{-1}\) is an integer vector. Thus \(r-TG^{-1}\) is smaller than \(UG^{-1}-TG^{-1}\).
+By definition, \(r\) is the closest integer vector to \(TG^{-1}\). But as \(U\) is an element of the lattice, \(UG^{-1}\) is an integer vector. Thus \(r-TG^{-1}\) is shorter than \(UG^{-1}-TG^{-1}\).
 
@@
 
-If we can prove that the left side of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.
+If we can prove that the left side \todo{right side?} of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.
 

### 2020-05-20 article/main.tex
retouche + preuve cas dur

@@
 There is no reason for \(T\) to be in the lattice, but as \(T-U\) is small, we know \(T\) is \emph{close} to \(U\).
-So, to find \(U\), we will solve CVP(\(L(\alpha,M,n)\),\(T\)).
 
-\subsection{A first way to solve the CVP : Babai's Round-off Algorithm}
+\subsection{A first way to find U : the Babai's rounding}
 
@@
 
-If we can prove that the left side \todo{right side?} of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.
+If we can prove that the right side \todo{je confonds ma gauche et ma droite et je vous zut :p} of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.
+
+ \subsection{A second way to find U : using an exact CVP solver}
+
+Solving exactly the CVP is a hard computational problem. But as we do not manipulate quantities too big, we still can use an exact CVP solver.
+
+We know that \(U\) is in the lattice, \(T\) is not and \(|T-U|\) is small.
+Then we can hope for \(U\) to be the closet vector of \(T\) which is in the lattice : the vector \(U\) may be the solution of \(CVP(L(\alpha,M,n),T)\).
+
+What can we say about \(|CVP(L(\alpha,M,n),T)-U|\) ?
+
+\begin{align*}
+|CVP(L(\alpha,M,n),T)-U| &= |CVP(L(\alpha,M,n),T)-T+T-U|\\
+& \leqslant |CVP(L(\alpha,M,n),T)-T|+|T-U|\\
+\end{align*}
+
+But, as the vector \(U\) is in the lattice and by definition of the CVP, \(|CVP(L(\alpha,M,n),T)-T| \leqslant |T-U| \), hence 
+\[|CVP(L(\alpha,M,n),T)-U|\leqslant 2|T-U|\]
+
+If we can prove that the right side of the inequality is smaller than the SVP of the lattice, then wa have proved that \(CVP(L(\alpha,M,n),T)\) is indeed the vector we were searching for.
 
@@
 
+\subsection{preuve cas difficile}
+\todo{À ranger au bon endroit}
+% Babai ça marche pas (genre zero)
+
+The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero.
+
+We call \(U\) the truncated vector : \(S'[\ell,128]\). We notice that \(U\) is in \(L(A,2^{128-\ell},n)\).
+
+We consider the case were we successfully guessed the 6 upper bits of each \(S_i\) and the \(\ell\) lower bits of \(S_0\). Then we know the 6 upper bits of \(U\).
+
+(maybe dessin)
+
+We consider the vector \(T'\) computed in step (numerodelastepenquestion). To recover \(U\), we use an exact CVP solver as explain in  (insererlasection) on \(T'\) and \(L(A,2^{128-\ell},n)\).
+
+As \(U\) and \(T'\) have exactly the same \(6\) upper bits  know that \(\lVert U-T \rVert \leqslant \sqrt{n2^{2(128-6-\ell)}}\).
+
+So, to be sure that our algorithm return the vector \(U\), we need to find sets of parameters \((n,\ell)\) such that :
+
+\[2\sqrt{n}^{(128-6-\ell)} \leqslant \lVert SVP(L(A,2^{128-\ell},n))\rVert \]
+
+\begin{itemize}
+	\item For \(\ell=11\), the smallest \(n\) we need is 39.
+
+	\item For \(\ell=12\), the smallest \(n\) we need is 38.
+
+	\item For \(\ell=13\), the smallest \(n\) we need is again 38 .
+
+\end{itemize}
+
 \begin{theorem}

### 2020-05-20 article/main.tex
dessin + début section 5

@@
-\documentclass[preprint]{iacrtrans}
+\documentclass[preprint,svgnames]{iacrtrans}
 
@@
 generator are $X_0, X_1, \dots$ where:
-\[
+\begin{equation}\label{eq:output}
   X_i =(S_i[0:64] \oplus S_i[64:128]) \ggg S_i[122:128].
-\]
+\end{equation}
 
@@
 
-\section{C connu}
+\section{Warm-up: A Prediction Algorithm for \textsf{PCG64} With Known Increment}
 
-\subsection{Réduction}
-Pour deviner l'état interne du générateur, une méthode consiste à procéder par brute-force.
-On devine les 64 bits supérieurs de $S_0$, on en déduit la rotation $r_0$ utilisée, on peut retrouver  les 64 bits inférieurs:
+We first consider the easier case where the ``increment'' (the $c$ term in the
+definition of the underlying linear congruential generator) is known --- recall
+that a default value is specified in case the user of the pseudo-random
+generator does not want to provide one.
 
-\begin{equation}
-   S_0[0:64] = S_0[64:128] \oplus \texttt{ROT}(X_0 , 64 - r_0) \\
+In this case, reconstructing the 128-bit internal state $S_i$ of the generator
+obis sufficient to produce the pseudo-random flow with 100\% accuracy (the
+generator can also be clocked backwards if necessary, so that the seed can be
+easily reconstructed). We therefore focus on $S_0$, the first internal state
+after seeding. A very simple strategy could be the following:
+\begin{enumerate}
+\item Guess the 64 upper bits of $S_0$.
+\item Compute the missing 64 lower bits using~\eqref{eq:output}, with:
+\[
+   S_0[0:64] = S_0[64:128] \oplus (X_0  \lll S[122:128]).
+\]
+\item If $X_1$ is computed correctly, output $S_0$.
+\end{enumerate}
+
+This requires $2^{64}$ iterations of a loop that does a dozen arithmetic
+operations. This is the baseline that we improve upon. An improved
+``guess-and-determine'' prediction algorithm is possible, which essentially
+amounts to expose a truncated version of the underlying linear congruential
+generator, and attack it using the well-known tools. This is possible by
+combining the following ingredients:
+\begin{itemize}
+\item The underlying linear congruential generator uses a power-of-two modulus,
+  therefore the $\ell$ low-order bits of $S_{i+1}$ are entirely determined by
+  the $\ell$ low-order bits of $S_i$. More precisely, we have:
+  \begin{equation}\label{eq:lcg}
+    S_{i+1} = aS_i + c \bmod 2^\ell, \qquad \text{for all } 0 \leq \ell \leq 128
   \end{equation}
+  Therefore, guessing low-order bits of $S_0$ yields a ``long-term advantage''
+  that holds for all subsequent states.
 
-Cette méthode demanderait de l'ordre de $2^{64}$ calculs, ce qui n'est pas acceptable.
-Nous pouvons cependant utiliser cette équation pour déduire des informations sur l'état interne de S.
-En effet,en devinant par brute-force les 6 bits du haut de S, et ses $\ell$ bits du bas, on récupère :
+\item Guessing a 6-bit rotation gives access to the XOR of the two halves of the
+  internal state. If a part of the state is known, then this transfers existing
+  knowledge to the other half.
+\end{itemize}
 
-\todo{schéma "relevant bits of the outputs"}
+In figure~\ref{fig:Cknown}, we see that guessing $S_0[0:\ell]$ and a few 6-bit
+rotations give access to $S_i[58:64+\ell]$ for the corresponding consecutive
+states. Therefore, we are facing a truncated linear congruential generator on
+$64+\ell$ bits, where we only know $6+\ell$ bits of output, for a few
+consecutive outputs. This is sufficient to reconstruct entirely the successive
+states of this truncated linear congruential generator. This reveals
+$S_0[0:64+\ell]$, and using~\eqref{eq:output} the entire $S_0$ can be
+reconstructed. The precise details follow.
 
-On réduit $S$ à une série géométrique plus simple.
-Notons $S'$ l'état interne auquel on retire sa composante en $C$, et en $S_0[0:\ell]$ (note de bas de page : le fait de retirer $S_0[0:\ell]$ n'est pas nécéssaire, mais nous permet de limiter la taille des instances utilisées à 64 bits, très utile pour la programmation en C).
-Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, nommé $\mathcal{L}(A,2^64,n)$.
-On se réduit donc au cas facile étudié dans la partie précédente.
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[yscale=0.66]
+  \path[red, use as bounding box] (-0.75, -9.5) rectangle (12.8, 2.25);
+
+  % S_0
+  \begin{scope}
+    % remplissage
+    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
+    \fill[fill=magenta] (0, 0) rectangle +(0.6, 1);
+    \fill[fill=magenta] (11, 0) rectangle  +(1.8, 1);
+
+    % bordures
+    \draw[thick]  (0, 0) rectangle (12.8, 1);
+    \draw  (0.6, 0) rectangle +(0, 1);
+    \draw[thick]  (6.4, 0) rectangle +(0, 1);
+    \draw  (7.0, 0) rectangle +(0, 1);
+    \draw  (4.6, 0) rectangle +(0, 1);
+    \draw  (11, 0) rectangle +(0, 1);
+
+    % déco autour
+    \node at (-0.5, 0.5) {$S_0$};
+    \draw[<->] (0, 1.5) -- node[above] {64 bits} +(6.375, 0);
+    \draw[<->] (6.425, 1.5) -- node[above] {64 bits} +(6.375, 0);
+  \end{scope}
+
+  % T_0
+  \begin{scope}[xshift=4.6cm, yshift=-1.5cm]    
+    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
+    \draw[thick]  (0, 0) rectangle +(2.4, 1);
+    \draw[]  (1.8, 0) rectangle +(0, 1);
+    \path (2.4, 0) rectangle node {$00000000 \cdots 00000000$} (6.4, 1);
+
+    % déco
+    \node at (-0.5, 0.5) {$T_0$};
+  \end{scope}
+
+  % flèches S_i --> T_i
+  \draw[ultra thick,->] (5.5, 0.5) -- +(0, -1.5);
+  \draw[ultra thick,->] (6.7, 0.5) -- +(0, -1.5);
+
+  %%%%%%%%%
+
+  % S_1
+  \begin{scope}[yshift=-3cm]
+    % remplissage
+    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
+    \fill[fill=magenta] (0, 0) rectangle +(0.6, 1);
+    \fill[fill=cyan] (11, 0) rectangle +(1.8, 1);
+
+    % bordures
+    \draw[thick]  (0, 0) rectangle (12.8, 1);
+    \draw  (0.6, 0) rectangle +(0, 1);
+    \draw[thick]  (6.4, 0) rectangle +(0, 1);
+    \draw  (7.0, 0) rectangle +(0, 1);
+    \draw  (4.6, 0) rectangle +(0, 1);
+    \draw  (11, 0) rectangle +(0, 1);
+
+    % déco autour
+    \node at (-0.5, 0.5) {$S_1$};
+
+    % flèches S_i --> T_i
+    \draw[ultra thick,->] (5.5, 0.5) -- +(0, -1.5);
+    \draw[ultra thick,->] (6.7, 0.5) -- +(0, -1.5);
+  \end{scope}
+
+  % T_1
+  \begin{scope}[xshift=4.6cm, yshift=-4.5cm]    
+    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
+    \draw[thick]  (0, 0) rectangle +(2.4, 1);
+    \draw[]  (1.8, 0) rectangle +(0, 1);
+    \path (2.4, 0) rectangle node {$00000000 \cdots 00000000$} (6.4, 1);
+
+    % déco
+    \node at (-0.5, 0.5) {$T_1$};
+  \end{scope}
+
+  %%%%%%%%%%%%%
+
+  % S_2
+  \begin{scope}[yshift=-6cm]
+    % remplissage
+    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
+    \fill[fill=magenta] (0, 0) rectangle +(0.6, 1);
+    \fill[fill=cyan] (11, 0) rectangle +(1.8, 1);
+
+    % bordures
+    \draw[thick]  (0, 0) rectangle (12.8, 1);
+    \draw  (0.6, 0) rectangle +(0, 1);
+    \draw[thick]  (6.4, 0) rectangle +(0, 1);
+    \draw  (7.0, 0) rectangle +(0, 1);
+    \draw  (4.6, 0) rectangle +(0, 1);
+    \draw  (11, 0) rectangle +(0, 1);
+
+    % déco autour
+    \node at (-0.5, 0.5) {$S_2$};
+
+    % flèches S_i --> T_i
+    \draw[ultra thick,->] (5.5, 0.5) -- +(0, -1.5);
+    \draw[ultra thick,->] (6.7, 0.5) -- +(0, -1.5);
+  \end{scope}
+
+  % T_2
+  \begin{scope}[xshift=4.6cm, yshift=-7.5cm]    
+    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
+    \draw[thick]  (0, 0) rectangle +(2.4, 1);
+    \draw[]  (1.8, 0) rectangle +(0, 1);
+    \path (2.4, 0) rectangle node {$00000000 \cdots 00000000$} (6.4, 1);
+
+    % déco
+    \node at (-0.5, 0.5) {$T_2$};
+    \draw[<->] (0, -0.5) -- node[below] {$\ell$ bits} +(1.775, 0);
+    \draw[<->] (1.8, -0.5) -- node[below] {6} +(0.6, 0);
+    \draw[<->] (0, -1.25) -- node[below] {64 bits} +(6.4, 0);
+    \draw[<->] (6.4, -1.25) -- node[below] {$\ell$ bits} +(1.775, 0);
+  \end{scope}
+\end{tikzpicture}
+
+\end{center}
+\caption{A guess-and-determine algorithm to reconstruct the first internal state
+  $S_0$. Magenta bits are guessed; cyan bits are obtained using the linear
+  congruence relation~\eqref{eq:lcg}; yellow bits are obtained
+  from the output and the guessed rotations using~\eqref{eq:output}.}
+\label{fig:Cknown}
+\end{figure}
+
+Set $S_i' = S_i - (S_0[0:\ell] + c) \sum_{j=0}^i a^i$. It is easy to check that
+$S'_{i+1} = a S'_i$, and therefore the successive $S'_i$'s follow a geometric
+progression\footnote{note de bas de page : le fait de retirer $S_0[0:\ell]$
+  n'est pas nécéssaire, mais nous permet de limiter la taille des instances
+  utilisées à 64 bits, très utile pour la programmation en C)}.
+
+Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, nommé $\mathcal{L}(A,2^64,n)$. On se réduit donc au cas facile étudié dans la partie précédente.
 \begin{equation}
@@
 
-\subsection{Algorithme}
 On calcule préalablement les matrices $G$ et $G^{-1}$, avec $G$ la matrice génératrice réduite du réseau $\mathcal{L}(A,2^64,n)$ par l'algorithme LLL.

### 2020-05-20 article/main.tex
Correction de la formule de S'
erreur de signe corrigée
Encore un petit truc que j'ai oublié de corriger
Rajout

@@
 
-Set $S_i' = S_i - (S_0[0:\ell] + c) \sum_{j=0}^i a^i$. It is easy to check that
+Set $S_i' = S_i - S_0[0:\ell] a^i - c (\sum_{j=0}^{i-1} a^i)$. It is easy to check that
 $S'_{i+1} = a S'_i$, and therefore the successive $S'_i$'s follow a geometric
-progression\footnote{note de bas de page : le fait de retirer $S_0[0:\ell]$
+progression. Indeed :
+\begin{align*}
+S_i = S_{i-1} a + c
+S_i = S_{i-2} a^2 + (a + 1) c
+S_i = S_0 a^i + (\sum_{j=0}^{i-1} a^i) c
+S_i - (\sum_{j=0}^{i-1} a^i) c - S_0[0:\ell] a^i = (S_0 - S_0[0:\ell]) a^i 
+\end{align*}
+\footnote{note de bas de page : le fait de retirer $S_0[0:\ell]$
   n'est pas nécéssaire, mais nous permet de limiter la taille des instances

### 2020-05-22 article/main.tex
algo c known v1.

--- 2020-05-20 article/main.tex
@@
 \section{Reconstructing Truncated Geometric Sequences}
+\label{sec:geometric}
+
 \todo{à fusionnier avec la section précédente!}
@@
 In this case, reconstructing the 128-bit internal state $S_i$ of the generator
-obis sufficient to produce the pseudo-random flow with 100\% accuracy (the
+is sufficient to produce the pseudo-random flow with 100\% accuracy (the
 generator can also be clocked backwards if necessary, so that the seed can be
-easily reconstructed). We therefore focus on $S_0$, the first internal state
-after seeding. A very simple strategy could be the following:
+easily reconstructed). We therefore focus on reconstructing $S_0$ (the seed)
+from $X_0, X_1, X_2, \dots$. A very simple strategy could be the following:
 \begin{enumerate}
@@
 In figure~\ref{fig:Cknown}, we see that guessing $S_0[0:\ell]$ and a few 6-bit
-rotations give access to $S_i[58:64+\ell]$ for the corresponding consecutive
-states. Therefore, we are facing a truncated linear congruential generator on
-$64+\ell$ bits, where we only know $6+\ell$ bits of output, for a few
-consecutive outputs. This is sufficient to reconstruct entirely the successive
-states of this truncated linear congruential generator. This reveals
-$S_0[0:64+\ell]$, and using~\eqref{eq:output} the entire $S_0$ can be
-reconstructed. The precise details follow.
+rotations give access to $S_i[58:64+\ell]$ for the corresponding
+states. Therefore, looking at $S_i[\ell:64+\ell]$, we are facing a truncated
+linear congruential generator on $64$ bits, where we have access to the most
+$6+\ell$ bits of each state, for a few consecutive states. This is sufficient to
+reconstruct entirely the successive states of this truncated linear congruential
+generator. This reveals $S_0[\ell:64+\ell]$, and using~\eqref{eq:output} the
+entire $S_0$ can be reconstructed. The precise details follow.
 
@@
     \draw[]  (1.8, 0) rectangle +(0, 1);
-    \path (2.4, 0) rectangle node {$00000000 \cdots 00000000$} (6.4, 1);
+    \path (2.4, 0) rectangle node {$???????? \cdots ????????$} (6.4, 1);
 
@@
   $S_0$. Magenta bits are guessed; cyan bits are obtained using the linear
-  congruence relation~\eqref{eq:lcg}; yellow bits are obtained
+  congruence relation~\eqref{eq:lcg} modulo $2^\ell$; yellow bits are obtained
   from the output and the guessed rotations using~\eqref{eq:output}.}
@@
 
-Set $S_i' = S_i - (S_0[0:\ell] + c) \sum_{j=0}^i a^i$. It is easy to check that
-$S'_{i+1} = a S'_i$, and therefore the successive $S'_i$'s follow a geometric
-progression\footnote{note de bas de page : le fait de retirer $S_0[0:\ell]$
-  n'est pas nécéssaire, mais nous permet de limiter la taille des instances
-  utilisées à 64 bits, très utile pour la programmation en C)}.
+Set $S_i' = S_i - S_0[0:\ell] a^i - c \sum_{j=0}^{i-1} a^i$. The sequence
+$(S'_i)$ describe the internal states we would observe if $c=0$ and
+$S_0[0:\ell] = 0$. It is easy to check that $S'_{i} = a^i (S_0 - S_0[0:\ell])$,
+and therefore the successive $S'_i$'s follow a geometric progression. % Indeed :
+% \begin{align*}
+% S_i &= a S_{i-1} + c\\
+%     &= a^2 S_{i-2} + (a + 1) c \\
+%     &\vdots  \\
+%     &= a^i S_0  + c \sum_{j=0}^{i-1} a^i \\
+%   S'_i  &= S_i - c \sum_{j=0}^{i-1} a^i - S_0[0:\ell] a^i \\
+%     &= a^i (S_0 - S_0[0:\ell])
+% \end{align*}
+The $\ell$ low-order bits of $S_0 - S_0[0:\ell]$ are all zero by construction;
+therefore set $T_i = \left(S'_i / 2^\ell\right) \bmod 2^{64}$ (the division by
+$2^\ell$ is exact). It follows that the $(T_i)$ is also a geometric progression
+with common ratio $a$, and we have access to the $\ell+6$ high order bits of
+$T_i$ if we guess the $i$-th rotation. We have seen in
+section~\ref{sec:geometric} that 19 known bits on 3 consecutive outputs are
+sufficient for a deterministic reconstruction of the full sequence. Therefore,
+we guess 19 low-order bits of the state and three consecutive rotations.
 
-Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, nommé $\mathcal{L}(A,2^64,n)$. On se réduit donc au cas facile étudié dans la partie précédente.
-\begin{equation}
-genMatrix =
-\begin{pmatrix} 
-1 & A & A ^2 & \dots & A ^{n- 1}\\
-0 & M & 0 & \dots & 0\\
-0 & 0 & M & \dots & 0\\
-\vdots & & & \ddots & \vdots\\
-0 & 0 & \dots & 0 & M
-\end{pmatrix}
-\end{equation}
+Therefore, the algorithm that reconstructs the internal state of the
+\textsf{PCG64} generator with known increment proceeds as follows:
 
-On calcule préalablement les matrices $G$ et $G^{-1}$, avec $G$ la matrice génératrice réduite du réseau $\mathcal{L}(A,2^64,n)$ par l'algorithme LLL.
-Pour tous les $r_i$ et les $S_0[0 : \ell]$ possibles:\\
-\begin{enumerate}
-\item On récupère le $T$ correspondant à nos suppositions.
-On calcule $T'$.
-($T'_i = (S_i[64 - 6 : 64 + \ell] - (A^i*S_0[0 : \ell] + C * \sum_{j = 0}^i A^j)[64 - 6 : 64 + \ell])[64 - 6 : 64 + \ell] * 2^{64 - 6 - \ell}$)
+\begin{center}\noindent\framebox{
+  \begin{minipage}{0.95\textwidth}
+\aalgbegin Algorithm K (State reconstruction with known increment). Reconstruct $S_0$ from three consecutive outputs $X_j$ ($0 \leq j < 3$). In all the algorithm, we have $j= 0, 1, 2$.
 
-\todo{schema T et T'}
+\algstep K1. [Guess state] For each $0 \leq w < 2^{19}$, set $S_0[0 : \ell] = w$ and do steps K2--K10.
 
-Notons cependant que les composantes en $C$ et $S[0:l]$ sont calculées en dehors de la boucle et stockées dans un tableau pour des raisons de performance.
-Ceci est fait de même dans tous les algorithmes suivants.
+\algstep K2. [Propagate state] Set $S_j[0:\ell] \gets a^j w \bmod 2^{19}$.
 
-\item We get the correspondant $S'[\ell : 64 + \ell]$ by using the babai rounding technique on $T'$ using $G$ and $G^{-1}$.
+\algstep K3. [Guess rotations] For each $0 \leq r_0, r_1, r_2 < 64$, do steps K4--K10.
 
-\item We test our solution by computing $S_0$ and testing if the following internals states are coherent with the PCG outputs. If the results are coherents, we return $S_0$.
+\algstep K4. [Undo rotations] Set $S_j[122:128] \gets r_j$ and $Y_j \gets X_j \lll r_j$.
 
-\end{enumerate}
+\algstep K5. [Propagate] Set $S_j[58:64] \gets r_j \oplus Y_j[58:64]$ and $S_j[64:83] \gets S_j[0:19] \oplus Y_j[0:19]$.
+
+\algstep K6. [Build partial $T$] Set $T_j \gets \left(S_j[58:83] - (a^j w + c \sum_{k = 0}^j a^k)[58:83]\right) \cdot 2^{39}$
+
+\algstep K7. [Reconstruct $T$] Set $(U_0, U_1, U_2) \gets \left\lfloor (T_0, T_1, T_2) \cdot \widetilde G^{-1} \right\rceil \cdot \widetilde G$.
+
+\algstep K8. [Reconstruct $S_0$] Set $S_0[19:83] \gets U_0$ and $S_0[83:122] \gets U_0[0:39] \oplus Y_0[19:58]$.
+
+\algstep K9. [Recompute $X_1$] Compute $\widehat S_1 \gets a S_0 + c$ and $\widehat X_1 =(\widehat S_1[0:64] \oplus \widehat S_1[64:128]) \ggg r_1$.
+
+\algstep K10. [Check] If $\widehat{X}_1 = X_1$, the output $S_0$ as a candidate internal state.
+\end{minipage}
+}
+\end{center}
 

### 2020-05-22 article/main.tex
algo v2

@@
     \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
-    \fill[fill=magenta] (0, 0) rectangle +(0.6, 1);
-    \fill[fill=magenta] (11, 0) rectangle  +(1.8, 1);
+    \fill[fill=magenta] (0, 0) rectangle node {$r_0$} +(0.6, 1);
+    \fill[fill=magenta] (11, 0) rectangle node {$w$} +(1.8, 1);
 
@@
     \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
-    \fill[fill=magenta] (0, 0) rectangle +(0.6, 1);
-    \fill[fill=cyan] (11, 0) rectangle +(1.8, 1);
+    \fill[fill=magenta] (0, 0) rectangle node {$r_1$} +(0.6, 1);
+    \fill[fill=cyan] (11, 0) rectangle node[text=white] {$w_1$} +(1.8, 1);
 
@@
     \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
-    \fill[fill=magenta] (0, 0) rectangle +(0.6, 1);
-    \fill[fill=cyan] (11, 0) rectangle +(1.8, 1);
+    \fill[fill=magenta] (0, 0) rectangle node {$r_2$} +(0.6, 1);
+    \fill[fill=cyan] (11, 0) rectangle node[text=white] {$w_2$} +(1.8, 1);
 
@@
   \begin{minipage}{0.95\textwidth}
-\aalgbegin Algorithm K (State reconstruction with known increment). Reconstruct $S_0$ from three consecutive outputs $X_j$ ($0 \leq j < 3$). In all the algorithm, we have $j= 0, 1, 2$.
+    \algbegin Algorithm K (State reconstruction with known
+    increment). Reconstruct $S_0$ from three consecutive outputs $X_j$
+    ($0 \leq j < 3$). In all steps of the algorithm, we have $j= 0, 1, 2$.
 
-\algstep K1. [Guess state] For each $0 \leq w < 2^{19}$, set $S_0[0 : \ell] = w$ and do steps K2--K10.
+\algstep K1. [Guess state] For each $0 \leq w < 2^{19}$, set $w_j = a^j w \bmod 2^{19}$ and do steps K2--K0.
 
-\algstep K2. [Propagate state] Set $S_j[0:\ell] \gets a^j w \bmod 2^{19}$.
+\algstep K2. [Guess rotations] For each $0 \leq r_0, r_1, r_2 < 64$, do steps K3--K9.
 
-\algstep K3. [Guess rotations] For each $0 \leq r_0, r_1, r_2 < 64$, do steps K4--K10.
+\algstep K3. [Undo rotations] Set $Y_j \gets X_j \lll r_j$. % ($Y_j$ is the $j$-th output on which the rotation has been undone).
 
-\algstep K4. [Undo rotations] Set $S_j[122:128] \gets r_j$ and $Y_j \gets X_j \lll r_j$.
+\algstep K4. [Build $T$] Set $T_j \gets \left(r_j \oplus Y_j[58:64]\right) \oplus \bigl( 64 \cdot \left(w_j \oplus Y_j[0:19]\right)\bigr)$.
 
-\algstep K5. [Propagate] Set $S_j[58:64] \gets r_j \oplus Y_j[58:64]$ and $S_j[64:83] \gets S_j[0:19] \oplus Y_j[0:19]$.
+\algstep K5. [Shift $T$] Set $T'_j \gets  T_j - (a^j w + c \sum_{k = 0}^j a^k)[58:83]$
 
-\algstep K6. [Build partial $T$] Set $T_j \gets \left(S_j[58:83] - (a^j w + c \sum_{k = 0}^j a^k)[58:83]\right) \cdot 2^{39}$
+\algstep K6. [Truncated LCG] Set $(U_0, U_1, U_2) \gets \left\lfloor (T_0, T_1, T_2) \cdot 2^{39} \cdot \widetilde G^{-1} \right\rceil \cdot \widetilde G$.
 
-\algstep K7. [Reconstruct $T$] Set $(U_0, U_1, U_2) \gets \left\lfloor (T_0, T_1, T_2) \cdot \widetilde G^{-1} \right\rceil \cdot \widetilde G$.
+\algstep K7. [Reconstruct $S_0$] Set $\widehat S_0[0:64] \gets w + 2^{19} \cdot U_0[0:45]$ and $\widehat S_0[64:128] \gets S_0[0:64] \oplus Y_0$.
 
-\algstep K8. [Reconstruct $S_0$] Set $S_0[19:83] \gets U_0$ and $S_0[83:122] \gets U_0[0:39] \oplus Y_0[19:58]$.
+\algstep K8. [Recompute $Y_1$] Set $\widehat S_1 \gets a \widehat S_0 + c$ and $\widehat Y_1 = \widehat S_1[0:64] \oplus \widehat S_1[64:128]$.
 
-\algstep K9. [Recompute $X_1$] Compute $\widehat S_1 \gets a S_0 + c$ and $\widehat X_1 =(\widehat S_1[0:64] \oplus \widehat S_1[64:128]) \ggg r_1$.
-
-\algstep K10. [Check] If $\widehat{X}_1 = X_1$, the output $S_0$ as a candidate internal state.
+\algstep K9. [Check] If $\widehat{Y}_1 = Y_1$, then output $\widehat S_0$ as a candidate internal state.
 \end{minipage}

### 2020-05-25 article/main.tex
modifications mineurs

@@
+% !TeX spellcheck = en
 \documentclass[preprint,svgnames]{iacrtrans}
@@
   mentionner}). The goal of designers then consists in designing the fastest
-possible generator that passes the day's favorite test suite.
+possible generator that passes the day's favourite test suite.
 
@@
 ``state'' and a 128-bit ``increment'', whose intended use is to provide several
-pseudo-random streams with the same seed (just as the initialization vectors do
+pseudo-random streams with the same seed (just as the initialisation vectors do
 in stream ciphers). A default increment is provided in case the end-user just
@@
 default situation in \textsf{NumPy}, where both the state and the increment are
-initialized using an external source of entropy. In this case, our prediction
+initialised using an external source of entropy. In this case, our prediction
 algorithm requires 3072 bits of pseudo-random stream ; it guesses between 51 and
@@
 
-Figure~\ref{pcg128out} summarizes the process. The overall design strategy is
+Figure~\ref{pcg128out} summaries the process. The overall design strategy is
 similar to that of a filtered LFSR: the successive states of a weak internal
@@
 
-\subsection{A first way to find U : the Babai's rounding}
+ \subsection{A first way to find U : using an exact CVP solver}
 
-We compute \(G\) the LLL-reduction of the generating matrix of \(L(\alpha,M,n)\).
+Solving exactly the CVP is a hard computational problem. The algorithms used are costly, but we can afford to use them up to dimension 50.
+
+We know that \(U\) is in the lattice, \(T\) is usually not and \(|T-U|\) is small.
+Then we can hope for \(U\) to be the closet vector of \(T\) which is in the lattice : the vector \(U\) may be the solution of \(CVP(L(\alpha,M,n),T)\).
+
+What can we say about \(|CVP(L(\alpha,M,n),T)-U|\) ?
+
+\begin{align*}
+|CVP(L(\alpha,M,n),T)-U| &= |CVP(L(\alpha,M,n),T)-T+T-U|\\
+& \leqslant |CVP(L(\alpha,M,n),T)-T|+|T-U|\\
+\end{align*}
+
+But, as the vector \(U\) is in the lattice and by definition of the CVP, \(|CVP(L(\alpha,M,n),T)-T| \leqslant |T-U| \), hence 
+\[|CVP(L(\alpha,M,n),T)-U|\leqslant 2|T-U|\]
+
+If we can prove that the right side of the inequality is smaller than the SVP of the lattice, then wa have proved that \(CVP(L(\alpha,M,n),T)\) is indeed the vector we were searching for.
+
+\subsection{A second way to find U : the Babai's rounding}
+
+As using an exact CVP solver is costly, we'll try a cheaper but less effective way to find the closest vector.
+
+We compute \(G\) the LLL-reduction of the generating matrix of \(L(\alpha,M,n)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
+\[\rVert G \lVert =  \max_{x \in \mathbb{R}}\frac{\lVert Ax \rVert_2}{\lVert x \rVert_2}\]
+In the case of the \(\lVert\rVert_2\) norm, \(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular value of \(G\).
+
+\medskip
 
@@
 
-If we can prove that the right side \todo{je confonds ma gauche et ma droite et je vous zut :p} of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.
+If we can prove that the right side of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.
 
- \subsection{A second way to find U : using an exact CVP solver}
-
-Solving exactly the CVP is a hard computational problem. But as we do not manipulate quantities too big, we still can use an exact CVP solver.
-
-We know that \(U\) is in the lattice, \(T\) is not and \(|T-U|\) is small.
-Then we can hope for \(U\) to be the closet vector of \(T\) which is in the lattice : the vector \(U\) may be the solution of \(CVP(L(\alpha,M,n),T)\).
-
-What can we say about \(|CVP(L(\alpha,M,n),T)-U|\) ?
-
-\begin{align*}
-|CVP(L(\alpha,M,n),T)-U| &= |CVP(L(\alpha,M,n),T)-T+T-U|\\
-& \leqslant |CVP(L(\alpha,M,n),T)-T|+|T-U|\\
-\end{align*}
-
-But, as the vector \(U\) is in the lattice and by definition of the CVP, \(|CVP(L(\alpha,M,n),T)-T| \leqslant |T-U| \), hence 
-\[|CVP(L(\alpha,M,n),T)-U|\leqslant 2|T-U|\]
-
-If we can prove that the right side of the inequality is smaller than the SVP of the lattice, then wa have proved that \(CVP(L(\alpha,M,n),T)\) is indeed the vector we were searching for.
+We see that  the right side of the inequality is often larger than in the first case.
+We will need to use larger parameters to find the vector we are looking for.
+But as the Babai rounding is largely cheaper than using an exact CVP solver, there will be cases where this method is far more practical.
+\todo{je trouve la formulation un peu moche}
 
@@
 The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero.
+As \(S'\) is in \(L(A,2^{128},n)\), \(S' \mod [2^{64+\ell}]\) is in \(L(A,2^{64+\ell},n)\)
 
-We call \(U\) the truncated vector : \(S'[\ell,64+\ell]\). We notice that \(U\) is in \(L(A,2^{64},n)\).
+We call \(U\) the truncated vector : \(S'[\ell,64+\ell]\).
+As the \(\ell\) lower bits of \(S'\) are zero, we have \(S' \equiv U \times 2^\ell\ \mod [2^{64+\ell}]\).
+Hence \(U\) is in \(L(A,2^{64},n)\).
 
@@
 
-(maybe dessin)
+\todo{maybe dessin}
 

### 2020-05-25 article/main.tex
algos + texte

--- 2020-05-22 article/main.tex
@@
-\documentclass[preprint,svgnames]{iacrtrans}
+\documentclass[nohyperref,preprint,svgnames]{iacrtrans}
 
@@
 \usepackage{subcaption}
-\usepackage{algorithm2e}
+%\usepackage{algorithm2e}
 \usepackage{tikz}
@@
 
-\usepackage[pdftex,bookmarks,bookmarksopen,bookmarksdepth=3]{hyperref}
-\hypersetup{colorlinks=true,citecolor=red,linkcolor=red,urlcolor=black}
+%\usepackage[pdftex,bookmarks,bookmarksopen,bookmarksdepth=3]{hyperref}
+%\hypersetup{colorlinks=true,citecolor=red,linkcolor=red,urlcolor=black}
 
 \DeclareMathSymbol{:}{\mathord}{operators}{"3A}
+\DeclareMathOperator{\LCG}{LCG}
+\DeclareMathOperator{\MCG}{MCG}
 
-\usepackage{algorithmic}
+\usepackage[noend]{algpseudocode}
+%\usepackage{algorithmicx}
 
 % knuth-style algos
-\newcommand{\slug}{\hbox{\kern1.5pt\vrule width2.5pt height6pt depth1.5pt\kern1.5pt}}
-\def\xskip{\hskip 7pt plus 3pt minus 4pt}
-\newdimen\algindent
-\newif\ifitempar \itempartrue % normally true unless briefly set false
-\def\algindentset#1{\setbox0\hbox{{\bf #1.\kern.25em}}\algindent=\wd0\relax}
-\def\algbegin #1 #2{\algindentset{#21}\alg #1 #2} % when steps all have 1 digit
-\def\aalgbegin #1 #2{\algindentset{#211}\alg #1 #2} % when 10 or more steps
-\def\aaalgbegin #1 #2{\algindentset{#2111}\alg #1 #2} % when 10 or more steps
-\def\alg#1(#2). {\medbreak % Usage: \algbegin Algorithm A (algname). This...
-  \noindent{\bf#1}({\it#2\/}).\xskip\ignorespaces}
-\def\algstep#1.{\ifitempar\smallskip\noindent\else\itempartrue
-  \hskip-\parindent\fi
-  \hbox to\algindent{\bf\hfil #1.\kern.25em}%
-  \hangindent=\algindent\hangafter=1\ignorespaces}
+% \newcommand{\slug}{\hbox{\kern1.5pt\vrule width2.5pt height6pt depth1.5pt\kern1.5pt}}
+% \def\xskip{\hskip 7pt plus 3pt minus 4pt}
+% \newdimen\algindent
+% \newif\ifitempar \itempartrue % normally true unless briefly set false
+% \def\algindentset#1{\setbox0\hbox{{\bf #1.\kern.25em}}\algindent=\wd0\relax}
+% \def\algbegin #1 #2{\algindentset{#21}\alg #1 #2} % when steps all have 1 digit
+% \def\aalgbegin #1 #2{\algindentset{#211}\alg #1 #2} % when 10 or more steps
+% \def\aaalgbegin #1 #2{\algindentset{#2111}\alg #1 #2} % when 10 or more steps
+% \def\alg#1(#2). {\medbreak % Usage: \algbegin Algorithm A (algname). This...
+%   \noindent{\bf#1}({\it#2\/}).\xskip\ignorespaces}
+% \def\algstep#1.{\ifitempar\smallskip\noindent\else\itempartrue
+%   \hskip-\parindent\fi
+%   \hbox to\algindent{\bf\hfil #1.\kern.25em}%
+%   \hangindent=\algindent\hangafter=1\ignorespaces}
 % end of borrowed macros
@@
 no value of $c$ is specified, then a default increment is provided:
-$c = 117397592171526113268558934119004209487$.
+$c = 117397592171526113268558934119004209487$. Note that $c$ must be odd.
 
@@
 
+\subsection{Reconstructing Linear Congruential Generators}
+
+Given a bit-size $k$, a fixed multiplier $a$ and an increment $c$, define the
+sequence $u_0 = x, u_{i+1} = a u_i + c \bmod 2^k$. The sequence $u$ form the
+successive internal states of a linear congruential generator (LCG). When the
+increment is zero, the sequence forms a geometric progression with common ratio
+$a$, and is sometimes referred to as the output of a multiplicative congruential
+generator (MCG).
+
+Let $\LCG_{n,k}(x, c)$ denote the vector
+$(u_0, \dots, u_{n-1}) \in \mathbb{Z}_{2^k}$ (we will omit $n$ and $k$ when they
+are obvious from the context).  It is easy to check that:
+\begin{align}
+  \label{eq:lcg-additive}
+  \LCG_{n,k}(x + y, c + d) &= \LCG_{n,k}(x, c) + \LCG_{n,k}(y, d),  \\
+  \label{eq:lcg-scalar}
+  \LCG_{n,k}(\lamba x, \lambda c) = \lambda \LCG_{n,k}(x, c).
+
+\end{align}
+
 \todo{à fusionnier avec la section précédente!}
@@
 \section{Warm-up: A Prediction Algorithm for \textsf{PCG64} With Known Increment}
+\label{sec:Cknown}
 
@@
 \]
-\item If $X_1$ is computed correctly, output $S_0$.
+\item If $X_1$ is computed correctly, then output $S_0$.
 \end{enumerate}
 
-This requires $2^{64}$ iterations of a loop that does a dozen arithmetic
-operations. This is the baseline that we improve upon. An improved
-``guess-and-determine'' prediction algorithm is possible, which essentially
-amounts to expose a truncated version of the underlying linear congruential
-generator, and attack it using the well-known tools. This is possible by
-combining the following ingredients:
+This ``baseline'' procedure requires $2^{64}$ iterations of a loop that does a
+dozen arithmetic operations; it always output the correct value of $S_0$, and
+may output a few other ones (they can be easily discarded by checking $X_2$). An
+improved ``guess-and-determine'' prediction algorithm is possible, which
+essentially amounts to expose a truncated version of the underlying linear
+congruential generator, and attack it using the well-known tools. This is
+possible by combining the following ingredients:
 \begin{itemize}
@@
 linear congruential generator on $64$ bits, where we have access to the most
-$6+\ell$ bits of each state, for a few consecutive states. This is sufficient to
-reconstruct entirely the successive states of this truncated linear congruential
-generator. This reveals $S_0[\ell:64+\ell]$, and using~\eqref{eq:output} the
-entire $S_0$ can be reconstructed. The precise details follow.
+$6+\ell$ bits of each state (denoted by $T$), for a few consecutive states. This
+is sufficient to reconstruct entirely the successive states of this truncated
+linear congruential generator. This reveals $S_0[\ell:64+\ell]$, and
+using~\eqref{eq:output} the entire $S_0$ can be reconstructed. The precise
+details follow.
 
@@
     \draw[]  (1.8, 0) rectangle +(0, 1);
-    \path (2.4, 0) rectangle node {$???????? \cdots ????????$} (6.4, 1);
+    \path (2.4, 0) rectangle node {$???????????????????$} (6.4, 1);
 
@@
 
-Set $S_i' = S_i - S_0[0:\ell] a^i - c \sum_{j=0}^{i-1} a^i$. The sequence
-$(S'_i)$ describe the internal states we would observe if $c=0$ and
-$S_0[0:\ell] = 0$. It is easy to check that $S'_{i} = a^i (S_0 - S_0[0:\ell])$,
-and therefore the successive $S'_i$'s follow a geometric progression. % Indeed :
-% \begin{align*}
-% S_i &= a S_{i-1} + c\\
-%     &= a^2 S_{i-2} + (a + 1) c \\
-%     &\vdots  \\
-%     &= a^i S_0  + c \sum_{j=0}^{i-1} a^i \\
-%   S'_i  &= S_i - c \sum_{j=0}^{i-1} a^i - S_0[0:\ell] a^i \\
-%     &= a^i (S_0 - S_0[0:\ell])
-% \end{align*}
-The $\ell$ low-order bits of $S_0 - S_0[0:\ell]$ are all zero by construction;
-therefore set $T_i = \left(S'_i / 2^\ell\right) \bmod 2^{64}$ (the division by
-$2^\ell$ is exact). It follows that the $(T_i)$ is also a geometric progression
-with common ratio $a$, and we have access to the $\ell+6$ high order bits of
-$T_i$ if we guess the $i$-th rotation. We have seen in
+We consider the first three internal states
+$\mathbf{S} = (S_0, S_1, S_2) = \LCG_{3, 128}(S_0, c)$. We will guess the $\ell$
+least-significant bits of $S_0$, therefore let us assume that their value is
+known and denote it by $w$. We define $\mathbf{S}' = \LCG_{3, 128}(S_0 - w,
+0)$. By~\eqref{eq:lcg-additive}, we have
+$\mathbf{S}' = \mathbf{S} - \LCG_{3, 128}(w, c)$. The point is that the elements
+of $\mathbf{S}'$ follow a geometric progression of common ratio $a$; in
+addition, the $\ell$ least significant bits of each components are zero. The
+crux of the reconstruction algorithm is to find $\mathbf{S}'$.
+
+For this, set $\mathbf{T}' = \left(\mathbf{S}' / 2^\ell\right) \bmod 2^{64}$
+(the division by $2^\ell$ is exact). It follows that the $\mathbf{T}'$ is also a
+geometric progression with common ratio $a$, and we have access to the $\ell+6$
+high order bits of $T'_i$ if we guess the $i$-th rotation. We have seen in
 section~\ref{sec:geometric} that 19 known bits on 3 consecutive outputs are
 sufficient for a deterministic reconstruction of the full sequence. Therefore,
-we guess 19 low-order bits of the state and three consecutive rotations.
+we guess $\ell = 19$ low-order bits of the state and three consecutive
+rotations.
 
@@
 
-\begin{center}\noindent\framebox{
-  \begin{minipage}{0.95\textwidth}
-    \algbegin Algorithm K (State reconstruction with known
-    increment). Reconstruct $S_0$ from three consecutive outputs $X_j$
-    ($0 \leq j < 3$). In all steps of the algorithm, we have $j= 0, 1, 2$.
+  \begin{algorithmic}[1]
+    \Procedure{ReconstructState${}_\ell$}{$X_0, X_1, X_2$}
+    \For{$0 \leq w < 2^{\ell}$} \Comment{Guess least-significant bits of $S_0$}
+    \State $w_j = a^j w \bmod 2^{\ell}$ \Comment{Propagate to $S_1, S_2$}
+    \State $S'_j \gets a^j w + c \sum_{k = 0}^j a^k \bmod 2^{128}$    \Comment{Shifted states}
 
-\algstep K1. [Guess state] For each $0 \leq w < 2^{19}$, set $w_j = a^j w \bmod 2^{19}$ and do steps K2--K0.
+    \For{$0 \leq r_0, r_1, r_2 < 64$} \Comment{Guess rotations}
 
-\algstep K2. [Guess rotations] For each $0 \leq r_0, r_1, r_2 < 64$, do steps K3--K9.
+    \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
 
-\algstep K3. [Undo rotations] Set $Y_j \gets X_j \lll r_j$. % ($Y_j$ is the $j$-th output on which the rotation has been undone).
+    \State $ T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(w_j \oplus Y_j[0:\ell]\right)$  \Comment{Truncated LCG output}
 
-\algstep K4. [Build $T$] Set $T_j \gets \left(r_j \oplus Y_j[58:64]\right) \oplus \bigl( 64 \cdot \left(w_j \oplus Y_j[0:19]\right)\bigr)$.
+    \State $ T'_j \gets    T_j -  S'_i[58:64+\ell]$  \Comment{Shifted Truncated LCG output}
 
-\algstep K5. [Shift $T$] Set $T'_j \gets  T_j - (a^j w + c \sum_{k = 0}^j a^k)[58:83]$
+    \State $(U_0, U_1, U_2) \gets \left\lfloor (T'_0, T'_1, T'_2) \cdot 2^{58-\ell} \cdot \widetilde G_3^{-1} \right\rceil \cdot \widetilde G_3$ \Comment{CVP via Babai rounding}
 
-\algstep K6. [Truncated LCG] Set $(U_0, U_1, U_2) \gets \left\lfloor (T_0, T_1, T_2) \cdot 2^{39} \cdot \widetilde G^{-1} \right\rceil \cdot \widetilde G$.
+    \State $ S_0[0:64] \gets  S'_0 + 2^{\ell} \cdot U_0[0:64-\ell]$ \Comment{Reconstruct $S_0$}
+    \State $ S_0[64:128] \gets S_0[0:64] \oplus Y_0$
 
-\algstep K7. [Reconstruct $S_0$] Set $\widehat S_0[0:64] \gets w + 2^{19} \cdot U_0[0:45]$ and $\widehat S_0[64:128] \gets S_0[0:64] \oplus Y_0$.
+    \State $ S_1 \gets a  S_0 + c$ \Comment{Check consistency of $X_1$}
+    \State $\widehat{Y}_1 =  S_1[0:64] \oplus  S_1[64:128]$
 
-\algstep K8. [Recompute $Y_1$] Set $\widehat S_1 \gets a \widehat S_0 + c$ and $\widehat Y_1 = \widehat S_1[0:64] \oplus \widehat S_1[64:128]$.
+    \If{$\widehat {Y}_1 = Y_1$}
+    \State output $ S_0$ as a candidate internal state.
+    \EndIf
+    \EndFor
+    \EndFor
+    \EndProcedure
+  \end{algorithmic}
 
-\algstep K9. [Check] If $\widehat{Y}_1 = Y_1$, then output $\widehat S_0$ as a candidate internal state.
-\end{minipage}
-}
-\end{center}
+\begin{theorem}
+  Algorithm \textsc{ReconstructState} is correct (it always return the correct internal state).
+\end{theorem}
 
-\subsection{preuve cas facile}
-
+\begin{proof} \todo{FIX ME!}
 The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero.
@@
 \end{itemize}
-
-[insert here] attack predicting algo description.
-
-\begin{theorem}
-  The algorithm described above works with probability.... in time ...
-\end{theorem}
+\end{proof}
 
 \section{Predicting PCG with Unknown Increment}
-We can't use the same algorithm for the general case. We don't known $C$, so we used an other geometric reduction.
-$(\Delta S_{i+1}S_i)_i$ est un suite géométrique de raison $A$.
+\label{sec:Cunknown}
 
-On calcule préalablement les matrices $G$ et $G^{-1}$.
+The same algorithm does not apply directly to the general case where the value
+of $c$ is unknown. The global strategy to obtain a prediction algorithm is
+essentially the same (target a truncated version of the underlying linear
+congruential generator), but is a bit more complicated to implement.
+
+Set $\Delta S_i = S{i+1} - S_i$; it is easily checked that $\Delta S_i$ is a
+geometric progression of common ratio $a$. Therefore, reconstructing both $S_0$
+and $\Delta S_0$ is sufficient to compute all subsequent states (and recover the
+unknown increment $c$). The prediction algorithm proceeds in three phases:
+\begin{enumerate}
+\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_4$.
+\item Fully reconstruct $\Delta S_0$ from this partial knowledge and $X_0, \dots, X_{39}$.
+\item Reconstruct $S_0$ from $\Delta S_0$ and $X_0, \dots, X_{32}$.
+\end{enumerate}
+
+\noindent Only the first phase is computationally intensive.
+
+\subsection{Phase 1: Partial Difference Reconstruction}
+
+In order to access to a part of $\Delta S_i$, we use the same strategy as in
+section~\ref{sec:Cknown}: we guess the least significant bits of $S_0$ and some
+rotations. The difference is that, in addition, to obtain the ``long-term
+advantage'' in the least significant bits of the states, we must also guess the
+least significant bits of $c$ as well. We must also guess one additional
+rotation to compute the differences $\Delta S_i$.
+
+Another problem is that of confirming that the guesses are correct. When $c$ is
+known, we could reconstruct the internal state; from there, filtering out the
+bad guesses is easy. When $c$ is unknown, a very strong consistency check can be
+implemented as follows: once we have found $\Delta S_0[0:64+\ell]$, we can
+compute $\Delta S_i[0:64+\ell]$ for any $i$; because we have guessed the first
+rotation and $\ell$ lest significant bits of the state, using~\eqref{eq:output}
+we gain access to $S_0[58:64+\ell]$; combined with the ``differences'', this
+reveals $S_i[58:64+\ell]$ for any $i$ (and we already had $S_i[0:\ell]$). This
+means that we can compute $S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$.
+
+Given a ``fresh'' output $X_i$, we can enumerate the 64 possible rotations
+$r_i$; for each one, we can check whether
+$(X_i \lll r_i)[0:\ell] = S_i[0:\ell] \oplus S_i[64:64+\ell]$. If none of the
+rotations yields a match, then the guesses made beforehand have to be wrong. As
+soon as $\ell \geq 6$, this test is efficient (and it can be done for any number
+of fresh $X_i$'s). As a consequence, bad guesses can be filtered deterministically.
+
+\begin{align*}
+  \mathbf{S} &= \LCG(S_0, c) \\
+             &= \LCG\left(S_0[0:\ell] + 2^\ell S_0[\ell:128], c[0:\ell] + 2^\ell c[\ell:128]\right) \\
+             &= \LCG\left(S_0[0:\ell], c[0:\ell]) + 2^\ell \LCG\left(S_0[\ell:128],  c[\ell:128]\right) \\
+  \Delta \mathbf{S} &= \LCG(S_1, c) - \LCG(S_0, c) \\
+             &= \LCG(S_1 - S_0, 0)
+\end{align*}
+
+\begin{algorithmic}[1]
+  \Procedure{ReconstructPartialDifference}{$X_0, \dots, X_9$}
+  \For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
+  \State $S'_j \gets a^j w_0 + (2c+1) \sum_{k = 0}^j a^k \bmod 2^{128}$ \Comment{Shifted states}
+  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
+  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
+  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(w_j \oplus Y_j[0:\ell]\right)$ \Comment{Truncated LCG}
+  \State $T'_j \gets T_j -  S'_i[58:64+\ell]$ \Comment{Shifted truncated LCG}
+  \State $\Delta T'_j = T'_{j+1} - T'_j$  \Comment{Difference}
+  \State $(\Delta U_0, \Delta U_1, \Delta U_2, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \Delta T'_1, \Delta T'_2, \Delta T'_3) \cdot 2^{39} \cdot \widetilde G_4^{-1} \right\rceil \cdot \widetilde G_4$ \Comment{CVP}
+  \State $\Delta S_0[0:64+\ell] \gets \left((w_1 - w_0) \bmod 2^{\ell}\right) + 2^{\ell} \cdot U_0[0:58-\ell]$
+  \State ...
+  \EndFor
+  \EndFor
+  \EndProcedure
+\end{algorithmic}
 
@@
 \subsection{Part 1 : Premier filtre}
+
 Pour tous les $r_i$, les $S_0[0 : \ell]$ et les $C[0 : \ell]$ possibles:

### 2020-05-25 article/main.tex

@@
-\documentclass[nohyperref,preprint,svgnames]{iacrtrans}
+% !TeX spellcheck = en
+\documentclass[preprint,svgnames]{iacrtrans}
 

### 2020-05-25 article/main.tex
maintenant ça compile

@@
   \label{eq:lcg-scalar}
-  \LCG_{n,k}(\lamba x, \lambda c) = \lambda \LCG_{n,k}(x, c).
-
+  \LCG_{n,k}(\lambda x, \lambda c) &= \lambda \LCG_{n,k}(x, c).   
 \end{align}
@@
 
-Given a ``fresh'' output $X_i$, we can enumerate the 64 possible rotations
-$r_i$; for each one, we can check whether
-$(X_i \lll r_i)[0:\ell] = S_i[0:\ell] \oplus S_i[64:64+\ell]$. If none of the
-rotations yields a match, then the guesses made beforehand have to be wrong. As
-soon as $\ell \geq 6$, this test is efficient (and it can be done for any number
-of fresh $X_i$'s). As a consequence, bad guesses can be filtered deterministically.
+Given a ``fresh'' output $X_i$, and assuming that the guesses are correct, then we should have:
+\[
+  S_i[0:\ell] \oplus S_i[64:64+\ell] \in \bigl\{ (X_i \lll r_i)[0:\ell]~|~r_i=0, 1, \dots, 63 \bigr\}.
+\]
+If none of the rotations yields a match, then the guesses made beforehand have
+to be wrong. As soon as $\ell \geq 6$, this test is efficient (and it can be
+done for any number of fresh $X_i$'s). As a consequence, bad guesses can be
+filtered deterministically.
 
@@
              &= \LCG\left(S_0[0:\ell] + 2^\ell S_0[\ell:128], c[0:\ell] + 2^\ell c[\ell:128]\right) \\
-             &= \LCG\left(S_0[0:\ell], c[0:\ell]) + 2^\ell \LCG\left(S_0[\ell:128],  c[\ell:128]\right) \\
+             &= \LCG\left(S_0[0:\ell], c[0:\ell]\right) + 2^\ell \LCG\left(S_0[\ell:128],  c[\ell:128]\right) \\
   \Delta \mathbf{S} &= \LCG(S_1, c) - \LCG(S_0, c) \\

### 2020-05-27 article/main.tex
algo step 1 cas C inconnu

@@
 Lagarias et Adi Shamir \cite{Frieze}.
-Plaçons nous dans le cas simple du Lehmer's generator\cite{Lehmer}.
+Plaçons nous dans le cas simple du Lehmer's generator \todo{FIXME}.
 It's a Linear Congruential Generator, a known multiplier $A$, a known modulus $2^128$ and no increment.
@@
 
-    \State $ S_0[0:64] \gets  S'_0 + 2^{\ell} \cdot U_0[0:64-\ell]$ \Comment{Reconstruct $S_0$}
+    \State $ S_0[0:64] \gets  S'_0[0:64] + 2^{\ell} \cdot U_0[0:64-\ell]$ \Comment{Reconstruct $S_0$}
     \State $ S_0[64:128] \gets S_0[0:64] \oplus Y_0$
@@
 
-Set $\Delta S_i = S{i+1} - S_i$; it is easily checked that $\Delta S_i$ is a
+Set $\Delta S_i = S_{i+1} - S_i$; it is easily checked that $\Delta S_i$ is a
 geometric progression of common ratio $a$. Therefore, reconstructing both $S_0$
 and $\Delta S_0$ is sufficient to compute all subsequent states (and recover the
-unknown increment $c$). The prediction algorithm proceeds in three phases:
+unknown increment $c$). We note that
+\begin{align*}
+  S_i - S_0 \equiv \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
+\end{align*}
+
+The prediction algorithm proceeds in three phases:
 \begin{enumerate}
-\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_4$.
-\item Fully reconstruct $\Delta S_0$ from this partial knowledge and $X_0, \dots, X_{39}$.
-\item Reconstruct $S_0$ from $\Delta S_0$ and $X_0, \dots, X_{32}$.
+\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{39}$.
+\item Fully reconstruct $\Delta S_0$ from this partial knowledge and the outputs.
+\item Reconstruct $S_0$ from $\Delta S_0$ and the outputs.
 \end{enumerate}
@@
 advantage'' in the least significant bits of the states, we must also guess the
-least significant bits of $c$ as well. We must also guess one additional
-rotation to compute the differences $\Delta S_i$.
+least significant bits of $c$ as well ($c$ is always odd; this makes one less
+bit to guess). We must also guess one additional rotation to compute the
+differences $\Delta S_i$.
 
@@
 Given a ``fresh'' output $X_i$, and assuming that the guesses are correct, then we should have:
-\[
+\begin{equation}\label{eq:consistency}
   S_i[0:\ell] \oplus S_i[64:64+\ell] \in \bigl\{ (X_i \lll r_i)[0:\ell]~|~r_i=0, 1, \dots, 63 \bigr\}.
-\]
+\end{equation}
 If none of the rotations yields a match, then the guesses made beforehand have
@@
 
+\begin{algorithmic}[1]
+  \Procedure{ConsistencyCheck}{$\Delta S_0, w_0, Y_0, X_5, \dots, X_k$}
+  \State // \emph{Implements the test given by eq.~\eqref{eq:consistency}}.
+  \State $x_0 = w_0 \oplus Y_0[0:\ell]$ \Comment{$x_0 = S_0[64:64+\ell]$}
+  \For{$i=5, \dots, k$}
+  \State $x_i = x_0 + \left(\Delta S_0 \frac{a^i-1}{a-1} \bmod 2^{64+\ell}\right)[64:64+\ell]$. \Comment{$x_i = S_i[64:64+\ell]$}
+%  \State $\Delta S_i[\ell:64+\ell] \gets a^i \Delta U_0 \bmod 2^{64}$
+  \State $\mathrm{Target} \gets S_i[0:\ell] \oplus x_i$
+  \If{$\mathrm{Target} \not\in \{ (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$}
+  % \State $\mathrm{Found} \gets \mathbf{False}$
+  % \For{$0 \leq r_i < 64$}
+  % \State $Y_i \gets X_i \lll r_i$
+  % \If{$Y_i[0:\ell] = \mathrm{Target}$}
+  % \State $\mathrm{Found} \gets \mathbf{True}$ 
+  % \EndIf
+  % \EndFor
+  % \If{not Found}
+  \State \Return \textbf{False} \Comment{Bad Guesses}
+  \EndIf
+  \EndFor
+  \State \Return \textbf{True} \Comment{Good  Guesses}
+  \EndProcedure
+
+\State
+
+\Procedure{ReconstructPartialDifference}{$X_0, \dots, X_k$}
+  \For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
+  \State $S'_j \gets a^j w_0 + (2c+1) \sum_{k = 0}^j a^k \bmod 2^{128}$ \Comment{Shifted states}
+  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
+  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
+  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(w_j \oplus Y_j[0:\ell]\right)$ \Comment{Truncated LCG}
+  \State $T'_j \gets (T_j -  S'_i[58:64+\ell]) \bmod 2^{6+\ell}$ \Comment{Shifted truncated LCG}
+  \State $\Delta T'_j = (T'_{j+1} - T'_j) \bmod 2^{6 + \ell}$  \Comment{Difference}
+  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \dots, \Delta T'_3) \cdot 2^{58-\ell} \cdot \widetilde G_4^{-1} \right\rceil \cdot \widetilde G_4$ \Comment{CVP}
+  \State $\Delta S_0[0:64+\ell] \gets \left((w_1 - w_0) \bmod 2^{\ell}\right) + 2^{\ell} \cdot \Delta U_0[0:58-\ell]$ \Comment{Check}
+  \If{$\Call{ConsistencyCheck}{\Delta_0, w_0, Y_0, X_5, \dots, X_k}$}
+  \State \Return $(w_0, c, r_0, \dots, r_4, \Delta_0)$.
+  \EndIf
+  \EndFor
+  \EndFor
+  \EndProcedure
+\end{algorithmic}
+
+\begin{theorem}
+  If $\ell \geq 13$, then algorithm \textsc{ReconstructPartialDifference} is
+  correct (it always returns the correct values).
+\end{theorem}
+
+\begin{proof}
+  \todo{WRITE-ME}
   \begin{align*}
@@
 \end{align*}
-
-\begin{algorithmic}[1]
-  \Procedure{ReconstructPartialDifference}{$X_0, \dots, X_9$}
-  \For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
-  \State $S'_j \gets a^j w_0 + (2c+1) \sum_{k = 0}^j a^k \bmod 2^{128}$ \Comment{Shifted states}
-  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
-  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
-  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(w_j \oplus Y_j[0:\ell]\right)$ \Comment{Truncated LCG}
-  \State $T'_j \gets T_j -  S'_i[58:64+\ell]$ \Comment{Shifted truncated LCG}
-  \State $\Delta T'_j = T'_{j+1} - T'_j$  \Comment{Difference}
-  \State $(\Delta U_0, \Delta U_1, \Delta U_2, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \Delta T'_1, \Delta T'_2, \Delta T'_3) \cdot 2^{39} \cdot \widetilde G_4^{-1} \right\rceil \cdot \widetilde G_4$ \Comment{CVP}
-  \State $\Delta S_0[0:64+\ell] \gets \left((w_1 - w_0) \bmod 2^{\ell}\right) + 2^{\ell} \cdot U_0[0:58-\ell]$
-  \State ...
-  \EndFor
-  \EndFor
-  \EndProcedure
-\end{algorithmic}
-
-\begin{tikzpicture}[yscale=0.66]
-  \draw[red, use as bounding box] (-0.75, -3.75) rectangle (12.8, 2.25);
-
-  % S_i
-  \begin{scope}
-    % remplissage
-    \fill[fill=lightgray] (4.6, 0) rectangle +(2.4, 1);
-    \fill[fill=lightgray] (0, 0) rectangle +(0.6, 1);
-    \fill[fill=lightgray] (11, 0) rectangle +(1.8, 1);
-
-    % bordures
-    \draw[thick]  (0, 0) rectangle (12.8, 1);
-    \draw  (0.6, 0) rectangle +(0, 1);
-    \draw[thick]  (6.4, 0) rectangle +(0, 1);
-    \draw  (7.0, 0) rectangle +(0, 1);
-    \draw  (4.6, 0) rectangle +(0, 1);
-    \draw  (11, 0) rectangle +(0, 1);
-
-    % déco autour
-    \node at (-0.5, 0.5) {$S_i$};
-    \draw[<->] (0, 1.5) -- node[above] {64 bits} +(6.375, 0);
-    \draw[<->] (6.425, 1.5) -- node[above] {64 bits} +(6.375, 0);
-  \end{scope}
-
-  % T_i
-  \begin{scope}[xshift=4.6cm, yshift=-1.5cm]    
-    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
-    \draw[thick]  (0, 0) rectangle +(2.4, 1);
-    \draw[]  (1.8, 0) rectangle +(0, 1);
-    \path (2.4, 0) rectangle node {00000000 \dots 00000000} (6.4, 1);
-
-    % déco
-    \node at (-0.5, 0.5) {$T_i$};
-    \draw[<->] (0, -0.5) -- node[below] {$\ell$ bits} +(1.775, 0);
-    \draw[<->] (1.8, -0.5) -- node[below] {6} +(0.6, 0);
-    \draw[<->] (0, -1.5) -- node[below] {64 bits} +(6.375, 0);
-  \end{scope}
-
-  % flèches S_i --> T_i
-  \draw[ultra thick,->] (5.5, 0.5) -- +(0, -1.5);
-  \draw[ultra thick,->] (6.7, 0.5) -- +(0, -1.5);
-
-\end{tikzpicture}
-
-\subsection{Part 1 : Premier filtre}
-
-Pour tous les $r_i$, les $S_0[0 : \ell]$ et les $C[0 : \ell]$ possibles:
-\begin{enumerate}
-  \item On récupère le $T$  correspondant à nos suppositions. On calcule $\Delta T$ :
-  \begin{equation}
-    \Delta T_i = (S_{i+1}[58 : 64 + \ell] - S_i[58 : 64 + \ell] - (S_0[0:\ell] * (A^{i+1} - A^i) + C[0:\ell] * A^i )[58 : 64 + \ell]) * 2^{64 - l - 6}
-  \end{equation}
-  Ou plus simplement $\Delta T_i = T'_{i+1} - T'_{i}$ avec $T'$ dont on aurait soustrait la composante en $C[0:\ell]$ et non en $C$.
-
-  \item We get the correspondant $\Delta S_0S_1[0 : 64 + \ell]$ by using the babai rounding technique on $\Delta T$ using $G$ and $G^{-1}$.
-
-  \item On teste s'il existe pour l'itération suivante au moins une rotation $r_n$ cohérente avec le $\Delta T$ trouvé. Pratiquement, pour chacune des 64 $r_n$ possibles, on teste si :
-  \begin{align}
-     \Delta S_0S_n[0:l] &= S_0[64:64 + \ell] \oplus \mathtt{ROT(X_n , 64 - r_n)[0:\ell]}\\
-     r_n &= S_0[58:64] \oplus \mathtt{ROT(X_n , 64 - r_n)[58:64]}
-  \end{align}
-  Si un de ces égalité n'est pas vérifiée, la solution n'est pas retenue, on passe à la prochaine itération.
-
-  \item On répète la dernière opération encore 3 fois.
-\end{enumerate}
-Cette première partie de l'algorithme comporte l'énorme majorité du temps de calcul, c'est donc sur elle que doit se porter la recherche de performance.
+\end{proof}
 

### 2020-05-27 article/main.tex
refonte section 3

@@
 
-\section{Algorithm of Frieze \textit{et al}}
-
-\todo{à fusionnier avec la section suivante!}
-Pour casser PCG, nous avons utilisé la méthode présentée par Alan M.
-Frieze, Johan Hastad, Ravi Kannan, Jeffray C.
-Lagarias et Adi Shamir \cite{Frieze}.
-Plaçons nous dans le cas simple du Lehmer's generator \todo{FIXME}.
-It's a Linear Congruential Generator, a known multiplier $A$, a known modulus $2^128$ and no increment.
-The outputs $X$ are the 64 upper-bits of $S$.
-For $n \in \mathbb{N}$, the $n$ firsts internal states form a lattice we will name $\mathcal{L}(A,2^{128},n)$.
-\todo{rajouter ref Lehmer's generator}
-
-This lattice is generated by the matrix :
-\begin{equation}
-genMatrix =
-\begin{pmatrix} 
-1 & A & A^2 & \dots & A^{n- 1}\\
-0 & M & 0 & \dots & 0\\
-0 & 0 & M & \dots & 0\\
-\vdots & & & \ddots & \vdots\\
-0 & 0 & \dots & 0 & M
-\end{pmatrix}
-\end{equation}
-
-The idea of this method is to consider a known vector $T$, close to $S$ the internal state we are looking for, and to search for the closest vector in $\mathcal{L}(A,2^{128},n)$.
-In the case of Lehmer's generator, we use $X * 2^{64}$ as an approximation of $S$.
-
-\todo{preuve que c'est bien le plus proche ici ?}
-
-Pour améliorer les performances de nos algorithmes les plus critiques, nous n'avons dans certains cas pas utilisé d'oracle CVP parfait, mais la méthode du babai-rounding.
-
-\todo{description babai rounding}
-\todo{et preuve ?}
-
 \section{Reconstructing Truncated Geometric Sequences}
@@
 
-Given a bit-size $k$, a fixed multiplier $a$ and an increment $c$, define the
-sequence $u_0 = x, u_{i+1} = a u_i + c \bmod 2^k$. The sequence $u$ form the
-successive internal states of a linear congruential generator (LCG). When the
-increment is zero, the sequence forms a geometric progression with common ratio
-$a$, and is sometimes referred to as the output of a multiplicative congruential
-generator (MCG).
+Given an integer \(k\), a fixed multiplier \(a\) and an increment \(c\), define the
+sequence \[u_0 = x, u_{i+1} = a u_i + c \bmod 2^k\] The sequence \(u\) form the
+successive internal states of a linear congruential generator (LCG).
 
-Let $\LCG_{n,k}(x, c)$ denote the vector
-$(u_0, \dots, u_{n-1}) \in \mathbb{Z}_{2^k}$ (we will omit $n$ and $k$ when they
-are obvious from the context).  It is easy to check that:
+Let $\LCG_{n,k}(x, c)$ denote the vector $(u_0, \dots, u_{n-1}) \in \mathbb{Z}_{2^k}$ (we will omit $n$ and $k$ when they are obvious from the context).  It is easy to check that:
 \begin{align}
@@
 \end{align}
+Hence those sets of vector have a lot of structure we can use.
 
-\todo{à fusionnier avec la section précédente!}
-\subsection{The lattice}
+The LCG have been largely studied.
+This is why we know techniques to recover internal states knowing only upper or lower bits of the output, see \cite{Boyar1989}.
+A particular LCG is the Lehmer's generator \cite{Lehmer} \todo{je sais pas écire la référence dans bibtex}, where the multiplier \(a\) is known and the increment is zero.
 
-One of the key point of the algorithm is, given a modulus \(M\) and a given common ration \(\alpha\), to find a modular geometric progression knowing only the most significant bits of its \(n\) first terms.
+In our algorithm to predict the PCG, we need to reconstruct several truncated Lehmer's generator.
+For that, we will present a classical lattice-based reconstruction for a truncated Lehmer's generator, as presented in \cite{Frieze}.
+\todo{C'est pas exactement celle là, on dit quoi du coup ?}
 
-We call \(L(\alpha,M,n)\) the set of n-tuples \(U = (u_0,\dots u_{n-1})\) satisfying :
+\subsection{The lattice associated to a Lehmer's generator}
 
-\[u_{i} = \alpha u_{i-1} \mod M \] for all \(i\) in \(\{1,\dots,n-1\}\).
+We fix a multiplier \(a\), integers \(k\), \(n\) and we consider \(L(n,k)\) the set of \(\LCG_{n,k}(x, 0)\), for \(x\) in \(\mathbb{Z}_{2^k}\). Then
 
-We easily notice that, if \(U\) and \(V\) are in \(L(\alpha,M,n)\), so are \(-U\) and \(U+V\), hence \(L(\alpha,M,n)\) is a lattice.
+\begin{itemize}
+	\item By \ref{eq:lcg-additive}, the vector \(\LCG(x_1,0) + \LCG(x_2,0) = \LCG(x_1+x_2,0)\), hence it is still in \(L(n,k)\).
+	\item By \ref{eq:lcg-scalar}, the vector \((-1)\times \LCG(x,0) = \LCG(-x,0)\) hence it is still in \(L(n,k)\).
+\end{itemize}
 
-The lattice \(L(\alpha,M,n)\) is spawned by the lines of the following \(n\times n\) matrix :
+Hence \(L(n,k)\) is a lattice. We easily notice that \(L(n,k)\) is spawned by the lines of the following \(n\times n\) matrix :
 
 \[\begin{pmatrix}
-1&\alpha&\alpha^2&\dots&\alpha^{n-1}\\
-0&M&0&\dots&0\\
-0&0&M&\dots&0\\
+1&a&a^2&\dots&a^{n-1}\\
+0&2^k&0&\dots&0\\
+0&0&2^k&\dots&0\\
 \dots&\dots&\dots&\dots&\dots\\
-0&0&0&\dots&M\\
+0&0&0&\dots&2^k\\
 \end{pmatrix}\]
+
+Once again, we will omit $n$ and $k$ when they are obvious from the context.
 
@@
 We know that \(U\) is in the lattice, \(T\) is usually not and \(|T-U|\) is small.
-Then we can hope for \(U\) to be the closet vector of \(T\) which is in the lattice : the vector \(U\) may be the solution of \(CVP(L(\alpha,M,n),T)\).
+Then we can hope for \(U\) to be the closet vector of \(T\) which is in the lattice : the vector \(U\) may be the solution of \(CVP(L,T)\).
 
-What can we say about \(|CVP(L(\alpha,M,n),T)-U|\) ?
+What can we say about \(|CVP(L,T)-U|\) ?
 
 \begin{align*}
-|CVP(L(\alpha,M,n),T)-U| &= |CVP(L(\alpha,M,n),T)-T+T-U|\\
-& \leqslant |CVP(L(\alpha,M,n),T)-T|+|T-U|\\
+|CVP(L,T)-U| &= |CVP(L,T)-T+T-U|\\
+& \leqslant |CVP(L,T)-T|+|T-U|\\
 \end{align*}
 
-But, as the vector \(U\) is in the lattice and by definition of the CVP, \(|CVP(L(\alpha,M,n),T)-T| \leqslant |T-U| \), hence 
-\[|CVP(L(\alpha,M,n),T)-U|\leqslant 2|T-U|\]
+But, as the vector \(U\) is in the lattice and by definition of the CVP, \(|CVP(L,T)-T| \leqslant |T-U| \), hence 
+\[|CVP(L,T)-U|\leqslant 2|T-U|\]
 
-If we can prove that the right side of the inequality is smaller than the SVP of the lattice, then wa have proved that \(CVP(L(\alpha,M,n),T)\) is indeed the vector we were searching for.
+If we can prove that the right side of the inequality is smaller than first minimum of the lattice \(\lambda_1(L)\), then we have proved that \(CVP(L,T)\) is indeed the vector we were searching for.
+
+In this article, we will only use the Babai's rounding when \(k = 128-\ell\) and when we know the \(6\) upper bit of our Lehmer's generator.
+Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{122-\ell} \).
+So, if we fix \(\ell\) we can search for the minimum \(n\) such that \[2\sqrt{n}2^{122-\ell} \leqslant \lambda_1(L(128-\ell,n))\].
+
+\begin{tabular}{|c|c|}
+	\hline
+	\(\ell\)  & minimum \(n\) \\
+	\hline
+	11 & 39 \\
+	12 & 38 \\
+	13 & 38 \\
+	\hline
+\end{tabular}
 
@@
 
-As using an exact CVP solver is costly, we'll try a cheaper but less effective way to find the closest vector.
+As using an exact CVP solver is costly, we'll try a cheaper but less precise way to find the closest vector.
 
-We compute \(G\) the LLL-reduction of the generating matrix of \(L(\alpha,M,n)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
-\[\rVert G \lVert =  \max_{x \in \mathbb{R}}\frac{\lVert Ax \rVert_2}{\lVert x \rVert_2}\]
+We compute \(G(n,k)\) the LLL-reduction of the generating matrix of \(L(n,k)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
+\[\rVert G \lVert =  \max_{x \in \mathbb{R}}\frac{\lVert Gx \rVert_2}{\lVert x \rVert_2}\]
 In the case of the \(\lVert\rVert_2\) norm, \(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular value of \(G\).
@@
 
-If we can prove that the right side of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.
+If we can prove that the right side of the inequality is smaller than the minimum of the lattice \(\lambda_1(L)\), then we have proved that \(rG\) is indeed the vector we were searching for.
 
-We see that  the right side of the inequality is often larger than in the first case.
-We will need to use larger parameters to find the vector we are looking for.
-But as the Babai rounding is largely cheaper than using an exact CVP solver, there will be cases where this method is far more practical.
-\todo{je trouve la formulation un peu moche}
+In this article, we will only use the Babai's rounding when \(k = 64\) and when we know the \(6+\ell\) upper bit of our Lehmer's generator.
+Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{58-\ell} \).
+So, if we fix \(n\) we can search for the minimum \(\ell\) such that \[(1+\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert)\sqrt{n}2^{58-\ell} \leqslant \lambda_1(L(n,64))\]
+
+\bigskip
+
+\begin{tabular}{|c|c|c|c|}
+	\hline
+	n & \(\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert\) & \( \lambda_1(L(n,64)) \) & minimum \(\ell\) \\
+	\hline
+	3 & 2.869065991049632 & \(4.08909120094543e^{12}\) & 19 \\
+	4 & 2.060717188416003 & \(2.43569932844650e^{14}\) & 13 \\
+	5 & 3.7692142154392747 & \(1.71669732658884e^{15}\) & 11 \\
+	6 & 2.6925925483728537 & \(1.02550589786192e^{16}\) & 8 \\
+	\hline
+\end{tabular}
 
@@
 
-\begin{proof} \todo{FIX ME!}
-The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero.
-As \(S'\) is in \(L(A,2^{128},n)\), \(S' \mod [2^{64+\ell}]\) is in \(L(A,2^{64+\ell},n)\)
-
-We call \(U\) the truncated vector : \(S'[\ell,64+\ell]\).
-As the \(\ell\) lower bits of \(S'\) are zero, we have \(S' \equiv U \times 2^\ell\ \mod [2^{64+\ell}]\).
-Hence \(U\) is in \(L(A,2^{64},n)\).
-
-We consider the case were we successfully guessed the 6 upper bits of each \(S_i\) and the \(\ell\) lower bits of \(S_0\). Then we know the \(6+\ell\) upper bits of \(U\).
-
-\todo{maybe dessin}
-
-We consider the vector \(T'\) computed in step (numerodelastepenquestion). To recover \(U\), we use the Babai's rounding explain in section (insererlasection) on \(T'\) and \(L(A,2^{64},n)\).
-
-As \(U\) and \(T'\) have exactly the same \(6+\ell\) upper bits  know that \(\lVert U-T \rVert \leqslant \sqrt{n2^{2(58-l)}}\).
-
-So, to be sure that our algorithm return the vector \(U\), we need to find sets of parameters \((n,\ell)\) such that :
-
-\[\sqrt{n2^{2(58-\ell)}} \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert) \leqslant \lVert SVP(L(A,2^{64},n))\rVert \]
-
-\begin{itemize}
-	\item For \(n=3\)
-
-	\(G = \begin{pmatrix}
-	-1241281756092&3827459685972&-728312298332\\
-	-5001120657083&-2117155768935&5479732607037\\
-	8655886039732&3303731088004&6319848582548\\
-	\end{pmatrix}\)
-
-	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 2.869065991049632\)
-
-	\(\lVert SVP(G) \rVert = 4.08909120094543e{^12}\)
-
-	Hence, the smallest l such that the inequality is satisfied is \(\ell = 19\) %17 avec borne CVP
-
-	\item For \(n=4\)
-
-	\(G = \begin{pmatrix}
-	-186304953996472&-126056243766680&7937589136904&93078431381544\\
-	-216211368070119&99587582169277&-214303762177807&-1707551230219\\
-	110964501361298&-5646098666150&-268280113597118&149382085707466\\
-	131252974561432&-233919070109448&-98716819647784&-134620659538888\\
-	\end{pmatrix}\)
-
-	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 2.060717188416003\)
-
-	\(\lVert SVP(G) \rVert = 2.43569932844650e^{14}\)
-
-	Hence, the smallest l such that the inequality is satisfied is \(\ell = 13\) %11 avec borne CVP
-
-	\item For \(n=5\)
-
-	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 3.7692142154392747\)
-
-	\(\lVert SVP(G) \rVert = 1.71669732658884e^{15}\)
-
-	Hence, the smallest l such that the inequality is satisfied is \(\ell = 11\)  %9 avec borne CVP
-
-	\item For \(n=6\)
-
-	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 2.6925925483728537\)
-
-	\(\lVert SVP(G) \rVert =1.02550589786192e^{16}\)
-
-	Hence, the smallest l such that the inequality is satisfied is \(\ell = 8\) %7 avec borne CVP
-\end{itemize}
-\end{proof}
+\todo{avec la nouvelle partie 3 et le blabla au dessus de l'algo, y'a plus vraiment besoin de preuve}
 

### 2020-05-28 challenges/lehmer.c
lehmer generator

@@ -1,0 +1,36 @@
+#include <inttypes.h>
+#include <stdio.h>
+#include <stdlib.h>
+
+__uint128_t state;
+
+uint64_t lehmer64() {
+  state *= 0xda942042e4dd58b5;
+  return state >> 64;
+}
+
+int main()
+{
+	FILE *f = fopen("/dev/urandom", "r");
+	if (f == NULL) {
+		perror("Something went wrong when opening /dev/urandom");
+		exit(EXIT_FAILURE);
+	}
+	if (fread(&state, sizeof(state), 1, f) != 1) {
+		perror("Something went wrong when reading /dev/urandom");
+		exit(EXIT_FAILURE);
+	}
+	state |= 1;
+	printf("Seed : %016" PRIx64 " %016" PRIx64 "\n\n", (uint64_t) (state >> 64), (uint64_t) state);
+
+	printf("Predictor input:\n");
+	for (int i = 0; i < 3; i++)
+		printf("X[%2d] = 0x%016" PRIx64 ";\n", i, lehmer64());
+	printf("\n");
+
+	printf("Remaining of the sequence (predictor output, in principle):\n");
+	for (int i = 3; i < 16; i++)
+		printf("X[%2d] = 0x%016" PRIx64 ";\n", i, lehmer64());
+
+	return EXIT_SUCCESS;
+}

### 2020-05-28 lehmer.py

@@ -1,0 +1,15 @@
+#Predictor input --- 3 consecutive calls to lehmer64()
+X = [0x1e1495fb7df7b3cf, 0xa267765f13b93bf9, 0xe5b3241c4f81c9a8]
+
+# Prediction algorithm
+a = 0xda942042e4dd58b5
+r = round(2.64929081169728e-7 * X[0] + 3.51729342107376e-7 * X[1] + 3.89110109147656e-8 * X[2])
+s = round(3.12752538137199e-7 * X[0] - 1.00664345453760e-7 * X[1] - 2.16685184476959e-7 * X[2])
+t = round(3.54263598631140e-8 * X[0] - 2.05535734808162e-7 * X[1] + 2.73269247090513e-7 * X[2])
+u = r * 1556524 + s * 2249380 + t * 1561981
+v = r * 8429177212358078682 + s * 4111469003616164778 + t * 3562247178301810180
+w = (a**3 * (a*u + v)) >> 64
+x = w & 0xffffffffffffffff
+
+# Predictor output --- what the fourth call to lehmer64() would produce
+print(hex(x))

### 2020-05-28 article/main.tex
retouche bib
tableaux flottants sections 3
Algos section 4 et 5

@@
 \usepackage{subcaption}
-%\usepackage{algorithm2e}
+\usepackage{algorithm}
 \usepackage{tikz}
@@
 \usepackage[noend]{algpseudocode}
-%\usepackage{algorithmicx}
+%\usepackage{algorithmic}
 
@@
 
+\begin{table}
+  \centering
     \begin{tabular}{|c|c|}
@@
 \end{tabular}
+  \caption{\todo{WRITE-ME}}
+  \label{tab:exact_cvp}
+\end{table}
 
@@
 
+\begin{table}
+  \centering
   \begin{tabular}{|c|c|c|c|}
@@
   \end{tabular}
+  \caption{\todo{WRITE-ME}}
+  \label{tab:babai}
+\end{table}
 
@@
 Therefore, the algorithm that reconstructs the internal state of the
-\textsf{PCG64} generator with known increment proceeds as follows:
+\textsf{PCG64} generator with known increment proceeds as shown in algorithm~\ref{algo:known}.
 
+\begin{algorithm}
 \begin{algorithmic}[1]
@@
 
-    \State $ S_1 \gets a  S_0 + c$ \Comment{Check consistency of $X_1$}
+    \State $ S_1 \gets a  S_0 + c$ \Comment{Recompute $X_1$}
     \State $\widehat{Y}_1 =  S_1[0:64] \oplus  S_1[64:128]$
 
-    \If{$\widehat {Y}_1 = Y_1$}
+    \If{$\widehat {Y}_1 = Y_1$} \Comment{Check consistency}
     \State output $ S_0$ as a candidate internal state.
@@
   \end{algorithmic}
+  \caption{State reconstruction Algorithm (case where $c$ is known)}
+  \label{algo:known}
+\end{algorithm}
 
-\begin{theorem}
-  Algorithm \textsc{ReconstructState} is correct (it always return the correct internal state).
-\end{theorem}
+% \begin{theorem}
+%   Algorithm \textsc{ReconstructState} is correct (it always return the correct internal state).
+% \end{theorem}
 
-\todo{avec la nouvelle partie 3 et le blabla au dessus de l'algo, y'a plus vraiment besoin de preuve}
+% \todo{avec la nouvelle partie 3 et le blabla au dessus de l'algo, y'a plus vraiment besoin de preuve}
+% Charles: je suis plutôt d'accord.
 
-\section{Predicting PCG with Unknown Increment}
+\section{Predicting PCG64 with Unknown Increment}
 \label{sec:Cunknown}
 
-The same algorithm does not apply directly to the general case where the value
-of $c$ is unknown. The global strategy to obtain a prediction algorithm is
-essentially the same (target a truncated version of the underlying linear
+The algorithm of section~\ref{sec:Cknown} does not apply directly to the general
+case where the value of $c$ is unknown. A ``baseline'' procedure would consist
+in guessing $S_0[64:128]$ and $S_1[64:128]$; using eq.~\eqref{eq:output}, this
+would reveal $S_0$ and $S_1$, from which the increment $c$ could be easily
+deduced. This would take $2^{128}$ iterations of a very simple procedure, but is
+completely infeasible.
+
+The global strategy to obtain a practical prediction algorithm is essentially
+the same as before (target a truncated version of the underlying linear
 congruential generator), but is a bit more complicated to implement.
@@
 and $\Delta S_0$ is sufficient to compute all subsequent states (and recover the
-unknown increment $c$). We note that
+unknown increment $c$). Let us set:
 \begin{align*}
-  S_i - S_0 \equiv \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
+  \nabla S_i \stackrel{def}{=} S_i - S_0 \equiv \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
 \end{align*}
-
-The prediction algorithm proceeds in three phases:
+Note that $\nabla S_0 = \Delta S_0$. The prediction algorithm we propose proceeds in three phases:
 \begin{enumerate}
@@
 
-\subsection{Phase 1: Partial Difference Reconstruction}
+\subsection{Partial Difference Reconstruction}
 
-In order to access to a part of $\Delta S_i$, we use the same strategy as in
-section~\ref{sec:Cknown}: we guess the least significant bits of $S_0$ and some
-rotations. The difference is that, in addition, to obtain the ``long-term
-advantage'' in the least significant bits of the states, we must also guess the
-least significant bits of $c$ as well ($c$ is always odd; this makes one less
-bit to guess). We must also guess one additional rotation to compute the
-differences $\Delta S_i$.
+In order to access to a part of $\Delta S_i$, we use the same
+``guess-and-determine'' strategy as in section~\ref{sec:Cknown}: we guess the
+least significant bits of $S_0$ and some rotations, then check consistency. The
+difference is that, since $c$ is unknown, we must in addition guess the least
+significant bits of $c$ to obtain the same ``long-term advantage'' ($c$ is
+always odd; this makes one less bit to guess). We must also guess $k+1$
+successive rotation to get information on $k$ successive differences
+$\Delta S_i$.
 
-Another problem is that of confirming that the guesses are correct. When $c$ is
+Another problem is that of confirming that the guesses are correct. When $c$ was
 known, we could reconstruct the internal state; from there, filtering out the
-bad guesses is easy. When $c$ is unknown, a very strong consistency check can be
-implemented as follows: once we have found $\Delta S_0[0:64+\ell]$, we can
-compute $\Delta S_i[0:64+\ell]$ for any $i$; because we have guessed the first
-rotation and $\ell$ lest significant bits of the state, using~\eqref{eq:output}
-we gain access to $S_0[58:64+\ell]$; combined with the ``differences'', this
-reveals $S_i[58:64+\ell]$ for any $i$ (and we already had $S_i[0:\ell]$). This
-means that we can compute $S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$.
+bad guesses was easy. When $c$ is unknown, the same strategy does not work, but
+a very strong consistency check can still be implemented. The essential idea is
+the following: once $\nabla S_0[0:64+\ell]$ is known, then all the rotations can
+be determined accurately and with little computational effort. This in turn
+allows a consistency check between the partial information deduced from the
+guesses and the actual outputs.
+
+Once we have found $\nabla S_0[0:64+\ell]$, we can compute
+$\nabla S_i[0:64+\ell]$ for any $i$ (geometric progression); because we have
+guessed the first rotation and the $\ell$ least significant bits of the state,
+using~\eqref{eq:output} we gain access to $S_0[58:64+\ell]$; combined with the
+``differences'' $\nabla S_i$, this reveals $S_i[58:64+\ell]$ for any $i$ (and we
+already had $S_i[0:\ell]$). This allows us to compute
+$S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$.
 
 Given a ``fresh'' output $X_i$, and assuming that the guesses are correct, then we should have:
+\begin{equation}\label{eq:find_rotation}
+  S_i[0:\ell] \oplus S_i[64:64+\ell] = (X_i \lll r_i)[0:\ell].
+\end{equation}
+If $X_i$ is known, then this gives a way to find the $i$-th rotation $r_i$ by
+trying the 64 possible values. As soon as $\ell \geq 6$, this should determine
+$r_i$ almost uniquely. In any case, if the guesses were correct, then we should have:
 \begin{equation}\label{eq:consistency}
-  S_i[0:\ell] \oplus S_i[64:64+\ell] \in \bigl\{ (X_i \lll r_i)[0:\ell]~|~r_i=0, 1, \dots, 63 \bigr\}.
+  S_i[0:\ell] \oplus S_i[64:64+\ell] \in \bigl\{ (X_i \lll r)[0:\ell]~|~0 \leq r < 64 \bigr\}.
 \end{equation}
-If none of the rotations yields a match, then the guesses made beforehand have
-to be wrong. As soon as $\ell \geq 6$, this test is efficient (and it can be
-done for any number of fresh $X_i$'s). As a consequence, bad guesses can be
-filtered deterministically.
+If none of the 64 possible rotations yields a match, then the guesses made
+beforehand have to be wrong. As a consequence, bad guesses can be filtered
+deterministically.
 
+More precisely, let us assume that we have guessed the $\ell$ least-significant
+bits of $S_0$ (we denote them by $w_0$) and the first rotation $r_0$. Set
+$Y_0 \stackrel{def}{=} X_0 \lll r_0$; there is an unknown value
+$x_0$ such that:
+\[
+  S_0 = w_0 + 2^\ell x + 2^{64} (w_0 \oplus Y_0[0:\ell]) + 2^{64+\ell} (x \oplus Y_0[\ell:64]) \qquad \left(0 \leq x < 2^{64-\ell}\right).
+\]
+The unknown value $x_0$ will not be recovered until the third phase of the whole
+procedure. We obtain the $i$-th state by
+$S_i \equiv \nabla S_i + S_0 \bmod 2^{64+\ell}$; however, because the ``middle''
+of $S_0$ is unknown, an unknown \emph{carry} may cross the 64-th during the
+addition and perturb $S_i[64:64+\ell]$. As a result, there are unknown
+quantities $0 \leq y_i < 2^{64-\ell}$ and $0 \leq z_i \leq 1$ such that:
+\begin{align*}
+  S_i[0:\ell] &\equiv \nabla S_i + w_0 \bmod 2^\ell \\
+  S_i[\ell:64] &\equiv y_i \\
+  S_i[64:64+\ell] &\equiv z_i + \nabla S_i[64:64+\ell] + (w_0 \oplus Y_0[0:\ell]) \bmod 2^\ell \\
+\end{align*}
+
+In algorithm~\ref{algo:unknown_1}, \textsc{ConsistencyCheck} uses
+eq.~\eqref{eq:consistency} combined with this observation to discard bad guesses.
+
+\begin{algorithm}
 \begin{algorithmic}[1]
   \Procedure{ConsistencyCheck}{$\Delta S_0, w_0, Y_0, X_5, \dots, X_k$}
-  \State // \emph{Implements the test given by eq.~\eqref{eq:consistency}}.
-  \State $x_0 = w_0 \oplus Y_0[0:\ell]$ \Comment{$x_0 = S_0[64:64+\ell]$}
+  \State $v_0 = w_0 \oplus Y_0[0:\ell]$ \Comment{$v_0 = S_0[64:64+\ell]$}
   \For{$i=5, \dots, k$}
-  \State $x_i = x_0 + \left(\Delta S_0 \frac{a^i-1}{a-1} \bmod 2^{64+\ell}\right)[64:64+\ell]$. \Comment{$x_i = S_i[64:64+\ell]$}
-%  \State $\Delta S_i[\ell:64+\ell] \gets a^i \Delta U_0 \bmod 2^{64}$
-  \State $\mathrm{Target} \gets S_i[0:\ell] \oplus x_i$
-  \If{$\mathrm{Target} \not\in \{ (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$}
-  % \State $\mathrm{Found} \gets \mathbf{False}$
-  % \For{$0 \leq r_i < 64$}
-  % \State $Y_i \gets X_i \lll r_i$
-  % \If{$Y_i[0:\ell] = \mathrm{Target}$}
-  % \State $\mathrm{Found} \gets \mathbf{True}$ 
-  % \EndIf
-  % \EndFor
-  % \If{not Found}
+  \State $u_i \gets \Delta S_0 (a^i-1)(a-1)^{-1} \bmod 2^{64+\ell}$ \Comment{$u_i = \nabla S_i[0:64+\ell]$}
+  \State $w_i = w_0 + u_i[0:\ell] \bmod 2^{\ell}$ \Comment{$w_i = S_i[0:\ell]$}
+  \State $v_i = v_0 + u_i[64:64+\ell] \bmod 2^{\ell}$ \Comment{$S_i[64:64+\ell] \in \{v_i, v'_i\}$}
+  \State $v'_i = v_i + 1 \bmod 2^{\ell}$
+  \State $\mathcal{C} \gets \{ w_i \oplus (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$ \Comment{Check eq.~\eqref{eq:consistency}}
+  \If{$\{v_i, v'_i\} \cap \mathcal{C} = \emptyset$}
   \State \Return \textbf{False} \Comment{Bad Guesses}
@@
   \EndFor
-  \State \Return \textbf{True} \Comment{Good  Guesses}
+  \State \Return \textbf{True} \Comment{No inconsistency}
   \EndProcedure
@@
 \end{algorithmic}
+\caption{Partial difference reconstruction algorithm (when $c$ is unknown).}
+\label{algo:unknown_1}
+\end{algorithm}
 
-\begin{theorem}
-  If $\ell \geq 13$, then algorithm \textsc{ReconstructPartialDifference} is
-  correct (it always returns the correct values).
-\end{theorem}
+The heart of the algorithm is again the reconstruction of a truncated geometric
+progression. Looking at table~\ref{tab:babai}, we see that the best choice
+consists in guessing 5 consecutive rotations and $\ell=13$ least-significant
+bits. Therefore, \textsc{ReconstructPartialDifference} does $2^{55}$ iterations
+of the inner loop, and succeeds deterministically.
 
-\begin{proof}
-  \todo{WRITE-ME}
+% \begin{theorem}
+%   If $\ell \geq 13$, then algorithm \textsc{ReconstructPartialDifference} is
+%   correct (it always returns the correct values).
+% \end{theorem}
+
+% \begin{proof}
+%   \todo{WRITE-ME}
+%   \begin{align*}
+%   \mathbf{S} &= \LCG(S_0, c) \\
+%              &= \LCG\left(S_0[0:\ell] + 2^\ell S_0[\ell:128], c[0:\ell] + 2^\ell c[\ell:128]\right) \\
+%              &= \LCG\left(S_0[0:\ell], c[0:\ell]\right) + 2^\ell \LCG\left(S_0[\ell:128],  c[\ell:128]\right) \\
+%   \Delta \mathbf{S} &= \LCG(S_1, c) - \LCG(S_0, c) \\
+%              &= \LCG(S_1 - S_0, 0)
+% \end{align*}
+% \end{proof}
+
+\subsection{Full Difference Reconstruction}
+
+Knowing the values of $\Delta S_0[0:64+\ell]$ as well as the $\ell$
+least-significant bits of $S_0$ and $c$ is sufficient to get rid the nastier
+feature \textsf{PCG64}: armed with this knowledge, we can determine all the
+subsequent rotations deterministically, at negligible cost, using
+eq~\eqref{eq:find_rotation}.
+
+For each $i$, it suffices to try the $64$ possible values of $r_i$; only one
+should satisfy the equation. Using $X_0, \dots, X_{40}$, we thus recover the 6
+most-significant bits of $S_0, \dots, S_{40}$. We therefore compute the 6 most
+significant bits of the differences $\Delta S_i$ between consecutive states (up
+to missing carries), and we are faced with the problem of reconstructing a
+128-bit geometric progression using 39 consecutive outputs truncated to their 6
+most-significant bits. This can be done by solving an instance of CVP in
+dimension $\leq 40$. We use the off-the-shelf CVP solver embedded in
+\texttt{fplll}: it runs in no measurable time.
+
+More precisely, there are still details to be dealt with regarding
+carries. Indeed, the \textsc{ReconstructRotations} returns the 6 most
+significant bits of any number of $S_i$'s (denoted by $r_i$). Therefore, there
+are unknown 122-bit quantities $x_i$ such that:
 \begin{align*}
-  \mathbf{S} &= \LCG(S_0, c) \\
-             &= \LCG\left(S_0[0:\ell] + 2^\ell S_0[\ell:128], c[0:\ell] + 2^\ell c[\ell:128]\right) \\
-             &= \LCG\left(S_0[0:\ell], c[0:\ell]\right) + 2^\ell \LCG\left(S_0[\ell:128],  c[\ell:128]\right) \\
-  \Delta \mathbf{S} &= \LCG(S_1, c) - \LCG(S_0, c) \\
-             &= \LCG(S_1 - S_0, 0)
+  S_i &= 2^{122} r_i + x_i \\
+  \Delta S_i &\equiv 2^{122} (r_{i+1} - r_i) + (x_{i+1} - x_{i}) \bmod 2^{128}
 \end{align*}
-\end{proof}
+
+\begin{algorithm}
+\begin{algorithmic}[1]
+  \Function{ReconstructRotations}{$\Delta S_0, v_0, i, k$}
+ \State // \emph{Return a list of potential} $\texttt{[}r_i, r_{i+1}, \dots, r_k\texttt{]}$; \emph{assume that $v_0 = S_0[64:64+\ell]$}
+  \If{$i > k$}
+  \State \Return \texttt{[]} \Comment{End recursion}
+  \EndIf
+  \State $\mathcal{T} \gets \Call{ReconstructRotations}{\Delta S_0, v_0, i + 1, k}$ \Comment{Find all the $(r_{i+1}, \dots, r_{k})$}
+  \State $\mathcal{H} \gets \texttt{[]}$ \Comment{List of possible $r_i$'s}
+
+  \State $u_i \gets \Delta S_0 (a^i-1)(a-1)^{-1} \bmod 2^{64+\ell}$ \Comment{$u_i = \nabla S_i[0:64+\ell]$}
+  \State $w_i = w_0 + u_i[0:\ell] \bmod 2^{\ell}$ \Comment{$w_i = S_i[0:\ell]$}
+  \State $v_i = v_0 + u_i[64:64+\ell] \bmod 2^{\ell}$ \Comment{$S_i[64:64+\ell] \in \{v_i, v'_i\}$}
+  \State $v'_i = v_i + 1 \bmod 2^{\ell}$
+
+  %\State $x_i = x_0 + \left(\Delta S_0 \frac{a^i-1}{a-1} \bmod 2^{64+\ell}\right)[64:64+\ell]$. \Comment{$x_i = S_i[64:64+\ell]$}
+  \For{$0 \leq r < 64$} \Comment{Try all rotations}
+  \If{$w_i \oplus (X_i \lll r)[0:\ell] \in \left\{v_i, v'_i\right\}$} \Comment{Check eq.~\eqref{eq:find_rotation}}
+  \State $\mathcal{H} \gets r \texttt{::} \mathcal{H}$ \Comment{New candidate $r_i$}
+  \EndIf
+  \EndFor
+  \State \Return $\left\{ h\texttt{::} t~|~ h \in \mathcal{H}, t \in \mathcal{T} \right\}$ \Comment{Return $\mathcal{H} \times \mathcal{T}$}
+  \EndFunction
+  \State
+  \Function{ReconstructFullDifference}{...}
+  \State \todo{WRITEME}
+  \EndFunction
+\end{algorithmic}
+\caption{Rotations and full difference reconstruction algorithm}
+\end{algorithm}
 
@@
 
-\section{Implementation of the Predictor and Practical Results}
+\section{Implementation and Practical Results}
 

### 2020-05-28 article/main.tex
l=0 + SVP

@@
 This is why we know techniques to recover internal states knowing only upper or lower bits of the output, see \cite{Boyar1989}.
-A particular LCG is the Lehmer's generator \cite{Lehmer} \todo{je sais pas écire la référence dans bibtex}, where the multiplier \(a\) is known and the increment is zero.
+A particular LCG is the Lehmer's generator \cite{Lehmer}, where the multiplier \(a\) is known and the increment is zero.
 
@@
 
-In this article, we will only use the Babai's rounding when \(k = 128-\ell\) and when we know the \(6\) upper bit of our Lehmer's generator.
+To compute \(\lambda_1(L)\), we find the shortest vector of L thanks to an exact SVP solver (SVP stands for Shortest Vector Problem) and we compute its norm.
+
+In this article, we will only use an exact CVP solver when \(k = 128-\ell\) and when we know the \(6\) upper bit of our Lehmer's generator.
 Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{122-\ell} \).
@@
 	\hline
+	0  & 42 \\
 	11 & 39 \\

### 2020-05-28 article/main.tex
tables aux bon endroits
début 5.3

--- 2020-05-28 article/main.tex
@@
 
-\subsection{Part 2 : Récupération de la différence complète et deuxième filtre}
-On calcule préalalement H matrice génératrice du réseau $\mathcal{L}(A,2^{128 - \ell},m)$ réduite par la méthode LLL.
-On a récupéré de la Part 1 une série de $S_0[0 : \ell]$, $r_i$ (i = 0, ..., n) et $C[0 : \ell]$ parmi lequels se trouve notre solution.
-\begin{enumerate}
-  \item recupérer toutes les $r_i$ possibles, pour $n \leq i \leq m$  par la même méthode que l'étape 3 de la Part 1, répétée $m-n$ fois. On garde en mémoire toutes les rotations possiles.
-  \item Pour chacune des $r$ possibles
-  \begin{enumerate}
-    \item We cumpute $r'$ et $\Delta r$ avec le même méthode que pour $\Delta T$ : 
-    \begin{align}
-      r'_i = r_i - (A^i*S_0[0 : \ell] + C[0 : \ell] * \sum_{j = 0}^i A^j)[64 - 6 : 64 + \ell])[122 : 128]
-      \Delta r_i &= r_{i+1} - r_i
-    \end{align}
+% \subsection{preuve cas difficile}
+% \todo{À ranger au bon endroit}
+% % Babai ça marche pas (genre zero)
 
-    \item We get the correspondant $\Delta S'_0S'_1$ ($S'$ dont on aurait retiré que la composante en $C[0 : \ell]$ et non en $C$ entier) computing $\texttt{CVP}(r * 2^{128 - \ell}, \mathcal{L}(A,2^{128 - \ell},m))$.
-We use the fpylll python library CVP function.
-On stocke le $DS_0S_1$ correspondant.
+% The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero.
 
-    \item \todo{parler du problème de retenue étrange}
+% We call \(U\) the truncated vector : \(S'[\ell,128]\). We notice that \(U\) is in \(L(A,2^{128-\ell},n)\).
 
-  \end{enumerate}
-\end{enumerate}
+% We consider the case were we successfully guessed the 6 upper bits of each \(S_i\) and the \(\ell\) lower bits of \(S_0\). Then we know the 6 upper bits of \(U\).
 
-\subsection{preuve cas difficile}
-\todo{À ranger au bon endroit}
-% Babai ça marche pas (genre zero)
+% (maybe dessin)
 
-The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero.
+% We consider the vector \(T'\) computed in step (numerodelastepenquestion). To recover \(U\), we use an exact CVP solver as explain in  (insererlasection) on \(T'\) and \(L(A,2^{128-\ell},n)\).
 
-We call \(U\) the truncated vector : \(S'[\ell,128]\). We notice that \(U\) is in \(L(A,2^{128-\ell},n)\).
+% As \(U\) and \(T'\) have exactly the same \(6\) upper bits  know that \(\lVert U-T \rVert \leqslant \sqrt{n2^{2(128-6-\ell)}}\).
 
-We consider the case were we successfully guessed the 6 upper bits of each \(S_i\) and the \(\ell\) lower bits of \(S_0\). Then we know the 6 upper bits of \(U\).
+% So, to be sure that our algorithm return the vector \(U\), we need to find sets of parameters \((n,\ell)\) such that :
 
-(maybe dessin)
+% \[2\sqrt{n}^{(128-6-\ell)} \leqslant \lVert SVP(L(A,2^{128-\ell},n))\rVert \]
 
-We consider the vector \(T'\) computed in step (numerodelastepenquestion). To recover \(U\), we use an exact CVP solver as explain in  (insererlasection) on \(T'\) and \(L(A,2^{128-\ell},n)\).
+% \begin{itemize}
+% 	\item For \(\ell=11\), the smallest \(n\) we need is 39.
 
-As \(U\) and \(T'\) have exactly the same \(6\) upper bits  know that \(\lVert U-T \rVert \leqslant \sqrt{n2^{2(128-6-\ell)}}\).
+% 	\item For \(\ell=12\), the smallest \(n\) we need is 38.
 
-So, to be sure that our algorithm return the vector \(U\), we need to find sets of parameters \((n,\ell)\) such that :
+% 	\item For \(\ell=13\), the smallest \(n\) we need is again 38 .
 
-\[2\sqrt{n}^{(128-6-\ell)} \leqslant \lVert SVP(L(A,2^{128-\ell},n))\rVert \]
+% \end{itemize}
 
-\begin{itemize}
-	\item For \(\ell=11\), the smallest \(n\) we need is 39.
+% \begin{theorem}
+%   The algorithm described above works with probability.... in time ...
+% \end{theorem}
 
-	\item For \(\ell=12\), the smallest \(n\) we need is 38.
+\subsection{Complete State Reconstruction}
 
-	\item For \(\ell=13\), the smallest \(n\) we need is again 38 .
-
-\end{itemize}
-
-\begin{theorem}
-  The algorithm described above works with probability.... in time ...
-\end{theorem}
-
+Once all the rotations have be recovered and $\Delta S_0$ has been found
+entirely, the only thing that remain is to actually find the entire $S_0$. For
+this, we use again eq.~\eqref{eq:output}. For this we use the fact that there is
+an (unknown) \emph{carry} $0\leq \gamma_i \leq 1$ such that:
+\begin{align*}
+  S_i[0:64] &= S_0[0:64] \boxplus \nabla S_i[0:64] \\
+  S_i[64:128] &= S_0[64:128] \boxplus \nabla S_i[64:128] \boxplus \gamma_i \\
+\end{align*}
+This therefore gives:
+\[
+  X_i  \lll r_i = \bigl(S_0[0:64] \boxplus \nabla S_i[0:64]\bigr) \oplus \bigl(S_0[64:128] \boxplus \nabla S_i[64:128] \boxplus \gamma_i\bigr).
+\]
 \section{Implementation and Practical Results}

### 2020-05-28 article/main.tex

@@
 
-\begin{table}
+\begin{table}[htbp!]
   \centering
@@
 
-\bigskip
-
-\begin{table}
+\begin{table}[htbp!]
   \centering
@@
 	\hline
-	3 & 2.869065991049632 & \(4.08909120094543e^{12}\) & 19 \\
-	4 & 2.060717188416003 & \(2.43569932844650e^{14}\) & 13 \\
-	5 & 3.7692142154392747 & \(1.71669732658884e^{15}\) & 11 \\
-	6 & 2.6925925483728537 & \(1.02550589786192e^{16}\) & 8 \\
+	3 & 2.869065991049632 & \(4.08909120094543e^{12} \simeq 2^{42}\) & 19 \\
+	4 & 2.060717188416003 & \(2.43569932844650e^{14} \simeq 2^{48}\) & 13 \\
+	5 & 3.7692142154392747 & \(1.71669732658884e^{15}\simeq 2^{51}\) & 11 \\
+	6 & 2.6925925483728537 & \(1.02550589786192e^{16}\simeq 2^{53}\) & 8 \\
 	\hline

### 2020-05-29 article/main.tex
fix typo
les chiffres signicatifs
refix typo
3 tonnes de TODO pour floette

@@
 
-    \draw (0.3, 1.1) -- (0.3, 1.5) -- (-1, 1.5) -- (-1, -4) -- (5.8, -4);
+    \draw (0.3, 1.1) -- (0.3, 1.5) -- (-1, 1.5) -- (-1, -4) -- node[above] {$r_i$} (5.8, -4);
     \draw (-1.1, -2.9) -- +(0.2, -0.2);
@@
 
-    \draw[->] (x) -- +(0, -1.5);
+    \draw[->] (x) -- node[right] {$Y_i$} +(0, -1.5);
     \node at (6.4, -6.25) (xi) {$X_i$};
@@
 
+This is a recurring theme is this paper, so we now state the problem in
+general. If $a, b$ are integers modulo $2^{128}$ and $0\leq i<j<128$, then there
+is a \emph{carry} $0 \leq \gamma \leq 1$ (resp. a \emph{borrow}
+$0 \leq 1 \leq \beta$) such that:
+\begin{align}
+  (a \boxplus b)[i:j]  &= a[i:j] \boxplus  b[i:j] \boxplus  \gamma, \label{eq:caryy}  \\
+  (a \boxminus b)[i:j] &= a[i:j] \boxminus b[i:j] \boxminus \beta.  \label{eq:borrow}
+\end{align}
+
 \subsection{Reconstructing Linear Congruential Generators}
@@
 
-Let $\LCG_{n,k}(x, c)$ denote the vector $(u_0, \dots, u_{n-1}) \in \mathbb{Z}_{2^k}$ (we will omit $n$ and $k$ when they are obvious from the context).  It is easy to check that:
+Let $\LCG_{k}(x, c)$ denote the vector $(u_0, u_1, u_2, \dots)$ of integers
+modulo $2^k$.  It is easy to check that:
 \begin{align}
 \label{eq:lcg-additive}
-\LCG_{n,k}(x + y, c + d) &= \LCG_{n,k}(x, c) + \LCG_{n,k}(y, d),  \\
+\LCG_{k}(x + y, c + d) &= \LCG_{k}(x, c) + \LCG_{k}(y, d),  \\
 \label{eq:lcg-scalar}
-\LCG_{n,k}(\lambda x, \lambda c) &= \lambda \LCG_{n,k}(x, c).   
+\LCG_{k}(\lambda x, \lambda c) &= \lambda \LCG_{k}(x, c).   
 \end{align}
@@
 The LCG have been largely studied.
-This is why we know techniques to recover internal states knowing only upper or lower bits of the output, see \cite{Boyar1989}.
+\todo{expand this} This is why we know techniques to recover internal states knowing only upper or lower bits of the output, see \cite{Boyar1989}.
 A particular LCG is the Lehmer's generator \cite{Lehmer}, where the multiplier \(a\) is known and the increment is zero.
@@
 \begin{itemize}
-	\item By \ref{eq:lcg-additive}, the vector \(\LCG(x_1,0) + \LCG(x_2,0) = \LCG(x_1+x_2,0)\), hence it is still in \(L(n,k)\).
-	\item By \ref{eq:lcg-scalar}, the vector \((-1)\times \LCG(x,0) = \LCG(-x,0)\) hence it is still in \(L(n,k)\).
+	\item By \eqref{eq:lcg-additive}, the vector \(\LCG(x_1,0) + \LCG(x_2,0) = \LCG(x_1+x_2,0)\), hence it is still in \(L(n,k)\).
+	\item By \eqref{eq:lcg-scalar}, the vector \((-1)\times \LCG(x,0) = \LCG(-x,0)\) hence it is still in \(L(n,k)\).
 \end{itemize}
@@
 
-\[\begin{pmatrix}
-1&a&a^2&\dots&a^{n-1}\\
-0&2^k&0&\dots&0\\
-0&0&2^k&\dots&0\\
-\dots&\dots&\dots&\dots&\dots\\
-0&0&0&\dots&2^k\\
-\end{pmatrix}\]
+\[
+  \begin{pmatrix}
+    1 & a   & a^2 & \dots & a^{n-1} \\
+    0 & 2^k & 0   & \dots & 0 \\
+    0 & 0   & 2^k & \dots & 0 \\
+    \dots & \dots & \dots & \dots & \dots\\
+    0 & 0 & 0 & \dots & 2^k \\
+  \end{pmatrix}
+\]
 
@@
 
-We want to find an element \(U\) in the lattice only knowing the most significant bits of each of its coordinates.
+\todo{NOTE: J'introduis $\ell$ ici. ATTENTION RETENUES MODULAIRES.}  We want to
+find an element \(U\) in the lattice only knowing the $\ell$ most significant
+bits of each of its coordinates.  More precisely, let $T_i = U_i[k-\ell:k]$
+denote the top $\ell$ bits of $U_i$. We assume that there is a ``noise vector''
+$N$ such that $N_i \in \{-1, 0, 1\}$, and we set
+$\widetilde{T}_i = T_i + N_i \bmod 2^{\ell}$. The problem we have is the
+following: recover $U$ from its noisy truncation $\widetilde{T}$.
 
-We decompose \(U = T + (T-U)\) where \(T\) contains only the upper bits and \(T-U\) the lowers one.
+In principle, $2^{k-\ell} \widetilde{T}$ should be ``close'' to a Lattice
+point. \todo{Mais il faut se dire que $2^{k-\ell} T$ et
+  $2^{k-\ell} \widetilde{T}$ peuvent être éloignés, surtout si $T_i = 0$ et
+  $N_i=-1$, à cause du $\bmod 2^\ell$... L'argument de Julia était que dans ce
+  cas-là, $2^{k-\ell} \widetilde{T}$ est proche d'un \emph{autre} point du
+  réseau, en ajoutant un $2^k$ sur la $i$-ème composante.}
+
+Decompose $U = T + (T-U)$ where $T$ contains only the upper bits and $T-U$ the lowers one.
 There is no reason for \(T\) to be in the lattice, but as \(T-U\) is small, we know \(T\) is \emph{close} to \(U\).
@@
 
+\todo{En fait, on va juste utiliser $k=128$, donc établir $n=42$ est suffisant. Faire disparaître $\ell$.}
 In this article, we will only use an exact CVP solver when \(k = 128-\ell\) and when we know the \(6\) upper bit of our Lehmer's generator.
@@
 
-As using an exact CVP solver is costly, we'll try a cheaper but less precise way to find the closest vector.
+As using an exact CVP solver is costly, we will try a cheaper but less precise way to find the closest vector.
 
 We compute \(G(n,k)\) the LLL-reduction of the generating matrix of \(L(n,k)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
-\[\rVert G \lVert =  \max_{x \in \mathbb{R}}\frac{\lVert Gx \rVert_2}{\lVert x \rVert_2}\]
-In the case of the \(\lVert\rVert_2\) norm, \(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular value of \(G\).
-
-\medskip
+\[\rVert G \lVert =  \sup_{x \in \mathbb{R}^n}\frac{\lVert Gx \rVert_2}{\lVert x \rVert_2}\]
+In the case of the \(\lVert \cdot \rVert_2\) norm,
+\(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular
+value of $G$ (equivalently, $\sigma(G)$ is the absolute value of the largest
+eigenvalue of $G^{t} G$ --- \todo{recheck this}).
 
@@
 
+\todo{Ce serait beaucoup plus naturel à mon avis de dire qu'on connaît les $\ell$ bits de poids fort. A ce stade, rien ne vient justifier le ``+6''.}
 In this article, we will only use the Babai's rounding when \(k = 64\) and when we know the \(6+\ell\) upper bit of our Lehmer's generator.
@@
 	\hline
-	3 & 2.869065991049632 & \(4.08909120094543e^{12} \simeq 2^{42}\) & 19 \\
-	4 & 2.060717188416003 & \(2.43569932844650e^{14} \simeq 2^{48}\) & 13 \\
-	5 & 3.7692142154392747 & \(1.71669732658884e^{15}\simeq 2^{51}\) & 11 \\
-	6 & 2.6925925483728537 & \(1.02550589786192e^{16}\simeq 2^{53}\) & 8 \\
+	3 & 2.87 & \(4.09e^{12} \simeq 2^{41.9}\) & 19 \\
+	4 & 2.06 & \(2.44e^{14} \simeq 2^{47.8}\) & 13 \\
+	5 & 3.77 & \(1.72e^{15}\simeq 2^{50.6}\) & 11 \\
+	6 & 2.69 & \(1.03e^{16}\simeq 2^{53.2}\) & 8 \\
 	\hline
@@
 In figure~\ref{fig:Cknown}, we see that guessing $S_0[0:\ell]$ and a few 6-bit
-rotations give access to $S_i[58:64+\ell]$ for the corresponding
+rotations $r_i$ give access to $S_i[58:64+\ell]$ for the corresponding
 states. Therefore, looking at $S_i[\ell:64+\ell]$, we are facing a truncated
@@
 
-We consider the first three internal states
-$\mathbf{S} = (S_0, S_1, S_2) = \LCG_{3, 128}(S_0, c)$. We will guess the $\ell$
+We consider the sequence of internal states
+$\mathbf{S} = (S_0, S_1, \dots) = \LCG_{128}(S_0, c)$. We will guess the $\ell$
 least-significant bits of $S_0$, therefore let us assume that their value is
-known and denote it by $w$. We define $\mathbf{S}' = \LCG_{3, 128}(S_0 - w,
-0)$. By~\eqref{eq:lcg-additive}, we have
-$\mathbf{S}' = \mathbf{S} - \LCG_{3, 128}(w, c)$. The point is that the elements
-of $\mathbf{S}'$ follow a geometric progression of common ratio $a$; in
-addition, the $\ell$ least significant bits of each components are zero. The
-crux of the reconstruction algorithm is to find $\mathbf{S}'$.
+known and denote it by $w$. We define $\mathbf{S}' = \LCG_{128}(S_0 - w, 0)$ and
+$\mathbf{K} = \LCG_{128}(w, c)$ (this is Known). By~\eqref{eq:lcg-additive}, we
+have $\mathbf{S}' = \mathbf{S} - \mathbf{K}$. The point is that the elements of
+$\mathbf{S}'$ follow a geometric progression of common ratio $a$; in addition,
+the $\ell$ least significant bits of each components are equal to zero. It
+follows that $\mathbf{S}'[\ell:64+\ell]$ follows a geometric progression of
+common ratio $a$, modulo $2^{64}$. The crux of the reconstruction algorithm is
+to find $\mathbf{S}'[\ell:64+\ell]$.
 
-For this, set $\mathbf{T}' = \left(\mathbf{S}' / 2^\ell\right) \bmod 2^{64}$
-(the division by $2^\ell$ is exact). It follows that the $\mathbf{T}'$ is also a
-geometric progression with common ratio $a$, and we have access to the $\ell+6$
-high order bits of $T'_i$ if we guess the $i$-th rotation. We have seen in
-section~\ref{sec:geometric} that 19 known bits on 3 consecutive outputs are
-sufficient for a deterministic reconstruction of the full sequence. Therefore,
-we guess $\ell = 19$ low-order bits of the state and three consecutive
-rotations.
+We know $K_i[58:64+\ell]$ for all $i$, and for each guessed rotation $r_i$ we
+have access to $T_i \stackrel{def}{=} S_i[58:64+\ell]$.  We want
+$T'_i \stackrel{def}{=} (S_i \boxminus K_i)[58:64+\ell]$, which is the
+truncation of $S'_i$. Thanks to~\eqref{eq:borrow}, we know that there is an
+unknown vector $\mathbf{B}$ of \emph{borrows}, whose components are either 0 or
+1, such that
+$\mathbf{T'} = \mathbf{S}[58:64+\ell] \boxminus \mathbf{K}[58:64+\ell] \boxminus
+\mathbf{B}$.
 
-Therefore, the algorithm that reconstructs the internal state of the
-\textsf{PCG64} generator with known increment proceeds as shown in algorithm~\ref{algo:known}.
+Because the borrow are unknown, we in fact compute
+$\widetilde{\mathbf{T}'} = \mathbf{S}[58:64+\ell] \boxminus
+\mathbf{K}[58:64+\ell]$, and clearly
+$\widetilde{\mathbf{T}'} = \mathbf{T}' \boxplus \mathbf{B}$.
+
+We will guess $n$ rotations and $\ell$ least-significant bits of the state, for
+a total of $2^{6n + \ell}$ guessed bits. Table~\ref{tab:babai} gives a
+lower-bound on $\ell$ given $n$, and we see that the total number of guessed
+bits reaches a minimum of $37$ when $n=3$ and $\ell=19$. Therefore, we guess
+$\ell = 19$ low-order bits of the state and three consecutive rotations.
+
+The algorithm that reconstructs the internal state of the \textsf{PCG64}
+generator with known increment proceeds as shown in
+algorithm~\ref{algo:known}. The point is that when the guesses are correct, then
+from the truncated geometric series $\mathbf{T'}$, the solution of the CVP
+instance reveals $U_j = S_j'[\ell:64+\ell]$. From there, the correction of the
+algorithm is easily established.
+
+The procedure is completely practical. More details are given in section~\ref{sec:implem}.
 
@@
   \Procedure{ReconstructState${}_\ell$}{$X_0, X_1, X_2$}
+  \State // \emph{Statement involving $j$ must be repeated for $j=0, 1, 2$.}
+  \State $\ell \gets 19$
   \For{$0 \leq w < 2^{\ell}$} \Comment{Guess least-significant bits of $S_0$}
-    \State $w_j = a^j w \bmod 2^{\ell}$ \Comment{Propagate to $S_1, S_2$}
-    \State $S'_j \gets a^j w + c \sum_{k = 0}^j a^k \bmod 2^{128}$    \Comment{Shifted states}
+%    \State $w_j = a^j w \bmod 2^{\ell}$ \Comment{Propagate to $S_1, S_2$}
+    \State $K_j \gets a^j w + c (a^j - 1)(a-1)^{-1} \bmod 2^{128}$    \Comment{Known part}
 
@@
 
-    \State $ T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(w_j \oplus Y_j[0:\ell]\right)$  \Comment{Truncated LCG output}
+    \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(K_j \oplus Y_j\right)[0:\ell]$  \Comment{Truncated LCG output}
 
-    \State $ T'_j \gets    T_j -  S'_i[58:64+\ell]$  \Comment{Shifted Truncated LCG output}
+    \State $\widetilde T'_j \gets T_j \boxminus K_j[58:64+\ell]$  \Comment{Truncated geometric series}
+    \State $(U_0, U_1, U_2) \gets \left\lfloor 2^{58-\ell} \cdot (\widetilde T'_0, \widetilde T'_1, \widetilde T'_2) \cdot  G_3^{-1} \right\rceil \cdot G_3$ \Comment{CVP (Babai rounding)}
 
-    \State $(U_0, U_1, U_2) \gets \left\lfloor (T'_0, T'_1, T'_2) \cdot 2^{58-\ell} \cdot \widetilde G_3^{-1} \right\rceil \cdot \widetilde G_3$ \Comment{CVP via Babai rounding}
+    \State $S_0[0:64] \gets  K_0[0:64] + 2^{\ell} \cdot U_0[0:64-\ell]$ \Comment{Reconstruct $S_0$}
+    \State $S_0[64:128] \gets S_0[0:64] \oplus Y_0$
 
-    \State $ S_0[0:64] \gets  S'_0[0:64] + 2^{\ell} \cdot U_0[0:64-\ell]$ \Comment{Reconstruct $S_0$}
-    \State $ S_0[64:128] \gets S_0[0:64] \oplus Y_0$
-
-    \State $ S_1 \gets a  S_0 + c$ \Comment{Recompute $X_1$}
+    \State $S_1 \gets a  S_0 + c$ \Comment{Recompute $X_1$}
     \State $\widehat{Y}_1 =  S_1[0:64] \oplus  S_1[64:128]$
@@
     \If{$\widehat {Y}_1 = Y_1$} \Comment{Check consistency}
-    \State output $ S_0$ as a candidate internal state.
+    \State output $S_0$ as a candidate internal state.
     \EndIf
@@
 
-% \begin{theorem}
-%   Algorithm \textsc{ReconstructState} is correct (it always return the correct internal state).
-% \end{theorem}
-
-% \todo{avec la nouvelle partie 3 et le blabla au dessus de l'algo, y'a plus vraiment besoin de preuve}
-% Charles: je suis plutôt d'accord.
-
 \section{Predicting PCG64 with Unknown Increment}
@@
 in guessing $S_0[64:128]$ and $S_1[64:128]$; using eq.~\eqref{eq:output}, this
-would reveal $S_0$ and $S_1$, from which the increment $c$ could be easily
-deduced. This would take $2^{128}$ iterations of a very simple procedure, but is
-completely infeasible.
+would reveal $S_0$ and $S_1$; from there, the increment $c$ is easily to obtain,
+and every secret information has been reconstructed. This would take $2^{128}$
+iterations of a very simple procedure, which is completely infeasible.
 
-The global strategy to obtain a practical prediction algorithm is essentially
-the same as before (target a truncated version of the underlying linear
-congruential generator), but is a bit more complicated to implement.
+% The global strategy to obtain a practical prediction algorithm is essentially
+% the same as before (target a truncated version of the underlying linear
+% congruential generator), but is a bit more complicated to implement.
 
@@
 and $\Delta S_0$ is sufficient to compute all subsequent states (and recover the
-unknown increment $c$). Let us set:
+unknown increment $c$). The global ``guess-and-determine'' strategy is
+essentially the same as before: gaining access to a truncated version of
+$\Delta S_i$, solving a small SVP instance, reconstructing $\Delta S_0$, then
+checking consistency.
+
+Let us set:
 \begin{align*}
-  \nabla S_i \stackrel{def}{=} S_i - S_0 \equiv \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
+  \nabla S_i \stackrel{def}{=} S_i \boxminus S_0 = \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
 \end{align*}
-Note that $\nabla S_0 = \Delta S_0$. The prediction algorithm we propose proceeds in three phases:
+Note that $\nabla S_1 = \Delta S_0$. The prediction algorithm we propose proceeds in three phases:
 \begin{enumerate}
@@
 
-Another problem is that of confirming that the guesses are correct. When $c$ was
-known, we could reconstruct the internal state; from there, filtering out the
-bad guesses was easy. When $c$ is unknown, the same strategy does not work, but
-a very strong consistency check can still be implemented. The essential idea is
-the following: once $\nabla S_0[0:64+\ell]$ is known, then all the rotations can
+Confirming that the guesses are correct is less immediate. When $c$ was known,
+we could reconstruct the internal state; from there, filtering out the bad
+guesses was easy. When $c$ is unknown, the same strategy does not work, but a
+very strong consistency check can still be implemented. The essential idea is
+the following: once $\Delta S_0[0:64+\ell]$ is known, then all the rotations can
 be determined accurately and with little computational effort. This in turn
@@
 
-Once we have found $\nabla S_0[0:64+\ell]$, we can compute
+Once we have found $\Delta S_0[0:64+\ell]$, we can compute
 $\nabla S_i[0:64+\ell]$ for any $i$ (geometric progression); because we have
@@
 The unknown value $x_0$ will not be recovered until the third phase of the whole
-procedure. We obtain the $i$-th state by
-$S_i \equiv \nabla S_i + S_0 \bmod 2^{64+\ell}$; however, because the ``middle''
-of $S_0$ is unknown, an unknown \emph{carry} may cross the 64-th during the
-addition and perturb $S_i[64:64+\ell]$. As a result, there are unknown
-quantities $0 \leq y_i < 2^{64-\ell}$ and $0 \leq z_i \leq 1$ such that:
+procedure. We obtain the $i$-th state by $S_i \equiv \nabla S_i \boxplus S_0$;
+however, because the ``middle'' of $S_0$ is unknown, we have the same problem as
+in eq.~\eqref{eq:diff_borrow}: an unknown \emph{carry} may cross the 64-th bit
+during the addition and perturb $S_i[64:64+\ell]$. As a result, there is an
+unknown vector $\mathbf{C}$, whose components are either 0 or 1, such that such
+that:
 \begin{align*}
-  S_i[0:\ell] &\equiv \nabla S_i + w_0 \bmod 2^\ell \\
-  S_i[\ell:64] &\equiv y_i \\
-  S_i[64:64+\ell] &\equiv z_i + \nabla S_i[64:64+\ell] + (w_0 \oplus Y_0[0:\ell]) \bmod 2^\ell \\
+  S_i[64:64+\ell] &= C_i \boxplus \nabla S_i[64:64+\ell] \boxplus (w_0 \oplus Y_0[0:\ell]) \bmod 2^\ell \\
 \end{align*}
@@
   \For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
-  \State $S'_j \gets a^j w_0 + (2c+1) \sum_{k = 0}^j a^k \bmod 2^{128}$ \Comment{Shifted states}
+  \State $K_j \gets a^j w_0 + (2c+1)(a^j - 1)(a-1)^{-1} \bmod 2^{128}$ \Comment{Known part}
   \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
   \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
-  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(w_j \oplus Y_j[0:\ell]\right)$ \Comment{Truncated LCG}
-  \State $T'_j \gets (T_j -  S'_i[58:64+\ell]) \bmod 2^{6+\ell}$ \Comment{Shifted truncated LCG}
+  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(K_j \oplus Y_j\right)[0:\ell]$ \Comment{Truncated LCG}
+  \State $T'_j \gets T_j \boxminus  K_i[58:64+\ell]$ \Comment{Truncated geometric series}
+  \State \todo{Attention double retenue : $\pm 1$.}
   \State $\Delta T'_j = (T'_{j+1} - T'_j) \bmod 2^{6 + \ell}$  \Comment{Difference}
@@
 
-% \subsection{preuve cas difficile}
-% \todo{À ranger au bon endroit}
-% % Babai ça marche pas (genre zero)
-
-% The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero.
-
-% We call \(U\) the truncated vector : \(S'[\ell,128]\). We notice that \(U\) is in \(L(A,2^{128-\ell},n)\).
-
-% We consider the case were we successfully guessed the 6 upper bits of each \(S_i\) and the \(\ell\) lower bits of \(S_0\). Then we know the 6 upper bits of \(U\).
-
-% (maybe dessin)
-
-% We consider the vector \(T'\) computed in step (numerodelastepenquestion). To recover \(U\), we use an exact CVP solver as explain in  (insererlasection) on \(T'\) and \(L(A,2^{128-\ell},n)\).
-
-% As \(U\) and \(T'\) have exactly the same \(6\) upper bits  know that \(\lVert U-T \rVert \leqslant \sqrt{n2^{2(128-6-\ell)}}\).
-
-% So, to be sure that our algorithm return the vector \(U\), we need to find sets of parameters \((n,\ell)\) such that :
-
-% \[2\sqrt{n}^{(128-6-\ell)} \leqslant \lVert SVP(L(A,2^{128-\ell},n))\rVert \]
-
-% \begin{itemize}
-% 	\item For \(\ell=11\), the smallest \(n\) we need is 39.
-
-% 	\item For \(\ell=12\), the smallest \(n\) we need is 38.
-
-% 	\item For \(\ell=13\), the smallest \(n\) we need is again 38 .
-
-% \end{itemize}
-
 % \begin{theorem}
@@
 entirely, the only thing that remain is to actually find the entire $S_0$. For
-this, we use again eq.~\eqref{eq:output}. For this we use the fact that there is
-an (unknown) \emph{carry} $0\leq \gamma_i \leq 1$ such that:
-\begin{align*}
-  S_i[0:64] &= S_0[0:64] \boxplus \nabla S_i[0:64] \\
-  S_i[64:128] &= S_0[64:128] \boxplus \nabla S_i[64:128] \boxplus \gamma_i \\
-\end{align*}
-This therefore gives:
+this, we use again eq.~\eqref{eq:output}, coupled with the ``differences'':
 \[
-  X_i  \lll r_i = \bigl(S_0[0:64] \boxplus \nabla S_i[0:64]\bigr) \oplus \bigl(S_0[64:128] \boxplus \nabla S_i[64:128] \boxplus \gamma_i\bigr).
+  X_i  \lll r_i = \bigl(S_0 \boxplus \nabla S_i)[0:64] \oplus \bigl(S_0 \boxplus \nabla S_i)[64:128].
 \]
+
 \section{Implementation and Practical Results}
+\label{sec:implem}
 

### 2020-05-29 article/main.tex
les retenus section 3

@@
 
-\section{Reconstructing Truncated Geometric Sequences}
+\section{Reconstructing Truncated Geometric progressions}
 \label{sec:geometric}
@@
 
-\subsection{The lattice associated to a Lehmer's generator}
+\subsection{The lattice associated to a geometric progression}\label{lattice}
 
@@
 
-\todo{NOTE: J'introduis $\ell$ ici. ATTENTION RETENUES MODULAIRES.}  We want to
-find an element \(U\) in the lattice only knowing the $\ell$ most significant
-bits of each of its coordinates.  More precisely, let $T_i = U_i[k-\ell:k]$
-denote the top $\ell$ bits of $U_i$. We assume that there is a ``noise vector''
-$N$ such that $N_i \in \{-1, 0, 1\}$, and we set
-$\widetilde{T}_i = T_i + N_i \bmod 2^{\ell}$. The problem we have is the
-following: recover $U$ from its noisy truncation $\widetilde{T}$.
+We want to find an element \(U\) in the lattice only knowing the $\ell$ most significant bits of each of its coordinates.
+More precisely, let $T_i = U_i[k-\ell:k]$ denote the top $\ell$ bits of $U_i$.
+We assume that there is a ``noise vector'' $N$ such that $N_i \in \{-1, 0, 1\}$, and we set $\widetilde{T}_i = T_i + N_i \bmod 2^{\ell}$.
+The problem we have is the  following: recover $U$ from its noisy truncation $\widetilde{T}$.
 
-In principle, $2^{k-\ell} \widetilde{T}$ should be ``close'' to a Lattice
-point. \todo{Mais il faut se dire que $2^{k-\ell} T$ et
-  $2^{k-\ell} \widetilde{T}$ peuvent être éloignés, surtout si $T_i = 0$ et
-  $N_i=-1$, à cause du $\bmod 2^\ell$... L'argument de Julia était que dans ce
-  cas-là, $2^{k-\ell} \widetilde{T}$ est proche d'un \emph{autre} point du
-  réseau, en ajoutant un $2^k$ sur la $i$-ème composante.}
+In principle, $2^{k-\ell} \widetilde{T}$ should be ``close'' to \(U\). We must distinguish three cases.
+\begin{itemize}
+	\item If for each \(i\) the integer \(\widetilde{T}_i\) satisfies \(1\leqslant\widetilde{T}_i\leqslant 2^{k-l}-2\) then \(\widetilde{T}_i = T_i + N_i\) (without modulo). So
 
-Decompose $U = T + (T-U)$ where $T$ contains only the upper bits and $T-U$ the lowers one.
-There is no reason for \(T\) to be in the lattice, but as \(T-U\) is small, we know \(T\) is \emph{close} to \(U\).
+	\begin{align*}
+	\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert &=\lvert 2^{k-\ell}(T_i + N_i)-U_i \rvert \\
+	 &= \lvert 2^{k-\ell}N_i - U_i[0,k-\ell] \rvert\\
+	 &\leqslant \lvert 2^{k-\ell}N_i\rvert + \lvert U_i[0,k-\ell] \rvert \\
+	 &\leqslant 2^{k-\ell} + 2^{k-\ell}-1\\
+	 &\leqslant 2^{k-\ell +1 }- 1
+	\end{align*}
+	Hence \(\lVert U -2^{k-\ell} \widetilde{T} \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \)
+
+	\item If there are indexes such that \(\widetilde{T}_i = 2^\ell-1\).
+	\begin{itemize}
+		\item If \(N_i \in \{0,1\}\), then \(\widetilde{T}_i = T_i + N_i\) (without modulo) and \[\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert \leqslant \leqslant 2^{k-\ell +1 }- 1\]
+		\item If \(T_i = 0\) and \(N_i = -1\), then \(U_i=U_i[0,k-\ell]\) and  \[\lvert 2^{k-\ell}\widetilde{T}_i-(U_i + 2^k) \rvert =\lvert 2^{k-\ell}(2^\ell -1) -(U_i + 2^k) \rvert = \lvert 2^{k-\ell} + U_i[0,k-\ell]\rvert \leqslant 2^{k-\ell + 1} -1 \]
+	\end{itemize}
+	We need a new vector \(U'\) defined as follow.
+	\begin{itemize}
+		\item If \(T_i = 0\) and \(N_i = -1\), \(U'_i=U_i+2^k\).
+		\item Else \(U'_i=U_i\).
+	\end{itemize}
+	In both case \(\lVert U' -2^{k-\ell} \widetilde{T} \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \)
+	\item If there are indexes such that \(\widetilde{T}_i = 0\).
+	\begin{itemize}
+		\item  If  \(N_i \in \{-1,0\}\), then \(\widetilde{T}_i = T_i + N_i\) (without modulo) and \[\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert \leqslant \leqslant 2^{k-\ell +1 }- 1\]
+		\item If \(N_i = 1\) we have a problem.
+	\end{itemize}
+	So, if there are indexes such that \(\widetilde{T}_i = 0\), we have to guess if \(N_i=1\) or not.
+
+\end{itemize}
 
@@
 
-We know that \(U\) is in the lattice, \(T\) is usually not and \(|T-U|\) is small.
-Then we can hope for \(U\) to be the closet vector of \(T\) which is in the lattice : the vector \(U\) may be the solution of \(CVP(L,T)\).
+We define \(U'\) as in subsection \ref{lattice}. We want to know if \(CVP(L,2^{k-\ell}\widetilde{T})\) is indeed \(U'\).
 
-What can we say about \(|CVP(L,T)-U|\) ?
+What can we say about \(|CVP(L,2^{k-\ell}\widetilde{T})-U'|\) ?
 
 \begin{align*}
-|CVP(L,T)-U| &= |CVP(L,T)-T+T-U|\\
-& \leqslant |CVP(L,T)-T|+|T-U|\\
+|CVP(L,T)-U| &= |CVP(L,2^{k-\ell}\widetilde{T})-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
+& \leqslant |CVP(L,T)-2^{k-\ell}\widetilde{T}|+|2^{k-\ell}\widetilde{T}-U'|\\
 \end{align*}
 
-But, as the vector \(U\) is in the lattice and by definition of the CVP, \(|CVP(L,T)-T| \leqslant |T-U| \), hence 
-\[|CVP(L,T)-U|\leqslant 2|T-U|\]
+But, as the vector \(U'\) is in the lattice and by definition of the CVP, \(|CVP(L,2^{k-\ell}\widetilde{T})-2^{k-\ell}\widetilde{T}| \leqslant |2^{k-\ell}\widetilde{T}-U'| \), hence 
+\[|CVP(L,2^{k-\ell}\widetilde{T})-U'|\leqslant 2|2^{k-\ell}\widetilde{T}-U'|\]
 
-If we can prove that the right side of the inequality is smaller than first minimum of the lattice \(\lambda_1(L)\), then we have proved that \(CVP(L,T)\) is indeed the vector we were searching for.
+If we can prove that the right side of the inequality is smaller than first minimum of the lattice \(\lambda_1(L)\), then we have proved that \(CVP(L,2^{k-\ell}\widetilde{T})\) is indeed the vector we were searching for.
 
@@
 
-\todo{En fait, on va juste utiliser $k=128$, donc établir $n=42$ est suffisant. Faire disparaître $\ell$.}
-In this article, we will only use an exact CVP solver when \(k = 128-\ell\) and when we know the \(6\) upper bit of our Lehmer's generator.
-Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{122-\ell} \).
-So, if we fix \(\ell\) we can search for the minimum \(n\) such that \[2\sqrt{n}2^{122-\ell} \leqslant \lambda_1(L(128-\ell,n))\].
+In this article, we will only use an exact CVP solver when \(k = 128\) and \(\ell = 6\).
+As discussed in the previous subsection, we suppose we are never in the situation\(\widetilde{T} = 2^\ell-1\) and \(N_i=1\).
+Hence \(\lVert T-U \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \).
+So, if we fix \(\ell = 6\) we can search for the minimum \(n\) such that \[2\sqrt{n}2^{122} \leqslant \lambda_1(L(128,n))\].
 
-\begin{table}[htbp!]
-  \centering
-    \begin{tabular}{|c|c|}
-	\hline
-	\(\ell\)  & minimum \(n\) \\
-	\hline
-	0  & 42 \\
-	11 & 39 \\
-	12 & 38 \\
-	13 & 38 \\
-	\hline
-\end{tabular}
-  \caption{\todo{WRITE-ME}}
-  \label{tab:exact_cvp}
-\end{table}
+After computation, we find that the minimal \(n\) is 42. \todo{a vérifier (snif snif)}
 
@@
 value of $G$ (equivalently, $\sigma(G)$ is the absolute value of the largest
-eigenvalue of $G^{t} G$ --- \todo{recheck this}).
+eigenvalue of $G^{t} G$).
 
-As \(T\) is not an element of the lattice, \(TG^{-1}\) is not an integer vector.
+As \(2^{k-\ell}\widetilde{T}\) is not an element of the lattice, \(2^{k-\ell}\widetilde{T}G^{-1}\) is not an integer vector.
 
-We fix \(r = rounding(TG^{-1}) \), then \(rG\) is an element of the lattice. What can we say about the distance between \(rG\) and \(U\) ?
+We fix \(r = rounding(2^{k-\ell}\widetilde{T}G^{-1}) \), then \(rG\) is an element of the lattice. What can we say about the distance between \(rG\) and \(U'\) ?
 
 \begin{align*}
-\lVert U - rG \rVert &= \lVert U - (r-TG^{-1} + TG^{-1})G \rVert\\
-&= \lVert U - T + (r-TG^{-1})G \rVert\\
-&\leqslant \lVert U - T \rVert + \lVert(r-TG^{-1})\rVert \times \lVert G\rVert\\	
+\lVert U' - rG \rVert &= \lVert U - (r-2^{k-\ell}\widetilde{T}G^{-1} + 2^{k-\ell}\widetilde{T}G^{-1})G \rVert\\
+&= \lVert U' - 2^{k-\ell}\widetilde{T} + (r-2^{k-\ell}\widetilde{T}G^{-1})G \rVert\\
+&\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(r-2^{k-\ell}\widetilde{T}G^{-1})\rVert \times \lVert G\rVert\\	
 \end{align*}
 
-By definition, \(r\) is the closest integer vector to \(TG^{-1}\). But as \(U\) is an element of the lattice, \(UG^{-1}\) is an integer vector. Thus \(r-TG^{-1}\) is shorter than \(UG^{-1}-TG^{-1}\).
+By definition, \(r\) is the closest integer vector to \(2^{k-\ell}\widetilde{T}G^{-1}\).
+But as \(U'\) is an element of the lattice, \(U'G^{-1}\) is an integer vector.
+Thus \(r-2^{k-\ell}\widetilde{T}G^{-1}\) is shorter than \(U'G^{-1}-2^{k-\ell}\widetilde{T}G^{-1}\).
 
@@
 \begin{align*}
-\lVert U - rG \rVert &\leqslant \lVert U - T \rVert + \lVert(r-TG^{-1})\rVert \times \lVert G\rVert\\	
-&\leqslant \lVert U - T \rVert + \lVert(UG^{-1}-TG^{-1})\rVert \times \lVert G\rVert\\	
-&\leqslant \lVert U - T \rVert + \lVert(U-Y)\rVert \times \lVert G^{-1} \rVert  \times \lVert G\rVert\\
-& 	\leqslant \lVert U - T \rVert \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert )\\
+\lVert U' - rG \rVert &\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(r-2^{k-\ell}\widetilde{T}G^{-1})\rVert \times \lVert G\rVert\\	
+&\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(U'G^{-1}-2^{k-\ell}\widetilde{T}G^{-1})\rVert \times \lVert G\rVert\\	
+&\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(U'-2^{k-\ell}\widetilde{T})\rVert \times \lVert G^{-1} \rVert  \times \lVert G\rVert\\
+& 	\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert )\\
 \end{align*}
@@
 
-\todo{Ce serait beaucoup plus naturel à mon avis de dire qu'on connaît les $\ell$ bits de poids fort. A ce stade, rien ne vient justifier le ``+6''.}
-In this article, we will only use the Babai's rounding when \(k = 64\) and when we know the \(6+\ell\) upper bit of our Lehmer's generator.
-Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{58-\ell} \).
-So, if we fix \(n\) we can search for the minimum \(\ell\) such that \[(1+\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert)\sqrt{n}2^{58-\ell} \leqslant \lambda_1(L(n,64))\]
+In this article, we will only use the Babai's rounding when \(k = 64\).
+Hence \(\lVert 2^{k-\ell}\widetilde{T}-U' \rVert \leqslant \sqrt{n}2^{64-\ell+1}-1 \).
+So, if we fix \(n\) we can search for the minimum \(\ell\)\todo{ATTENTION ici \(\ell\) c'est le nombre de bit devinés, donc l'ancien \(6+\ell\)} such that \[(1+\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert)\sqrt{n}2^{64-\ell+1}-1 \leqslant \lambda_1(L(n,64))\]
 
-\begin{table}[htbp!]
+\begin{table}[htbp!] \todo{a vérifier}
   \centering
@@
   \end{tabular}
-  \caption{\todo{WRITE-ME}}
+  \caption{minimal \(\ell\) needed for a given \(n\)}
   \label{tab:babai}

### 2020-05-30 Cknown/main.c
C known: comments

@@ -3,7 +3,6 @@
 #include <stdio.h>
 #include <time.h>
 
-// avec ICC, ce truc ne marche pas sur ppti-gpu-1 si on met -xHost (zmm?). Il faut investiguer...
 
 void result_found(pcg128_t S, double start)
 {
@@ -38,9 +37,7 @@ int main()
     u64 done = 0;
     
     #pragma omp parallel for
-    for (u64 W0 = 0; W0 < (1<<known_low) ; W0++){
-	
-	/*Variables privées*/
+    for (u64 W0 = 0; W0 < (1<<known_low) ; W0++){	
 	pcg128_t S[nbiter];
 	int rot[nbiter] = {0, 0, 0};
 	struct task_t task;
@@ -53,7 +50,7 @@ int main()
 
 
 	for (int r = 0 ; r < 1 << (nbiter*6) ; r++) {
-	    /***** Modification de rot et unrotX *****/
+		/* bump rotations */
 	    rot[0] = (rot[0] + 1) % 64;
 	    int i = 0;
 	    while (rot[i] == 0 && i < nbiter) {
@@ -63,7 +60,6 @@ int main()
 	    if (i > 0)
 	    	refresh_task(rot, &task);
 
-	    /***** Résolution *****/
 	    if (solve(S, rot, &task))
 		result_found(S[0], start);
 	}

### 2020-05-30 article/main.tex
section 5.3

@@
 % !TeX spellcheck = en
-\documentclass[preprint,svgnames]{iacrtrans}
+\documentclass[submission,svgnames,journal=tosc]{iacrtrans}
 
@@
 
-%\usepackage[noend]{algpseudocode}
-
-%\usepackage[pdftex,bookmarks,bookmarksopen,bookmarksdepth=3]{hyperref}
-%\hypersetup{colorlinks=true,citecolor=red,linkcolor=red,urlcolor=black}
-
 \DeclareMathSymbol{:}{\mathord}{operators}{"3A}
@@
 \DeclareMathOperator{\MCG}{MCG}
+\DeclareMathOperator{\MAJ}{MAJ}
 
 \usepackage[noend]{algpseudocode}
-%\usepackage{algorithmic}
-
-% knuth-style algos
-% \newcommand{\slug}{\hbox{\kern1.5pt\vrule width2.5pt height6pt depth1.5pt\kern1.5pt}}
-% \def\xskip{\hskip 7pt plus 3pt minus 4pt}
-% \newdimen\algindent
-% \newif\ifitempar \itempartrue % normally true unless briefly set false
-% \def\algindentset#1{\setbox0\hbox{{\bf #1.\kern.25em}}\algindent=\wd0\relax}
-% \def\algbegin #1 #2{\algindentset{#21}\alg #1 #2} % when steps all have 1 digit
-% \def\aalgbegin #1 #2{\algindentset{#211}\alg #1 #2} % when 10 or more steps
-% \def\aaalgbegin #1 #2{\algindentset{#2111}\alg #1 #2} % when 10 or more steps
-% \def\alg#1(#2). {\medbreak % Usage: \algbegin Algorithm A (algname). This...
-%   \noindent{\bf#1}({\it#2\/}).\xskip\ignorespaces}
-% \def\algstep#1.{\ifitempar\smallskip\noindent\else\itempartrue
-%   \hskip-\parindent\fi
-%   \hbox to\algindent{\bf\hfil #1.\kern.25em}%
-%   \hangindent=\algindent\hangafter=1\ignorespaces}
-% end of borrowed macros
 
@@
 
-\title{Predicting the PCG PSeudo-Random Number Generator In Practice}
+\title[Predicting the PCG Pseudo-Random Number Generator in Practice]{Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice}
 
@@
 \item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{39}$.
-\item Fully reconstruct $\Delta S_0$ from this partial knowledge and the outputs.
-\item Reconstruct $S_0$ from $\Delta S_0$ and the outputs.
+\item Reconstruct all rotations $r_i$ from this partial knowledge.
+\item Fully reconstruct $\Delta S_0$ from the rotations.
+\item Reconstruct $S_0$ from $\Delta S_0$ and the rotations.
 \end{enumerate}
@@
 
+\todo{check indices nabla}
+
 \begin{algorithm}
@@
 this, we use again eq.~\eqref{eq:output}, coupled with the ``differences'':
+\begin{align*}
+  S_i &= S_0 \boxplus \nabla S_i \\
+  Y_i &= S_i[0:64] \oplus S_i[64:128].
+\end{align*}
+
+The $Y_i$ and $\nabla S_i$ are known, $\nabla S_0 = 0$, and the problem consists
+in recovering $S_0$. We could probably encode it as an instance of \textsf{SAT},
+feed it to a SAT-solver and be done with it.
+
+Nevertheless, here is a detailed recovery procedure which obtain all bits of
+$S_0$, from right to left, by exploiting the non-linearity of modular
+addition. It takes negligible time. Let $\mathcal{C}_i$ the vector of (incoming)
+carries generated during the addition of $S_0$ and $\nabla S_i$:
+\begin{align*}
+  S_i[j]           &= S_0[j] \oplus \nabla S_i[j] \oplus \mathcal{C}_i[j]\\
+  \mathcal{C}_i[j] &= \begin{cases}
+    0 &\text{if } j = 0 \\
+    \MAJ(S_0[j-1], \nabla S_i[j-1], \mathcal{C}_i[j-1]) &\text{if } j > 0
+  \end{cases}
+\end{align*}
+
+Combining all the above, we have:
+\begin{equation}\label{eq:master}
+  Y_i[j] = Y_0[j] \oplus \bigl( \nabla S_i[j] \oplus \nabla S_i[64+j] \bigr) \oplus \bigl(\mathcal{C}_i[j] \oplus \mathcal{C}_i[64+j]\bigr)\\
+\end{equation}
+This useful equation enables an induction process.
+
+\begin{itemize}
+\item When $j=0$, the $0$-th carries are zero, and therefore eq.~\eqref{eq:master}
+reveals the 64-th carries:
 \[
-  X_i  \lll r_i = \bigl(S_0 \boxplus \nabla S_i)[0:64] \oplus \bigl(S_0 \boxplus \nabla S_i)[64:128].
+  \mathcal{C}_i[64+j] = \bigl( Y_0[j] \oplus Y_i[j] \bigr)  \oplus \bigl( \nabla S_i[j] \oplus \nabla S_i[64+j] \bigr).
 \]
+
+\item Next, suppose that $\mathcal{C}_i[0:j]$, $S_0[0:j-1]$, $\mathcal{C}_i[64:64+j]$
+and $S_0[64:64+j-1]$ are known, for all $i$. We can compute
+$\mathcal{C}_i[j] \oplus \mathcal{C}_i[64+j]$ for any $i$ using
+eq.~\eqref{eq:master}. We then look a a specific index $\mathfrak{i} > 0$ such that
+\[
+   \nabla S_{\mathfrak{i}}[j-1] \neq \mathcal{C}_{\mathfrak{i}}[j-1] \quad \text{and} \quad \nabla S_{\mathfrak{i}}[64 + j-1] = \mathcal{C}_{\mathfrak{i}}[64 + j-1].
+\]
+The point is that, thanks to the majority functions,
+$\mathcal{C}_{\mathfrak{i}}[j] = S_0[j-1]$ and
+$\mathcal{C}_{\mathfrak{i}}[64+j] = \nabla S_{\mathfrak{i}}[64 + j-1]$. It follows
+that:
+\[
+  S_0[j-1] = Y_0[j-1] \oplus Y_{\mathfrak{i}}[j-1] \oplus \bigl( \nabla S_{\mathfrak{i}}[j-1] \oplus \nabla S_{\mathfrak{i}}[64+j-1] \oplus \nabla S_{\mathfrak{i}}[64 + j-1]\bigr)
+\]
+From there, we also have $S_0[64+j-1] = Y_0[64+j-1] \oplus S_0[j-1]$, and the
+$j$-th carry bits can be computed normally.
+\end{itemize}
+
+The whole procedure is shown in algorithm~\ref{algo:last}.
+
+\begin{algorithm}
+\begin{algorithmic}[1]
+  \Function{ReconstructState}{$\Delta S_0, r_0, \dots, r_k, X_0, \dots, X_k$}
+  \State \todo{check indices nabla}
+  \For{$i = 0, 1, \dots, k$} \Comment{Setup}
+  \State $\nabla S_i \gets \Delta S_0 (a^i-1)(a-1)^{-1} \bmod 2^{128}$
+  \State $Y_i \gets X_i \lll r_i$ \Comment{Undo rotations}
+  \State $\mathcal{C}_i[0] \gets 0$ \Comment{Bootstrap induction}
+  \State $\mathcal{C}_i[64] \gets \bigl( Y_i[0] \oplus Y_i[j] \bigr) \oplus \bigl( \nabla S_i[j] \oplus \nabla S_i[64+j]$
+  \EndFor
+  \For{$j=1, 2, \dots, 64$} \Comment{Induction}
+  \State $\mathfrak{i} \gets \bot$ \Comment{Find good index}
+  \For{$k = 1, 2, \dots, k$}
+  \If{$\nabla S_{k}[j-1] \neq \mathcal{C}_{k}[j-1] \wedge \nabla S_{k}[64 + j-1] = \mathcal{C}_{k}[64 + j-1]$}
+  \State $\mathfrak{i} \gets k$
+  \EndIf
+  \EndFor
+  \If{$\mathfrak{i} = \bot$} \Comment{No suitable indice found?}
+  \State \textbf{Abort} with Failure
+  \EndIf
+  \State  \Comment{Compute next state bit}
+  \State $S_0[j-1] \gets Y_0[j-1] \oplus Y_{\mathfrak{i}}[j-1] \oplus \bigl( \nabla S_{\mathfrak{i}}[j-1] \oplus \nabla S_{\mathfrak{i}}[64+j-1] \oplus \nabla S_{\mathfrak{i}}[64 + j-1]\bigr)$
+  \State $S_0[64 + j-1] \gets Y_0[64 + j-1] \oplus S_0[j-1]$
+  \For{$i = 0, 1, \dots, k$} \Comment{Compute next carries}
+  \State $\mathcal{C}_i[j] \gets \MAJ(S_0[j-1], \nabla S_i[j-1], \mathcal{C}_i[j-1])$
+  \State $\mathcal{C}_i[64 + j] \gets \MAJ(S_0[64+j-1], \nabla S_i[64+j-1], \mathcal{C}_i[64+j-1])$
+  \EndFor
+  \EndFor
+  \State \Return $S_0$
+  \EndFunction
+\end{algorithmic}
+\caption{Full state reconstruction algorithm}
+\label{algo:last}
+\end{algorithm}
 
@@
 \label{sec:implem}
+
+We have implemented the state reconstruction algorithms described above using a
+mixture of \textsf{C} (for the computationally expensive parts) and
+\textsf{Python} (for the rest). We used the \texttt{fplll} library~\cite{fplll}
+to solve CVP instances exactly in dimension 40.
 

### 2020-05-30 article/main.tex
écriture section 6

@@
 \usepackage{tikz}
+\usepackage{minted}
 \usepackage{cellspace}
@@
   \State $v'_i = v_i + 1 \bmod 2^{\ell}$
-  \State $\mathcal{C} \gets \{ w_i \oplus (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$ \Comment{Check eq.~\eqref{eq:consistency}}
-  \If{$\{v_i, v'_i\} \cap \mathcal{C} = \emptyset$}
+  \State $\mathcal{C}_i \gets \{ w_i \oplus (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$ \Comment{Check eq.~\eqref{eq:consistency}}
+  \If{$\{v_i, v'_i\} \cap \mathcal{C}_i = \emptyset$}
   \State \Return \textbf{False} \Comment{Bad Guesses}
@@
 \caption{Rotations and full difference reconstruction algorithm}
+\label{algo:unknown_2}
 \end{algorithm}
@@
 \textsf{Python} (for the rest). We used the \texttt{fplll} library~\cite{fplll}
-to solve CVP instances exactly in dimension 40.
+to solve CVP instances exactly in dimension 42.
+
+In this section, we briefly outline important aspects of our implementations and
+present practical results. Our codes are available online at:
+\[
+  % \url{https://github.com/cbouilla/pcg/}
+  \verb|[----REDACTED FOR PEER REVIEW----]|
+\]
+(and in the supplementary material).
+
+The designer of \textsf{PCG} was kind enough to send us challenge inputs. She
+generated random seeds and provided us with the first outputs of the
+pseudo-random generator. We were able to reconstruct the seed with an extremely
+high confidence level, because they re-generate the same outputs. We emailed
+back the seeds and received confirmation that they were indeed correct.
+
+\subsection{Known Increment}
+
+When the increment $c$ is known, algorithm~\ref{algo:known} is all it takes to
+reconstruct the internal state of the generator and predict it (or output the
+seed). We implemented it in \textsf{C}, using \textsf{OpenMP} to parallelize the
+outer loop that guess the least-significant bits of the state. This yields a
+simple multi-core implementation. We used the \texttt{gcc 8.3.0} compiler.
+
+We ran it on a server equipped with two 16-core \textsf{Intel Xeon Gold 6130 CPU
+  @ 2.10GHz} ``\emph{Skylake}'' CPUs. The inner loop does $2^{37}$ iterations
+and terminates in 42.3s, which makes 23 core minutes.
+
+These processors operate at a different frequency depending on the number of
+cores used and the type of instructions executed. Our code uses only scalar
+instructions, so the CPUs runs at the highest frequency tier when executing
+it. Using a single software thread per physical core (each core presents two
+hardware execution contexts, commercially called \emph{HyperThreads}) allows the
+CPU to run at 2.8Ghz, the maximum ``Turbo'' frequency on all cores. Using one
+software thread per hardware thread reduces the frequency to $\approx$ 2.6Ghz,
+but allows to better saturate the execution units of the CPU and yields a nearly
+20\% speedup overall.
+
+Therefore the algorithm requires $2^{41.67}$ CPU cycles in total; this makes
+less than 5 cycles per iteration of the inner loop. We used several
+implementation tricks to reach this level of efficiency:
+
+\begin{itemize}
+\item We used the \verb|__uint128_t| type provided by most \textsf{C} compilers
+  to do 128-bit arithmetic when computing $S_1$ from $S_0$. Apart from that, the
+  algorithm has been designed to do mostly 64-bit arithmetic, for the sake of
+  efficiency.
+
+\item Looking at the algorithm, it is clear that $U_1$ and $U_2$ are actually
+  not needed, so we just don't compute them.
+
+\item $\widetilde{T}_J$ is a function of $w, j$ and $r_J$ (with
+  $j=0,1,2$). therefore, for each new value of $w$, we precompute once and for
+  all an array indexed by $(J, R_J)$ of the 192 possible values of
+  $\widetilde{T}_J$.
+
+\item Pushing the same idea a bit further, we precompute parts of the
+  matrix-vector product inside the rounding: this computes a linear combination
+  of the rows of $G_3^{-1}$, in which $\widetilde{T}_j$ is the coefficient of
+  the $j$-th row. So we precompute the 576 possible products $\widetilde{T}_j \cdot G_3^{-1}[j, k]$.
+
+\item We enumerate the possible rotations in lexicographic order. This means
+  that $\widetilde{T}_0$ changes in each iteration while $\widetilde{T}_1$ (resp
+  $\widetilde{T}_2$) changes every 64 (resp 4096) iterations. Therefore, in 98\%
+  of the iterations, two-thirds of the matrix-vector product inside the rounding
+  are the same as from the previous iteration. Therefore, we fully compute the
+  matrix-vector product only when $r_1$ changes and only update it when $r_0$
+  changes.
+
+\item The rounding operation, when done naively by writing
+  $\texttt{llround(x)}$, is actually a bottleneck: it calls a library function
+  that accounted for about 20\% of the total running time. We instead used the
+  following technique, which correctly returns $\lfloor x \rceil$ whenever
+  $|x| < 2^{51}$:
+  \begin{minted}{C}
+long long fast_round(double x)
+{
+    union { double d; long long l; } magic; 
+    magic.d = x + 6755399441055744.0;
+    magic.l <<= 13; 
+    magic.l >>= 13;
+    return magic.l;
+}
+\end{minted}
+This hack exploits the IEEE754 representation of double-precision floats: the
+mantissa lies in bits $[0:52]$ while the sign bit and the exponents take the 12
+most significant bits. Adding $2^{52}+2^{51}$ forces the mantissa to shift to
+the correct position and inserts an extra 1 bit at position 51. The two shifts
+clear the extra bit and the exponent, while correctly expanding the sign bit.
+\end{itemize}
+
+\subsection{Unknown Increment}
+
+When the increment $c$ is known, the internal state of \textsf{PCG64} can be
+practically reconstructed from $X_0, \dots, X_{42}$ using the algorithms shows
+in section~\ref{sec:Cunknown}. Only algorithm~\ref{algo:unknown_1} is
+computationally expensive; we implemented it in \textsf{C}, while we implemented
+algorithms~\ref{algo:unknown_2} and~\ref{algo:last} in \textsf{Python}.
+
+We have shown that algorithm~\ref{algo:unknown_1} is correct when $\ell=13$. In
+fact, we empirically observed that it returns the right result with probability
+$\approx 66\%$ when $\ell=11$ and with probability $99\%$ when
+$\ell=12$. Decrementing $\ell$ reduces the workload by a factor of 4. Therefore,
+if $T$ denotes the running time when $\ell=13$, we actually used the following
+strategy:
+\begin{enumerate}
+\item Run with $\ell=11$ (if it succeeds, stop).
+\item Run with $\ell=11$ with different inputs.
+\item Run with $\ell=11$ with different inputs.
+\item Run with $\ell=12$.
+\item Run with $\ell=13$.
+\end{enumerate}
+It is not difficult to check that the expected running time of this procedure is
+$\frac{1079}{10800} T \approx 0.1 T$. In our implementation, $T = 200,000$ CPU
+hours, so the expected running time of the reconstruction procedure is about
+$20,000$ CPU hours. In fact we were lucky: on the challenge input, the first
+attempt with $\ell=11$ succeeded and took only $12,500$ CPU hours.
+
+The whole process actually ran in 35 wall-clock minutes using 512 cluster nodes,
+each equipped with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz}
+(``\emph{Cascade Lake}''). The actual machine is the
+\verb|[----REDACTED FOR PEER REVIEW----]| located at the
+\verb|[----ALSO REDACTED----]| computation center. Note that on this particular
+parallel computer, running the algorithm with $\ell=13$ would take 10 hours
+using the same amount of resources, so the whole procedure \emph{is} practical,
+even in the absolute worst case.
+
+The outer loop of algorithm~\ref{algo:unknown_1} makes $2^{2\ell - 1}$
+iterations while the inner loop makes $2^{30}$ iterations. Using a single
+hardware execution context, we measured that one of the outer loop takes between
+41.5s and 44s (apparently not all nodes of the cluster are running at exactly
+the same speed, potentially because of ``turbo boost'' adjustments and thermal
+constraints; most of the time, they take exactly... 42 seconds). Because of this
+variability, we implemented a master-slave work distribution pattern, in which a
+master process dispatches iterations of the outer loop to slave processes. This
+also made checkpointing very easy. We used \textsf{MPI} for inter-process
+communication.
+
+With $\ell=11$, the whole process took $2^{56.74}$ CPU cycles, which makes less
+than 6 cycles per iteration of the inner loop. We used essentially the same
+implementation tricks discussed above. However, this time we had to additionally
+implement the \textsc{ConsistencyCheck} procedure, which is called in the inner
+loop. We observed that the set of possible candidate values $\mathcal{C}$ only
+depends on $w_0$ (the variable of the outer loop). Therefore, before entering
+the inner loop, we precompute a bit field of size $2^\ell$ describing
+$\mathcal{C}_i$. To simplify the implementation, we flatten them by computing
+$\mathcal{C} = \cup_i \mathcal{C}_i$. This slightly increase the probability of
+false positives.
 

### 2020-05-30 article/main.tex
fixes

@@
 \]
-Where $a = 47026247687942121848144207491837523525$ is a fixed 126-bit
-constant. The first initial state $S_0$ is the seed of the generator. The
-increment $c$ may be specified by the user of the PRNG to produce different
-output streams with the same seed (just as the IV acts in a stream cipher). If
-no value of $c$ is specified, then a default increment is provided:
-$c = 117397592171526113268558934119004209487$. Note that $c$ must be odd.
+
+Where is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
+the generator. The increment $c$ may be specified by the user of the PRNG to
+produce different output streams with the same seed (just as the IV acts in a
+stream cipher). If no value of $c$ is specified, then a default increment is
+provided. Note that $c$ must be odd. The default values are:
+\begin{align*}
+  a &= \phantom{1}47026247687942121848144207491837523525 & \text{(fixed)}\\
+  c &= 117397592171526113268558934119004209487           & \text{(default value, user-definable)}
+\end{align*}
 
@@
 \begin{enumerate}
-\item Guess the 64 upper bits of $S_0$.
+\item Guess the 64 upper bits of $S_0$ (this includes the rotation).
 \item Compute the missing 64 lower bits using~\eqref{eq:output}, with:
@@
 \]
-\item If $X_1$ is computed correctly, then output $S_0$.
+\item Compute $S_1$ then extract $X_1$; if $X_1$ is correct, then output $S_0$.
 \end{enumerate}
@@
   \end{equation}
-  Therefore, guessing low-order bits of $S_0$ yields a ``long-term advantage''
-  that holds for all subsequent states.
+  Therefore, guessing the least-significant bits of $S_0$ yields a ``long-term
+  advantage'' that holds for all subsequent states.
 
@@
 
-Because the borrow are unknown, we in fact compute
+Because the borrows are unknown, we in fact compute
 $\widetilde{\mathbf{T}'} = \mathbf{S}[58:64+\ell] \boxminus
@@
 
-    \State $\widetilde T'_j \gets T_j \boxminus K_j[58:64+\ell]$  \Comment{Truncated geometric series}
+    \State $\widetilde T'_j \gets T_j \boxminus K_j[58:64+\ell]$  \Comment{Truncated geometric series on $6+\ell$ bits}
     \State $(U_0, U_1, U_2) \gets \left\lfloor 2^{58-\ell} \cdot (\widetilde T'_0, \widetilde T'_1, \widetilde T'_2) \cdot  G_3^{-1} \right\rceil \cdot G_3$ \Comment{CVP (Babai rounding)}
@@
 
-% The global strategy to obtain a practical prediction algorithm is essentially
-% the same as before (target a truncated version of the underlying linear
-% congruential generator), but is a bit more complicated to implement.
-
-Set $\Delta S_i = S_{i+1} - S_i$; it is easily checked that $\Delta S_i$ is a
+Set $\Delta S_i = S_{i+1} \boxminus S_i$; it is easily checked that $\Delta S_i$ is a
 geometric progression of common ratio $a$. Therefore, reconstructing both $S_0$
@@
 Let us set:
-\begin{align*}
-  \nabla S_i \stackrel{def}{=} S_i \boxminus S_0 = \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
-\end{align*}
-Note that $\nabla S_1 = \Delta S_0$. The prediction algorithm we propose proceeds in three phases:
+\begin{equation}\label{eq:nabla}
+  \nabla S_i \stackrel{def}{=} S_i - S_0 = \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
+\end{equation}
+Note that $\nabla S_0 = 0$ and $\nabla S_1 = \Delta S_0$. The prediction algorithm we propose proceeds in three phases:
 \begin{enumerate}
@@
 
-\noindent Only the first phase is computationally intensive.
+\noindent Only the first phase is computationally intensive. The four steps are discussed in the next four subsections.
 
@@
 guesses was easy. When $c$ is unknown, the same strategy does not work, but a
-very strong consistency check can still be implemented. The essential idea is
-the following: once $\Delta S_0[0:64+\ell]$ is known, then all the rotations can
-be determined accurately and with little computational effort. This in turn
-allows a consistency check between the partial information deduced from the
-guesses and the actual outputs.
+very strong consistency check can still be implemented. % The essential idea is
+% the following: once $\Delta S_0[0:64+\ell]$ is known, then all the rotations can
+% be determined accurately and with little computational effort. This in turn
+% allows a consistency check between the partial information deduced from the
+% guesses and the actual outputs.
 
 Once we have found $\Delta S_0[0:64+\ell]$, we can compute
-$\nabla S_i[0:64+\ell]$ for any $i$ (geometric progression); because we have
-guessed the first rotation and the $\ell$ least significant bits of the state,
-using~\eqref{eq:output} we gain access to $S_0[58:64+\ell]$; combined with the
-``differences'' $\nabla S_i$, this reveals $S_i[58:64+\ell]$ for any $i$ (and we
-already had $S_i[0:\ell]$). This allows us to compute
-$S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$.
-
-Given a ``fresh'' output $X_i$, and assuming that the guesses are correct, then we should have:
+$\nabla S_i[0:64+\ell]$ for any $i$ because eq.~\eqref{eq:nabla} holds modulo
+$2^{64+\ell}$; because we have guessed the first rotation and the $\ell$ least
+significant bits of the state, using~\eqref{eq:output} we gain access to
+$S_0[58:64+\ell]$; combined with the ``differences'' $\nabla S_i$, this reveals
+$S_i[58:64+\ell]$ for any $i$ (and we already had $S_i[0:\ell]$). This allows us
+to compute $S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$. Given a ``fresh''
+output $X_i$, and assuming that the guesses are correct, then we should have:
 \begin{equation}\label{eq:find_rotation}
@@
 \end{equation}
-If $X_i$ is known, then this gives a way to find the $i$-th rotation $r_i$ by
-trying the 64 possible values. As soon as $\ell \geq 6$, this should determine
-$r_i$ almost uniquely. In any case, if the guesses were correct, then we should have:
+% If $X_i$ is known, then this gives a way to find the $i$-th rotation $r_i$ by
+% trying the 64 possible values. As soon as $\ell \geq 6$, this should determine
+% $r_i$ almost uniquely. 
+In particular, if the guesses were correct, then we should have for any $i$:
 \begin{equation}\label{eq:consistency}
@@
 If none of the 64 possible rotations yields a match, then the guesses made
-beforehand have to be wrong. As a consequence, bad guesses can be filtered
-deterministically.
-
-More precisely, let us assume that we have guessed the $\ell$ least-significant
-bits of $S_0$ (we denote them by $w_0$) and the first rotation $r_0$. Set
-$Y_0 \stackrel{def}{=} X_0 \lll r_0$; there is an unknown value
-$x_0$ such that:
-\[
-  S_0 = w_0 + 2^\ell x + 2^{64} (w_0 \oplus Y_0[0:\ell]) + 2^{64+\ell} (x \oplus Y_0[\ell:64]) \qquad \left(0 \leq x < 2^{64-\ell}\right).
-\]
+beforehand have to be wrong. As a consequence, bad guesses can be filtered with
+an arbitrarily low probability of false positives, by trying several indices
+$i$.
+
+A few details still need to be fleshed out. To be precise, let us assume that we
+have guessed the $\ell$ least-significant bits of $S_0$ (we denote them by
+$w_0$) and the first rotation $r_0$. Set $Y_0 = X_0 \lll r_0$. % there is an
+% unknown value $x_0$ such that:
+% \[
+%   S_0 = w_0 + 2^\ell x + 2^{64} (w_0 \oplus Y_0[0:\ell]) + 2^{64+\ell} (x \oplus Y_0[\ell:64]) \qquad \left(0 \leq x < 2^{64-\ell}\right).
+% \]
 The unknown value $x_0$ will not be recovered until the third phase of the whole

### 2020-05-30 article/main.tex
intro ferrenberg

@@
 
-\todo{Mentionner Ferrenberg 92}
+Failures in cryptographic pseudo-random generators have catastrophic security
+implications. Let us mention for instance the well-known problem in Debian
+Linux, where a bug in the OpenSSL package led to insufficient entropy gathering
+and to practical attacks on the SSH and SSL protocols (the only remaining source
+of entropy comes from the PID of the process, i.e. 16 bits or less of effective
+entropy)~\cite{YilekRSES09}.
+
+However, problems in non-cryptographic random number generators can also have
+problematic consequences (barring the obvious case where they are used in lieu
+of their cryptographic counterparts). When they are used in scientific computing
+for Monte-Carlo simulations, their defects have the potential to alter the
+results of the simulations. Ferrenberg et al.~\cite{Ferrenberg92} ran a
+classical Ferromagnetism Ising model Monte-Carlo simulation, in a special case
+where the exact results could be computed analytically, and compared the results
+of the simulation with the ``true'' answer. They used several conventional
+pseudo-random generators: a 32-bit linear congruential generator, two LFSRs,
+various combinations thereof, etc. They observed that changing the source of
+random numbers significantly altered the outcome of the physical
+simulation. Different generators produced different biases: in particular a
+given LFSRs yielded energy levels that were systematically too low and critical
+temperatures that were always too high, while another kind of generator yielded
+the opposite.
 
@@
 ciphers in counter mode~\cite{Salmon11}. The need for speed then leads to the
-use of weakened cryptographic primitives.
-
-In most cases, it is easy to see that a non-cryptographic PRG does not meet the
-cryptographic notion of pseudo-randomness, and there are few exceptions. In this
-paper, we study the \textsf{PCG} family of non-cryptographic pseudo-random
-generators~\cite{melissapaper,melissaweb}.
+use of weakened cryptographic primitives (round-reduced AES or custom and
+presumably weak block-ciphers)
+
+In most cases, it is fairly easy to see that a specific non-cryptographic PRG
+does not meet the cryptographic notion of pseudo-randomness, and there are few
+exceptions. Most are fairly easy to predict, meaning that after having observed
+a prefix of the output, it is easy to produce the next ``pseudo-random''
+bits.
+
+In this paper, we study the \textsf{PCG} family of non-cryptographic
+pseudo-random generators~\cite{melissapaper,melissaweb}. Its designer claimed
+that predicting it ought to be ``challenging'', even though cryptographic
+strength is not claimed.
 
@@
 to predict the pseudo-random stream deterministically and clock the generator
-backwards. The original seeds can also easily be reconstructed. The algorithm is
-practical and we have executed it in practice. It follows that predicting the
-output of the \textsf{PCG} should not be considered challenging anymore.
+backwards. The original seeds can also easily be reconstructed. The state
+reconstruction algorithm is practical and we have executed it in practice. It
+follows that predicting the output of the \textsf{PCG} should not be considered
+challenging anymore.
+
+While the \textsf{PCG} pseudo-random generator is not meant as a cryptographic
+primitive, obtaining an actual prediction algorithm requires the use of (fairly
+conventional) cryptanalytic techniques.
 
@@
 algorithm requires 3072 bits of pseudo-random stream ; it guesses between 51 and
-55 bits, then for each guess it solves and instance of CVP in dimension 4 (using
+55 bits, then for each guess it solves an instance of CVP in dimension 4 (using
 about 75 arithmetic operations). This recovers 64 more bits of information about
@@
 
-\paragraph{Related Work.} \todo{Mentionner Knuth et les LCG + Frieze + Boyar + Stern...}
+\paragraph{Related Work.} Knuth considered the problem of recovering the
+internal state of a truncated linear congruential generator
+$x_{i+1} = ax_i + c \bmod 2^k$ when $a$ and $c$ are unknown~\cite{Knuth85}; he
+gave an algorithm exponential in the number of truncated bits.
+
+\todo{Mentionner Knuth et les LCG + Frieze + Boyar + Stern...}
 
@@
 
-The designer of \textsf{PCG} was kind enough to send us challenge inputs. She
-generated random seeds and provided us with the first outputs of the
-pseudo-random generator. We were able to reconstruct the seed with an extremely
-high confidence level, because they re-generate the same outputs. We emailed
-back the seeds and received confirmation that they were indeed correct.
+The designer of \textsf{PCG} was kind enough to send us two sets challenge
+inputs: one with the default (known) increment and one with a random secret
+increment. She generated random seeds and provided us with the first outputs of
+the pseudo-random generator. We were able to reconstruct the seed with an
+extremely high confidence level, because they re-generate the same outputs. We
+emailed back the seeds and received confirmation that they were indeed correct.
+
+We have therefore successfully taken the challenge of predicting the output of
+the \textsf{PCG64} generator.
 

### 2020-05-31 article/main.tex
cas retenue

@@
 	\end{align*}
-	Hence \(\lVert U -2^{k-\ell} \widetilde{T} \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \)
 
@@
 		\item If \(N_i \in \{0,1\}\), then \(\widetilde{T}_i = T_i + N_i\) (without modulo) and \[\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert \leqslant \leqslant 2^{k-\ell +1 }- 1\]
-		\item If \(T_i = 0\) and \(N_i = -1\), then \(U_i=U_i[0,k-\ell]\) and  \[\lvert 2^{k-\ell}\widetilde{T}_i-(U_i + 2^k) \rvert =\lvert 2^{k-\ell}(2^\ell -1) -(U_i + 2^k) \rvert = \lvert 2^{k-\ell} + U_i[0,k-\ell]\rvert \leqslant 2^{k-\ell + 1} -1 \]
+		\item If \(N_i = -1\), then \(U_i=U_i[0,k-\ell]\) and  \[\lvert 2^{k-\ell}\widetilde{T}_i-(U_i + 2^k) \rvert =\lvert 2^{k-\ell}(2^\ell -1) -(U_i + 2^k) \rvert = \lvert 2^{k-\ell} + U_i[0,k-\ell]\rvert \leqslant 2^{k-\ell + 1} -1 \]
+	\end{itemize}
+
+	\item If there are indexes such that \(\widetilde{T}_i = 0\).
+	\begin{itemize}
+		\item  If  \(N_i \in \{-1,0\}\), then \(\widetilde{T}_i = T_i + N_i\) (without modulo) and \[\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert \leqslant \leqslant 2^{k-\ell +1 }- 1\]
+		\item If \(N_i = 1\) , then \(U_i= 2^k-2^(k-l) + U_i[0,k-\ell]\) and  \[\lvert 2^{k-\ell}\widetilde{T}_i-(U_i - 2^k) \rvert =\lvert 0 -(U_i - 2^k) \rvert = \lvert 2^k-2^(k-l) + U_i[0,k-\ell] - 2^k\rvert \leqslant 2^{k-\ell + 1} -1 \]
+	\end{itemize}
 \end{itemize}
@@
 	\item If \(T_i = 0\) and \(N_i = -1\), \(U'_i=U_i+2^k\).
+	\item If \(T_i = 2^\ell - 1\) and \(N_i = +1\), \(U'_i=U_i-2^k\).
 	\item Else \(U'_i=U_i\).
 \end{itemize}
-	In both case \(\lVert U' -2^{k-\ell} \widetilde{T} \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \)
-	\item If there are indexes such that \(\widetilde{T}_i = 0\).
+In all this cases \(\lVert U' -2^{k-\ell} \widetilde{T} \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \), and \(U' \equiv U \mod 2^k \).
+
+\todo{C'est un lemme mais pour l'instant y'a pas de commande lemme}
+\begin{theorem}
+	If $\widetilde{T}$ is the noisy truncation of \(U\) of \(\ell\) bits, then there exists \(U'\) such that :
 	\begin{itemize}
-		\item  If  \(N_i \in \{-1,0\}\), then \(\widetilde{T}_i = T_i + N_i\) (without modulo) and \[\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert \leqslant \leqslant 2^{k-\ell +1 }- 1\]
-		\item If \(N_i = 1\) we have a problem.
+		\item \(U'\equiv U \mod 2^k\)
+		\item \(\lVert U' -2^{k-\ell} \widetilde{T} \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \)
 	\end{itemize}
-	So, if there are indexes such that \(\widetilde{T}_i = 0\), we have to guess if \(N_i=1\) or not.
-
-\end{itemize}
+\end{theorem}
 
@@
 In this article, we will only use an exact CVP solver when \(k = 128\) and \(\ell = 6\).
-As discussed in the previous subsection, we suppose we are never in the situation\(\widetilde{T} = 2^\ell-1\) and \(N_i=1\).
-Hence \(\lVert T-U \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \).
-So, if we fix \(\ell = 6\) we can search for the minimum \(n\) such that \[2\sqrt{n}2^{122} \leqslant \lambda_1(L(128,n))\].
-
-After computation, we find that the minimal \(n\) is 42. \todo{a vérifier (snif snif)}
+So, if we fix \(\ell = 6\) we can search for the minimum \(n\) such that \[2\sqrt{n}(2^{128-6+1}-1) \leqslant \lambda_1(L(128,n))\].
+
+After computation, we find that the minimal \(n\) is 63.
 
@@
 
-\begin{table}[htbp!] \todo{a vérifier}
+\begin{table}[htbp!]
   \centering
@@
 	\hline
-	3 & 2.87 & \(4.09e^{12} \simeq 2^{41.9}\) & 19 \\
-	4 & 2.06 & \(2.44e^{14} \simeq 2^{47.8}\) & 13 \\
-	5 & 3.77 & \(1.72e^{15}\simeq 2^{50.6}\) & 11 \\
-	6 & 2.69 & \(1.03e^{16}\simeq 2^{53.2}\) & 8 \\
+	3 & 2.87 & \(4.09e^{12} \simeq 2^{41.9}\) & 26 \\
+	4 & 2.06 & \(2.44e^{14} \simeq 2^{47.8}\) & 20 \\
+	5 & 3.77 & \(1.72e^{15}\simeq 2^{50.6}\) & 18 \\
+	6 & 2.69 & \(1.03e^{16}\simeq 2^{53.2}\) & 15 \\
 	\hline

### 2020-05-31 article/main.tex
intro

--- 2020-05-30 article/main.tex
@@
 \usepackage{minted}
+\usepackage{epigraph}
 \usepackage{cellspace}
@@
 
-\todo{Ca, c'est du blabla...}
+\epigraph{Any one who considers arithmetical methods of producing random digits is, of course, in a state of sin.}{John von Neumann, 1949}
+
+%\todo{Ca, c'est du blabla...}
 
@@
 cryptographic notion of pseudo-randomness asks that it passes \emph{any}
-efficient test. There are \textit{de facto} statistical test suites (\todo{les
-  mentionner}). The goal of designers then consists in designing the fastest
-possible generator that passes the day's favourite test suite.
+efficient test. There are \textit{de facto} statistical test suites; an initial
+battery of randomness tests for RNGs was suggested by Knuth in the 1969 first
+edition of The Art of Computer Programming. In 1996, Knuth's tests were then
+supplanted by Marsaglia's \texttt{Diehard} tests. In 2007, L'Ecuyer proposed
+the \texttt{TestU01}~\cite{LEcuyerS07} library, whose ``BigCrush'' test is
+considered state-of-the-art by the relevant community, to the best of our
+knowledge. In 2010, the NIST proposed its own statistical test suite (Special
+Publication 800-22), to which improvements were later
+suggested~\cite{ZhuMLZJ16}. We understand that the \texttt{PractRand} test suite
+also has a good reputation.
+
+In any case, The goal of designers then consists in designing the simplest and
+fastest possible pseudo-random generator that passes the day's favourite test
+suite. A few selected ones are shown in Fig.~\ref{fig:prngs}. Deterministic
+pseudo-random generators can be traced back to the work of Von Neumann
+
+Lehmer proposed
+linear congruential generators in 1949 for use on the ENIAC computer~\cite{Lehmer}
+
+\texttt{Xorshift}~\cite{Marsaglia03},
+\texttt{Xorshift128+}~\cite{Vigna17}.
+
+\begin{figure}
+  \begin{minipage}[t]{5cm}
+\begin{minted}[fontsize=\small,frame=single]{C}
+u128 a;
+u64 m = 0xda942042e4dd58b5;
+
+u64_t lehmer64() 
+{
+    a *= m;
+    return a >> 64;
+}
+\end{minted}
+  \end{minipage}
+    \begin{minipage}[t]{4cm}
+% 128-bit LFSR
+\begin{minted}[fontsize=\small,frame=single]{C}
+u32 a, b, c, d;
+
+u32 xorshift128()
+{
+    u32 t = d;
+    u32 s = a;
+    d = c;
+    c = b;
+    b = s;
+    t ^= t << 11;
+    t ^= t >> 8;
+    s ^= s >> 19;
+    a = t ^ s;
+    return a;
+}
+\end{minted}
+    \end{minipage}
+    \begin{minipage}[t]{4cm}
+    % xorshift128+ (as in google's V8)
+\begin{minted}[fontsize=\small,frame=single]{C}
+u64 a, b;
+
+u64 xorshift128plus()
+{
+    u64 s1 = a
+    u64 s0 = b;
+    a = s0;
+    s1 ^= s1 << 23;        
+    s1 ^= s1 >> 17;        
+    s1 ^= s0;        
+    s1 ^= s0 >> 26;
+    b = s1;
+    return a + b;
+}
+\end{minted}
+    \end{minipage}
+    \caption{Some conventional pseudo-random generators, designed for speed and simplicity.}
+    \label{fig:prngs}
+  \end{figure}
 
@@
 However, problems in non-cryptographic random number generators can also have
-problematic consequences (barring the obvious case where they are used in lieu
-of their cryptographic counterparts). When they are used in scientific computing
-for Monte-Carlo simulations, their defects have the potential to alter the
-results of the simulations. Ferrenberg et al.~\cite{Ferrenberg92} ran a
+dire consequences (barring the obvious case where they are used in lieu of their
+cryptographic counterparts). When they are used in scientific computing for
+Monte-Carlo simulations, their defects have the potential to alter the results
+of the numerical simulation. Ferrenberg et al.~\cite{Ferrenberg92} ran a
 classical Ferromagnetism Ising model Monte-Carlo simulation, in a special case
@@
 temperatures that were always too high, while another kind of generator yielded
-the opposite.
+the opposite (in many, repeated, trials).
 
@@
 primitive, obtaining an actual prediction algorithm requires the use of (fairly
-conventional) cryptanalytic techniques.
+conventional) cryptanalytic techniques. Making it practical requires in addition
+a non-trivial implementation effort.
 
@@
 simplified prediction algorithm recovers the internal state from 192 bits of
-pseudo-random stream. The process runs in 20 CPU minutes (on a single core of a
-server processor). It guesses 36 bits of the internal state, then solves an
-instance of the \textsc{Closest Vector Problem} (CVP) in a 3-dimensional
-euclidean lattice. This requires about 50 arithmetic operations in total and
-reveals the entire internal state if the guesses are correct.
+pseudo-random stream. The process runs in 20 CPU minutes. It guesses 37 bits of
+the internal state, then solves an instance of the \textsc{Closest Vector
+  Problem} (CVP) in a 3-dimensional euclidean lattice. This requires about 50
+arithmetic operations in total and reveals the entire internal state if the
+guesses are correct.
 

### 2020-05-31 article/main.tex
re-intro

@@
 
-\todo{mentioner stream ciphers}
+It is well-known that pseudo-random generators can be turned into symmetric
+encryption algorithm, by generating ``random'' masks to be used in the one-time
+pad. This is precisely what stream ciphers do.
 
@@
 applications, it is not necessarily necessary: to be used in Monte-Carlo
-simulations or generate random choices in games, a relaxed, non-cryptographic
-notion of pseudo-randomness may be sufficient. This allows for faster
-algorithms. For instance, the \texttt{Math.Random} function in Google's V8
-open-source JavaScript engine uses the \textsf{XorShift128} generator, which is
-a Linear Feedback Shift Register tailored for high software speed. On 64-bit
-processors, it produces 64 pseudo-random bits with 3 shifts and 4 XORs
-instructions. The \textsf{python} standard library's \texttt{random} module uses
-the Mersenne Twister. The \textsf{C} library that comes along \texttt{gcc} (the
-\texttt{glibc}) uses a (poor) truncated linear congruential generator by default
-to implement the \texttt{rand} function.
+numerical simulations or generate random choices in games, a relaxed,
+non-cryptographic notion of pseudo-randomness may be sufficient. This allows for
+faster algorithms. For instance, \textsf{python} standard library's
+\texttt{random} module uses the Mersenne Twister~\cite{MatsumotoN98}. The
+\textsf{C} library that comes along \texttt{gcc} (the \texttt{glibc}) uses a
+(poor) truncated linear congruential generator by default to implement the
+\texttt{rand} function.
 
@@
 
-In any case, The goal of designers then consists in designing the simplest and
-fastest possible pseudo-random generator that passes the day's favourite test
-suite. A few selected ones are shown in Fig.~\ref{fig:prngs}. Deterministic
-pseudo-random generators can be traced back to the work of Von Neumann
-
-Lehmer proposed
-linear congruential generators in 1949 for use on the ENIAC computer~\cite{Lehmer}
-
-\texttt{Xorshift}~\cite{Marsaglia03},
-\texttt{Xorshift128+}~\cite{Vigna17}.
+In any case, designers of conventional pseudo-random generators try to obtain
+the simplest and fastest algorithm that passes the day's favourite test suite. A
+few selected ones are shown in Fig.~\ref{fig:prngs}. \texttt{Lehmer64} is a
+truncated linear congruential generator, touted as ``the fastest PRNG that
+passes \texttt{BigCrush}''~\cite{LemireLehmer}. \texttt{Xorshift128} is a clever
+implementation of a 128-bit LFSR with period $2^{128} - 1$ due to
+Marsaglia~\cite{Marsaglia03}, using only a few simple 32-bit
+operations. \texttt{Xorshift128+} is a improved version due to
+Vigna~\cite{Vigna17} that returns the sum of two consecutive outputs of a
+Xorshift LFSR; it passes the ``BigCrush'' test suite and is the default PRNG in
+many Javascript implementations, including that in Google's V8 engine (Chrome),
+Firefox and Safari.
 
@@
 implications. Let us mention for instance the well-known problem in Debian
-Linux, where a bug in the OpenSSL package led to insufficient entropy gathering
-and to practical attacks on the SSH and SSL protocols (the only remaining source
-of entropy comes from the PID of the process, i.e. 16 bits or less of effective
-entropy)~\cite{YilekRSES09}.
+Linux from 2008, where a bug in the OpenSSL package led to insufficient
+entropy gathering and to practical attacks on the SSH and SSL protocols (the
+only remaining source of entropy comes from the PID of the process,
+\textit{i.e.} 16 bits or less of effective entropy)~\cite{YilekRSES09}.
 
@@
 cryptographic counterparts). When they are used in scientific computing for
-Monte-Carlo simulations, their defects have the potential to alter the results
-of the numerical simulation. Ferrenberg et al.~\cite{Ferrenberg92} ran a
-classical Ferromagnetism Ising model Monte-Carlo simulation, in a special case
-where the exact results could be computed analytically, and compared the results
-of the simulation with the ``true'' answer. They used several conventional
+Monte-Carlo methods, their defects have the potential to alter the results of
+numerical simulations. Ferrenberg et al.~\cite{Ferrenberg92} ran a classical
+Ferromagnetism Ising model Monte-Carlo simulation, in a special case where the
+exact results could be computed analytically, and compared the results of the
+simulation with the ``true'' answer. They used several conventional
 pseudo-random generators: a 32-bit linear congruential generator, two LFSRs,
 various combinations thereof, etc. They observed that changing the source of
-random numbers significantly altered the outcome of the physical
+random numbers significantly altered the outcome of the numerical
 simulation. Different generators produced different biases: in particular a
@@
 a prefix of the output, it is easy to produce the next ``pseudo-random''
-bits.
+bits. This makes a good source of exercises for cryptography students.
 
 In this paper, we study the \textsf{PCG} family of non-cryptographic
-pseudo-random generators~\cite{melissapaper,melissaweb}. Its designer claimed
-that predicting it ought to be ``challenging'', even though cryptographic
-strength is not claimed.
+pseudo-random generators proposed by O'Neil~\cite{melissapaper,melissaweb}. She
+did not claim that the algorithm has cryptographic strength, but that predicting
+its output ought to be ``challenging''. We therefore took up the challenge.
 
 \textsf{PCG} stands for ``Permuted Congruential Generator'': it essentially
-consists in applying a non-linear filtering function on top of a (fairly weak)
-linear congruential generator (in a way reminiscent to the venerable filtered
-LFSRs). The resulting combination is fast and passes current test suites. Its
-designer claimed that distinguishing the pseudo-random stream produced would be
-``challenging''. The \textsf{PCG} family contains many members, but we focus on
-its strongest member, named either \textsf{PCG64} or \textsf{PCG-XSL-RR}. It has
-a 128-bit internal state and produces 64 bits when clocked. It is the default
+consists in applying a non-linear filtering function on top of a linear
+congruential generator (in a way reminiscent to the venerable filtered
+LFSRs). The resulting combination is fast and passes current statistical test
+suites. The \textsf{PCG} family contains many members, but we focus on the
+strongest one, named either \textsf{PCG64} or \textsf{PCG-XSL-RR}. It has a
+128-bit internal state and produces 64 bits when clocked. It is the default
 pseudo-random number generator in the popular \textsf{NumPy} scientific
@@
 While the \textsf{PCG} pseudo-random generator is not meant as a cryptographic
-primitive, obtaining an actual prediction algorithm requires the use of (fairly
-conventional) cryptanalytic techniques. Making it practical requires in addition
-a non-trivial implementation effort.
-
-Our algorithm reconstruct the internal state using the ``guess-and-determine''
-technique: some bits of the internal state are guessed ; assuming the guesses
-are correct, some other information is computed ; a consistency check discards
-bad guesses early on ; then candidate internal states are computed and fully
+primitive, obtaining an actual prediction algorithm requires the use of
+cryptanalytic techniques. Making it practical requires in addition a non-trivial
+implementation effort.
+
+Our algorithm reconstruct the internal state using a ``guess-and-determine''
+approach: some bits of the internal state are guessed ; assuming the guesses are
+correct, some other information is computed ; a consistency check discards bad
+guesses early on ; then candidate internal states are computed and fully
 tested. The problem actually come in two distinct flavors.
@@
 
-\paragraph{Related Work.} Knuth considered the problem of recovering the
-internal state of a truncated linear congruential generator
-$x_{i+1} = ax_i + c \bmod 2^k$ when $a$ and $c$ are unknown~\cite{Knuth85}; he
-gave an algorithm exponential in the number of truncated bits.
+\paragraph{Related Work.} Deterministic pseudo-random generators can be traced
+back to the work of Von Neumann and Metropolis on the ENIAC
+computer~\cite{vonNeumann1951}; they suggested around 1946 to use the
+``middle-square'' method: if $u_n$ is a $k$-digit number, form $u_{n+1}$ by
+taking the square of the $\frac{k}{2}$ middle digits of $u_n$. This is a
+venerable precursor of the Blum-Blum-Shub ``provably secure'' PRNG. The main
+problem of this method is that it produces sequences that quickly enters short
+cycles.
+
+Lehmer later proposed linear congruential generators in 1949, also for use on
+the ENIAC computer~\cite{Lehmer}. He gave the sequence defined by
+$u_0 = 47594118, u_{n+1} = 23 u_n \bmod 10^8 + 1$ and proved that it had period
+5882353, a clear improvement compared to the middle-square approach. More
+details on early pseudo-random generators can be found in~\cite{Knuth}.
+
+Knuth considered the problem of recovering the internal state of a truncated
+linear congruential generator $x_{i+1} = ax_i + c \bmod 2^k$ when $a$ and $c$
+are unknown~\cite{Knuth85}; he gave an algorithm exponential in the number of
+truncated bits.
 
 \todo{Mentionner Knuth et les LCG + Frieze + Boyar + Stern...}
-
-\todo{mentionner Vigna ?}
-
-% The combination between being highly commended (it achieved 5th place out of 29
-% in a recent performance test survey\cite{survey}) and a well presented
-% website\cite{melissaweb} giving an feeling of professionalism can lead to the
-% impression that PCG is a sufficiently secure family of generators.
-
-% Some attacks against PCG have been carried out by Sebastiano
-% Vigna\cite{vignaweb}, but this didn't yield any significant results, given that
-% he only tried to break an incomplete version in a fairly straightforward way. He
-% worked on a very specific version of PCG. He considers the result without
-% modulo, which is a big simplification. Moreover, he didn't study the 128 bits
-% long generator, but only the 64 bits long\cite{vignacode}.
 
@@
 strategy:
-\begin{enumerate}
-\item Run with $\ell=11$ (if it succeeds, stop).
-\item Run with $\ell=11$ with different inputs.
-\item Run with $\ell=11$ with different inputs.
-\item Run with $\ell=12$.
-\item Run with $\ell=13$.
-\end{enumerate}
+\begin{algorithmic}[1]
+\State Run with $\ell=11$ (if it succeeds, stop).
+\State Run with $\ell=11$ with different inputs.
+\State Run with $\ell=11$ with different inputs.
+\State Run with $\ell=12$.
+\State Run with $\ell=13$.
+\end{algorithmic}
 It is not difficult to check that the expected running time of this procedure is
@@
 (``\emph{Cascade Lake}''). The actual machine is the
-\verb|[----REDACTED FOR PEER REVIEW----]| located at the
+\verb|[----REDACTED FOR PEER REVIEW----]| computer located at the
 \verb|[----ALSO REDACTED----]| computation center. Note that on this particular
@@
 $\mathcal{C} = \cup_i \mathcal{C}_i$. This slightly increase the probability of
-false positives.
+false positives, but makes our code slightly simpler.
 

### 2020-05-31 article/main.tex
blabla section 3

--- 2020-05-31 article/main.tex
@@
 \usepackage{minted}
-\usepackage{epigraph}
 \usepackage{cellspace}
@@
 
-\epigraph{Any one who considers arithmetical methods of producing random digits is, of course, in a state of sin.}{John von Neumann, 1949}
-
-%\todo{Ca, c'est du blabla...}
+\todo{Ca, c'est du blabla...}
 
@@
 cryptographic notion of pseudo-randomness asks that it passes \emph{any}
-efficient test. There are \textit{de facto} statistical test suites; an initial
-battery of randomness tests for RNGs was suggested by Knuth in the 1969 first
-edition of The Art of Computer Programming. In 1996, Knuth's tests were then
-supplanted by Marsaglia's \texttt{Diehard} tests. In 2007, L'Ecuyer proposed
-the \texttt{TestU01}~\cite{LEcuyerS07} library, whose ``BigCrush'' test is
-considered state-of-the-art by the relevant community, to the best of our
-knowledge. In 2010, the NIST proposed its own statistical test suite (Special
-Publication 800-22), to which improvements were later
-suggested~\cite{ZhuMLZJ16}. We understand that the \texttt{PractRand} test suite
-also has a good reputation.
-
-In any case, The goal of designers then consists in designing the simplest and
-fastest possible pseudo-random generator that passes the day's favourite test
-suite. A few selected ones are shown in Fig.~\ref{fig:prngs}. Deterministic
-pseudo-random generators can be traced back to the work of Von Neumann
-
-Lehmer proposed
-linear congruential generators in 1949 for use on the ENIAC computer~\cite{Lehmer}
-
-\texttt{Xorshift}~\cite{Marsaglia03},
-\texttt{Xorshift128+}~\cite{Vigna17}.
-
-\begin{figure}
-  \begin{minipage}[t]{5cm}
-\begin{minted}[fontsize=\small,frame=single]{C}
-u128 a;
-u64 m = 0xda942042e4dd58b5;
-
-u64_t lehmer64() 
-{
-    a *= m;
-    return a >> 64;
-}
-\end{minted}
-  \end{minipage}
-    \begin{minipage}[t]{4cm}
-% 128-bit LFSR
-\begin{minted}[fontsize=\small,frame=single]{C}
-u32 a, b, c, d;
-
-u32 xorshift128()
-{
-    u32 t = d;
-    u32 s = a;
-    d = c;
-    c = b;
-    b = s;
-    t ^= t << 11;
-    t ^= t >> 8;
-    s ^= s >> 19;
-    a = t ^ s;
-    return a;
-}
-\end{minted}
-    \end{minipage}
-    \begin{minipage}[t]{4cm}
-    % xorshift128+ (as in google's V8)
-\begin{minted}[fontsize=\small,frame=single]{C}
-u64 a, b;
-
-u64 xorshift128plus()
-{
-    u64 s1 = a
-    u64 s0 = b;
-    a = s0;
-    s1 ^= s1 << 23;        
-    s1 ^= s1 >> 17;        
-    s1 ^= s0;        
-    s1 ^= s0 >> 26;
-    b = s1;
-    return a + b;
-}
-\end{minted}
-    \end{minipage}
-    \caption{Some conventional pseudo-random generators, designed for speed and simplicity.}
-    \label{fig:prngs}
-  \end{figure}
+efficient test. There are \textit{de facto} statistical test suites (\todo{les
+  mentionner}). The goal of designers then consists in designing the fastest
+possible generator that passes the day's favourite test suite.
 
@@
 However, problems in non-cryptographic random number generators can also have
-dire consequences (barring the obvious case where they are used in lieu of their
-cryptographic counterparts). When they are used in scientific computing for
-Monte-Carlo simulations, their defects have the potential to alter the results
-of the numerical simulation. Ferrenberg et al.~\cite{Ferrenberg92} ran a
+problematic consequences (barring the obvious case where they are used in lieu
+of their cryptographic counterparts). When they are used in scientific computing
+for Monte-Carlo simulations, their defects have the potential to alter the
+results of the simulations. Ferrenberg et al.~\cite{Ferrenberg92} ran a
 classical Ferromagnetism Ising model Monte-Carlo simulation, in a special case
@@
 temperatures that were always too high, while another kind of generator yielded
-the opposite (in many, repeated, trials).
+the opposite.
 
@@
 primitive, obtaining an actual prediction algorithm requires the use of (fairly
-conventional) cryptanalytic techniques. Making it practical requires in addition
-a non-trivial implementation effort.
+conventional) cryptanalytic techniques.
 
@@
 simplified prediction algorithm recovers the internal state from 192 bits of
-pseudo-random stream. The process runs in 20 CPU minutes. It guesses 37 bits of
-the internal state, then solves an instance of the \textsc{Closest Vector
-  Problem} (CVP) in a 3-dimensional euclidean lattice. This requires about 50
-arithmetic operations in total and reveals the entire internal state if the
-guesses are correct.
+pseudo-random stream. The process runs in 20 CPU minutes (on a single core of a
+server processor). It guesses 36 bits of the internal state, then solves an
+instance of the \textsc{Closest Vector Problem} (CVP) in a 3-dimensional
+euclidean lattice. This requires about 50 arithmetic operations in total and
+reveals the entire internal state if the guesses are correct.
 
@@
 
-The LCG have been largely studied.
-\todo{expand this} This is why we know techniques to recover internal states knowing only upper or lower bits of the output, see \cite{Boyar1989}.
-A particular LCG is the Lehmer's generator \cite{Lehmer}, where the multiplier \(a\) is known and the increment is zero.
-
-In our algorithm to predict the PCG, we need to reconstruct several truncated Lehmer's generator.
-For that, we will present a classical lattice-based reconstruction for a truncated Lehmer's generator, as presented in \cite{Frieze}.
-\todo{C'est pas exactement celle là, on dit quoi du coup ?}
+The LCG have been largely studied, and considered as cryptographically insecure.
+
+A reconstruction algorithm has been proposed by D.
+Knuth in 1985\cite{Knuth85}.
+The LCG he considered revealed only the upper bits of each internal state.
+The modulus \(2^k\) was known while the multiplier \(a\) and the increment \(c\) weren't.
+This algorithm could reconstruct the LCG in \(\mathcal{O}(2^l)\) where \(l\) is the number of non revealed bit of each internal state.
+
+In 1989, J.
+Boyar improve the result in a first way \cite{Boyar1989}.
+He presented an algorithm which could predict a LCG where all the parameters (multiplier, increment and initial state) where unknown under the condition that the number of bit unrevealed was really small in comparison to the modulus.
+
+In 1988, A.
+Frieze \textit{et al} \cite{Frieze} improve the result in a second way.
+They supposed that the multiplier \(a\) and the modulus \(2^k\) were known and used lattice-based technique to recover the LCG more efficiently.
+
+Here we will use techniques presented in \cite{Frieze} to reconstruct some Lehmer's generator (LCG where the increment \(c\) is zero) \cite{Lehmer}.
 

### 2020-06-01 article/main.tex
ack
sections 3, 4

@@
 \usepackage{multirow}
-%\usepackage{parskip}
 \usetikzlibrary{patterns}
+\usepackage[noend]{algpseudocode}
 
@@
 \DeclareMathOperator{\MAJ}{MAJ}
-
-\usepackage[noend]{algpseudocode}
+\DeclareMathOperator{\CVP}{CVP}
+
+% horrible hack
+% Math symbol font matha
+\DeclareFontFamily{U}{matha}{\hyphenchar\font45}
+\DeclareFontShape{U}{matha}{m}{n}{
+      <5> <6> <7> <8> <9> <10> gen * matha
+      <10.95> matha10 <12> <14.4> <17.28> <20.74> <24.88> matha12
+      }{}
+\DeclareSymbolFont{matha}{U}{matha}{m}{n}
+\DeclareFontSubstitution{U}{matha}{m}{n}
+
+% Math symbol font mathb
+\DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
+\DeclareFontShape{U}{mathx}{m}{n}{
+      <5> <6> <7> <8> <9> <10>
+      <10.95> <12> <14.4> <17.28> <20.74> <24.88>
+      mathx10
+      }{}
+\DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
+\DeclareFontSubstitution{U}{mathx}{m}{n}
+
+% Symbol definition
+\DeclareMathDelimiter{\vvvert}{0}{matha}{"7E}{mathx}{"17}
 
@@
 
-\todo{Mentionner Knuth et les LCG + Frieze + Boyar + Stern...}
+The LCG have been largely studied, and considered as cryptographically insecure.
+
+A reconstruction algorithm has been proposed by D.
+Knuth in 1985\cite{Knuth85}.
+The LCG he considered revealed only the upper bits of each internal state.
+The modulus $2^k$ was known while the multiplier $a$ and the increment $c$ weren't.
+This algorithm could reconstruct the LCG in $\mathcal{O}(2^l)$ where $l$ is the number of non revealed bit of each internal state.
+
+In 1989, J.
+Boyar improve the result in a first way \cite{Boyar1989}.
+She presented an algorithm which could predict a LCG where all the parameters (multiplier, increment and initial state) where unknown under the condition that the number of bit unrevealed was really small in comparison to the modulus.
+
+In 1988, A.
+Frieze \textit{et al} \cite{Frieze} improve the result in a second way.
+They supposed that the multiplier $a$ and the modulus $2^k$ were known and used lattice-based technique to recover the LCG more efficiently.
+
+Here we will use techniques presented in \cite{Frieze} to reconstruct some Lehmer's generator (LCG where the increment $c$ is zero) \cite{Lehmer}.
 
@@
 
-This section describes the \textsf{PCG64} non-cryptographic pseudo-random number
-generator (more precisely \textsf{PCG-XSL-RR} in the designer's terminology). It
-has an internal state of 128-bit, which operate as a linear congruential
-generator modulo~$2^{128}$. More precisely:
+This section introduces some notations and describes the \textsf{PCG64}
+non-cryptographic pseudo-random number generator (a.k.a. \textsf{PCG-XSL-RR} in
+the designer's terminology).
+
+If $x \in \{0, 1\}^n$ is an $n$-bit string, then $x[i:j]$ denotes the bit string
+$x_ix_{i+1}\dots x_{j-2}x_{j-1}$, where $x = x_0\dots x_{n-1}$ (this is the
+``slice notation'' used in \textsf{Python}). The set $\mathbb{Z}_{2^k}$ of
+integers modulo $2^k$ is seen as the set of $k$-bit strings. If $x$ is a
+floating point number, then $\lfloor x \rceil$ denotes the nearest integer
+(using the ``rounding half to even'' tie-breaking rule --- this is the default
+in IEEE754 arithmetic). If $U$ is a vector or a sequence, then $U_i$ is the
+$i$-th element (we use capital letters for these). If $U$ is such a sequence, we
+denote by $U \bmod M$ the sequence $(U_0 \bmod M, U_1 \bmod M, \dots)$. The XOR
+operation is denoted $\oplus$, left and right rotations are denote $\lll$ and
+$\ggg$ respectively. Modular addition is denoted $+$ (or $\boxplus$ to make it
+even more explicit).
+
+\textsf{PCG64} has an internal state of 128-bit, which operate as a linear
+congruential generator modulo~$2^{128}$. More precisely:
 \[
@@
 \]
-
 Where is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
@@
 64-bit halves of the internal state are XORed together, and this 64-bit result
-is rotated right by $r$ positions. The successive 64-bit outputs of the
-generator are $X_0, X_1, \dots$ where:
+is rotated right by $r$ positions.
+
+The successive 64-bit outputs of the generator are $X_0, X_1, \dots$ where:
 \begin{equation}\label{eq:output}
-  X_i =(S_i[0:64] \oplus S_i[64:128]) \ggg S_i[122:128].
+  X_i =(\underbrace{S_i[0:64] \oplus S_i[64:128]}_{Y_i}) \ggg \underbrace{S_i[122:128]}_{r_i}.
 \end{equation}
-
-Figure~\ref{pcg128out} summaries the process. The overall design strategy is
+For the sake of convenience, we denote by $Y_i$ the XOR of the two halves of the
+state (before the rotation) and by $r_i$ the number of shifts of the ``$i$-th
+rotation''.
+
+Fig.~\ref{pcg128out} summaries the process. The overall design strategy is
 similar to that of a filtered LFSR: the successive states of a weak internal
@@
 Updating the internal state requires a $128 \times 128 \rightarrow 128$
-multiplication operation. In fact, this can be done with two
-$64 \times 64 \rightarrow 64$ multiplications, one
+multiplication operation. In fact, this can be done with three
 $64 \times 64 \rightarrow 128$ multiplication and two 64-bit additions. High-end
 desktop CPUs all implement these operations in hardware, so the generator is
-relatively fast on these platforms.
+quite fast on these platforms.
 
@@
 
-\section{Reconstructing Truncated Geometric progressions}
+\section{Tools}
 \label{sec:geometric}
 
-This is a recurring theme is this paper, so we now state the problem in
-general. If $a, b$ are integers modulo $2^{128}$ and $0\leq i<j<128$, then there
-is a \emph{carry} $0 \leq \gamma \leq 1$ (resp. a \emph{borrow}
+In the rest of this paper, we often perform arithmetic operations on integers
+where only some bits are known. This leads to generation of unknown carries. If
+$a, b$ are integers modulo $2^{128}$ and $0\leq i<j<128$, then there is a
+\emph{carry} $0 \leq \gamma \leq 1$ (resp. a \emph{borrow}
 $0 \leq 1 \leq \beta$) such that:
@@
 
-\subsection{Reconstructing Linear Congruential Generators}
-
-Given an integer \(k\), a fixed multiplier \(a\) and an increment \(c\), define the
-sequence \[u_0 = x, u_{i+1} = a u_i + c \bmod 2^k\] The sequence \(u\) form the
-successive internal states of a linear congruential generator (LCG).
-
-Let $\LCG_{k}(x, c)$ denote the vector $(u_0, u_1, u_2, \dots)$ of integers
-modulo $2^k$.  It is easy to check that:
+\subsection{Linear Congruential Generators and Lattices}
+
+Given an integer $k$, a fixed multiplier $a$, an increment $c$ and a ``seed''
+$x$, define the sequence:
+\[
+  U_0 = x, U_{i+1} = a U_i + c.
+\]
+When reduced modulo $2^k$, the sequence $U$ form the successive states of a
+linear congruential generator (LCG). Let $\LCG_{k}(x, c)$ denote the vector
+$(u_0, u_1, u_2, \dots)$ of integers modulo $2^k$. It is easy to check that:
 \begin{align}
@@
 \end{align}
-Hence those sets of vector have a lot of structure we can use.
-
-The LCG have been largely studied, and considered as cryptographically insecure.
-
-A reconstruction algorithm has been proposed by D.
-Knuth in 1985\cite{Knuth85}.
-The LCG he considered revealed only the upper bits of each internal state.
-The modulus \(2^k\) was known while the multiplier \(a\) and the increment \(c\) weren't.
-This algorithm could reconstruct the LCG in \(\mathcal{O}(2^l)\) where \(l\) is the number of non revealed bit of each internal state.
-
-In 1989, J.
-Boyar improve the result in a first way \cite{Boyar1989}.
-He presented an algorithm which could predict a LCG where all the parameters (multiplier, increment and initial state) where unknown under the condition that the number of bit unrevealed was really small in comparison to the modulus.
-
-In 1988, A.
-Frieze \textit{et al} \cite{Frieze} improve the result in a second way.
-They supposed that the multiplier \(a\) and the modulus \(2^k\) were known and used lattice-based technique to recover the LCG more efficiently.
-
-Here we will use techniques presented in \cite{Frieze} to reconstruct some Lehmer's generator (LCG where the increment \(c\) is zero) \cite{Lehmer}.
-
-\subsection{The lattice associated to a geometric progression}\label{lattice}
-
-We fix a multiplier \(a\), integers \(k\), \(n\) and we consider \(L(n,k)\) the set of \(\LCG_{n,k}(x, 0)\), for \(x\) in \(\mathbb{Z}_{2^k}\). Then
-
-\begin{itemize}
-	\item By \eqref{eq:lcg-additive}, the vector \(\LCG(x_1,0) + \LCG(x_2,0) = \LCG(x_1+x_2,0)\), hence it is still in \(L(n,k)\).
-	\item By \eqref{eq:lcg-scalar}, the vector \((-1)\times \LCG(x,0) = \LCG(-x,0)\) hence it is still in \(L(n,k)\).
-\end{itemize}
-
-Hence \(L(n,k)\) is a lattice. We easily notice that \(L(n,k)\) is spawned by the lines of the following \(n\times n\) matrix :
-
+
+Let $\mathcal{L}$ denote the euclidean lattice spawned by the rows of the
+following $n\times n$ matrix:
 \[
-  \begin{pmatrix}
+  G_{n,k} = \begin{pmatrix}
     1 & a   & a^2 & \dots & a^{n-1} \\
@@
 \]
-
-Once again, we will omit $n$ and $k$ when they are obvious from the context.
-
-We want to find an element \(U\) in the lattice only knowing the $\ell$ most significant bits of each of its coordinates.
-More precisely, let $T_i = U_i[k-\ell:k]$ denote the top $\ell$ bits of $U_i$.
-We assume that there is a ``noise vector'' $N$ such that $N_i \in \{-1, 0, 1\}$, and we set $\widetilde{T}_i = T_i + N_i \bmod 2^{\ell}$.
-The problem we have is the  following: recover $U$ from its noisy truncation $\widetilde{T}$.
-
-In principle, $2^{k-\ell} \widetilde{T}$ should be ``close'' to \(U\). We must distinguish three cases.
+This lattice contains all $n$-terms geometric progressions of common ratio $a$
+modulo $2^k$; therefore the first $n$ terms of the sequence $\LCG_{k}(x, 0)$
+give the coordinates of a point of this lattice.
+
+Reconstructing the state of a truncated linear congruential generator can
+generally be seen as the problem of finding a point of this lattice given only
+an approximation thereof, for instance when the least-significant bits of each
+components have been dropped. The ``lattice approach'' to truncated linear
+congruential generators is due to Frieze \textit{et al.}~\cite{Frieze}.
+
+Let $U$ be a vector of $n$ integers modulo $2^k$ such that
+$U_{i+1} = a U_i \bmod 2^k$; it is a geometric progression of common ratio $a$,
+modulo $2^k$, therefore $(U_0, \dots, U_{n-1}) \in \mathcal{L}$. Let
+$T_i = U_i[k-\ell:k]$ denote the top $\ell$ bits of $U_i$, and let $N$ denote an
+arbitrary ``noise vector'' such that $N_i \in \{-1, 0, 1\}$. Finally, set
+$\widetilde{T}_i = T_i + N_i \bmod 2^{\ell}$.  We will be facing the following
+problem (``reconstructing noisy truncated geometric series'') several times:
+\begin{description}
+\item[INPUT]  $\widetilde{T} = (\widetilde{T}_0, \dots, \widetilde{T}_{n-1}) \in \left(\mathbb{Z}_{2^\ell}\right)^n$, a (noisy) version of $U$ truncated to the top $\ell$ bits.
+\item[OUTPUT] $U_0 \bmod 2^k$, the first term of the (non-truncated) geometric sequence.
+\end{description}
+
+We will be facing a ``high-dimension'' instance in section~\ref{sec:big_cvp}
+and many ``low-dimension'' in sections~\ref{sec:Cknown}
+and~\ref{sec:Cunknown}. The rest of this section discusses algorithmic tools to
+solve these problems. We first claim that $2^{k-\ell} \widetilde{T}$ is
+``close'' to a point of the lattice $\mathcal{L}_{n,k}$.
+
+\begin{lemma}\label{thm:close}
+  There exists $U' \in \mathcal{L}$ such that $U'_i \equiv U_i \bmod 2^k$
+  and $\left\lVert U' -2^{k-\ell} \widetilde{T} \right\rVert \leq 2 \sqrt{n} 2^{k-\ell}$.
+\end{lemma}
+
+\begin{proof}
+  We first observe that $U$ belongs to the lattice $\mathcal{L}$. We start by setting $U' \gets U$, and we examine all coordinates of $U$:
   \begin{itemize}
-	\item If for each \(i\) the integer \(\widetilde{T}_i\) satisfies \(1\leqslant\widetilde{T}_i\leqslant 2^{k-l}-2\) then \(\widetilde{T}_i = T_i + N_i\) (without modulo). So
-
+  \item If $\widetilde{T}_i = T_i + N_i$ (without modulo), then we have:
     \begin{align*}
-	\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert &=\lvert 2^{k-\ell}(T_i + N_i)-U_i \rvert \\
-	 &= \lvert 2^{k-\ell}N_i - U_i[0,k-\ell] \rvert\\
-	 &\leqslant \lvert 2^{k-\ell}N_i\rvert + \lvert U_i[0,k-\ell] \rvert \\
-	 &\leqslant 2^{k-\ell} + 2^{k-\ell}-1\\
-	 &\leqslant 2^{k-\ell +1 }- 1
+      \left| 2^{k-\ell}\widetilde{T}_i-U_i \right| &= \left| 2^{k-\ell}(T_i + N_i)-U_i \right| 
+                                                     = \left| 2^{k-\ell}N_i - U_i[0,k-\ell] \right| \\
+                                                   &\leq \left| 2^{k-\ell}N_i\rvert + \lvert U_i[0,k-\ell] \right| 
+                                                   % &\leq 2^{k-\ell} + 2^{k-\ell} - 1\\
+      \leq 2^{k-\ell +1 }.
     \end{align*}
 
-	\item If there are indexes such that \(\widetilde{T}_i = 2^\ell-1\).
+  \item Otherwise, there are two possible ``wraparound'' cases:
     \begin{itemize}
-		\item If \(N_i \in \{0,1\}\), then \(\widetilde{T}_i = T_i + N_i\) (without modulo) and \[\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert \leqslant \leqslant 2^{k-\ell +1 }- 1\]
-		\item If \(N_i = -1\), then \(U_i=U_i[0,k-\ell]\) and  \[\lvert 2^{k-\ell}\widetilde{T}_i-(U_i + 2^k) \rvert =\lvert 2^{k-\ell}(2^\ell -1) -(U_i + 2^k) \rvert = \lvert 2^{k-\ell} + U_i[0,k-\ell]\rvert \leqslant 2^{k-\ell + 1} -1 \]
-	\end{itemize}
-
-	\item If there are indexes such that \(\widetilde{T}_i = 0\).
-	\begin{itemize}
-		\item  If  \(N_i \in \{-1,0\}\), then \(\widetilde{T}_i = T_i + N_i\) (without modulo) and \[\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert \leqslant \leqslant 2^{k-\ell +1 }- 1\]
-		\item If \(N_i = 1\) , then \(U_i= 2^k-2^(k-l) + U_i[0,k-\ell]\) and  \[\lvert 2^{k-\ell}\widetilde{T}_i-(U_i - 2^k) \rvert =\lvert 0 -(U_i - 2^k) \rvert = \lvert 2^k-2^(k-l) + U_i[0,k-\ell] - 2^k\rvert \leqslant 2^{k-\ell + 1} -1 \]
+    \item Either $T_i = 0$ and $N_i = -1$, which leads to $\widetilde{T}_i = 2^\ell-1$. 
+      In this case, we have $U_i = U_i[0,k-\ell]$ and we set $U'_i = U_i + 2^{k}$ (note that this amount to adding a lattice vector to $U'$, so $U'$ stays in the lattice). We have:
+      \[
+        \left| 2^{k-\ell}\widetilde{T}_i-(U_i + 2^k) \right| = %\left| 2^{k-\ell}(2^\ell -1) -(U_i + 2^k) \right| =
+        \left| 2^{k-\ell} + U_i[0,k-\ell]\right| \leq 2^{k-\ell + 1}. 
+      \]
+
+    \item Or $T_i = 2^{\ell}-1$ and $N_i = +1$, which leads to
+      $\widetilde{T}_i = 0$.  This implies that
+      $U_i= 2^k-2^{k-\ell} + U_i[0,k-\ell]$; we set $U'_i = U_i - 2^k$ (again,
+      with this modification $U'$ stays in the lattice), and we find:
+      \[
+        \left| 2^{k-\ell}\widetilde{T}_i-(U_i - 2^k) \right| %=\left| 0 -(U_i - 2^k) \right|
+        = \left| 2^k-2^{k-\ell} + U_i[0,k-\ell] - 2^k\right| \leq 2^{k-\ell + 1}.
+      \]
     \end{itemize}
   \end{itemize}
-	We need a new vector \(U'\) defined as follow.
-\begin{itemize}
-	\item If \(T_i = 0\) and \(N_i = -1\), \(U'_i=U_i+2^k\).
-	\item If \(T_i = 2^\ell - 1\) and \(N_i = +1\), \(U'_i=U_i-2^k\).
-	\item Else \(U'_i=U_i\).
-\end{itemize}
-In all this cases \(\lVert U' -2^{k-\ell} \widetilde{T} \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \), and \(U' \equiv U \mod 2^k \).
-
-\todo{C'est un lemme mais pour l'instant y'a pas de commande lemme}
-\begin{theorem}
-	If $\widetilde{T}$ is the noisy truncation of \(U\) of \(\ell\) bits, then there exists \(U'\) such that :
-	\begin{itemize}
-		\item \(U'\equiv U \mod 2^k\)
-		\item \(\lVert U' -2^{k-\ell} \widetilde{T} \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \)
-	\end{itemize}
-\end{theorem}
-
-\subsection{A first way to find U : using an exact CVP solver}
-
-Solving exactly the CVP is a hard computational problem. The algorithms used are costly, but we can afford to use them up to dimension 50.
-
-We define \(U'\) as in subsection \ref{lattice}. We want to know if \(CVP(L,2^{k-\ell}\widetilde{T})\) is indeed \(U'\).
-
-What can we say about \(|CVP(L,2^{k-\ell}\widetilde{T})-U'|\) ?
-
+  In the end, we have
+  $\left\| U' -2^{k-\ell} \widetilde{T} \right\| \leq 2 \sqrt{n} 2^{k-\ell}$,
+  and $U' \equiv U \mod 2^k $.
+\end{proof}
+
+\subsection{Reconstruction in ``High'' Dimension Using an Exact CVP Solver}
+
+Lemma~\ref{thm:close} tells us that the approximation of a geometric sequence
+obtained by dropping least-significant bits cannot be arbitrarily far from a
+lattice point which reveals $U_0 \bmod 2^k$. Therefore, we may possibly
+reconstruct truncated geometric series by finding the lattice vector closest to
+the approximation we have. This means solving instances of the well-known
+\textsf{Closest Vector Problem} (CVP), a fundamental hard problem on
+lattices. It is NP-hard, and all known algorithms are exponential in the
+dimension of the lattice, yet they can be fairly practical up to dimension
+$\approx 70$.
+
+Let $\CVP(\mathcal{L}, x)$ denote the vector of $\mathcal{L}$ closest to the
+input vector $x$. Using the same notations as above, we want to know if
+$\CVP\left(\mathcal{L}, 2^{k-\ell}\widetilde{T}\right)$ is indeed $U'$. This
+will necessarily be the case when $\|2^{k-\ell}\widetilde{T} - U'\|$ is smaller
+than the length of the shortest non-zero vector of $\mathcal{L}$ --- this quantity,
+the \emph{first minimum} of the lattice, is denoted by $\lambda_1(\mathcal{L})$.
+%What can we say about $|CVP(L,2^{k-\ell}\widetilde{T})-U'|$ ?
+By the triangular inequality, we have:
 \begin{align*}
-|CVP(L,T)-U| &= |CVP(L,2^{k-\ell}\widetilde{T})-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
-& \leqslant |CVP(L,T)-2^{k-\ell}\widetilde{T}|+|2^{k-\ell}\widetilde{T}-U'|\\
+|\CVP(\mathcal{L}, T) - U'| %&= \left| \CVP\left( \mathcal{L}, 2^{k-\ell}\widetilde{T} \right) \riht| + \left|2^{k-\ell}\widetilde{T} - U'\right| %-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
+                             & \leq \left| \CVP\left(\mathcal{L}, 2^{k-\ell}\widetilde{T} \right) - 2^{k-\ell}\widetilde{T}\right | + \left|2^{k-\ell}\widetilde{T} - U'\right|.
 \end{align*}
-
-But, as the vector \(U'\) is in the lattice and by definition of the CVP, \(|CVP(L,2^{k-\ell}\widetilde{T})-2^{k-\ell}\widetilde{T}| \leqslant |2^{k-\ell}\widetilde{T}-U'| \), hence 
-\[|CVP(L,2^{k-\ell}\widetilde{T})-U'|\leqslant 2|2^{k-\ell}\widetilde{T}-U'|\]
-
-If we can prove that the right side of the inequality is smaller than first minimum of the lattice \(\lambda_1(L)\), then we have proved that \(CVP(L,2^{k-\ell}\widetilde{T})\) is indeed the vector we were searching for.
-
-To compute \(\lambda_1(L)\), we find the shortest vector of L thanks to an exact SVP solver (SVP stands for Shortest Vector Problem) and we compute its norm.
-
-In this article, we will only use an exact CVP solver when \(k = 128\) and \(\ell = 6\).
-So, if we fix \(\ell = 6\) we can search for the minimum \(n\) such that \[2\sqrt{n}(2^{128-6+1}-1) \leqslant \lambda_1(L(128,n))\].
-
-After computation, we find that the minimal \(n\) is 63.
-
-\subsection{A second way to find U : the Babai's rounding}
-
-As using an exact CVP solver is costly, we will try a cheaper but less precise way to find the closest vector.
-
-We compute \(G(n,k)\) the LLL-reduction of the generating matrix of \(L(n,k)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
-\[\rVert G \lVert =  \sup_{x \in \mathbb{R}^n}\frac{\lVert Gx \rVert_2}{\lVert x \rVert_2}\]
-In the case of the \(\lVert \cdot \rVert_2\) norm,
-\(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular
-value of $G$ (equivalently, $\sigma(G)$ is the absolute value of the largest
-eigenvalue of $G^{t} G$).
-
-As \(2^{k-\ell}\widetilde{T}\) is not an element of the lattice, \(2^{k-\ell}\widetilde{T}G^{-1}\) is not an integer vector.
-
-We fix \(r = rounding(2^{k-\ell}\widetilde{T}G^{-1}) \), then \(rG\) is an element of the lattice. What can we say about the distance between \(rG\) and \(U'\) ?
-
+But, as the vector $U'$ belongs to the lattice, by definition of the closest
+vector and by lemma~\ref{thm:close}:
+% $|CVP(L,2^{k-\ell}\widetilde{T})-2^{k-\ell}\widetilde{T}| \leq
+% |2^{k-\ell}\widetilde{T}-U'| $, hence
+\[
+  |\CVP(\mathcal{L}, T) - U'| \leq 2 \left|2^{k-\ell} \widetilde{T} - U' \right| \leq 4 \sqrt{n} 2^{k-\ell}.
+\]
+
+If we can prove that the right side of this inequality is smaller than the first
+minimum of the lattice $\lambda_1(\mathcal{L})$, then we would have proved that
+$\CVP(\mathcal{L}, T)$ indeed reveals $U_0 \bmod 2^k$.
+
+In section~\ref{sec:big_cvp}, we will be facing the problem of reconstructing a
+geometric sequence modulo $2^{128}$ given arbitrarily many (noisy versions of
+the) most-significant $6$ bits of successive elements of the sequence. Therefore
+we have $k=128$ and $\ell=6$, and we wish to determine the required number of
+samples, \textit{i.e.} the value of $n$. This means finding the values of $n$
+such that $\sqrt{n} 2^{124} \leq \lambda_1\left(\mathcal{L}\right)$.
+
+Starting from $n=\lceil 122/6\rceil\rceil$, we computed the length of the
+shortest vector of the lattice spanned by $G_{n, 128}$ for each successive $n$
+until the condition holds. The \textsf{Shortest Vector Problem} is another
+well-known lattice NP-hard problem; we used the (almost) off-the-shelf
+\texttt{G6K} library~\cite{AlbrechtDHKPS19}, which gave results very quickly by
+sieving. \texttt{fplll}~\cite{fplll} was too slow above dimension 50, in the
+default settings.
+
+After this computation, we found that the minimal possible $n$ is 63: with
+$n=63$, the shortest vector of $\mathcal{L}$ has length greater than
+$2^{127.02}$, which is high enough. This vector can be obtained by bootstrapping
+the geometric sequence with
+\[
+  U_0 = 12144252875850345479015002205241987363
+\]
+then reducing the terms modulo $2^{128}$ in zero-centered representation
+(subtracting $2^{128}$ to $U_i$ if $U_i > 2^{127}$). It follows that when
+$n\geq 63, k=128$ and $\ell=6$, any CVP oracle will return a vector congruent to
+the original $U$ when given $\widetilde{T}$.
+
+\subsection{Reconstruction in Low Dimension Using Babai's Rounding}
+
+In sections~\ref{sec:Cknown} and~\ref{sec:cvp_small_2} we will need to
+reconstruct billions of noisy truncated geometric series modulo $2^{64}$ with
+very few terms, of which a large fraction of most-significant bits are known. In
+this setting, the CVP problem becomes much easier. This enables us to use faster
+and more \textit{ad hoc} methods, such as Babai's rounding
+algorithm~\cite{Babai86}.
+
+If $M$ is a square matrix, we denote by $\vvvert M \vvvert$ the induced matrix norm :
+\[
+  \vvvert M \vvvert =  \sup_{x \in \mathbb{R}^n} \frac{\| xM \|}{\| x \|}
+\]
+In the case of the $\| \cdot \|_2$ norm used throughout this paper, $\| M \|$ is
+the largest singular value of $G$; equivalently, it is the square root of the
+absolute value of the largest eigenvalue of $M^{t} M$.
+
+Denote again by $\mathcal{L}$ the $n$-dimensional lattice spanned by the rows of
+$G_{n,64}$, and let $H$ denote the LLL-reduction of $G_{n,64}$. The same lattice
+is also spanned by the rows of $H$. For instance, with $n=3$:
+\[
+  H = \left(\begin{array}{rrr}
+                -1241281756092 & 3827459685972 & -728312298332 \\
+                -5001120657083 & -2117155768935 & 5479732607037 \\
+                8655886039732 & 3303731088004 & 6319848582548
+              \end{array}\right)
+\]
+Set $S = 2^{k-\ell} \widetilde{T}H^{-1}$; as $2^{k-\ell}\widetilde{T}$ is not
+\textit{a priori} an element of the lattice, $S$ need not be an integer
+vector. Let then $R$ denote the rounding of $S$, \textit{i.e.}
+$R_i = \left\lfloor S_i \right\rceil$. Then $R H$ is an element of the
+lattice. Under the right conditions, it will be the vector of the lattice
+closest to $2^{k-\ell} \widetilde{T}$. Indeed:
 \begin{align*}
-\lVert U' - rG \rVert &= \lVert U - (r-2^{k-\ell}\widetilde{T}G^{-1} + 2^{k-\ell}\widetilde{T}G^{-1})G \rVert\\
-&= \lVert U' - 2^{k-\ell}\widetilde{T} + (r-2^{k-\ell}\widetilde{T}G^{-1})G \rVert\\
-&\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(r-2^{k-\ell}\widetilde{T}G^{-1})\rVert \times \lVert G\rVert\\	
+\lVert U' - R H \rVert &= \left\| U' - \left(R - 2^{k-\ell}\widetilde{T}H^{-1} + 2^{k-\ell}\widetilde{T}H^{-1}\right)H \right\| \\
+&= \left\| U' - 2^{k-\ell}\widetilde{T} - \left(R - 2^{k-\ell} \widetilde{T}H^{-1}\right )H \right\| \\
+&\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\| R - 2^{k-\ell}\widetilde{T}H^{-1} \right\| \times \vvvert H \vvvert.
 \end{align*}
-
-By definition, \(r\) is the closest integer vector to \(2^{k-\ell}\widetilde{T}G^{-1}\).
-But as \(U'\) is an element of the lattice, \(U'G^{-1}\) is an integer vector.
-Thus \(r-2^{k-\ell}\widetilde{T}G^{-1}\) is shorter than \(U'G^{-1}-2^{k-\ell}\widetilde{T}G^{-1}\).
-
-Hence :
+By definition, $R$ is the closest integer vector to
+$2^{k-\ell}\widetilde{T}H^{-1}$. But as $U'$ is an element of the lattice,
+$U'H^{-1}$ is an integer vector. Thus $R-2^{k-\ell}\widetilde{T}G^{-1}$ is
+shorter than $U'G^{-1}-2^{k-\ell}\widetilde{T}G^{-1}$. Hence :
 \begin{align*}
-\lVert U' - rG \rVert &\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(r-2^{k-\ell}\widetilde{T}G^{-1})\rVert \times \lVert G\rVert\\	
-&\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(U'G^{-1}-2^{k-\ell}\widetilde{T}G^{-1})\rVert \times \lVert G\rVert\\	
-&\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(U'-2^{k-\ell}\widetilde{T})\rVert \times \lVert G^{-1} \rVert  \times \lVert G\rVert\\
-& 	\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert )\\
+  \left\| U' - RH \right\| &\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\|R - 2^{k-\ell}\widetilde{T}H^{-1}\right\| \times \vvvert H \vvvert \\	
+                        &\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\|U'H^{-1}-2^{k-\ell}\widetilde{T}H^{-1}\right\| \times \vvvert H \vvvert \\	
+                        &\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\|U' - 2^{k-\ell}\widetilde{T}\right\| \times \left\vvvert H^{-1} \right\vvvert  \times \vvvert H \vvvert \\
+                        & 	\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| \times \left(1 + \vvvert H^{-1} \vvvert  \times \vvvert H \vvvert \right).
 \end{align*}
 
-If we can prove that the right side of the inequality is smaller than the minimum of the lattice \(\lambda_1(L)\), then we have proved that \(rG\) is indeed the vector we were searching for.
-
-In this article, we will only use the Babai's rounding when \(k = 64\).
-Hence \(\lVert 2^{k-\ell}\widetilde{T}-U' \rVert \leqslant \sqrt{n}2^{64-\ell+1}-1 \).
-So, if we fix \(n\) we can search for the minimum \(\ell\)\todo{ATTENTION ici \(\ell\) c'est le nombre de bit devinés, donc l'ancien \(6+\ell\)} such that \[(1+\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert)\sqrt{n}2^{64-\ell+1}-1 \leqslant \lambda_1(L(n,64))\]
+If we can prove that the right side of the inequality is smaller than the first
+minimum of the lattice, then we would have proved that $RH$ is indeed the
+closest vector we were searching for. Because we have fixed $k = 64$, by
+lemma~\ref{thm:close} we have
+$\left\| 2^{k-\ell}\widetilde{T} - U' \right\| \leq 2 \sqrt{n}2^{64-\ell}-1
+$. So, if we fix $n$ we can search for the minimum number $\ell$ of known
+most-significant bits such that:
+\[
+  \left(1 + \vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert \right) 2 \sqrt{n} 2^{64-\ell} \leq \lambda_1(\mathcal{L})
+\]
 
@@
   \centering
-  \begin{tabular}{|c|c|c|c|}
+  \begin{tabular}{|c|Sc|Sc|c|}
 	\hline
-	n & \(\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert\) & \( \lambda_1(L(n,64)) \) & minimum \(\ell\) \\
+	n & $\vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert$ & $ \lambda_1(\mathcal{L}) $ & minimum $\ell$ \\
 	\hline
-	3 & 2.87 & \(4.09e^{12} \simeq 2^{41.9}\) & 26 \\
-	4 & 2.06 & \(2.44e^{14} \simeq 2^{47.8}\) & 20 \\
-	5 & 3.77 & \(1.72e^{15}\simeq 2^{50.6}\) & 18 \\
-	6 & 2.69 & \(1.03e^{16}\simeq 2^{53.2}\) & 15 \\
+	3 & 2.87 & $4.09e^{12} \simeq 2^{41.9}$ & 26 \\
+	4 & 2.06 & $2.44e^{14} \simeq 2^{47.8}$ & 20 \\
+	5 & 3.77 & $1.72e^{15}\simeq 2^{50.6}$ & 18 \\
+	6 & 2.69 & $1.03e^{16}\simeq 2^{53.2}$ & 15 \\
 	\hline
   \end{tabular}
-  \caption{minimal \(\ell\) needed for a given \(n\)}
+  \caption{minimal $\ell$ needed for a given $n$}
   \label{tab:babai}
@@
 
-\section{Warm-up: A Prediction Algorithm for \textsf{PCG64} With Known Increment}
+When $\ell$ is greater than the values given in table~\ref{tab:babai}, then
+Babai's rounding technique will always return the closest vector, and allow us
+to reconstruct a truncated geometric series.
+
+\subsection{Application to the \texttt{Lehmer64} generator}
+
+Adapting the previous reasoning enables an efficient state reconstruction
+algorithm for the \texttt{Lehmer64} generator shown in
+Fig.~\ref{fig:prngs}. When clocked, it outputs the top 64 bits of a geometric
+sequence ($k=128$ and $\ell=64$). Three successive outputs are sufficient to
+reconstruct the internal state using Babai's rounding technique. This yields the
+following reconstruction algorithm:
+\begin{minted}[fontsize=\scriptsize]{python}
+def reconstruct(X):
+    """
+    Produce the internal state of the generator given three consecutive outputs.
+    16 multiplications, 1 division, 11 additions and 3 roundings only.
+    """
+    a = 0xda942042e4dd58b5
+    r = round(2.64929081169728e-7 * X[0] + 3.51729342107376e-7 * X[1] + 3.89110109147656e-8 * X[2])
+    s = round(3.12752538137199e-7 * X[0] - 1.00664345453760e-7 * X[1] - 2.16685184476959e-7 * X[2])
+    t = round(3.54263598631140e-8 * X[0] - 2.05535734808162e-7 * X[1] + 2.73269247090513e-7 * X[2])
+    u = r * 1556524 + s * 2249380 + t * 1561981
+    v = r * 8429177212358078682 + s * 4111469003616164778 + t * 3562247178301810180
+    state = (a*u + v) % (2**128)
+    return state
+\end{minted}
+
+But enough fooling around. Let's go back to \textsf{PCG64}.
+
+\section{State Reconstruction for \textsf{PCG64} With Known Increment}
 \label{sec:Cknown}
@@
 may output a few other ones (they can be easily discarded by checking $X_2$). An
-improved ``guess-and-determine'' prediction algorithm is possible, which
-essentially amounts to expose a truncated version of the underlying linear
-congruential generator, and attack it using the well-known tools. This is
-possible by combining the following ingredients:
+improved ``guess-and-determine'' state reconstruction algorithm is possible,
+which essentially amounts to expose a truncated version of the underlying linear
+congruential generator, and attack it using the tools exposed in
+section~\ref{sec:geometric}. This is possible by combining the following
+ingredients:
 \begin{itemize}
@@
 
-\item Guessing a 6-bit rotation gives access to the XOR of the two halves of the
-  internal state. If a part of the state is known, then this transfers existing
-  knowledge to the other half.
+\item Guessing a 6-bit rotation $r_i$ gives access to $Y_i$ (the XOR of the two
+  halves of the internal state). Thus, if a part of the state is known, then
+  this transfers existing knowledge to the other half.
 \end{itemize}
@@
 We consider the sequence of internal states
-$\mathbf{S} = (S_0, S_1, \dots) = \LCG_{128}(S_0, c)$. We will guess the $\ell$
+$S = (S_0, S_1, \dots) = \LCG_{128}(S_0, c)$. We will guess the $\ell$
 least-significant bits of $S_0$, therefore let us assume that their value is
-known and denote it by $w$. We define $\mathbf{S}' = \LCG_{128}(S_0 - w, 0)$ and
-$\mathbf{K} = \LCG_{128}(w, c)$ (this is Known). By~\eqref{eq:lcg-additive}, we
-have $\mathbf{S}' = \mathbf{S} - \mathbf{K}$. The point is that the elements of
-$\mathbf{S}'$ follow a geometric progression of common ratio $a$; in addition,
-the $\ell$ least significant bits of each components are equal to zero. It
-follows that $\mathbf{S}'[\ell:64+\ell]$ follows a geometric progression of
-common ratio $a$, modulo $2^{64}$. The crux of the reconstruction algorithm is
-to find $\mathbf{S}'[\ell:64+\ell]$.
+known and denote it by $w$. We define $S' = \LCG_{128}(S_0 - w, 0)$ and
+$K = \LCG_{128}(w, c)$ --- this is Known. By~\eqref{eq:lcg-additive}, we have
+$S' = S - K$. The point is that the elements of $S'$ follow a geometric
+progression of common ratio $a$; in addition, the $\ell$ least significant bits
+of each components are equal to zero. It follows that $S'[\ell:64+\ell]$ also
+follows a geometric progression of common ratio $a$, this time modulo
+$2^{64}$. The crux of the reconstruction algorithm is to find
+$S'[\ell:64+\ell]$.
 
@@
 truncation of $S'_i$. Thanks to~\eqref{eq:borrow}, we know that there is an
-unknown vector $\mathbf{B}$ of \emph{borrows}, whose components are either 0 or
-1, such that
-$\mathbf{T'} = \mathbf{S}[58:64+\ell] \boxminus \mathbf{K}[58:64+\ell] \boxminus
-\mathbf{B}$.
-
-Because the borrows are unknown, we in fact compute
-$\widetilde{\mathbf{T}'} = \mathbf{S}[58:64+\ell] \boxminus
-\mathbf{K}[58:64+\ell]$, and clearly
-$\widetilde{\mathbf{T}'} = \mathbf{T}' \boxplus \mathbf{B}$.
+unknown vector $B$ of \emph{borrows}, whose components are either 0 or 1, such
+that $T' = S[58:64+\ell] \boxminus K[58:64+\ell] \boxminus B$. Because the
+borrows are unknown, we in fact compute
+$\widetilde{T'} = S[58:64+\ell] \boxminus K[58:64+\ell]$, and clearly
+$\widetilde{T'} = T' \boxplus B$. We are thus in the context of the problem
+discussed in section~\ref{sec:geometric}, namely reconstructing a geometric
+sequence given $6+\ell$ (noisy) most-significant bits. The ``noise'' is the
+unknown vector $B$ of borrows.
 
@@
 lower-bound on $\ell$ given $n$, and we see that the total number of guessed
-bits reaches a minimum of $37$ when $n=3$ and $\ell=19$. Therefore, we guess
-$\ell = 19$ low-order bits of the state and three consecutive rotations.
+bits reaches a minimum of $38$ when $n=3$ and $\ell=20$. Therefore, success is
+guaranteed if we guess $\ell = 20$ low-order bits of the state and three
+consecutive rotations.
 
@@
 
-The procedure is completely practical. More details are given in section~\ref{sec:implem}.
+The procedure is completely practical. More details are given in
+section~\ref{sec:implem}. Let us just mention that the procedure often works
+(twice faster) with $\ell=19$ or even four times faster with $\ell = 18$ (with a
+reduced success probablity).
 
@@
   \State // \emph{Statement involving $j$ must be repeated for $j=0, 1, 2$.}
-  \State $\ell \gets 19$
+  \State $H \gets$ LLL reduction of $G_{3,64}$
+  \State $\ell \gets 20$
   \For{$0 \leq w < 2^{\ell}$} \Comment{Guess least-significant bits of $S_0$}
@@
     \State $\widetilde T'_j \gets T_j \boxminus K_j[58:64+\ell]$  \Comment{Truncated geometric series on $6+\ell$ bits}
-    \State $(U_0, U_1, U_2) \gets \left\lfloor 2^{58-\ell} \cdot (\widetilde T'_0, \widetilde T'_1, \widetilde T'_2) \cdot  G_3^{-1} \right\rceil \cdot G_3$ \Comment{CVP (Babai rounding)}
+    \State $(U_0, U_1, U_2) \gets \left\lfloor 2^{58-\ell} \cdot (\widetilde T'_0, \widetilde T'_1, \widetilde T'_2) \cdot  H^{-1} \right\rceil \cdot H$ \Comment{CVP (Babai rounding)}
 
@@
 \end{equation}
-Note that $\nabla S_0 = 0$ and $\nabla S_1 = \Delta S_0$. The prediction algorithm we propose proceeds in three phases:
+Note that $\nabla S_0 = 0$ and $\nabla S_1 = \Delta S_0$. Therefore, knowledge
+of $\Delta_0$ entails that of the whole sequence of $\nabla S_i$. The prediction
+algorithm we propose proceeds in three phases:
 \begin{enumerate}
-\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{39}$.
+\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{63}$.
 \item Reconstruct all rotations $r_i$ from this partial knowledge.
@@
 \subsection{Partial Difference Reconstruction}
+\label{sec:cvp_small_2}
 
@@
 $S_i[58:64+\ell]$ for any $i$ (and we already had $S_i[0:\ell]$). This allows us
-to compute $S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$. Given a ``fresh''
+to compute $Y_i[0:\ell] = S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$. Given a ``fresh''
 output $X_i$, and assuming that the guesses are correct, then we should have:
@@
 % \]
-The unknown value $x_0$ will not be recovered until the third phase of the whole
-procedure. We obtain the $i$-th state by $S_i \equiv \nabla S_i \boxplus S_0$;
-however, because the ``middle'' of $S_0$ is unknown, we have the same problem as
-in eq.~\eqref{eq:diff_borrow}: an unknown \emph{carry} may cross the 64-th bit
-during the addition and perturb $S_i[64:64+\ell]$. As a result, there is an
-unknown vector $\mathbf{C}$, whose components are either 0 or 1, such that such
-that:
+%The unknown value $x_0$ will not be recovered until the third phase of the whole
+%procedure.
+We obtain the $i$-th state by $S_i \equiv \nabla S_i \boxplus S_0$; however,
+because the ``middle'' of $S_0$ is unknown, then an unknown \emph{carry} may
+cross the 64-th bit during the addition and perturb $S_i[64:64+\ell]$. As a
+result, there is an unknown vector $C$, whose components are either 0 or 1, such
+that such that:
 \begin{align*}
-  S_i[64:64+\ell] &= C_i \boxplus \nabla S_i[64:64+\ell] \boxplus (w_0 \oplus Y_0[0:\ell]) \bmod 2^\ell \\
+  S_i[64:64+\ell] &= C_i \boxplus \nabla S_i[64:64+\ell] \boxplus (w_0 \oplus Y_0[0:\ell]) \\
 \end{align*}
@@
 \subsection{Full Difference Reconstruction}
+\label{sec:big_cvp}
 
@@
 
+% pas dans la version anonyme.
+% \paragraph{Acknowledgements.} The authors thank Damien Vergnaud and Vincent
+% Zucca for helpful discussions. We also thank Léo Ducas for helping with the
+% \texttt{G6K} library.
+
 \bibliographystyle{alpha}

### 2020-06-01 article/main.tex
section 5 en cours

@@
 \subsection{Reconstruction in ``High'' Dimension Using an Exact CVP Solver}
+\label{sec:cvp_big}
 
@@
 \subsection{Reconstruction in Low Dimension Using Babai's Rounding}
+\label{sec:cvp_small}
 
@@
 sequence given $6+\ell$ (noisy) most-significant bits. The ``noise'' is the
-unknown vector $B$ of borrows.
+unknown vector $B$ of borrows. %Notice that the noise is asymmetric.
 
@@
 algorithm~\ref{algo:known}. The point is that when the guesses are correct, then
-from the truncated geometric series $\mathbf{T'}$, the solution of the CVP
-instance reveals $U_j = S_j'[\ell:64+\ell]$. From there, the correction of the
-algorithm is easily established.
+from the truncated geometric series $T'$, the solution of the CVP instance
+reveals $U_j = S_j'[\ell:64+\ell]$. From there, the correction of the algorithm
+is easily established.
 
@@
 
+We consider again the sequence of internal states
+$S = (S_0, S_1, \dots) = \LCG_{128}(S_0, c)$. We will guess the $\ell$
+least-significant bits of $S_0$ and of $c$, therefore let us assume that their
+value is known and denote it by $w_0$ and $c_0$. We define
+$S' = \LCG_{128}(S_0 - w, c - c_0)$ and $K = \LCG_{128}(w_0, c_0)$ --- again,
+$K$ is Known and $S' = S - K$. This time, the components of $S'$ do \emph{not}
+follow a geometric progression; but we still have that the $\ell$ least
+significant bits of each $S'_i$ are zero. Set
+$\Delta S'_i \stackrel{def}{=} S'_{i+1} - S'_i$; $\Delta S'[\ell:64+\ell]$
+follows a geometric progression of common ratio $a$ modulo $2^{64}$
+(again). This time, we have to find $\Delta S'_0[\ell:64+\ell]$.
+
+As in section~\ref{sec:Cknown}, we have access to
+$T_i \stackrel{def}{=} S_i[58:64+\ell]$. We want to subtract the known part to
+obtain $T'_i \stackrel{def}{=} (S_i \boxminus K_i)[58:64+\ell]$, which is the
+truncation of $S'_i$. This again introduces an unknown vector $B$ of
+\emph{borrows}, and in fact we can only compute
+$\widetilde{T'} = S[58:64+\ell] \boxminus K[58:64+\ell]$, with
+$\widetilde{T'} = T' \boxplus B$. As explained above, to access a geometric
+sequence, we would like to obtain
+$\Delta T'_i \stackrel{def}{=} T'_{i+1} - T'_i$, but we can only compute:
+\[
+  \Delta \widetilde{T}'_i \stackrel{def}{=} \widetilde{T}'_{i+1} - \widetilde{T}'_i
+  =  (T'_{i+1} \boxminus T'_{i}) \boxplus (B_{i+1} \boxminus B_{i})
+\]
+
+We are thus still in the context of the problem discussed in
+section~\ref{sec:geometric}, but this time the ``noise'' caused by the carries
+is given by $B_{i+1} - B_i$. When the guesses are correct, then Babai's rounding
+will reconstruct $\Delta \widetilde{S}'[\ell:64+\ell]$ from
+$\Delta \widetilde{T}'$. This in turn yields $\Delta S_0[0:64+\ell]$.
+
 Once we have found $\Delta S_0[0:64+\ell]$, we can compute
@@
 $S_i[58:64+\ell]$ for any $i$ (and we already had $S_i[0:\ell]$). This allows us
-to compute $Y_i[0:\ell] = S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$. Given a ``fresh''
-output $X_i$, and assuming that the guesses are correct, then we should have:
+to compute $Y_i[0:\ell] = S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$. Given
+a ``fresh'' output $X_i$, and assuming that the guesses are correct, then we
+should have:
 \begin{equation}\label{eq:find_rotation}
@@
 \begin{align*}
-  S_i[64:64+\ell] &= C_i \boxplus \nabla S_i[64:64+\ell] \boxplus (w_0 \oplus Y_0[0:\ell]) \\
+  S_i[64:64+\ell] &= C_i \boxplus \nabla S_i[64:64+\ell] \boxplus (\underbrace{w_0 \oplus Y_0[0:\ell]}_{S_0[64:64+\ell]}) \\
 \end{align*}
@@
   \State $u_i \gets \Delta S_0 (a^i-1)(a-1)^{-1} \bmod 2^{64+\ell}$ \Comment{$u_i = \nabla S_i[0:64+\ell]$}
-  \State $w_i = w_0 + u_i[0:\ell] \bmod 2^{\ell}$ \Comment{$w_i = S_i[0:\ell]$}
-  \State $v_i = v_0 + u_i[64:64+\ell] \bmod 2^{\ell}$ \Comment{$S_i[64:64+\ell] \in \{v_i, v'_i\}$}
-  \State $v'_i = v_i + 1 \bmod 2^{\ell}$
+  \State $w_i = w_0 \boxplus u_i[0:\ell]$ \Comment{$w_i = S_i[0:\ell]$}
+  \State $v_i = v_0 \boxplus u_i[64:64+\ell]$ \Comment{$S_i[64:64+\ell] \in \{v_i, v'_i\}$}
+  \State $v'_i = v_i \boxplus 1$
   \State $\mathcal{C}_i \gets \{ w_i \oplus (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$ \Comment{Check eq.~\eqref{eq:consistency}}
@@
 \Procedure{ReconstructPartialDifference}{$X_0, \dots, X_k$}
-  \For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
-  \State $K_j \gets a^j w_0 + (2c+1)(a^j - 1)(a-1)^{-1} \bmod 2^{128}$ \Comment{Known part}
+  \State // \emph{Statement involving $j$ must be repeated for $j=0, 1, 2, 3, 4$.}
+  \State $H \gets$ LLL reduction of $G_{4,64}$
+  \State $\ell \gets 14$
+\For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c_0 < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
+\State $K_j \gets a^j w_0 + (2c_0 + 1)(a^j - 1)(a-1)^{-1} \bmod 2^{128}$ \Comment{Known part}
   \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
@@
   \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(K_j \oplus Y_j\right)[0:\ell]$ \Comment{Truncated LCG}
-  \State $T'_j \gets T_j \boxminus  K_i[58:64+\ell]$ \Comment{Truncated geometric series}
-  \State \todo{Attention double retenue : $\pm 1$.}
-  \State $\Delta T'_j = (T'_{j+1} - T'_j) \bmod 2^{6 + \ell}$  \Comment{Difference}
-  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \dots, \Delta T'_3) \cdot 2^{58-\ell} \cdot \widetilde G_4^{-1} \right\rceil \cdot \widetilde G_4$ \Comment{CVP}
-  \State $\Delta S_0[0:64+\ell] \gets \left((w_1 - w_0) \bmod 2^{\ell}\right) + 2^{\ell} \cdot \Delta U_0[0:58-\ell]$ \Comment{Check}
+  \State $\widetilde{T}'_j \gets T_j \boxminus  K_i[58:64+\ell]$ \Comment{Cancel known part}
+  \State $\widetilde{\Delta T'}_j = \widetilde{T'}_{j+1} \boxminus \widetilde{T}'_j$  \Comment{Difference (truncated geom. seq.)}
+  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \dots, \Delta T'_3) \cdot 2^{58-\ell} \cdot \widetilde H^{-1} \right\rceil \cdot \widetilde H$ \Comment{CVP}
+  \State $\Delta S_0[0:64+\ell] \gets \left(K_1 \boxminus K_0\right)[0:\ell]
+  + 2^{\ell} \cdot \Delta U_0[0:64]$ \Comment{Check}
   \If{$\Call{ConsistencyCheck}{\Delta_0, w_0, Y_0, X_5, \dots, X_k}$}
@@
 progression. Looking at table~\ref{tab:babai}, we see that the best choice
-consists in guessing 5 consecutive rotations and $\ell=13$ least-significant
-bits. Therefore, \textsc{ReconstructPartialDifference} does $2^{55}$ iterations
+consists in guessing 5 consecutive rotations and $\ell=14$ least-significant
+bits. Therefore, \textsc{ReconstructPartialDifference} does $2^{57}$ iterations
 of the inner loop, and succeeds deterministically.
 
-% \begin{theorem}
-%   If $\ell \geq 13$, then algorithm \textsc{ReconstructPartialDifference} is
-%   correct (it always returns the correct values).
-% \end{theorem}
-
-% \begin{proof}
-%   \todo{WRITE-ME}
-%   \begin{align*}
-%   \mathbf{S} &= \LCG(S_0, c) \\
-%              &= \LCG\left(S_0[0:\ell] + 2^\ell S_0[\ell:128], c[0:\ell] + 2^\ell c[\ell:128]\right) \\
-%              &= \LCG\left(S_0[0:\ell], c[0:\ell]\right) + 2^\ell \LCG\left(S_0[\ell:128],  c[\ell:128]\right) \\
-%   \Delta \mathbf{S} &= \LCG(S_1, c) - \LCG(S_0, c) \\
-%              &= \LCG(S_1 - S_0, 0)
-% \end{align*}
-% \end{proof}
-
-\subsection{Full Difference Reconstruction}
-\label{sec:big_cvp}
+\subsection{Predicting all the Rotations}
 
@@
 subsequent rotations deterministically, at negligible cost, using
-eq~\eqref{eq:find_rotation}.
-
-For each $i$, it suffices to try the $64$ possible values of $r_i$; only one
-should satisfy the equation. Using $X_0, \dots, X_{40}$, we thus recover the 6
-most-significant bits of $S_0, \dots, S_{40}$. We therefore compute the 6 most
-significant bits of the differences $\Delta S_i$ between consecutive states (up
-to missing carries), and we are faced with the problem of reconstructing a
-128-bit geometric progression using 39 consecutive outputs truncated to their 6
-most-significant bits. This can be done by solving an instance of CVP in
-dimension $\leq 40$. We use the off-the-shelf CVP solver embedded in
-\texttt{fplll}: it runs in no measurable time.
-
-More precisely, there are still details to be dealt with regarding
-carries. Indeed, the \textsc{ReconstructRotations} returns the 6 most
-significant bits of any number of $S_i$'s (denoted by $r_i$). Therefore, there
-are unknown 122-bit quantities $x_i$ such that:
-\begin{align*}
-  S_i &= 2^{122} r_i + x_i \\
-  \Delta S_i &\equiv 2^{122} (r_{i+1} - r_i) + (x_{i+1} - x_{i}) \bmod 2^{128}
-\end{align*}
-
-\todo{check indices nabla}
+eq~\eqref{eq:find_rotation}. For each index $i$, it suffices to try the $64$
+possible values of $r_i$; only one should satisfy
+eq~\eqref{eq:find_rotation}. The complete pseudo-code is shown in
+algorithm~\ref{algo:unknown_2}.
+
+It is unlikely that several possible values of $r_i$ match: each value is
+``checked'' on $\ell$ bits, so an accidental match happens with probability
+$2^{\ell-6}$. The total number of lists returned by
+\textsc{ReconstructRotations} then follows a binomial distribution of parameters
+$2^{\ell-6}, k$. With $\ell=14$ and $k=64$, then only one rotation vector should
+pass the test for $0 \leq i < 64$ on average.
 
@@
   \EndFunction
-  \State
-  \Function{ReconstructFullDifference}{...}
-  \State \todo{WRITEME}
-  \EndFunction
 \end{algorithmic}
@@
 
-% \begin{theorem}
-%   The algorithm described above works with probability.... in time ...
-% \end{theorem}
+\subsection{Full Difference Reconstruction}
+\label{sec:big_cvp}
+
+Using $X_0, X_1, \dots, X_{63}$, we recover all rotations and thus we recover
+the 6 most-significant bits of $S_0, S_1, \dots, S_{63}$. This allows us to
+compute the 6 most significant bits of the differences $\Delta S_i$ between
+consecutive states (up to missing carries), and we are faced with the problem of
+reconstructing a 128-bit geometric progression using 63 consecutive outputs
+truncated to their 6 most-significant bits. There is again an unknown vector of
+borrows $B$ such that $\Delta S_i[122:128] \boxplus C_i= r_{i+1} \boxminus r_i$.
+
+Reconstructing $\Delta S_0$ from the $r_i$ is exactly the problem discussed in
+section~\ref{sec:cvp_big}. This can be done by solving an instance of CVP in
+dimension $63$. We use the off-the-shelf CVP solver embedded in \texttt{fplll}:
+it runs in no measurable time.
 

### 2020-06-01 article/main.tex
fin section 5

@@
 \begin{equation}\label{eq:nabla}
-  \nabla S_i \stackrel{def}{=} S_i - S_0 = \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
+  \nabla S_i \stackrel{def}{=} S_i - S_0 \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
 \end{equation}
@@
   \State $\widetilde{T}'_j \gets T_j \boxminus  K_i[58:64+\ell]$ \Comment{Cancel known part}
-  \State $\widetilde{\Delta T'}_j = \widetilde{T'}_{j+1} \boxminus \widetilde{T}'_j$  \Comment{Difference (truncated geom. seq.)}
-  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \dots, \Delta T'_3) \cdot 2^{58-\ell} \cdot \widetilde H^{-1} \right\rceil \cdot \widetilde H$ \Comment{CVP}
+  \State $\Delta \widetilde{T}'_j = \widetilde{T}'_{j+1} \boxminus \widetilde{T}'_j$  \Comment{Difference (truncated geom. seq.)}
+  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta \widetilde{T}'_0, \dots, \Delta \widetilde{T}'_3) \cdot 2^{58-\ell} \cdot \widetilde H^{-1} \right\rceil \cdot \widetilde H$ \Comment{CVP}
   \State $\Delta S_0[0:64+\ell] \gets \left(K_1 \boxminus K_0\right)[0:\ell]
@@
   \If{$\Call{ConsistencyCheck}{\Delta_0, w_0, Y_0, X_5, \dots, X_k}$}
-  \State \Return $(w_0, c, r_0, \dots, r_4, \Delta_0)$.
+  \State \Return $(w_0, c_0, r_0, \dots, r_4, \Delta S_0)$.
   \EndIf
@@
 
-The whole procedure is shown in algorithm~\ref{algo:last}.
+The whole procedure is shown in algorithm~\ref{algo:last}. Note that once $S_0$
+has been found, then all subsequent states can be computed with error using
+$S_i = S_0 \boxplus \nabla S_i$. In particular, computing $S_1$ gives $c$ by
+$c \gets S_1 \boxminus a S_0$. This complete the reconstruction procedure for
+\textsf{PCG64}.
 
@@
 \textsf{Python} (for the rest). We used the \texttt{fplll} library~\cite{fplll}
-to solve CVP instances exactly in dimension 42.
+to solve CVP instances exactly in dimension 63.
 
 In this section, we briefly outline important aspects of our implementations and
-present practical results. Our codes are available online at:
+present practical results. Our codes are available in the supplementary material
+as well as online at:
 \[
@@
 \]
-(and in the supplementary material).
-
 The designer of \textsf{PCG} was kind enough to send us two sets challenge

### 2020-06-01 article/main.tex
des todos par milliers

--- 2020-06-01 article/main.tex
@@
 
-\keywords{keywords}
+\keywords{\todo{keyword}}
 
 \begin{abstract}
-  blablabla
+  \todo{blabla}
 \end{abstract}
@@
 Not all pseudo-random generators are of cryptographic strength. In some
-applications, it is not necessarily necessary: to be used in Monte-Carlo
+applications, it is not necessarily necessary \todo{lourd}: to be used in Monte-Carlo
 numerical simulations or generate random choices in games, a relaxed,
@@
 family to send us ``challenge'' pseudo-random streams ; we ran our code and
-emailed back the seeds used to generate the challenge streams the next day.
+emailed back the seeds used to generate the challenge streams the next day.\todo{et on avait tout juste, enfin j'espère}
 
@@
 
-The LCG have been largely studied, and considered as cryptographically insecure.
-
-A reconstruction algorithm has been proposed by D.
-Knuth in 1985\cite{Knuth85}.
-The LCG he considered revealed only the upper bits of each internal state.
-The modulus $2^k$ was known while the multiplier $a$ and the increment $c$ weren't.
-This algorithm could reconstruct the LCG in $\mathcal{O}(2^l)$ where $l$ is the number of non revealed bit of each internal state.
-
 In 1989, J.
-Boyar improve the result in a first way \cite{Boyar1989}.
+Boyar studied further the problem \cite{Boyar1989}.
 She presented an algorithm which could predict a LCG where all the parameters (multiplier, increment and initial state) where unknown under the condition that the number of bit unrevealed was really small in comparison to the modulus.
@@
 In 1988, A.
-Frieze \textit{et al} \cite{Frieze} improve the result in a second way.
+Frieze \textit{et al} \cite{Frieze} improved the efficiency of reconstruction algorithms in simpler cases.
 They supposed that the multiplier $a$ and the modulus $2^k$ were known and used lattice-based technique to recover the LCG more efficiently.
@@
 \]
-Where is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
+Where \todo{il manque une lettre, je suppose que c'est a} is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
 the generator. The increment $c$ may be specified by the user of the PRNG to
@@
 \emph{carry} $0 \leq \gamma \leq 1$ (resp. a \emph{borrow}
-$0 \leq 1 \leq \beta$) such that:
+$0 \leq 1 \leq \beta$ \todo{je pense pas que ce soit ça}) such that:
 \begin{align}
@@
 \begin{align*}
-|\CVP(\mathcal{L}, T) - U'| %&= \left| \CVP\left( \mathcal{L}, 2^{k-\ell}\widetilde{T} \right) \riht| + \left|2^{k-\ell}\widetilde{T} - U'\right| %-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
+|\CVP(\mathcal{L}, 2^{k-\ell}\widetilde{T}) - U'| %&= \left| \CVP\left( \mathcal{L}, 2^{k-\ell}\widetilde{T} \right) \riht| + \left|2^{k-\ell}\widetilde{T} - U'\right| %-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
                              & \leq \left| \CVP\left(\mathcal{L}, 2^{k-\ell}\widetilde{T} \right) - 2^{k-\ell}\widetilde{T}\right | + \left|2^{k-\ell}\widetilde{T} - U'\right|.
@@
 \[
-  |\CVP(\mathcal{L}, T) - U'| \leq 2 \left|2^{k-\ell} \widetilde{T} - U' \right| \leq 4 \sqrt{n} 2^{k-\ell}.
+  |\CVP(\mathcal{L}, 2^{k-\ell}\widetilde{T}) - U'| \leq 2 \left|2^{k-\ell} \widetilde{T} - U' \right| \leq 4 \sqrt{n} 2^{k-\ell}.
 \]
@@
 minimum of the lattice $\lambda_1(\mathcal{L})$, then we would have proved that
-$\CVP(\mathcal{L}, T)$ indeed reveals $U_0 \bmod 2^k$.
+$\CVP(\mathcal{L}, 2^{k-\ell}\widetilde{T})$ indeed reveals $U_0 \bmod 2^k$.
 
@@
 \section{Conclusion}
-
+\todo{y'a pas de conclusion}
 % pas dans la version anonyme.

### 2020-06-01 Cknown/fonctions.c
success proba

@@ -35,7 +35,7 @@ void init_var_globales(){
     //increment polynome polC OK !
     polC[0] = 0;
     pcg128_t powA =1;
-    for (int i = 1; i < nbiter ; i++){
+    for (int i = 1; i < 3 ; i++){
         polC[i] = polC[i-1] + powA * c;
         powA *= a;
     }
@@ -65,7 +65,7 @@ static u64 rot(u64 x, int i)
 void getPolW(pcg128_t *polW, u64 W0)
 { //OK !
     polW[0] = W0;
-    for(int i = 1 ; i < nbiter ; i++){
+    for(int i = 1 ; i < 3 ; i++){
         polW[i] = polW[i-1] * a;
     }
 }
@@ -73,7 +73,7 @@ void getPolW(pcg128_t *polW, u64 W0)
 void getSumPol(u64* sumPol, u64* sumPolY, const pcg128_t* polW)
 {
     pcg128_t sum;
-    for(int i = 0 ; i < nbiter ; i++){
+    for(int i = 0 ; i < 3 ; i++){
         sum = polC[i] + polW[i];
         sumPol[i] = sum;
         sumPolY[i] = (sum >> 58) % (1 << (known_low + 6));
@@ -98,9 +98,9 @@ static inline long long crazy_round(double x)
 
 void setup_task(u64 W0, const u64 *urX, struct task_t *task)
 {
-        pcg128_t polW[nbiter]; // temporary byproduct
-        u64 sumPol[nbiter]; // temporary byproduct
-        u64 sumPolY[nbiter]; // temporary byproduct
+        pcg128_t polW[3]; // temporary byproduct
+        u64 sumPol[3]; // temporary byproduct
+        u64 sumPolY[3]; // temporary byproduct
         
         // initialise polW à partir de W0
         getPolW(polW, W0);
@@ -108,8 +108,8 @@ void setup_task(u64 W0, const u64 *urX, struct task_t *task)
         // initialise sumPol et sumPolY à partir de polW
         getSumPol(sumPol, sumPolY, polW);
 
-        u64 Yprime[nbiter][64];
-        for (int i = 0; i < nbiter; i++)
+        u64 Yprime[3][64];
+        for (int i = 0; i < 3; i++)
                 for (int j = 0; j < 64; j++) {
                         u64 X = rot(urX[i], j);
                         u64 Y = (((sumPol[i] ^ X) % (1 << known_low)) << 6) + (j ^ (X >> 58));
[…]

### 2020-06-01 Cknown/success_proba.c

@@ -1,0 +1,81 @@
+#include <assert.h>
+#include <stdlib.h>
+#include <stdio.h>
+#include <time.h>
+
+#include "fonctions.h"
+
+FILE *f; 
+
+bool run_test()
+{
+	pcg128_t S[nbiter];
+
+	if (fread(&S[0], sizeof(pcg128_t), 1, f) != 1) {
+		perror("Something went wrong when reading /dev/urandom");
+		exit(EXIT_FAILURE);
+	} 	
+        
+	// prepare input
+	u64 X[nbiter];
+    	for (int i = 1; i < nbiter; i++)
+    		S[i] = a * S[i-1] + c;
+    	for (int i = 0; i < nbiter; i++)
+    		X[i] = pcg_output_xsl_rr_128_64(S[i]);
+
+    	// cheat
+      	u64 W0 = (u64) (S[0] % (1 <<known_low));
+	int rot[nbiter];
+	for (int i = 0; i < nbiter; i++)
+		rot[i] = (int) (S[i] >> 122);
+	
+      	// for (int i = 0; i < nbiter; i++)
+      	// 	printf("X[%d] = %016llx\n", i, X[i]);
+      	// printf("W0 = %llx\n", W0);
+
+
+	struct task_t task;
+	setup_task(W0, X, &task);
+
+	//int rot[nbiter] = {0, 0, 0};
+	
+	//for (int r = 0 ; r < 1 << (nbiter*6) ; r++) {
+	//	/* bump rotations */
+	//    rot[0] = (rot[0] + 1) % 64;
+	//    int i = 0;
+	//    while (rot[i] == 0 && i < nbiter) {
+	//	i++;
+	//	rot[i] = (rot[i] + 1) % 64;
+	//    }
+	//    if (i > 0)
+	    	refresh_task(rot, &task);
+
+	    pcg128_t S_target[nbiter];
+	    if (solve(S_target, rot, &task)) {
+	    	assert(S_target[0] == S[0]);
+		return true;
+	    }
+	//}
+	return false;
+}
+
+
+int main()
+{
+    
+    /*  INITIALISATION DES PARAMETRES  */    
+    init_var_globales();
+    f = fopen("/dev/urandom", "r");
+    if (f == NULL) {
+		perror("Something went wrong when opening /dev/urandom");
+		exit(EXIT_FAILURE);
+    }
+
+    int good = 0;
+    for (int foobar = 0; foobar < 1000000; foobar++) {
[…]

### 2020-06-01 Cunknown/fonctions.c
success_proba + more stuff

@@ -26,9 +26,9 @@ double invG[16] =
 void init_var_globales()
 {
     //multiplier a OK !
-    a = (((pcg128_t) 2549297995355413924) << 64) + ((pcg128_t) 4865540595714422341);
+    a = (((pcg128_t) 2549297995355413924ull) << 64) + ((pcg128_t) 4865540595714422341ull);
 
-    a_inv = (((pcg128_t) 566787436162029664) << 64) + ((pcg128_t) 11001107174925446285);
+    a_inv = (((pcg128_t) 566787436162029664ull) << 64) + ((pcg128_t) 11001107174925446285ull);
 
     //increment polynome polC OK !
     polA[0] = 0;

### 2020-06-01 Cunknown/fonctions.h

@@ -11,7 +11,7 @@ typedef long long          i64;
 typedef unsigned long long u64;
 
 /***** Macro et Variables globales *****/
-#define known_low 11 //à relancer 3 fois, sur des X differents (X0 - X4 ,X5 - X9,...) puis lancer avec 12 et enfin 13
+#define known_low 10 //à relancer 3 fois, sur des X differents (X0 - X4 ,X5 - X9,...) puis lancer avec 12 et enfin 13
 #define nbiter 5
 #define nbtest 4
 #define nboutput (nbiter + nbtest)
@@ -27,7 +27,7 @@ struct task_t {
     u64 lowSumPol[nbiter + nbtest];
     u64 sumPolY[nbiter];
     u64 sumPolTest[nbtest];
-    u64 tabTmp[k * nbiter];
+    u64 tabTmp[64 * nbiter];
     int rot[nbiter];
     char *goodY;
 };

### 2020-06-01 Cunknown/main.c

@@ -128,12 +128,12 @@ void do_task(u64 current, struct task_t *task, const u64 (*X)[nboutput])
     u64 WC = 1 + 2 * (current % (1 << (known_low - 1)));
     
     prepare_task(X, W0, WC, task);
-    for (u64 r = 0; r < 1 << (nbiter * known_up); r++) {
-	task->rot[0] = (task->rot[0] + 1) % k;
+    for (u64 r = 0; r < 1 << (nbiter * 6); r++) {
+	task->rot[0] = (task->rot[0] + 1) % 64;
 	int i = 0;
 	while (task->rot[i] == 0 && i < nbiter) {
 	    i++;
-	    task->rot[i] = (task->rot[i] + 1) % k;
+	    task->rot[i] = (task->rot[i] + 1) % 64;
 	}
 	if (solve_isgood(task))
 	    result_found(X, W0, WC, r);

### 2020-06-01 Cunknown/success_proba.c

@@ -1,0 +1,80 @@
+#include <assert.h>
+#include <stdlib.h>
+#include <stdio.h>
+#include <time.h>
+
+#include "fonctions.h"
+
+FILE *f; 
+struct task_t task;
+
+
+bool run_test()
+{
+ 	
+	pcg128_t S[nboutput];
+	pcg128_t c;
+
+	if (fread(&S[0], sizeof(pcg128_t), 1, f) != 1) {
+		perror("Something went wrong when reading /dev/urandom");
+		exit(EXIT_FAILURE);
+	} 	
+        if (fread(&c, sizeof(pcg128_t), 1, f) != 1) {
+		perror("Something went wrong when reading /dev/urandom");
+		exit(EXIT_FAILURE);
+	}
+	c |= 1;
+
+	// prepare input
+	u64 X[nboutput];
+    	for (int i = 1; i < nboutput; i++)
+    		S[i] = a * S[i-1] + c;
+    	for (int i = 0; i < nboutput; i++)
+    		X[i] = pcg_output_xsl_rr_128_64(S[i]);
+
+    	// cheat
+      	u64 W0 = (u64) (S[0] % (1 <<known_low));
+      	u64 Wc = (u64) (c    % (1 <<known_low));
+	   assert((Wc % 2) == 1);
+
+	prepare_task(&X, W0, Wc, &task);
+	 
+    for (int i = 0; i < nbiter; i++)
+        task.rot[i] = (int) (S[i] >> 122);
+
+	 /* check */
+	bool ok = false;
+        if (solve_isgood(&task)) {
+            u64 DS640;
+            u64 Y0;
+            solve(&task, &DS640, &Y0);
+            ok = true;
+        }
+        
+        /* reset goodY */
+    finish_task(&X, &task);
+	return ok;
+}
+
+
+int main()
+{
+    
+    /*  INITIALISATION DES PARAMETRES  */    
+    init_var_globales();
+
+    f = fopen("/dev/urandom", "r");
+    if (f == NULL) {
+		perror("Something went wrong when opening /dev/urandom");
+		exit(EXIT_FAILURE);
+    }
+    init_task(&task);
+
+    int good = 0;
+    for (int foobar = 0; foobar < 1000000; foobar++) {
+    	if (run_test())
+    		good++;
+    }
+    printf("Total good = %d\n", good);
+    return EXIT_SUCCESS;
+}

### 2020-06-01 challenges/pcg64-check.c

@@ -1,0 +1,28 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <stdint.h>
+
+#include "pcg_variants.h"
+
+int main()
+{ 
+	pcg128_t seed = PCG_128BIT_CONSTANT(0xffbb5e70239d0d0dULL, 0x5a21d22cdb7dfb4cULL);
+	pcg128_t increment = PCG_128BIT_CONSTANT(0x3d2204fa591b526aULL, 0x07cc1987323f4721ULL);
+ 
+	printf("increment : %016" PRIx64 " %016" PRIx64 "\n", (uint64_t) (increment >> 64), (uint64_t) increment);
+	printf("Seed      : %016" PRIx64 " %016" PRIx64 "\n\n", (uint64_t) (seed >> 64), (uint64_t) seed);
+
+	pcg64_random_t rng;
+	pcg64_srandom_r(&rng, seed, increment);
+
+	printf("Predictor input:\n");
+	for (int i = 0; i < 48; i++) {
+		printf("X[%2d] = 0x%016" PRIx64 ";\n", i, pcg64_random_r(&rng));
+	}
+	printf("\n");
+	printf("Remaining of the sequence:\n");
+	for (int i = 48; i < 64; i++)
+		printf("X[%2d] = 0x%016" PRIx64 ";\n", i, pcg64_random_r(&rng));
+
+	return EXIT_SUCCESS;
+}

### 2020-06-01 challenges/pcg64s-check.c

@@ -1,0 +1,25 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <stdint.h>
+
+#include "pcg_variants.h"
+
+int main()
+{
+	pcg128_t seed = (((pcg128_t) 0x5b3d418b8b51364b) << 64) + ((pcg128_t) 0x74eed710223e763f);
+	printf("Seed : %016" PRIx64 " %016" PRIx64 "\n\n", (uint64_t) (seed >> 64), (uint64_t) seed);
+
+	pcg64s_random_t rng;
+	pcg64s_srandom_r(&rng, seed);
+
+	printf("Predictor input:\n");
+	for (int i = 0; i < 3; i++)
+		printf("X[%2d] = 0x%016" PRIx64 ";\n", i, pcg64s_random_r(&rng));
+	printf("\n");
+
+	printf("Remaining of the sequence :\n");
+	for (int i = 3; i < 16; i++)
+		printf("X[%2d] = 0x%016" PRIx64 ";\n", i, pcg64s_random_r(&rng));
+
+	return EXIT_SUCCESS;
+}

### 2020-06-01 good_big_matrix.py

@@ -1,0 +1,107 @@
+from math import log
+import random
+import fpylll
+
+# this python program produces a good basis for
+# the "large" lattice in dimension 64, modulo 2**128;
+# The shortest vector has norm 2^127.03...
+
+a = 47026247687942121848144207491837523525
+
+def getG(n, mod):
+    """
+    Naive basis
+    """
+    G = fpylll.IntegerMatrix(n, n)
+    for i in range(0, n):
+        G[0, i] = pow(a, i, mod)
+    for i in range(1, n):
+        G[i, i] = mod
+    return G
+
+G = getG(64, 2**128)
+
+L = [
+ 12144252875850345479015002205241987363,
+ -15348615718011256064636590606166444173,
+ -9084835538682141851035496395127502261,
+ 17234466561204322281445746373216450637,
+ -30709520135048357458223997075862325767,
+ 13273365585194684751192188342361392061,
+ -16657707705878774883995565342082094357,
+ 10551835762264010302610847026218529611,
+ 9741843352863159575387669322076694498,
+ 30401545415842458720824638334153655996,
+ -21301422832830108630138234390597720063,
+ 12943082679906251720732751543854447750,
+ -26608736029397995753447945248664058244,
+ 16504008779817367542456313016700742074,
+ -10623665749400971081452814639205027657,
+ -42221955640534425567937724106639808954,
+ 30891705174976345231868365741077473210,
+ 36583589311333508006721438259000555949,
+ -7050528015382194186611161130462441554,
+ 29023698504320799821125567462694089388,
+ -28780651977143078032165325525310692006,
+ 80641828999898464226011186926508194344,
+ -56574857747725737575017746952734912354,
+ 69356676566431307205578324735925446877,
+ -33869683024546744201772555262569839986,
+ -26565235055559932529672543662504791838,
+ 12570516463211853078121725511001272792,
+ -3545191593186327315454855798676360051,
+ 34483716896778783922111688009948767715,
+ -20939456075167816323854415536161525816,
+ 37054527231762903007893323872329597868,
+ -65113113872795261158497215763575544978,
+ -27970657614360739023057572629108089923,
+ -9380443740630786183893558576010683559,
+ -31684372116213316987059381973832378607,
+ 6514548643534369904269689195712545814,
+ 14847729039548998339264458667105218977,
+ 20177225057138896843222830352311122871,
+ -10200019003678439599482650259589205069,
[…]

### 2020-06-01 article/main.tex
section 6

--- 2020-06-01 article/main.tex
@@
 
+The analysis of section~\ref{sec:geometric} yields parameters that guarantee
+that the reconstruction procedure \emph{always} succeeds. In most cases, these
+parameters are pessimistic. We ran a series of experiments to determine more
+practical choices: using smaller-than-guaranteed values of $\ell$ (the number of
+guessed least-significant bits), we measured the success probability of the
+state reconstruction procedure. The results are shown in table~\ref{tab:empirical}.
+
+\begin{table}
+  \centering
+\begin{tabular}{|c|c|c|c|c|}
+  \cline{1-2}\cline{4-5}
+  \multicolumn{2}{|c|}{$n=3$ (section~\ref{sec:Cknown})} & &  \multicolumn{2}{c|}{$n=4$ (section~\ref{sec:cvp_small_2})} \\
+  \cline{1-2}\cline{4-5}
+  $\ell$ & Success proba. && $\ell$ & Success proba. \\
+  \cline{1-2}\cline{4-5}
+  \noalign{\vskip\doublerulesep
+  \vskip-\arrayrulewidth}
+  \cline{1-2}\cline{4-5}
+  16 & $\approx 0.125$     && 10 &  $\approx 0.12$ \\
+  \cline{1-2}\cline{4-5}
+  17 & $\approx 0.25$     && 11 &  $\approx 0.64$ \\
+  \cline{1-2}\cline{4-5}
+  18 & $\approx 0.5$      && 12 &  $\approx 0.995$ \\
+  \cline{1-2}\cline{4-5}
+  19 & 1                  && 13 &  1 \\
+  \cline{1-2}\cline{4-5}
+  20 & 1 (proved)         && 14 &  1 (proved) \\
+  \cline{1-2}\cline{4-5}
+\end{tabular}
+\caption{Empirical success probabilities with smaller parameters.}
+\label{tab:empirical}
+\end{table}
+
 \subsection{Known Increment}
@@
 
+From section~\ref{sec:cvp_small}, we know that guessing $\ell=20$
+least-significant bits ensures deterministic success. However, we observed
+empirically that $\ell=19$ works with probability $\approx 1$, and runs twice as
+fast. $\ell=18$ and $\ell=17$ run with probability $\approx 1/2$ and
+$\approx 1/4$ respectively, therefore are much less useful. In practice, we used
+$\ell=19$.
+
 We ran it on a server equipped with two 16-core \textsf{Intel Xeon Gold 6130 CPU
-  @ 2.10GHz} ``\emph{Skylake}'' CPUs. The inner loop does $2^{37}$ iterations
+  @ 2.10GHz} (``\emph{Skylake}'') CPUs. The inner loop does $2^{37}$ iterations
 and terminates in 42.3s, which makes 23 core minutes.
@@
 When the increment $c$ is known, the internal state of \textsf{PCG64} can be
-practically reconstructed from $X_0, \dots, X_{42}$ using the algorithms shows
+practically reconstructed from $X_0, \dots, X_{63}$ using the algorithms shown
 in section~\ref{sec:Cunknown}. Only algorithm~\ref{algo:unknown_1} is
@@
 
-We have shown that algorithm~\ref{algo:unknown_1} is correct when $\ell=13$. In
-fact, we empirically observed that it returns the right result with probability
-$\approx 66\%$ when $\ell=11$ and with probability $99\%$ when
-$\ell=12$. Decrementing $\ell$ reduces the workload by a factor of 4. Therefore,
-if $T$ denotes the running time when $\ell=13$, we actually used the following
-strategy:
-\begin{algorithmic}[1]
-\State Run with $\ell=11$ (if it succeeds, stop).
-\State Run with $\ell=11$ with different inputs.
-\State Run with $\ell=11$ with different inputs.
-\State Run with $\ell=12$.
-\State Run with $\ell=13$.
-\end{algorithmic}
-It is not difficult to check that the expected running time of this procedure is
-$\frac{1079}{10800} T \approx 0.1 T$. In our implementation, $T = 200,000$ CPU
+We have shown that algorithm~\ref{algo:unknown_1} is correct when $\ell=14$. The
+procedure does $2^{29 + 2\ell}$ iterations of the inner loop, so decreasing
+$\ell$ would really be interesting. Looking at table~\ref{tab:empirical}, we
+settle for $\ell=13$ in the worst case; let $T$ denotes the running time when
+$\ell=13$.
+
+It seems that the most promising strategy consists in choosing $\ell=11$; if the
+reconstruction procedure fails, then we try again with different inputs. The
+expected running time of this approach number of trials is
+$T/(16 \times 0.64) \approx T / 10.25$. In our implementation, $T = 200,000$ CPU
 hours, so the expected running time of the reconstruction procedure is about
 $20,000$ CPU hours. In fact we were lucky: on the challenge input, the first
-attempt with $\ell=11$ succeeded and took only $12,500$ CPU hours.
-
-The whole process actually ran in 35 wall-clock minutes using 512 cluster nodes,
-each equipped with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz}
-(``\emph{Cascade Lake}''). The actual machine is the
-\verb|[----REDACTED FOR PEER REVIEW----]| computer located at the
-\verb|[----ALSO REDACTED----]| computation center. Note that on this particular
-parallel computer, running the algorithm with $\ell=13$ would take 10 hours
-using the same amount of resources, so the whole procedure \emph{is} practical,
-even in the absolute worst case.
+attempt with $\ell=11$ succeeded, so the whole process took only $12,500$ CPU
+hours.
+
+It actually ran in 35 wall-clock minutes using 512 cluster nodes, each equipped
+with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz} (``\emph{Cascade
+  Lake}''). The actual machine is the \verb|[----REDACTED FOR PEER REVIEW----]|
+computer located at the \verb|[----ALSO REDACTED----]| computation center. Note
+that on this particular parallel computer, running the algorithm with $\ell=13$
+would take 10 hours using the same amount of resources, so the whole procedure
+\emph{is} practical, even in the absolute worst case.
 
@@
 the same speed, potentially because of ``turbo boost'' adjustments and thermal
-constraints; most of the time, they take exactly... 42 seconds). Because of this
-variability, we implemented a master-slave work distribution pattern, in which a
-master process dispatches iterations of the outer loop to slave processes. This
-also made checkpointing very easy. We used \textsf{MPI} for inter-process
-communication.
+constraints). Because of this variability, we implemented a master-slave work
+distribution pattern, in which a master process dispatches iterations of the
+outer loop to slave processes. This also made checkpointing very easy. We used
+\textsf{MPI} for inter-process communication.
 

### 2020-06-01 Cunknown/step2.py
torture fplll avec n=64

@@ -11,7 +11,7 @@ known_up  = 6
 known_low  = 11
 a = 2549297995355413924 * 2**64 + 4865540595714422341
 nbiter = 5
-nboutput = 40
+nboutput = 64
 N = 2**128
 K = 2**64
 LOW = 2**known_low
@@ -199,8 +199,10 @@ def findDS(rot, W0, WC): #OK!
     for i in range(nboutput):
         rotprim.append((rot[i] - ((powA[i] * W0 + polA[i] * WC) >> 122)) % 64)
     tmp = tuple([(rotprim[i+1] - rotprim[i]) << (122 - known_low) for i in range(nboutput - 1)])
-    return fpylll.CVP.closest_vector(Greduite2, tmp, method = "proved")
-    
+    print("running fpylll")
+    return fpylll.CVP.closest_vector(Greduite2, tmp, method = "fast")
+    print("done")
+
 def findDSdebug(rot, W0, WC): #OK!
     """
     >>> rot = (36, 11, 26, 47, 27, 43, 40, 4, 6, 14, 46, 7, 26, 15, 29, 52, 5, 27, 59, 51, 50, 10, 18, 12, 20, 13, 41, 11, 33, 29, 11, 34, 57, 43, 57, 9, 4, 44, 5, 61)
@@ -219,8 +221,9 @@ def findDSdebug(rot, W0, WC): #OK!
     rotprim[1] = (rotprim[1] - 1) % 64
     # fin différence
     tmp = tuple([(rotprim[i+1] - rotprim[i]) << (122 - known_low) for i in range(nboutput - 1)])
-    return fpylll.CVP.closest_vector(Greduite2, tmp, method = "proved")
-
+    print("running fpylll")
+    return fpylll.CVP.closest_vector(Greduite2, tmp, method = "fast")
+    print("done")
 
 def full_DS(X, W0, WC, rots):
     uX = unrotateX(X,rot)    
@@ -243,7 +246,7 @@ def full_DS(X, W0, WC, rots):
 
 
 if __name__ == '__main__':
-    X = [0] * 48
+    X = [0] * nboutput
     X[ 0] = 0x46b9d66bb1e9fbf0;
     X[ 1] = 0xc0b7382496f0e363;
     X[ 2] = 0x78bf5fb5f4bbc09e;
[…]

### 2020-06-01 article/main.tex
micro stuff

@@
 
-\subsection{Application to the \texttt{Lehmer64} generator}
+\subsection{Application to the \texttt{lehmer64} generator}
 
 Adapting the previous reasoning enables an efficient state reconstruction
-algorithm for the \texttt{Lehmer64} generator shown in
+algorithm for the \texttt{lehmer64()} generator shown in
 Fig.~\ref{fig:prngs}. When clocked, it outputs the top 64 bits of a geometric
@@
     """
-    Produce the internal state of the generator given three consecutive outputs.
+    Produce the internal state of the generator given three consecutive outputs of lehmer64().
     16 multiplications, 1 division, 11 additions and 3 roundings only.
@@
 
-But enough fooling around. Let's go back to \textsf{PCG64}.
+But enough fooling around; let's go back to \textsf{PCG64}.
 

### 2020-06-01 article/main.tex
conclusion + numpy + stuff

@@
 
-\section{Introduction} %ce qu'on avait écrit pour le projet, peut-être pas incroyable...
+\section{Introduction}
 
 \epigraph{Any one who considers arithmetical methods of producing random digits is, of course, in a state of sin.}{John von Neumann, 1949}
-
-%\todo{Ca, c'est du blabla...}
 
@@
 Not all pseudo-random generators are of cryptographic strength. In some
-applications, it is not necessarily necessary \todo{lourd}: to be used in Monte-Carlo
+applications, they are simply not necessary: to be used in Monte-Carlo
 numerical simulations or generate random choices in games, a relaxed,
@@
 the simplest and fastest algorithm that passes the day's favourite test suite. A
-few selected ones are shown in Fig.~\ref{fig:prngs}. \texttt{Lehmer64} is a
+few selected ones are shown in Fig.~\ref{fig:prngs}. \texttt{lehmer64} is a
 truncated linear congruential generator, touted as ``the fastest PRNG that
-passes \texttt{BigCrush}''~\cite{LemireLehmer}. \texttt{Xorshift128} is a clever
+passes \texttt{BigCrush}''~\cite{LemireLehmer}. \texttt{xorshift128} is a clever
 implementation of a 128-bit LFSR with period $2^{128} - 1$ due to
 Marsaglia~\cite{Marsaglia03}, using only a few simple 32-bit
-operations. \texttt{Xorshift128+} is a improved version due to
+operations. \texttt{xorshift128+} is a improved version due to
 Vigna~\cite{Vigna17} that returns the sum of two consecutive outputs of a
@@
 128-bit internal state and produces 64 bits when clocked. It is the default
-pseudo-random number generator in the popular \textsf{NumPy} scientific
-computing package for \textsf{Python}.
+pseudo-random number generator in the popular \textsf{NumPy}~\cite{Numpy}
+scientific computing package for \textsf{Python}.
 
@@
 initialised using an external source of entropy. In this case, our prediction
-algorithm requires 3072 bits of pseudo-random stream ; it guesses between 51 and
+algorithm requires 4096 bits of pseudo-random stream ; it guesses between 51 and
 55 bits, then for each guess it solves an instance of CVP in dimension 4 (using
@@
 
-We implemented the algorithms in \textsf{C}, then asked the designer of the PCG
-family to send us ``challenge'' pseudo-random streams ; we ran our code and
-emailed back the seeds used to generate the challenge streams the next day.\todo{et on avait tout juste, enfin j'espère}
+We implemented the algorithms in \textsf{C}, then asked the designer of the
+\textsf{PCG} family to send us ``challenge'' pseudo-random streams ; we ran our
+code and emailed back the (correct) seeds used to generate the challenge streams the next
+day.
 
@@
 
-Knuth considered the problem of recovering the internal state of a truncated
-linear congruential generator $x_{i+1} = ax_i + c \bmod 2^k$ when $a$ and $c$
-are unknown~\cite{Knuth85}; he gave an algorithm exponential in the number of
-truncated bits.
-
-In 1989, J.
-Boyar studied further the problem \cite{Boyar1989}.
-She presented an algorithm which could predict a LCG where all the parameters (multiplier, increment and initial state) where unknown under the condition that the number of bit unrevealed was really small in comparison to the modulus.
-
-In 1988, A.
-Frieze \textit{et al} \cite{Frieze} improved the efficiency of reconstruction algorithms in simpler cases.
-They supposed that the multiplier $a$ and the modulus $2^k$ were known and used lattice-based technique to recover the LCG more efficiently.
-
-Here we will use techniques presented in \cite{Frieze} to reconstruct some Lehmer's generator (LCG where the increment $c$ is zero) \cite{Lehmer}.
+Knuth discussed whether \emph{truncated} linear congruential generators could be
+good stream ciphers; he therefore studied the problem of recovering the internal
+state of a truncated linear congruential generator
+$x_{i+1} = ax_i + c \bmod 2^k$ when $a$ and $c$ are unknown~\cite{Knuth85}; he
+gave an algorithm exponential in the number of truncated bits.
+
+Boyar studied further the problem~\cite{Boyar1989} and presented an algorithm
+which could predict a linear congruential generator when all the parameters
+(multiplier, increment, modulus and initial state) are unknown; she extended her
+idea to the case of truncated linear congruential generators, under the
+condition that the number of bit unrevealed is really small in comparison to the
+size of the modulus.
+
+Frieze \textit{et al}~\cite{Frieze} improved the efficiency of reconstruction
+algorithms in simpler cases. They supposed that the multiplier $a$ and the
+modulus $2^k$ were known and used lattice-based techniques to recover a
+truncated linear congruential generator more efficiently, with more truncated
+bits.  more efficiently.
+
+Later on, Joux and Stern extended this result to the case where the multiplier
+$a$ and the modulus $2^k$ are unknown, also using lattice
+techniques~\cite{JouxS98}.
 
@@
 \]
-Where \todo{il manque une lettre, je suppose que c'est a} is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
+Where the ``multiplier'' $a$ is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
 the generator. The increment $c$ may be specified by the user of the PRNG to
@@
 \emph{carry} $0 \leq \gamma \leq 1$ (resp. a \emph{borrow}
-$0 \leq 1 \leq \beta$ \todo{je pense pas que ce soit ça}) such that:
+$0 \leq \beta \leq 1$) such that:
 \begin{align}
@@
 modulo $2^k$; therefore the first $n$ terms of the sequence $\LCG_{k}(x, 0)$
-give the coordinates of a point of this lattice.
+give the coordinates of a vector in this lattice.
 
@@
 
-If we can prove that the right side of the inequality is smaller than the first
+Note that $\vvvert H^{-1} \vvvert \times \vvvert H \vvvert$ is the condition
+number of the matrix $H$. Lattice reduction has the side effect of reducing the
+condition number, therefore it makes sense to use an LLL-reduced basis. If we
+can prove that the right side of the inequality is smaller than the first
 minimum of the lattice, then we would have proved that $RH$ is indeed the
@@
 
-But enough fooling around; let's go back to \textsf{PCG64}.
+%\noindent But enough fooling around; let's go back to \textsf{PCG64}.
 
@@
 
-\section{Predicting PCG64 with Unknown Increment}
+\section{State Reconstruction for \textsf{PCG64} With Secret Increment}
 \label{sec:Cunknown}
@@
   \Function{ReconstructState}{$\Delta S_0, r_0, \dots, r_k, X_0, \dots, X_k$}
-  \State \todo{check indices nabla}
   \For{$i = 0, 1, \dots, k$} \Comment{Setup}
@@
 with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz} (``\emph{Cascade
-  Lake}''). The actual machine is the \verb|[----REDACTED FOR PEER REVIEW----]|
-computer located at the \verb|[----ALSO REDACTED----]| computation center. Note
-that on this particular parallel computer, running the algorithm with $\ell=13$
-would take 10 hours using the same amount of resources, so the whole procedure
-\emph{is} practical, even in the absolute worst case.
+  Lake}''). The actual machine is the \verb|[----REDACTED----]| computer located
+at the \verb|[----ALSO REDACTED----]| computation center. Note that on this
+particular parallel computer, running the algorithm with $\ell=13$ would take 10
+hours using the same amount of resources, so the whole procedure \emph{is}
+practical, even in the absolute worst case.
 
@@
 \section{Conclusion}
-\todo{y'a pas de conclusion}
+
+We have presented a practical state reconstruction algorithm for the
+\textsf{PCG64} conventional pseudo-random number generator, the default in the
+\texttt{NumPy} library. In the worst case, we recovers all the secret
+information using 512 consecutive output bytes, using 2.3 CPU years of
+computation. We have executed the algorithm in practice using a large parallel
+computer. The \textsf{PCG64} generator is fast and not intended for
+cryptographic purposes; we have shown that, in practice, this comes at the price
+of strong pseudo-randomness. It should absolutely not be used when
+unpredictability of the random numbers is required, for fear of practical
+attacks.
+
+On the other hand, our results do not mean that \textsf{PCG64} should be
+deprecated for scientific computing. But they do mean that its output has
+detectable properties. Whether these properties may affect the results of
+Monte-Carlo numerical simulations is another matter entirely.
+
 % pas dans la version anonyme.
@@
 \bibliography{biblio}
-
-%\appendix
 

### 2020-06-01 article/main.tex
détails

@@
 
-\keywords{\todo{keyword}}
+\keywords{Pseudo-random number generator, guess-and-determine attack, truncated congruential generator, euclidean lattices, closest vector problem, practical attack}
 
@@
 
+\epigraph{Any one who considers arithmetical methods of producing random digits is, of course, in a state of sin.}{John von Neumann, 1949}
+
 \section{Introduction}
-
-\epigraph{Any one who considers arithmetical methods of producing random digits is, of course, in a state of sin.}{John von Neumann, 1949}
 
@@
 Not all pseudo-random generators are of cryptographic strength. In some
-applications, they are simply not necessary: to be used in Monte-Carlo
-numerical simulations or generate random choices in games, a relaxed,
-non-cryptographic notion of pseudo-randomness may be sufficient. This allows for
-faster algorithms. For instance, \textsf{python} standard library's
-\texttt{random} module uses the Mersenne Twister~\cite{MatsumotoN98}. The
-\textsf{C} library that comes along \texttt{gcc} (the \texttt{glibc}) uses a
-(poor) truncated linear congruential generator by default to implement the
-\texttt{rand} function.
+applications, it is simply not necessary: to be used in Monte-Carlo numerical
+simulations or generate random choices in games, a relaxed, non-cryptographic
+notion of pseudo-randomness may be sufficient. This allows for faster
+algorithms. For instance, \textsf{python} standard library's \texttt{random}
+module uses the Mersenne Twister~\cite{MatsumotoN98}. The \textsf{C} library
+that comes along \texttt{gcc} (the \texttt{glibc}) uses a (poor) truncated
+linear congruential generator by default to implement the \texttt{rand}
+function.
 
@@
 battery of randomness tests for RNGs was suggested by Knuth in the 1969 first
-edition of The Art of Computer Programming. In 1996, Knuth's tests were then
+edition of \emph{The Art of Computer Programming}. In 1996, Knuth's tests were then
 supplanted by Marsaglia's \texttt{Diehard} tests. In 2007, L'Ecuyer proposed
@@
 
-In most cases, it is fairly easy to see that a specific non-cryptographic PRG
-does not meet the cryptographic notion of pseudo-randomness, and there are few
+In most cases, it is fairly easy to see that a given conventional PRG does not
+meet the cryptographic notion of pseudo-randomness, and there are few
 exceptions. Most are fairly easy to predict, meaning that after having observed
 a prefix of the output, it is easy to produce the next ``pseudo-random''
-bits. This makes a good source of exercises for cryptography students.
+bits. This makes a good source of exercises for cryptology students.
 
@@
 reconstruction algorithm is practical and we have executed it in practice. It
-follows that predicting the output of the \textsf{PCG} should not be considered
-challenging anymore.
+follows that predicting the output of the \textsf{PCG} should be considered
+practically feasible.
 
@@
 
-We implemented the algorithms in \textsf{C}, then asked the designer of the
-\textsf{PCG} family to send us ``challenge'' pseudo-random streams ; we ran our
-code and emailed back the (correct) seeds used to generate the challenge streams the next
+We implemented our algorithms, then asked the designer of the \textsf{PCG}
+family to send us ``challenge'' pseudo-random streams ; we ran our code and
+emailed back the (correct) seeds used to generate the challenge streams the next
 day.
@@
 modulus $2^k$ were known and used lattice-based techniques to recover a
-truncated linear congruential generator more efficiently, with more truncated
-bits.  more efficiently.
+truncated linear congruential generator with more truncated bits.
 
@@
 the approximation we have. This means solving instances of the well-known
-\textsf{Closest Vector Problem} (CVP), a fundamental hard problem on
+\textsc{Closest Vector Problem} (CVP), a fundamental hard problem on
 lattices. It is NP-hard, and all known algorithms are exponential in the
@@
 shortest vector of the lattice spanned by $G_{n, 128}$ for each successive $n$
-until the condition holds. The \textsf{Shortest Vector Problem} is another
+until the condition holds. The \textsc{Shortest Vector Problem} (SVP) is another
 well-known lattice NP-hard problem; we used the (almost) off-the-shelf

### 2020-06-01 article/main.tex
Abstract

@@
 
-\newcommand{\todo}[1]{\textcolor{red}{TODO:[#1]}}
-
-\title[Predicting the PCG Pseudo-Random Number Generator in Practice]{Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice}
-
-\author{Charles Bouillaguet\inst{1} \and Julia Sauvage\inst{2} \and Florette Martinez\inst{3}}
+%\newcommand{\todo}[1]{\textcolor{red}{TODO:[#1]}}
+
+\title{Predicting the PCG Pseudo-Random Number Generator In Practice}
+
+\author{Charles Bouillaguet\inst{1} \and Florette Martinez\inst{3} \and Julia Sauvage\inst{2}}
 
@@
 \begin{abstract}
-  \todo{blabla}
+  The Permuted Congruential Generators (\textsf{PCG}) are popular conventional
+  (non-cryptographic) pseudo-random generators designed in 2014. They are used
+  by default in the \texttt{NumPy} scientific computing package. Even though
+  they are not of cryptographic strength, their designer stated that predicting
+  their output should be nevertheless be "challenging".
+
+  In this article, we present a practical algorithm that recovers all the hidden
+  parameters and reconstructs the successive internal states of the
+  generator. This enables us to predict the next ``random'' numbers, and output
+  the seeds of the generator. We have successfully executed the reconstruction
+  algorithm using 512 bytes of challenge input; in the worst case, the process
+  takes 20 000 CPU hours.
+
+  This reconstruction algorithm makes use of cryptanalytic techniques, both
+  symmetric and lattice-based. In particular, the most computationally expensive
+  part is a guess-and-determine procedure that solves about $2^{52}$ instances
+  of the \textsc{Closest Vector Problem} on a very small lattice.
 \end{abstract}

### 2020-06-01 Cknown/main.c
more stuff before closing up

@@ -19,10 +19,7 @@ void result_found(pcg128_t S, double start)
 
 
 int main()
-{
-    
-    /*  INITIALISATION DES PARAMETRES  */
-    
+{    
     init_var_globales();
 	
     /********** INPUT ***********/

### 2020-06-01 Cknown/success_proba.c

@@ -5,6 +5,9 @@
 
 #include "fonctions.h"
 
+/* auxiliary script to determine the success probability of the reconstruction procedure
+   with small ell */
+
 FILE *f; 
 
 bool run_test()
@@ -29,41 +32,21 @@ bool run_test()
 	for (int i = 0; i < nbiter; i++)
 		rot[i] = (int) (S[i] >> 122);
 	
-      	// for (int i = 0; i < nbiter; i++)
-      	// 	printf("X[%d] = %016llx\n", i, X[i]);
-      	// printf("W0 = %llx\n", W0);
-
-
 	struct task_t task;
 	setup_task(W0, X, &task);
+	refresh_task(rot, &task);
 
-	//int rot[nbiter] = {0, 0, 0};
-	
-	//for (int r = 0 ; r < 1 << (nbiter*6) ; r++) {
-	//	/* bump rotations */
-	//    rot[0] = (rot[0] + 1) % 64;
-	//    int i = 0;
-	//    while (rot[i] == 0 && i < nbiter) {
-	//	i++;
-	//	rot[i] = (rot[i] + 1) % 64;
-	//    }
-	//    if (i > 0)
-	    	refresh_task(rot, &task);
-
-	    pcg128_t S_target[nbiter];
-	    if (solve(S_target, rot, &task)) {
+	pcg128_t S_target[nbiter];
+	if (solve(S_target, rot, &task)) {
 	    	assert(S_target[0] == S[0]);
 		return true;
-	    }
-	//}
+	}
 	return false;
 }
 
 
 int main()
 {
-    
-    /*  INITIALISATION DES PARAMETRES  */    
     init_var_globales();
     f = fopen("/dev/urandom", "r");
     if (f == NULL) {

### 2020-06-01 Cunknown/main.c

@@ -191,13 +191,6 @@ void master(const u64 (*X)[nboutput])
 	    break;
 	}
 
-#if 0
-	/* DEBUG */
-	int W_0 = 0x01d0;
-	int W_c = 0x035b;
-	task_ptr = (W_0 << (known_low - 1)) + (W_c - 1) / 2;
-#endif
-
 	FILE *journal_file = fopen(journal_filename, "a");
 	if (journal_file == NULL)
 		err(1, "Impossible to open journal file %s for append\n", journal_filename);	
@@ -306,6 +299,7 @@ int main(int argc, char **argv)
 
 	init_var_globales();
 	
+	// challenge input from M. O'Neil
 	u64 X[nboutput];
 	X[ 0] = 0x46b9d66bb1e9fbf0;
 	X[ 1] = 0xc0b7382496f0e363;

### 2020-06-01 Cunknown/simpler_matrix.py

@@ -1,0 +1,91 @@
+a = 47026247687942121848144207491837523525
+
+def getG(n, mod):
+    """
+    Naive basis
+    """
+    G = matrix(ZZ, n, n)
+    for i in range(0, n):
+        G[0, i] = pow(a, i, mod)
+    for i in range(1, n):
+        G[i, i] = mod
+    return G
+
+G = getG(64, 2**128)
+
+L = [
+ 12144252875850345479015002205241987363,
+ -15348615718011256064636590606166444173,
+ -9084835538682141851035496395127502261,
+ 17234466561204322281445746373216450637,
+ -30709520135048357458223997075862325767,
+ 13273365585194684751192188342361392061,
+ -16657707705878774883995565342082094357,
+ 10551835762264010302610847026218529611,
+ 9741843352863159575387669322076694498,
+ 30401545415842458720824638334153655996,
+ -21301422832830108630138234390597720063,
+ 12943082679906251720732751543854447750,
+ -26608736029397995753447945248664058244,
+ 16504008779817367542456313016700742074,
+ -10623665749400971081452814639205027657,
+ -42221955640534425567937724106639808954,
+ 30891705174976345231868365741077473210,
+ 36583589311333508006721438259000555949,
+ -7050528015382194186611161130462441554,
+ 29023698504320799821125567462694089388,
+ -28780651977143078032165325525310692006,
+ 80641828999898464226011186926508194344,
+ -56574857747725737575017746952734912354,
+ 69356676566431307205578324735925446877,
+ -33869683024546744201772555262569839986,
+ -26565235055559932529672543662504791838,
+ 12570516463211853078121725511001272792,
+ -3545191593186327315454855798676360051,
+ 34483716896778783922111688009948767715,
+ -20939456075167816323854415536161525816,
+ 37054527231762903007893323872329597868,
+ -65113113872795261158497215763575544978,
+ -27970657614360739023057572629108089923,
+ -9380443740630786183893558576010683559,
+ -31684372116213316987059381973832378607,
+ 6514548643534369904269689195712545814,
+ 14847729039548998339264458667105218977,
+ 20177225057138896843222830352311122871,
+ -10200019003678439599482650259589205069,
+ 10164570837071212387748750871984064078,
+ -16485770795775188591072402663394972722,
+ -10182226575367101730098853924492492198,
+ -21298248172336954279037664114356234573,
+ -2151417390168764172560507076988402637,
[…]

### 2020-06-01 Cunknown/step2.py

@@ -57,24 +57,10 @@ Greduite1 = getGreduite(nbiter - 1, 2**64)
 Greduite2 = getGreduite(nboutput - 1, 2**(128 - known_low))
 
 
-#### Récupération des données ####
-#def recupDonnees():
-#   listfiles = os.listdir('results')
-#   print(listfiles)
-#   listDonnees = []
-#   for i in range(listfiles):
-#       listDonnees.append[]
-#       f = open(listfiles[i], "r")
-#       while(1):
-#           s = f.readline()
-            
-
 
 def sortiesGenerateur():#OK !
     c = random.randrange(N) | 1
     S = [random.randrange(N)]
-    #c=6364136223846793005 * 2^64 + 1442695040888963407#increment par defaut de pcg (connu)
-    #S=[8487854484825256858 + 11929896274893053136 * 2^64]
     for i in range (nboutput - 1):
         S.append((S[i] * a + c) % N)
     X = []
@@ -246,6 +232,7 @@ def full_DS(X, W0, WC, rots):
 
 
 if __name__ == '__main__':
+    # challenge input from M. O'Neil
     X = [0] * nboutput
     X[ 0] = 0x46b9d66bb1e9fbf0;
     X[ 1] = 0xc0b7382496f0e363;

### 2020-06-01 logs/parser.py

@@ -1,0 +1,25 @@
+import re
+
+pattern = re.compile(r' \[(?P<time>[0-9]+\.[0-9])s\]')
+
+match = pattern.search('Done task 53a (W_0=0001 / W_c=0275) [41.6s]')
+assert match
+assert match['time'] == '41.6'
+
+T = []
+with open("breaker1352922.out") as f:
+    for line in f:
+        match = pattern.search(line) 
+        if match:
+            T.append(float(match.group('time')))
+
+d = {}
+for t in T:
+    x = round(10 * t)
+    if x not in d:
+        d[x] = 1
+    else:
+        d[x] += 1
+
+for t in d:
+    print("{}: {}".format(t, d[t]))

### 2020-06-01 logs/render.py

@@ -1,0 +1,47 @@
+from matplotlib import pyplot as plt
+
+d = {41.4: 8405, 
+     41.5: 7858, 
+     41.6: 574570, 
+     41.7: 92711, 
+     41.8: 349415, 
+     41.9: 11835, 
+     42.0: 2763, 
+     42.1: 8848, 
+     42.2: 644201, 
+     42.3: 4384, 
+     42.4: 13976, 
+     42.5: 133008, 
+     42.6: 241950, 
+     42.7: 1324, 
+     43.4: 491, 
+     54.7: 364, 
+     54.8: 531, 
+     43.6: 50, 
+     54.9: 54, 
+     43.5: 403, 
+     43.3: 5, 
+     55.0: 2, 
+     43.7: 2, 
+     40.0: 1, 
+     43.8: 1
+}
+
+X = sorted(d.keys())
+Y = [d[k] for k in X]
+
+fig, ax = plt.subplots(1, 1, figsize=(16, 10))
+ax.bar(X, Y, color='violet', width=0.08)
+    
+ax.set_yscale('log', basey=2)
+#ax.axis(xmin=0, xmax=npasses+1, ymin=ymin, ymax=N*1.15)
+#ax.set_yticks(yticks)
+ax.set_xticks([i for i in range(39, 57)])
+fig.set_tight_layout(True)
+#ax.legend(loc='best')
+ax.set_ylabel('# Tasks')
+ax.set_xlabel('Time per Task (s)')
+#ax.yaxis.set_ticks_position('both')
+
+plt.show()
+#fig.savefig(outfile)

### 2020-06-01 minL.py

@@ -1,0 +1,91 @@
+import fpylll
+from math import sqrt
+
+from g6k import Siever
+from g6k.algorithms.pump import pump
+from g6k.utils.stats import dummy_tracer
+
+a = 47026247687942121848144207491837523525
+
+def getG(n, mod):
+    G = fpylll.IntegerMatrix(n, n)
+    for i in range(0, n):
+        G[0, i] = pow(a, i, mod)
+    for i in range(1, n):
+        G[i, i] = mod
+    fpylll.LLL.reduction(G) # in place
+    return G
+
+def svp_fpylll(G):
+    """
+    Returns the shortest vector of the lattice spanned by the rows of G.
+    """
+    x = fpylll.SVP.shortest_vector(G, method = "fast", flags=fpylll.SVP.VERBOSE)
+    return x
+
+
+def svp_g6k(G):
+    """
+    Returns the shortest vector of the lattice spanned by the rows of G.
+    """
+    g6k = Siever(G)
+    pump(g6k, dummy_tracer, 0, G.nrows, 0)
+    return G[0]
+
+def norm(v):
+    return sqrt(sum(x**2 for x in v))
+    
+def show_targets(ell):
+    for n in range(30, 70):
+        target = 2 * sqrt(n) * (2**(123-ell) - 1)
+        print("{}: {}".format(n, target))
+
+def min_m_CVP(ell):
+    for n in range(60, 64, 1):
+        G = getG(n, 2**(128-ell))
+        shortest = svp_g6k(G)
+        lambda1 = norm(shortest)
+        target = 2 * sqrt(n) * (2**(123 - ell) - 1)
+        print("{}: {} vs {}".format(n, target, lambda1))
+        if target <= lambda1:
+            break
+    print("SUCCESS")
+    return n
+
+# k = 128
+## n = 42 -> 7.10958185672e+37
+## n = 48 -> 9.40694696471e+37
+## n = 54 -> 1.23309364376e+38
+## n = 60 -> 1.48734754252e+38
+## n = 
+
+if __name__ == '__main__':
+    show_targets(0)
+    #min_m_CVP(0)
+
+    n = 64
+    G = getG(n, 2**128)
+    print(sum([norm(x) for x in G])/n)
+    #baseline = norm(G[0])
+    g6k = Siever(G)
+    #print("baseline = {}".format(baseline))
+    for i in range(64):
+        pump(g6k, dummy_tracer, 0, n, 0)
+        print([norm(x) for x in G[0:8]])
+    
+    #print(G[0])
+
[…]

### 2020-06-05 article/main.tex
version preprint

@@
 % !TeX spellcheck = en
-\documentclass[submission,svgnames,journal=tosc]{iacrtrans}
+\documentclass[preprint,svgnames,journal=tosc]{iacrtrans}
 
@@
 
-\author{Charles Bouillaguet\inst{1} \and Florette Martinez\inst{3} \and Julia Sauvage\inst{2}}
-
-\institute{% 
-University of Lille, France \\ 
+\author{Charles Bouillaguet\inst{1} \and Florette Martinez\inst{2} \and Julia Sauvage\inst{3}}
+
+\institute{% ordre alphabétique
+Univ. Lille, CNRS, Centrale Lille, UMR 9189 - CRIStAL - Centre de Recherche en Informatique Signal et Automatique de Lille, F-59000 Lille, France \\ 
 \email{charles.bouillaguet@univ-lille.fr}
 \and 
-Sorbonne University \\
+Sorbonne Université, CNRS, LIP6, F-75005 Paris, France \\
+\email{florette.martinez@lip6.fr}
+\and 
+Sorbonne Université \\
 \email{julia.sauvage@etu.upmc.fr}
-\and 
-LIP6, CNRS, SU ? \\
-\email{florette.martinez@lip6.fr}
-
 }
@@
 as well as online at:
-\[
-  % \url{https://github.com/cbouilla/pcg/}
-  \verb|[----REDACTED FOR PEER REVIEW----]|
-\]
+\begin{center}
+  \url{https://github.com/cbouilla/pcg/}
+  %\verb|[----REDACTED FOR PEER REVIEW----]|
+\end{center}
 The designer of \textsf{PCG} was kind enough to send us two sets challenge
@@
 with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz} (``\emph{Cascade
-  Lake}''). The actual machine is the \verb|[----REDACTED----]| computer located
-at the \verb|[----ALSO REDACTED----]| computation center. Note that on this
-particular parallel computer, running the algorithm with $\ell=13$ would take 10
-hours using the same amount of resources, so the whole procedure \emph{is}
-practical, even in the absolute worst case.
+  Lake}''). The actual machine is the \texttt{jean-zay} computer located at the
+IDRIS national computation center. Note that on this particular parallel
+computer, running the algorithm with $\ell=13$ would take 10 hours using the
+same amount of resources, so the whole procedure \emph{is} practical, even in
+the absolute worst case.
 
@@
 % pas dans la version anonyme.
-% \paragraph{Acknowledgements.} The authors thank Damien Vergnaud and Vincent
-% Zucca for helpful discussions. We also thank Léo Ducas for helping with the
-% \texttt{G6K} library.
+\paragraph{Acknowledgements.} The authors thank Thomas Espitau, Damien Vergnaud
+and Vincent Zucca for helpful discussions. We also thank Léo Ducas for helping
+us with the \texttt{G6K} library and Melissa O'Neil for kindly indulging our
+request for challenge inputs.
+
+This work was granted access to the HPC resources of IDRIS under the allocation
+2019-A0060610749 made by GENCI.
 

### 2020-06-05 Cunknown/step3.py

@@ -59,6 +59,8 @@ def sortiesGenerateur(S0, c, k):
 
 
 if __name__ == '__main__':
+
+    # challenge input from M. O'Neil
     X = [0] * 48
     X[ 0] = 0x46b9d66bb1e9fbf0;
     X[ 1] = 0xc0b7382496f0e363;

### 2020-06-05 challenges/pcg64s-challenge.c

@@ -17,7 +17,6 @@ int main()
 		perror("Something went wrong when reading /dev/urandom");
 		exit(EXIT_FAILURE);
 	}
-	seed = (((pcg128_t) 0x5b3d418b8b51364b) << 64) + ((pcg128_t) 0x74eed710223e763f);
 	printf("Seed : %016" PRIx64 " %016" PRIx64 "\n\n", (uint64_t) (seed >> 64), (uint64_t) seed);
 
 	pcg64s_random_t rng;

### 2020-06-05 good_big_matrix.py

@@ -105,3 +105,6 @@ for i in range(64):
 
 #target = [G[0, i] + random.randrange(-2**120, 2**120) for i in range(64)]
 print(log(GG[0].norm(), 2))
+
+# SVP = 1h20 pour GG (approx.  HKZ-réduite)
+#     = ???? pour G (seulement LLL-réduite)

### 2020-08-19 article/main.tex
remarques

@@
 % !TeX spellcheck = en
-\documentclass[preprint,svgnames,journal=tosc]{iacrtrans}
+\documentclass[submission,svgnames,journal=tosc]{iacrtrans}
 
@@
 In the realm of non-cryptographic random generators, a PRG is deemed ``good
-enough'' it is passes \emph{some} efficient statistical tests --- whereas the
+enough'' if it passes \emph{some} efficient statistical tests --- whereas the
 cryptographic notion of pseudo-randomness asks that it passes \emph{any}
@@
 
-Our algorithm reconstruct the internal state using a ``guess-and-determine''
+Our algorithm reconstructs the internal state using a ``guess-and-determine''
 approach: some bits of the internal state are guessed ; assuming the guesses are
@@
 guesses early on ; then candidate internal states are computed and fully
-tested. The problem actually come in two distinct flavors.
+tested. The problem actually comes in two distinct flavors.
 
@@
 modulo $2^k$, therefore $(U_0, \dots, U_{n-1}) \in \mathcal{L}$. Let
-$T_i = U_i[k-\ell:k]$ denote the top $\ell$ bits of $U_i$, and let $N$ denote an
+$T_i = U_i[k-t:k]$ denote the top $t$ bits of $U_i$, and let $N$ denote an
 arbitrary ``noise vector'' such that $N_i \in \{-1, 0, 1\}$. Finally, set
-$\widetilde{T}_i = T_i + N_i \bmod 2^{\ell}$.  We will be facing the following
+$\widetilde{T}_i = T_i + N_i \bmod 2^{t}$.  We will be facing the following
 problem (``reconstructing noisy truncated geometric series'') several times:
 \begin{description}
-\item[INPUT]  $\widetilde{T} = (\widetilde{T}_0, \dots, \widetilde{T}_{n-1}) \in \left(\mathbb{Z}_{2^\ell}\right)^n$, a (noisy) version of $U$ truncated to the top $\ell$ bits.
+\item[INPUT]  $\widetilde{T} = (\widetilde{T}_0, \dots, \widetilde{T}_{n-1}) \in \left(\mathbb{Z}_{2^t}\right)^n$, a (noisy) version of $U$ truncated to the top $t$ bits.
 \item[OUTPUT] $U_0 \bmod 2^k$, the first term of the (non-truncated) geometric sequence.
@@
 and~\ref{sec:Cunknown}. The rest of this section discusses algorithmic tools to
-solve these problems. We first claim that $2^{k-\ell} \widetilde{T}$ is
-``close'' to a point of the lattice $\mathcal{L}_{n,k}$.
+solve these problems. We first claim that $2^{k-t} \widetilde{T}$ is
+``close'' to a point of the lattice $\mathcal{L}$.
 
@@
   There exists $U' \in \mathcal{L}$ such that $U'_i \equiv U_i \bmod 2^k$
-  and $\left\lVert U' -2^{k-\ell} \widetilde{T} \right\rVert \leq 2 \sqrt{n} 2^{k-\ell}$.
+  and $\left\lVert U' -2^{k-t} \widetilde{T} \right\rVert \leq 2 \sqrt{n} 2^{k-t}$.
 \end{lemma}
@@
     \begin{align*}
-      \left| 2^{k-\ell}\widetilde{T}_i-U_i \right| &= \left| 2^{k-\ell}(T_i + N_i)-U_i \right| 
-                                                     = \left| 2^{k-\ell}N_i - U_i[0,k-\ell] \right| \\
-                                                   &\leq \left| 2^{k-\ell}N_i\rvert + \lvert U_i[0,k-\ell] \right| 
+      \left| 2^{k-t}\widetilde{T}_i-U_i \right| &= \left| 2^{k-t}(T_i + N_i)-U_i \right| 
+                                                     = \left| 2^{k-t}N_i - U_i[0,k-t] \right| \\
+                                                   &\leq \left| 2^{k-t}N_i\rvert + \lvert U_i[0,k-t] \right| 
                                                    % &\leq 2^{k-\ell} + 2^{k-\ell} - 1\\
-      \leq 2^{k-\ell +1 }.
+      \leq 2^{k-t +1 }.
     \end{align*}
@@
     \begin{itemize}
-    \item Either $T_i = 0$ and $N_i = -1$, which leads to $\widetilde{T}_i = 2^\ell-1$. 
-      In this case, we have $U_i = U_i[0,k-\ell]$ and we set $U'_i = U_i + 2^{k}$ (note that this amount to adding a lattice vector to $U'$, so $U'$ stays in the lattice). We have:
+    \item Either $T_i = 0$ and $N_i = -1$, which leads to $\widetilde{T}_i = 2^t-1$. 
+      In this case, we have $U_i = U_i[0,k-t]$ and we set $U'_i = U_i + 2^{k}$ (note that this amount to adding a lattice vector to $U'$, so $U'$ stays in the lattice). We have:
       \[
-        \left| 2^{k-\ell}\widetilde{T}_i-(U_i + 2^k) \right| = %\left| 2^{k-\ell}(2^\ell -1) -(U_i + 2^k) \right| =
-        \left| 2^{k-\ell} + U_i[0,k-\ell]\right| \leq 2^{k-\ell + 1}. 
+        \left| 2^{k-t}\widetilde{T}_i-(U_i + 2^k) \right| = %\left| 2^{k-t}(2^t -1) -(U_i + 2^k) \right| =
+        \left| 2^{k-t} + U_i[0,k-t]\right| \leq 2^{k-t + 1}. 
       \]
 
-    \item Or $T_i = 2^{\ell}-1$ and $N_i = +1$, which leads to
+    \item Or $T_i = 2^{t}-1$ and $N_i = +1$, which leads to
       $\widetilde{T}_i = 0$.  This implies that
-      $U_i= 2^k-2^{k-\ell} + U_i[0,k-\ell]$; we set $U'_i = U_i - 2^k$ (again,
+      $U_i= 2^k-2^{k-t} + U_i[0,k-t]$; we set $U'_i = U_i - 2^k$ (again,
       with this modification $U'$ stays in the lattice), and we find:
       \[
-        \left| 2^{k-\ell}\widetilde{T}_i-(U_i - 2^k) \right| %=\left| 0 -(U_i - 2^k) \right|
-        = \left| 2^k-2^{k-\ell} + U_i[0,k-\ell] - 2^k\right| \leq 2^{k-\ell + 1}.
+        \left| 2^{k-t}\widetilde{T}_i-(U_i - 2^k) \right| %=\left| 0 -(U_i - 2^k) \right|
+        = \left| 2^k-2^{k-t} + U_i[0,k-t] - 2^k\right| \leq 2^{k-t + 1}.
       \]
@@
   In the end, we have
-  $\left\| U' -2^{k-\ell} \widetilde{T} \right\| \leq 2 \sqrt{n} 2^{k-\ell}$,
+  $\left\| U' -2^{k-t} \widetilde{T} \right\| \leq 2 \sqrt{n} 2^{k-t}$,
   and $U' \equiv U \mod 2^k $.
@@
 input vector $x$. Using the same notations as above, we want to know if
-$\CVP\left(\mathcal{L}, 2^{k-\ell}\widetilde{T}\right)$ is indeed $U'$. This
-will necessarily be the case when $\|2^{k-\ell}\widetilde{T} - U'\|$ is smaller
+$\CVP\left(\mathcal{L}, 2^{k-t}\widetilde{T}\right)$ is indeed $U'$. This
+will necessarily be the case when $\|2^{k-t}\widetilde{T} - U'\|$ is smaller
 than the length of the shortest non-zero vector of $\mathcal{L}$ --- this quantity,
@@
 \begin{align*}
-|\CVP(\mathcal{L}, 2^{k-\ell}\widetilde{T}) - U'| %&= \left| \CVP\left( \mathcal{L}, 2^{k-\ell}\widetilde{T} \right) \riht| + \left|2^{k-\ell}\widetilde{T} - U'\right| %-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
-                             & \leq \left| \CVP\left(\mathcal{L}, 2^{k-\ell}\widetilde{T} \right) - 2^{k-\ell}\widetilde{T}\right | + \left|2^{k-\ell}\widetilde{T} - U'\right|.
+|\CVP(\mathcal{L}, 2^{k-t}\widetilde{T}) - U'| %&= \left| \CVP\left( \mathcal{L}, 2^{k-\ell}\widetilde{T} \right) \riht| + \left|2^{k-\ell}\widetilde{T} - U'\right| %-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
+                             & \leq \left| \CVP\left(\mathcal{L}, 2^{k-t}\widetilde{T} \right) - 2^{k-t}\widetilde{T}\right | + \left|2^{k-t}\widetilde{T} - U'\right|.
 \end{align*}
@@
 \[
-  |\CVP(\mathcal{L}, 2^{k-\ell}\widetilde{T}) - U'| \leq 2 \left|2^{k-\ell} \widetilde{T} - U' \right| \leq 4 \sqrt{n} 2^{k-\ell}.
+  |\CVP(\mathcal{L}, 2^{k-t}\widetilde{T}) - U'| \leq 2 \left|2^{k-t} \widetilde{T} - U' \right| \leq 4 \sqrt{n} 2^{k-t}.
 \]
@@
 minimum of the lattice $\lambda_1(\mathcal{L})$, then we would have proved that
-$\CVP(\mathcal{L}, 2^{k-\ell}\widetilde{T})$ indeed reveals $U_0 \bmod 2^k$.
+$\CVP(\mathcal{L}, 2^{k-t}\widetilde{T})$ indeed reveals $U_0 \bmod 2^k$.
 
@@
 the) most-significant $6$ bits of successive elements of the sequence. Therefore
-we have $k=128$ and $\ell=6$, and we wish to determine the required number of
+we have $k=128$ and $t=6$, and we wish to determine the required number of
 samples, \textit{i.e.} the value of $n$. This means finding the values of $n$
@@
 
-Starting from $n=\lceil 122/6\rceil\rceil$, we computed the length of the
+Starting from $n=\lceil 122/6\rceil$, we computed the length of the
 shortest vector of the lattice spanned by $G_{n, 128}$ for each successive $n$
@@
 (subtracting $2^{128}$ to $U_i$ if $U_i > 2^{127}$). It follows that when
-$n\geq 63, k=128$ and $\ell=6$, any CVP oracle will return a vector congruent to
+$n\geq 63, k=128$ and $t=6$, any CVP oracle will return a vector congruent to
 the original $U$ when given $\widetilde{T}$.
@@
 \]
-Set $S = 2^{k-\ell} \widetilde{T}H^{-1}$; as $2^{k-\ell}\widetilde{T}$ is not
-\textit{a priori} an element of the lattice, $S$ need not be an integer
+Set $S = 2^{k-t} \widetilde{T}H^{-1}$; as $2^{k-t}\widetilde{T}$ does not belong
+\textit{a priori} to the lattice, $S$ does not have to be an integer
 vector. Let then $R$ denote the rounding of $S$, \textit{i.e.}
@@
 lattice. Under the right conditions, it will be the vector of the lattice
-closest to $2^{k-\ell} \widetilde{T}$. Indeed:
+closest to $2^{k-t} \widetilde{T}$. Indeed:
 \begin{align*}
-\lVert U' - R H \rVert &= \left\| U' - \left(R - 2^{k-\ell}\widetilde{T}H^{-1} + 2^{k-\ell}\widetilde{T}H^{-1}\right)H \right\| \\
-&= \left\| U' - 2^{k-\ell}\widetilde{T} - \left(R - 2^{k-\ell} \widetilde{T}H^{-1}\right )H \right\| \\
-&\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\| R - 2^{k-\ell}\widetilde{T}H^{-1} \right\| \times \vvvert H \vvvert.
+\lVert U' - R H \rVert &= \left\| U' - \left(R - 2^{k-t}\widetilde{T}H^{-1} + 2^{k-t}\widetilde{T}H^{-1}\right)H \right\| \\
+&= \left\| U' - 2^{k-t}\widetilde{T} - \left(R - 2^{k-t} \widetilde{T}H^{-1}\right )H \right\| \\
+&\leq \left\| U' - 2^{k-t}\widetilde{T} \right\| + \left\| R - 2^{k-t}\widetilde{T}H^{-1} \right\| \times \vvvert H \vvvert.
 \end{align*}
 By definition, $R$ is the closest integer vector to
-$2^{k-\ell}\widetilde{T}H^{-1}$. But as $U'$ is an element of the lattice,
-$U'H^{-1}$ is an integer vector. Thus $R-2^{k-\ell}\widetilde{T}G^{-1}$ is
-shorter than $U'G^{-1}-2^{k-\ell}\widetilde{T}G^{-1}$. Hence :
+$2^{k-t}\widetilde{T}H^{-1}$. But as $U'$ is an element of the lattice,
+$U'H^{-1}$ is an integer vector. Thus $R-2^{k-t}\widetilde{T}H^{-1}$ is
+shorter than $U'H^{-1}-2^{k-t}\widetilde{T}H^{-1}$. Hence :
 \begin{align*}
-  \left\| U' - RH \right\| &\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\|R - 2^{k-\ell}\widetilde{T}H^{-1}\right\| \times \vvvert H \vvvert \\	
-                        &\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\|U'H^{-1}-2^{k-\ell}\widetilde{T}H^{-1}\right\| \times \vvvert H \vvvert \\	
-                        &\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| + \left\|U' - 2^{k-\ell}\widetilde{T}\right\| \times \left\vvvert H^{-1} \right\vvvert  \times \vvvert H \vvvert \\
-                        & 	\leq \left\| U' - 2^{k-\ell}\widetilde{T} \right\| \times \left(1 + \vvvert H^{-1} \vvvert  \times \vvvert H \vvvert \right).
+  \left\| U' - RH \right\| &\leq \left\| U' - 2^{k-t}\widetilde{T} \right\| + \left\|R - 2^{k-t}\widetilde{T}H^{-1}\right\| \times \vvvert H \vvvert \\	
+                        &\leq \left\| U' - 2^{k-t}\widetilde{T} \right\| + \left\|U'H^{-1}-2^{k-t}\widetilde{T}H^{-1}\right\| \times \vvvert H \vvvert \\	
+                        &\leq \left\| U' - 2^{k-t}\widetilde{T} \right\| + \left\|U' - 2^{k-t}\widetilde{T}\right\| \times \left\vvvert H^{-1} \right\vvvert  \times \vvvert H \vvvert \\
+                        & 	\leq \left\| U' - 2^{k-t}\widetilde{T} \right\| \times \left(1 + \vvvert H^{-1} \vvvert  \times \vvvert H \vvvert \right).
 \end{align*}
@@
 lemma~\ref{thm:close} we have
-$\left\| 2^{k-\ell}\widetilde{T} - U' \right\| \leq 2 \sqrt{n}2^{64-\ell}-1
-$. So, if we fix $n$ we can search for the minimum number $\ell$ of known
+$\left\| 2^{k-t}\widetilde{T} - U' \right\| \leq 2 \sqrt{n}2^{64-t}-1
+$. So, if we fix $n$ we can search for the minimum number $t$ of known
 most-significant bits such that:
 \[
-  \left(1 + \vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert \right) 2 \sqrt{n} 2^{64-\ell} \leq \lambda_1(\mathcal{L})
+  \left(1 + \vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert \right) 2 \sqrt{n} 2^{64-t} \leq \lambda_1(\mathcal{L})
 \]
@@
 	\hline
-	n & $\vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert$ & $ \lambda_1(\mathcal{L}) $ & minimum $\ell$ \\
+	n & $\vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert$ & $ \lambda_1(\mathcal{L}) $ & minimum $t$ \\
 	\hline
@@
   \end{tabular}
-  \caption{minimal $\ell$ needed for a given $n$}
+  \caption{minimal $t$ needed for a given $n$}
   \label{tab:babai}
@@
 
-When $\ell$ is greater than the values given in table~\ref{tab:babai}, then
+When $t$ is greater than the values given in table~\ref{tab:babai}, then
 Babai's rounding technique will always return the closest vector, and allow us
@@
 Fig.~\ref{fig:prngs}. When clocked, it outputs the top 64 bits of a geometric
-sequence ($k=128$ and $\ell=64$). Three successive outputs are sufficient to
+sequence ($k=128$ and $t=64$). Three successive outputs are sufficient to
 reconstruct the internal state using Babai's rounding technique. This yields the
@@
 This ``baseline'' procedure requires $2^{64}$ iterations of a loop that does a
-dozen arithmetic operations; it always output the correct value of $S_0$, and
+dozen arithmetic operations; it always outputs the correct value of $S_0$, and
 may output a few other ones (they can be easily discarded by checking $X_2$). An
@@
 known and denote it by $w$. We define $S' = \LCG_{128}(S_0 - w, 0)$ and
-$K = \LCG_{128}(w, c)$ --- this is Known. By~\eqref{eq:lcg-additive}, we have
+$K = \LCG_{128}(w, c)$ --- this is known. By~\eqref{eq:lcg-additive}, we have
 $S' = S - K$. The point is that the elements of $S'$ follow a geometric
@@
 discussed in section~\ref{sec:geometric}, namely reconstructing a geometric
-sequence given $6+\ell$ (noisy) most-significant bits. The ``noise'' is the
+sequence given $t = 6+\ell$ (noisy) most-significant bits. The ``noise'' is the
 unknown vector $B$ of borrows. %Notice that the noise is asymmetric.
@@
 in guessing $S_0[64:128]$ and $S_1[64:128]$; using eq.~\eqref{eq:output}, this
-would reveal $S_0$ and $S_1$; from there, the increment $c$ is easily to obtain,
+would reveal $S_0$ and $S_1$; from there, the increment $c$ is easy to obtain,
 and every secret information has been reconstructed. This would take $2^{128}$
@@
 Note that $\nabla S_0 = 0$ and $\nabla S_1 = \Delta S_0$. Therefore, knowledge
-of $\Delta_0$ entails that of the whole sequence of $\nabla S_i$. The prediction
+of $\Delta S_0$ entails that of the whole sequence of $\nabla S_i$. The prediction
 algorithm we propose proceeds in three phases:
@@
 always odd; this makes one less bit to guess). We must also guess $k+1$
-successive rotation to get information on $k$ successive differences
+successive rotations to get information on $k$ successive differences
 $\Delta S_i$.
@@
 value is known and denote it by $w_0$ and $c_0$. We define
-$S' = \LCG_{128}(S_0 - w, c - c_0)$ and $K = \LCG_{128}(w_0, c_0)$ --- again,
-$K$ is Known and $S' = S - K$. This time, the components of $S'$ do \emph{not}
+$S' = \LCG_{128}(S_0 - w_0, c - c_0)$ and $K = \LCG_{128}(w_0, c_0)$ --- again,
+$K$ is known and $S' = S - K$. This time, the components of $S'$ do \emph{not}
 follow a geometric progression; but we still have that the $\ell$ least
@@
 The heart of the algorithm is again the reconstruction of a truncated geometric
-progression. Looking at table~\ref{tab:babai}, we see that the best choice
+progression knowing the \(t=\ell+6\) upper bits. Looking at table~\ref{tab:babai}, we see that the best choice
 consists in guessing 5 consecutive rotations and $\ell=14$ least-significant

### 2020-08-19 article/main.tex
encoreremarques

@@
 
-\title{Predicting the PCG Pseudo-Random Number Generator In Practice}
+\title{Practical seed-recovery for the PCG Pseudo-Random Number Generator}
 
@@
   \centering
-  \begin{tabular}{|c|Sc|Sc|c|}
+  \begin{tabular}{|c|Sc|Sc|c|Sc|}
 	\hline
-	n & $\vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert$ & $ \lambda_1(\mathcal{L}) $ & minimum $t$ \\
+	n & $\vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert$ & $ \lambda_1(\mathcal{L}) $ & minimum $t$ & \(\left(1 + \vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert \right) 2 \sqrt{n} 2^{64-t}\)\\
 	\hline
-	3 & 2.87 & $4.09e^{12} \simeq 2^{41.9}$ & 26 \\
-	4 & 2.06 & $2.44e^{14} \simeq 2^{47.8}$ & 20 \\
-	5 & 3.77 & $1.72e^{15}\simeq 2^{50.6}$ & 18 \\
-	6 & 2.69 & $1.03e^{16}\simeq 2^{53.2}$ & 15 \\
+	3 & 2.87 & $4.09e^{12} \simeq 2^{41.9}$ & 26 & \(3.69e^{12}\) \\
+	4 & 2.06 & $2.44e^{14} \simeq 2^{47.8}$ & 20 &\(2.15e^{14}\)\\
+	5 & 3.77 & $1.72e^{15}\simeq 2^{50.6}$ & 18 &\(1.5e^{15}\)\\
+	6 & 2.69 & $1.03e^{16}\simeq 2^{53.2}$ & 15 &\(1.02e^{16}\)\\
 	\hline
@@
 The heart of the algorithm is again the reconstruction of a truncated geometric
-progression knowing the \(t=\ell+6\) upper bits. Looking at table~\ref{tab:babai}, we see that the best choice
+progression knowing the \(t=\ell+6\) upper bits of four consecutive terms. Looking at table~\ref{tab:babai}, we see that the best choice
 consists in guessing 5 consecutive rotations and $\ell=14$ least-significant
@@
 Knowing the values of $\Delta S_0[0:64+\ell]$ as well as the $\ell$
-least-significant bits of $S_0$ and $c$ is sufficient to get rid the nastier
-feature \textsf{PCG64}: armed with this knowledge, we can determine all the
+least-significant bits of $S_0$ and $c$ is sufficient to get rid of the nastier
+feature of \textsf{PCG64}: armed with this knowledge, we can determine all the
 subsequent rotations deterministically, at negligible cost, using
@@
 dimension $63$. We use the off-the-shelf CVP solver embedded in \texttt{fplll}:
-it runs in no measurable time.
+it runs in negligible time.
 

### 2020-08-19 article/main.tex
hiddennumberprobleme

@@
 
-We will be facing a ``high-dimension'' instance in section~\ref{sec:big_cvp}
-and many ``low-dimension'' in sections~\ref{sec:Cknown}
+This is a particular case of the hidden number problem. We will be facing a ``high-dimension'' instance in section~\ref{sec:big_cvp}
+and many ``low-dimension'' instances in sections~\ref{sec:Cknown}
 and~\ref{sec:Cunknown}. The rest of this section discusses algorithmic tools to

### 2020-08-28 article/main.tex
dernieresmodifs

@@
 % !TeX spellcheck = en
-\documentclass[submission,svgnames,journal=tosc]{iacrtrans}
+\documentclass[journal=tosc,final]{iacrtrans}
 
@@
 \and 
-Sorbonne Université \\
+Sorbonne Université, F-75005 Paris, France \\
 \email{julia.sauvage@etu.upmc.fr}
@@
 \begin{document}
+\input{settings.tosc}
 \maketitle
@@
   they are not of cryptographic strength, their designer stated that predicting
-  their output should be nevertheless be "challenging".
+  their output should nevertheless be "challenging".
 
@@
 Pseudo-random generators (PRG) are well-studied primitives in symmetric
-cryptography. A PRG is an efficient deterministic algorithm that stretch a small
+cryptography. A PRG is an efficient deterministic algorithm that stretches a small
 random seed into a longer pseudo-random stream. To achieve cryptographic-grade
@@
 It is well-known that pseudo-random generators can be turned into symmetric
-encryption algorithm, by generating ``random'' masks to be used in the one-time
+encryption algorithms, by generating ``random'' masks to be used in the one-time
 pad. This is precisely what stream ciphers do.
@@
 simulation. Different generators produced different biases: in particular a
-given LFSRs yielded energy levels that were systematically too low and critical
+given LFSR yielded energy levels that were systematically too low and critical
 temperatures that were always too high, while another kind of generator yielded
@@
 denote by $U \bmod M$ the sequence $(U_0 \bmod M, U_1 \bmod M, \dots)$. The XOR
-operation is denoted $\oplus$, left and right rotations are denote $\lll$ and
+operation is denoted $\oplus$, left and right rotations are denoted $\lll$ and
 $\ggg$ respectively. Modular addition is denoted $+$ (or $\boxplus$ to make it
@@
 \]
-Where the ``multiplier'' $a$ is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
+where the ``multiplier'' $a$ is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
 the generator. The increment $c$ may be specified by the user of the PRNG to
@@
 multiplication operation. In fact, this can be done with three
-$64 \times 64 \rightarrow 128$ multiplication and two 64-bit additions. High-end
+$64 \times 64 \rightarrow 128$ multiplications and two 64-bit additions. High-end
 desktop CPUs all implement these operations in hardware, so the generator is
@@
 \]
-In the case of the $\| \cdot \|_2$ norm used throughout this paper, $\| M \|$ is
-the largest singular value of $G$; equivalently, it is the square root of the
+In the case of the $\| \cdot \|_2$ norm used throughout this paper, $\vvvert M \vvvert$ is
+the largest singular value of $M$; equivalently, it is the square root of the
 absolute value of the largest eigenvalue of $M^{t} M$.
@@
 Babai's rounding technique will always return the closest vector, and allow us
-to reconstruct a truncated geometric series.
+to reconstruct a truncated geometric serie.
 
@@
 states. Therefore, looking at $S_i[\ell:64+\ell]$, we are facing a truncated
-linear congruential generator on $64$ bits, where we have access to the most
-$6+\ell$ bits of each state (denoted by $T$), for a few consecutive states. This
+linear congruential generator on $64$ bits, where we have access to the
+$6+\ell$ most-significant bits of each state (denoted by $T$), for a few consecutive states. This
 is sufficient to reconstruct entirely the successive states of this truncated
@@
     % remplissage
-    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
+    \fill[fill=yellow] (4.6, 0) rectangle +(2.4, 1);
     \fill[fill=magenta] (0, 0) rectangle node {$r_0$} +(0.6, 1);
@@
     % remplissage
-    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
+    \fill[fill=yellow] (4.6, 0) rectangle +(2.4, 1);
     \fill[fill=magenta] (0, 0) rectangle node {$r_1$} +(0.6, 1);
@@
     % remplissage
-    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
+    \fill[fill=yellow] (4.6, 0) rectangle +(2.4, 1);
     \fill[fill=magenta] (0, 0) rectangle node {$r_2$} +(0.6, 1);
@@
 a total of $2^{6n + \ell}$ guessed bits. Table~\ref{tab:babai} gives a
-lower-bound on $\ell$ given $n$, and we see that the total number of guessed
+lower-bound on $t=6+\ell$ given $n$, and we see that the total number of guessed
 bits reaches a minimum of $38$ when $n=3$ and $\ell=20$. Therefore, success is
@@
 \]
-The point is that, thanks to the majority functions,
+The point is that, thanks to the majority function,
 $\mathcal{C}_{\mathfrak{i}}[j] = S_0[j-1]$ and
@@
 that the reconstruction procedure \emph{always} succeeds. In most cases, these
-parameters are pessimistic. We ran a series of experiments to determine more
+parameters are pessimistic. We ran a serie of experiments to determine more
 practical choices: using smaller-than-guaranteed values of $\ell$ (the number of
@@
   \cline{1-2}\cline{4-5}
-  19 & 1                  && 13 &  1 \\
+  19 & \(\approx 1\)                  && 13 &  \(\approx 1\) \\
   \cline{1-2}\cline{4-5}
@@
 seed). We implemented it in \textsf{C}, using \textsf{OpenMP} to parallelize the
-outer loop that guess the least-significant bits of the state. This yields a
+outer loop that guesses the least-significant bits of the state. This yields a
 simple multi-core implementation. We used the \texttt{gcc 8.3.0} compiler.

### 2020-11-01 article/main.tex
Corrections article

@@
 \usepackage{graphicx}
-\usepackage{url}
+\usepackage[hyphens]{url}
 \usepackage{xspace}
@@
 Sorbonne Université, F-75005 Paris, France \\
-\email{julia.sauvage@etu.upmc.fr}
+\email{julia.sauvage@etu.sorbonne-universite.fr}
 }
@@
 u128 a;
-u64 m = 0xda942042e4dd58b5;
 
@@
 {
-    a *= m;
+    a *= 0xda942042e4dd58b5;
     return a >> 64;
@@
 Therefore the algorithm requires $2^{41.67}$ CPU cycles in total; this makes
-less than 5 cycles per iteration of the inner loop. We used several
+less than 26 cycles per iteration of the inner loop. We used several
 implementation tricks to reach this level of efficiency:
@@
 With $\ell=11$, the whole process took $2^{56.74}$ CPU cycles, which makes less
-than 6 cycles per iteration of the inner loop. We used essentially the same
+than 54 cycles per iteration of the inner loop. We used essentially the same
 implementation tricks discussed above. However, this time we had to additionally
@@
 % pas dans la version anonyme.
-\paragraph{Acknowledgements.} The authors thank Thomas Espitau, Damien Vergnaud
-and Vincent Zucca for helpful discussions. We also thank Léo Ducas for helping
-us with the \texttt{G6K} library and Melissa O'Neil for kindly indulging our
-request for challenge inputs.
+\paragraph{Acknowledgements.} The authors thank Thomas Espitau, Virgile Tellier,
+Damien Vergnaud and Vincent Zucca for helpful discussions. We also thank Léo
+Ducas for helping us with the \texttt{G6K} library and Melissa O'Neil for kindly
+indulging our request for challenge inputs. We finally thank Alexandra Assanovna
+Elbakyan for helping scientists to access the works of their colleagues.
 

### 2020-11-01 slides/slides/6_implementation(Cknown).tex
slides

\begin{frame}{Implementation (Easy case, known increment)}

%With \(n=3\) and \(\ell=20\), we prove that the Babai rounding solve exactly the CVP in our case.

%In practice, the Babai Rounding is still exact for \(\ell=19\)

  \begin{block}{Summary}
    \begin{itemize}
    \item Guess 37 bits:
      \begin{itemize}
      \item \(n=3\) successive rotations (6 bits each),
      \item \(\ell=19\) least significant bits of \(S_0\),
      \end{itemize}
    \item Solve \(2^{37}\) instances of CVP in dimension 3 (Babai Rounding).
    \item Reconstruct initial state, check.
    \end{itemize}
  \end{block}

  \begin{alertblock}{Caveat}
    Attack proved correct for $\ell=20$, works fine for $\ell=19$...
  \end{alertblock}

  \begin{exampleblock}{Concretely...}
    \begin{itemize}
    \item $25$ CPU cycles per guess, 23 CPU-minutes in total.
    \end{itemize}
  \end{exampleblock}

\end{frame}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:

### 2020-11-01 slides/slides/8_implementation(Cunknown).tex
slides

\begin{frame}[label=bragging]{Implementation (Hard case, unknown increment)}
  \begin{block}{Summary}
    \begin{itemize}
    \item Guess $k=$51--55 bits:
      \begin{itemize}
      \item $n=5$ successive rotations (6 bits each),
      \item $\ell=$ 11--13 least significant bits of \(S_0\) \textbf{and} \alert{$c$}.
      \end{itemize}
    \item Solve \(2^{k}\) instances of CVP in dimension 4 (Babai Rounding).
    \item Consistency Check.
    \end{itemize}
  \end{block}

  \begin{alertblock}{Caveat}
    \begin{itemize}
    \item Attack proved correct for $\ell=14$.
    \item Works fine for $\ell=13$.
    \item OK with $p=0.66$ with $\ell=11$.
    \end{itemize}
  \end{alertblock}

  \begin{exampleblock}{Concretely...}
    \begin{itemize}
    \item $55$ CPU cycles per guess, 12.5k--20k CPU-hours in total.
    \end{itemize}
  \end{exampleblock}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}[label=bragging]{Doing it for Real}
  \begin{center}
    \includegraphics[width=0.7\linewidth]{pictures/Jean_Zay.jpg}
  \end{center}

  \begin{itemize}
    \item Nodes: 2$\times$ 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz}, ``\emph{Cascade Lake}''
    \item Running time: 35 minutes on 512 nodes.
    \item Complete recovery using 512 consecutive output bytes.
    \end{itemize}

\end{frame}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:

### 2020-11-02 slides/slides/8_implementation(Cunknown).tex
bosse sur les slides

@@
     \begin{itemize}
+    \item Observe 64 outputs (4096 bits).
     \item Guess $k=$51--55 bits:
@@
     \begin{itemize}
-    \item Attack proved correct for $\ell=14$.
-    \item Works fine for $\ell=13$.
-    \item OK with $p=0.66$ with $\ell=11$.
+    \item Attack proved correct for $\ell=14$ (works fine for $\ell=13$).
+    \item Succeeds with $p=0.66$ with $\ell=11$.
     \end{itemize}
@@
 \begin{frame}[label=bragging]{Doing it for Real}
-  \begin{center}
-    \includegraphics[width=0.7\linewidth]{pictures/Jean_Zay.jpg}
-  \end{center}
+  \begin{columns}
+    \begin{column}[T]{3.5cm}
+      \includegraphics[height=0.6\textheight]{pictures/jean-zay.jpg}
+    \end{column}
+    \begin{column}[T]{6cm}
+      \includegraphics[width=\textwidth]{pictures/jean_Zay.jpg}
+    \end{column}
+  \end{columns}
 
   \begin{itemize}
-    \item Nodes: 2$\times$ 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz}, ``\emph{Cascade Lake}''
-    \item Running time: 35 minutes on 512 nodes.
-    \item Complete recovery using 512 consecutive output bytes.
+    \item Used 512 nodes
+      \begin{itemize}
+      \item 2$\times$20-core \textsf{Xeon Gold 6248 @ 2.5Ghz}
+      \end{itemize}
+    \item Running time: 35 minutes.
+    \end{itemize}
+  \end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+  \begin{frame}{Conclusion}
+
+  \begin{itemize}
+    \item Reconstructing the seed for PCG is \textbf{practical}.
+    \item PCG is \textbf{not} cryptographically secure (never claimed to be).
+    \item \textbf{Don't} use \textsf{Numpy} to generate nonces...
     \end{itemize}
@@
 \end{frame}
+
 %%% Local Variables:

### 2020-11-12 Cknown/main.c
repo polishing

@@ -3,6 +3,11 @@
 #include <stdio.h>
 #include <time.h>
 
+/* 
+ * This program reconstructs the seed of PCG when the default increment is used.
+ * It requires 3 consecutive outputs (given in X in the main function). 
+ * The program runs in less than 25 minutes on a single core.
+ */
 
 void result_found(pcg128_t S, double start)
 {
@@ -26,9 +31,9 @@ int main()
     u64 X[nbiter];
 
     // challenge output given by M. O'Neil
-    X[ 0] = 0x6ec191a37a421087;
-    X[ 1] = 0xec140ace169176fc;
-    X[ 2] = 0x85994d489913af70;
+    X[0] = 0x6ec191a37a421087;
+    X[1] = 0xec140ace169176fc;
+    X[2] = 0x85994d489913af70;
 
     double start = wtime();
     u64 done = 0;

### 2020-11-13 lehmer.py
funny lehmer variant

@@ -1,15 +1,68 @@
-#Predictor input --- 3 consecutive calls to lehmer64()
-X = [0x1e1495fb7df7b3cf, 0xa267765f13b93bf9, 0xe5b3241c4f81c9a8]
+from itertools import islice
+import random
 
-# Prediction algorithm
-a = 0xda942042e4dd58b5
-r = round(2.64929081169728e-7 * X[0] + 3.51729342107376e-7 * X[1] + 3.89110109147656e-8 * X[2])
-s = round(3.12752538137199e-7 * X[0] - 1.00664345453760e-7 * X[1] - 2.16685184476959e-7 * X[2])
-t = round(3.54263598631140e-8 * X[0] - 2.05535734808162e-7 * X[1] + 2.73269247090513e-7 * X[2])
-u = r * 1556524 + s * 2249380 + t * 1561981
-v = r * 8429177212358078682 + s * 4111469003616164778 + t * 3562247178301810180
-w = (a**3 * (a*u + v)) >> 64
-x = w & 0xffffffffffffffff
+# actual implementation of the mystery function.
+state = 0x123456deadbeefbadcafe
+for i in range(16):
+    state = (state * 0xda942042e4dd58b5) % (2**128)
+    print("{}   :   {:016x}".format(state, state >> 64))
+    if i == 7:
+        print('----------------------------')
+    
+print('=================================')
 
-# Predictor output --- what the fourth call to lehmer64() would produce
-print(hex(x))
+
+def mystery(a, b, c, d, e, f, g, h):
+    """
+    >>> seed = [0xa872551a0848efe8, 0xeb568cb9eddd0de1, 0x72582f39ab1530c7, 0xf13af01e3a960ad8, \
+                0xc6bfb791661c6d3f, 0x71c294663911d20e, 0x41dc534da47e7130, 0x141e26c46484e8f2]
+    >>> stream = mystere(*seed)
+    >>> for x in islice(stream, 0, 8):
+    ...     print(hex(x))
+    0x3f76377c69c0d314
+    0xc8a788c2474ec97
+    0x8ee93c62db44939b
+    0xf385b6d382ab0e42
+    0x597e35963c650290
+    0xd8ca9a66f7b08d1a
+    0x2b05a43500bcbe58
+    0x157cd33ef08dfeac
+    """
+    prime = 2**64 - 59
+    while True:
[…]
